### 2017-11-02 article/article.tex
started the vim article

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}

\begin{document}
\title{On the Vim editor:\\
vim features and tweaks after 10 years of usage}

\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
The Vim editor is very rich in capabilities
and thus complex.
This article is a description of the Vim text editor
and a set of enhancements proposed for it.
It is the result of more than ten years of experience
in using Vim for writing and editing various types of documents,
mostly:
Python, C++, JavaScript, ChucK, etc, programs;
and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
binary files.
It is said that it takes about ten years to master this
text editor, and I find that other experienced users
have a different view of Vim and that they use a different
set of tools.
Therefore, this document exposes my insights in order
to confront my usage with the praxis of the Vim users community
and to make available a reference document with which new users
can grasp an overview by reading it and the discussions that it may generate.
Also, it should be useful for users of any degree of experience,
including me, as a compendium of commands, namespaces and tweaks.
Upon feedback, and maturing of my Vim usage,
this document might be enhanced or receive additional
material.
\end{abstract}

\section{Introduction}
Vim is a very complex editor,
considered by the Linux community to be
matched only by Emacs.
They both are the standard advanced text editors
of the open source community
and have beed developed for decades.
This document describes the Vim editor
and proposes a set of enhancements of the user
experience which
are implemented as Vim Plugins.
The editor has a very mature documentation
and the contents herein presented is a
report on the overall understandings I
have of Vim after a little bit more
of ten years using it.
The purposes of this document are:
\begin{itemize}
  \item to help new users in grasping Vim essentials
  and convenient practices.
  \item To record the view of the editor that
  an user (me) has after $\approx 10$ years of usage.
  \item To confront my usage with that of other experienced
  users. This is helpful for me, but also for the other users
  as they might benefit from this content and from discussions
  that might arise from it.
\end{itemize}

I used other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writting and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited database files and other types os files.
With Vim, I mostly write software (web and scientific),
music, poems and short stories.
It is very useful because:
\begin{itemize}
  \item it is meant to be a plain text (e.g. ascii, utf8) editor
  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
  \item It has a powerful architecture and set of commands.
  \item It is highly configurable and most often the users
  has a set of commands.
\end{itemize}

The standard capabilities of the editor
should become clear in Section~\ref{basics}.
Vim is constantly evolving and there are many plugins,
some of them very popular for both general and specific
type of users.
Accordingly, there are many possible enhancements,
and Section~\ref{issues} report the most prominent of them
for me and potential workarounds made available as plugins.

\subsection{Historical note}
Vim was first released publicly in 1991.
It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.

% Current stable, alpha, beta and development version
% no articles on Vim

\section{Basics}\label{basics}
Vim's interface is text-based.
In the GUI mode (gVim),
there are convenient menus and toolbars
but all functionalities are still available though
the command line mode.
Vimscript is the internal language of Vim,
and is often used for scripting by users
although other languages might be used 
(e.g. Python, Perl, Lua, Racker, Ruby and Tcl).

Vim has some basic modes of usage:
\begin{itemize}
  \item Navigation mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A code goal of the navigation mode it to allow fast
  navigation of the document while allowing
  the typist to maintain its fingers on the home row
  (i.e. on the center of the keyboard).
  \item Insert mode:
  \item Command mode:
  \item Ex mode:
\end{itemize}

\subsection{Help commands, files, tutor, and usr\_toc}

\subsection{Using Vim's modes}
\subsubsection{Navigating}
hjkl keys
crtl+[dubf]
jumps
markers
Most important help files on Netrw

\subsubsection{Insertion}
ctrl+o

\subsection{Navigation + insertion}
Commands that bridge from Navigation to Insertion:
csrCSR
Default, line and block visual selection:
  and AiIcCdD

\subsection{Netrw}
No insert mode.

Edit directory
Sexplore
Create and delete files and directories
open files: on window, split, preview
Change display of files
Bookmarks
Most important help files on Netrw

\subsection{Standard configuration files and directories and My .vim/vimrc}
\subsection{Spell and spelllang (en and pt\_br)}
\subsection{Tabs, splits, buffers and namespaces}
\& \% \$ and the following
\subsection{Mappings and abbreviations}
\subsection{Macros and registers for copy and paste}
\subsection{History of commands}
\subsection{List of markers}
\subsection{Undo}
\subsection{Scripting, Functions, Vimscript and Python}
\subsection{Plugins}
\subsubsection{Standard features}
\subsubsection{Writting plugins}
\subsubsection{Plugin systems: usage for using and writing plugins}
\subsection{Colour}
\subsubsection{Standard, 8 and 16 bits, and true color, Screen/Byobu}
\subsubsection{Gruvebox, Solarize and other colourschemes}
\subsection{Fonts}
\subsubsection{Cools nd popular fonts}
\subsubsection{How to set fonts in xterm, gnome-terminal and GVim.}
\subsection{Verbosing, logs and possibilities of using it to study your own}
\subsection{age (often used commands and typed sequences).}
\subsection{Highlighting}
\subsection{Bash and Vim commands}
\subsection{Compiling, standard features and plugins}
\subsection{Quickfix}
\section{Issues and plugins}\label{issues}

\section{Final words and further work}

Potential enhancements to this document are:
\begin{itemize}
  \item The inclusion of reading emails and connecting over ssh.
\end{itemize}

\end{document}

### 2017-11-03 article/article.tex
working on the modes

@@
 \usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{hyperref} 	%% Use to fix Figure or Table: ex: \begin{table}[H]
+\usepackage{hyperref}
 \hypersetup{
@@
 and to make available a reference document with which new users
-can grasp an overview by reading it and the discussions that it may generate.
+can grasp a sound overview by reading it and the discussions that
+it might generate.
 Also, it should be useful for users of any degree of experience,
@@
 of the open source community
-and have beed developed for decades.
+and have been developed for decades.
 This document describes the Vim editor
@@
   and convenient practices.
+  \item To make a sound overview of the editor available.
   \item To record the view of the editor that
-  an user (me) has after $\approx 10$ years of usage.
+  an user (me) has after $10$ years of usage.
   \item To confront my usage with that of other experienced
@@
 
-I used other editors, e.g. Kate, gedit, and Notepad2.
+I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
 I used Vim for writting and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
@@
 
+% Vi and Vim
 \subsection{Historical note}
@@
 It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
-
-% Current stable, alpha, beta and development version
-% no articles on Vim
+Vim's development is coordinated (and performed) since the beginning
+mainly by Bram Moolenaar.
+Today, current stable version is reported with 8.0.1173
+in the commit messages but tagged with 8.0.1257.
+I found no explicit alpha or beta versions
+and found no scientific article on Vim (this might be the first one).
 
@@
 
+\subsection{The bare minimum}\label{minimum}
+You open a file with Vim by executing
+the command: \texttt{vim <filename>}.
+Inside Vim, you start in the normal
+mode, and might want to move around using
+\texttt{h-j-k-l} for left-down-up-right.
+To insert characters, move your
+cursor to the desired location an press i,
+which puts Vim in the insert mode.
+Go back to normal mode by pressing
+\texttt{<ESC>}.
+You save the file by typing \texttt{:w<CR>},
+you exit vim by typing \texttt{:q<CR>}.
+
+\subsection{Vim help}
+You can find help for vim in various places.
+The standard resource is the Vim help resources.
+They are accessed by typing \texttt{:h <anything><CR>}
+in normal mode.
+Examples of such \texttt{<anything>} are:
+colour, navigation, :vs, vimtutor.
+Type 
+\texttt{:h usr\_toc<CR>}
+to access the official Vim User Manual,
+which is considerably lengthy and complex
+and is usually not read by users for a few years.
+In learning Vim, one
+might want to run the \texttt{vimtutor} command
+(outside Vim) to start the Vim Tutor.
+
+There are good resources on the Web for learning
+and tweaking Vim:
+\begin{itemize}
+  \item ``Vim Adventures'' is an online RPG game for practicing
+  and memorizing Vim commands while having fun.
+  This game is quite famous among Vim users.
+  \item There are official and semi-official Vim sites:
+  \url{www.vim.org}, \url{https://www.vi-improved.org} and
+  \url{http://vim.wikia.com/}.
+  \item Many hacks, understandings and general issues
+  (e.g. how to make such a move) are asked and answered
+  in online forums (e.g. Quora, Stack Overflow, Stack Exchange, Reddit).
+  One often finds these links through a search engine.
+  \item You can find many videos about Vim.
+  One traditional site is \url{http://vimcasts.org},
+  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+\end{itemize}
+
+There are mailing lists and IRC channels listed in these
+sites for getting help on Vim usage and development.
+
+% any other commands? files?
+
+\subsection{Using Vim's modes}
 Vim has some basic modes of usage:
 \begin{itemize}
-  \item Navigation mode: used for changing
+  \item Normal mode: used for changing
   the position of the cursor or the text displayed
   at the window.
-  A code goal of the navigation mode it to allow fast
+  A code goal of the normal mode it to allow fast
   navigation of the document while allowing
@@
   (i.e. on the center of the keyboard).
-  \item Insert mode:
-  \item Command mode:
-  \item Ex mode:
+  The mode is also used for manupulating text
+  (e.g. copy, paste, delete, change case).
+  \item Insert mode: for inserting text.
+  \item Command-line mode: for entering Ex commands.
+  \item Ex mode: similar to command-line mode,
+  but more specialized for running various Ex commands.
+  \item Visual mode: for making, manipulating and navigating
+  selections of texts.
+  \item Select mode: similar to visual mode but
+  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
 \end{itemize}
 
-\subsection{Help commands, files, tutor, and usr\_toc}
+There are seven additional modes which are mostly subordinate 
+to the basic modes and will be described when convenient.
+The manual page for Vim modes can be accessed by typing
+\texttt{:h vim-mode}.
 
-\subsection{Using Vim's modes}
-\subsubsection{Navigating}
-hjkl keys
-crtl+[dubf]
-jumps
-markers
-Most important help files on Netrw
+\subsubsection{Normal mode}
+Sometimes also called navigation or command mode,
+the normal mode is most powerful for
+navigating, manipulating texts and changing to other modes.
+
+The simplest of these three is changing to other modes:
+type any of these letters to change to insert mode:
+\texttt{iIaAoOsScC}. More on the transition between
+normal and insert mode on Section~\ref{navIn}.
+Type any of these characters to change to command-line mode:
+\texttt{:/?}.
+Type \texttt{Q} to enter Ex mode.
+Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
+
+For most basic and naive navigation, one should check Section~\ref{minimum}.
+Most often, one uses:
+\begin{itemize}
+  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
+and whole page down and up, although these commands might
+be set to scroll a different number of lines.
+  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
+  \item \texttt{(w,b,e)} to move to the next, previous and end-of-next word. There motions iterate over sequences of characters separated
+    by special characters (e.g. punctuation and parenthesis) as
+    specified by the output of \texttt{:se iskeyword}.
+    To interate over space-separated tokens, use \texttt{W,B,E}).
+    To move to the end of last word, one might use \texttt{be}
+    or \texttt{ge}.
+  \item \texttt{(fX,tX)} to move to or just before any X character.
+  \item \texttt{\},\{} for moving to the next blank line.
+  \item Search with / or ?, although these are in truth command-line 
+  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
+    jump list.
+  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
+    the first non-blank character of the line.
+    A mark is registered by the user in any cursor position
+    by typing \texttt{mX}, where X is any letter.
+    If X is lowercase, the mark is local to the buffer (the file),
+    if it is uppercase or numeric, it is global to the Vim session.
+\end{itemize}
+
+There are many more facilities to navigate Vim
+as explained in \texttt{:h navigation}.
+E.g. one might find useful the \texttt{),(} commands
+for iterating through sentences.
+
+For changing the text, usual commands include:
+\begin{itemize}
+  \item \texttt{d\{motion\}} to delete the characters
+    involved in the motion command.
+  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
+  \item \texttt{x,X} to delete the character under or before the cursor  \item \texttt{~} to swap the case of a character.
+  \item \texttt{gu\{motion\},gU\{motion\},g~\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
+\end{itemize}
+
+There are way more commands to change the texts.
+Some of them are discussed in Section~\ref{navIn}
+because they involve a transition to the insertion mode.
+A thorough consideration of the commands in the command mode
+is found by executing \texttt{:h navigation} and 
+\texttt{:h change.txt}.
 
 \subsubsection{Insertion}
-ctrl+o
+Once in the insertion mode, the character keys
+input the characters at the cursor position.
+One can exit insert mode by pressing \texttt{<Esc>},
+and Vim will be put in normal mode.
+Most used commands in insert mode include:
+\begin{itemize}
+  \item \texttt{Ctrl+o} to execute one and only command in normal mode.
+  \item \texttt{asdjn}
+\end{itemize}
 
-\subsection{Navigation + insertion}
+\subsubsection{Command-line mode}
+% enter by typing :, ?, / in normal or visual modes.
+
+\subsection{Navigation + insertion}\label{navIn}
 Commands that bridge from Navigation to Insertion:
-csrCSR
+csrCSR, iws, etc
 Default, line and block visual selection:
@@
 \subsection{Netrw}
+Most important help files on Netrw
 No insert mode.
@@
 \subsection{History of commands}
-\subsection{List of markers}
+\subsection{Lists of Vim state variables}
+% of markers, jumps, registers, commands (hist)
+% :
 \subsection{Undo}
@@
 \subsection{Quickfix}
+\subsection{Persistence and vimrc}
+% markers, undo, buffers, registers, bookmarked directories,
+% vimrc for settings
 \section{Issues and plugins}\label{issues}
@@
 
+\appendix
+\section{Example of usage session}
+% open file, split, tab, find, replace char,
+% append to line, visual block, etc.
+\section{Key naming in Vim documentation}
+% CTRL, <C-X> <C-S-X>
+% Special chars..
+\section{My vimrc file}
+In this Appendix is my vimrc file.
+Although it has comments, one should
+look for the options that (s)he does
+not understand, as a thorough explanation
+of the file is tedious and out of the scope
+of this document.
+
+\section{Overview of my usage of Vim}
+% historical usage, other editors,
+% most often edited file types
+
 \end{document}

### 2017-11-04 article/article.tex
added useful Vim src/Makefile. Used for the article.

@@
   that might arise from it.
+  \item To propose some enhancements to Vim through simple plugins.
 \end{itemize}
 
+Advanced users might just focus on the proposed enhancements and
+in some of the basic topics, such as in using numerous terminals 
+(it works pretty well!) and using sessions (Section~\ref{sessions}).
+These make Vim very convenient e.g. for users that edit multiple files,
+run scripts and compile programs and latex files.
+I used for years multiple terminal abs and windows,
+and sometimes a byobu/screen terminal multiplexer.
+Nowadays I am using standard Vim windows as terminals
+such as described in Section~\ref{terminals}.
+
 I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
-I used Vim for writting and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
+I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
 Eventually, I edited database files and other types os files.
@@
 \subsection{Using Vim's modes}
-Vim has some basic modes of usage:
+Vim has some basic and fully implemented modes of usage:
 \begin{itemize}
@@
 \end{itemize}
+
+There is another basic mode, but it is not fully implemented:
+the Terminal-Job mode.
 
@@
     To interate over space-separated tokens, use \texttt{W,B,E}).
-    To move to the end of last word, one might use \texttt{be}
+    To move to the end of last word, one might bumba use \texttt{be}
     or \texttt{ge}.
@@
   \item \texttt{Ctrl+o} to execute one and only command in normal mode.
+	  This enter a secondary mode (see Section~\ref{modes})
   \item \texttt{asdjn}
@@
 % enter by typing :, ?, / in normal or visual modes.
+
+\subsubsection{Terminal-Job mode}\label{terminal}
 
@@
 \subsection{Persistence and vimrc}
+\subsubsection{Sessions}\label{sessions}
 % markers, undo, buffers, registers, bookmarked directories,
@@
 \section{Issues and plugins}\label{issues}
+\subsection{Sessions}
+\subsection{AA messages}
+\subsection{Slick Vim}
+% tab line and status line and ruler
 
@@
 % Special chars..
-\section{My vimrc file}
+\section{My vimrc file and usage}
 In this Appendix is my vimrc file.
@@
 
+Mostly, I toggle the status line with \texttt{\\\\B}
+and the tab line with \texttt{\\\\T}.
+Save and close windows with \texttt{\\w} and \texttt{\\q}.
+
 \section{Overview of my usage of Vim}
+% sessions, terminal for ipython3 and evince
+% alternative with byobu or many terminals
 % historical usage, other editors,

### 2017-11-08 article/article.tex
bunch of things done in the article

@@
 \usepackage[utf8]{inputenc}
+\newcommand{\ttt}[1] {
+	\texttt{<#1>}}
+\newcommand{\tttt}[1] {
+	\texttt{#1}}
 
 \begin{document}
-\title{On the Vim editor:\\
-vim features and tweaks after 10 years of usage}
-
+\title{An anthropological account of the Vim editor:\\
+ouroboros, features and tweaks after 10 years of usage}
+% Anthropological computation
 \author{Renato Fabbri\\
@@
 in using Vim for writing and editing various types of documents,
-mostly:
-Python, C++, JavaScript, ChucK, etc, programs;
+e.g.:
+Python, C++, JavaScript, ChucK, programs;
 and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
 binary files.
-It is said that it takes about ten years to master this
-text editor, and I find that other experienced users
+It is said that it takes about ten years to master (or start mastering)
+this text editor, and I find that other experienced users
 have a different view of Vim and that they use a different
@@
 Therefore, this document exposes my insights in order
-to confront my usage with the praxis of the Vim users community
+to confront my usage with that of other Vim users
 and to make available a reference document with which new users
@@
 They both are the standard advanced text editors
-of the open source community
+of the free software and open source communities
 and have been developed for decades.
@@
   and convenient practices.
-  \item To make a sound overview of the editor available.
+  \item To attain a sound overall description of the editor.
   \item To record the view of the editor that
-  an user (me) has after $10$ years of usage.
+  a user (me) has after $10$ years of usage.
   \item To confront my usage with that of other experienced
@@
 
-Advanced users might just focus on the proposed enhancements and
-in some of the basic topics, such as in using numerous terminals 
-(it works pretty well!) and using sessions (Section~\ref{sessions}).
-These make Vim very convenient e.g. for users that edit multiple files,
-run scripts and compile programs and latex files.
-I used for years multiple terminal abs and windows,
-and sometimes a byobu/screen terminal multiplexer.
-Nowadays I am using standard Vim windows as terminals
-such as described in Section~\ref{terminals}.
+Advanced users might just skim through the reference presented
+at the basic Section~\ref{basics},
+and focus on the proposed enhancements.
 
@@
   \item It is highly configurable and most often the users
-  has a set of commands.
+  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
 \end{itemize}
@@
 
+This document is written is a DRY KISS
+(Don't Repeat Yourself, Keep It Simple Stupid) style.
+Complex is to master the use of Vim
+and one finds sound
+references in help files and a nice vimrc.
+Therefore the following content should be kept
+as uncomplicated and original as possible.
+Also, because of Vim's complexity and entailed
+bond of this document to my usage,
+there is an anthropological component
+which is evdent in the use of the first person
+in sentences.
+This can be understood as anthropological computation
+or physics~\cite{anPh,anPh2} and observed to help
+in the technological groundwork of civil society.
+
 % Vi and Vim
@@
 mainly by Bram Moolenaar.
-Today, current stable version is reported with 8.0.1173
-in the commit messages but tagged with 8.0.1257.
-I found no explicit alpha or beta versions
-and found no scientific article on Vim (this might be the first one).
+Today, current bleeding-edge version is 8.0.1257.
+I found no explicit stable, alpha or beta versions
+and found no scientific article on Vim 
+(this might be the first one).
 
@@
 (e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
+Each line of a Vimscript is a command on the
+command-line mode.
 
@@
 Go back to normal mode by pressing
-\texttt{<ESC>}.
+\texttt{<ESC>} for \texttt{<C-C}.
 You save the file by typing \texttt{:w<CR>},
@@
 \texttt{:h usr\_toc<CR>}
-to access the official Vim User Manual,
+to access the official User Manual,
 which is considerably lengthy and complex
@@
 % any other commands? files?
+\subsection{Namespaces and state lists}
+Vim is a text editor ouroboros because
+text and writting alters text and writting.
+You have namespaces where tokens have scalar
+or complex values.
+In Vimscript, the colon commands (also Ex or command-line commands) are related through spaces, punctuations and keywords (see \texttt{:h script}.
+
+\subsubsection{Commands and mappinds}
+There are commands, typing sequences which trigger automated actions,
+for each mode:
+\begin{itemize}
+	\item in Normal mode all keys are mapped to commands.
+		There redundancy and additional commands
+		using Ctrl and Shift keys.
+		Some keys expect a second key,
+		and have available possibilities,
+		specially the z and g.
+	\item in the other modes, the sequences available for mappings are more obvious. One should look at \ttt{:h index} to know about all the mappings.
+\end{itemize}
+
+\ttt{C-\textbackslash} is always reserved for extensions,
+which make it a safe namespace to use (while there are no
+such extensions).
+
+A colon command can be written as a string
+and executed by the \tttt{:execute} colon command.
+E.g. \tttt{:execute 'vs afile.txt'}.
+As there are colon commands that execute commands in other
+modes, e.g. \tttt{:normal ?\^def },
+the \tttt{:execute} is the way to build commands in any mode:
+\tttt{:execute 'normal' tabn 'gt'}.
+
+\subsubsection{Variables}
+There are some types of variables in Vim:
+\begin{itemize}
+	\item Environment variables: names start with \$ an hold system
+		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
+	\item Option variables: names start with a \& and are meant to control the behavior of the editor.
+		One might change a value through set or let, e.g.
+		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
+	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
+	\item Internal variables are created with let and preferably have a prefix:
+	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
+		are local to the buffer, window, tab page, function, and
+		sourced Vim file, respectively.
+ \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
+		\tttt{a:} is for function arguments.
+		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
+		More about internal variables in \tttt{:h internal-variables}.
+	\item The values of variables can be scalars, strings, lists, dictionaries, function references, etc (\tttt{:h eval}).
+\end{itemize}
+You can echo any of such variables or use in expressions.
+Notice that you will only be able to echo a \tttt{b:} variable inside
+the buffer where it is defined.
+For all the Vimscript capabilities, including loops, conditionals,
+and builting functions, refer to \ttt{:h vim-script}.
+Rudimentary classes are possible through dictionaries,
+but the language is otherwise overall quite powerful in
+dealing this text and commands.
+
+\subsubsection{State lists}
+Vim keeps a number of lists which expresses the state of the editor.
+These are examples of useful lists:
+\begin{itemize}
+	\item The entered commands are accessed through \tttt{:hist a}. the tokens a / s : can be used for specific types of commands, such
+		are search and colon commands.
+	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load it to the window with \tttt{:b <num or token in file name>}.
+	\item The windows open are listed is \tttt{:ls} with markers and in \tttt{:tabs}.
+	\item Tabs list can be reached through \tttt{:tabs}.
+		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
+	\item Jumps are available through \tttt{:jumps}.
+		One repositions the cursor at each jump position through \ttt{C-o} and \ttt{C-i}.
+	\item An undo list can be reached with \tttt{:changes}.
+	\item Quickfix and Location lists which are populated through \tttt{:make} and \tttt{:grep}.
+
+	\item Tags list
+	\item Argument list
+\end{itemize}
+
+A file with information about the state of the editor
+can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
+and the file has the commands to access various settings of Vim.
+Another good list of commands to know about Vim's state is kept on
+\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
+I would specially highlight the \tttt{:syntax} command because
+it displays the tokens and related type of meanings when run
+inside e.g. a \tttt{.vim} or help file.
+
+You can keep track of states though commands and persistent visual
+cues, specially the tabs bar, the status line, and the line
+reserved for the command-line.
+For state persistense, one might keep an undo file for each file as
+in~\cite{vimrc}.
+Sessions are easy to manage, enabling one to save and load the
+editor's state, with the opened windows, tabs, buffers, etc.
+I use the mappings in~\cite{vimrc} because they keep the sessions
+in a reasonable directory and makes it easier to remember and tweak
+then the standard commands to deal with sessions.
+More information in \tttt{:h sessions}.
+One might use \tttt{:h views} to keep the state of one window,
+but sessions keep all the states from all windows.
+This entails a strategy to deal with Vim that is similar to
+the use of Byobu/Tmux/Screen.
+The main limitation I found to this approach is that
+Vim is not keeping track of the terminals opened.
+If you open a terminal inside Vim with the \tttt{:term}
+command, you will save the session as usual, but when loading
+you get dummy empty windows for them and an error message.
+Screen and Tmux are the most popular terminal multiplexers.
+Byobu, built on top of them, has awesome keyboard shortcuts
+for managing sessions, windows and splits of terminals.
+Byoby/Screen/Tmux keep the state for future load.
+
+Because browsing the interface in Vim is so fast,
+and it favors copy and paste of text,
+I tend to keeep a tab with some terminals:
+one with an IPython shell, another two for compiling latex
+and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
+One enters normal mode in terminal with \tttt{C-W N}
+and it is very corfortable to copy and browse the bash history of the
+terminal.
+I found the mappings on~\cite{vimrc} very helpful for directing
+editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
+which I make available across terminal and normal modes.
+But I've been thinking on using \ttt{C-} commands also
+for tabs (not only for splits).
+
+A good strategy I find is to have selective visual cues of the state to make persistent or hide. A mapping to toggle each of them.
+Now I toggle byobu/screen/tmux bar with \ttt{F5},
+status line and tabs bar with \ttt{localleader-TB} according to script~\cite{vimrc}.
+I am mostly using the cleanest setting, toggling the tabs bar sometimes. Numbering is always there. I rarely turn them off but keep the mappings \ttt{leader-nN} just in case.
+Instead of keeping the status bar, I use \ttt{C-g} to know about the
+file and \ttt{gC-g} to know more and rarily.
+It seems not possible to remove the statusbar between splits.
+After asking around and and experiments, I realized that it seems
+reasonable to keep at least one line dividing the windows, so
+if it comes to it, I just \tttt{set statusline=-}.
+Unfortunatelly, as far as I could dig, one will need to enter
+Vim code to enable a horizontal split without losing a line.
+My ideal of this feature would be to have a visual cue of the
+first and/or last line of the windows in the splits in the number
+column, or complete the spaces and empty chars with \$\$\$\$ or so.
 
@@
 \subsection{Mappings and abbreviations}
+\texttt{:h index} shows all the mappings.
+Every letter and character in the keyboard is used.
+In Normal mode: <TAB> is the same as <C-I>,
+<BS> and <C-H> is the same as h (<C-H> don't sowk here).
+<C-J> and <C-N> is the same as j.
+C-P is the same as k.
+Space is same as l
+CTRL-[ <Esc> are not used
+C-\ a-z reserved for extensions
+CTRL-\_ not used
+
+learn how to use filters ! and !!
+% .%#: registers
+\url{https://github.com/vim/vim/blob/master/runtime/doc/index.txt}.
+
+my usage of <C-(HJKL)> for jumping through windows.
+<C-L> is lost though, and require :redraw or :e
+
++ same as <CR> which is not that useful.
+
+Many ][ combinations are not used, e.g.
+with abhjklfg
+\_ might be used as a more powerful \^, leaving it free for mappings.
+
+Directions, home, end, page up and down, insert, all have mappings
+in more centraly located keys.
+unbounded: g,z,[]
+
+Del is same as x
+
+text objects: redundancy between a/i(),{},[],<>
+
+many entries available through the g commands
+z for folds, spell checking and some movements (mainly when wrap is set).
+
+\subsubsection{Insert mode}
+All standard char keys are used for entering chars on text.
+<C-G>jk can be achieved by <C-O>jk which is more powerful
+in moving through multiple lines.
+j-[ same as ESC
+		CTRL-\ a - z	reserved for extensions
+		CTRL-\ others	not used
+
+<C-J> and <C-M> are <CR>
+
 \subsection{Macros and registers for copy and paste}
@@
 
+Potential next steps in using Vim:
+\begin{itemize}
+  \item Measure the performance text mining routines in Vim against those implemented in C or Python.
+  \item Make a functionalities to list all the mappings available in each mode and the typing combinations which are available.
+\end{itemize}
+
 \appendix
@@
 \section{Key naming in Vim documentation}
-% CTRL, <C-X> <C-S-X>
+% CTRL, <C-X> <C-S-X>s 1250
 % Special chars..
@@
 \section{Overview of my usage of Vim}
+% move to usage appendix TTM
+in some of the basic topics, such as in using numerous terminals 
+(it works pretty well!) and using sessions (Section~\ref{sessions}).
+These make Vim very convenient e.g. for users that edit multiple files,
+run scripts and compile programs and latex files.
+I used for years multiple terminal abs and windows,
+and sometimes a byobu/screen terminal multiplexer.
+Nowadays I am using standard Vim windows as terminals
+such as described in Section~\ref{terminals}.
+
 % sessions, terminal for ipython3 and evince

### 2017-11-09 article/article.tex
finalizando revisao da parte principal

@@
 This article is a description of the Vim text editor
-and a set of enhancements proposed for it.
+and a set of considerations on its usage and design.
 It is the result of more than ten years of experience
-in using Vim for writing and editing various types of documents,
+with Vim for writing and editing various types of documents,
 e.g.:
@@
 have a different view of Vim and that they use a different
-set of tools.
+set of features.
 Therefore, this document exposes my insights in order
-to confront my usage with that of other Vim users
-and to make available a reference document with which new users
+to confront my usage with that of other Vim users.
+Another goal is
+to make available a reference document with which new users
 can grasp a sound overview by reading it and the discussions that
@@
 
-\section{Introduction}
+\section{Introduction}\label{intro}
 Vim is a very complex editor,
@@
 and proposes a set of enhancements of the user
-experience which
-are implemented as Vim Plugins.
+experience through simple tweaks.
 The editor has a very mature documentation
@@
   that might arise from it.
-  \item To propose some enhancements to Vim through simple plugins.
+  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
 \end{itemize}
 
-Advanced users might just skim through the reference presented
-at the basic Section~\ref{basics},
-and focus on the proposed enhancements.
+Advanced users might just skim throughj
+Section~\ref{basics} and consider more carefully
+Section~\ref{namespaces}.
+The concluding remarks and proposed enhancements
+might also deserve some attention.
 
@@
 some of them very popular for both general and specific
-type of users.
+users.
 Accordingly, there are many possible enhancements,
-and Section~\ref{issues} report the most prominent of them
+and Section~\ref{conc} report the most prominent of them
 for me and potential workarounds made available as plugins.
@@
 or physics~\cite{anPh,anPh2} and observed to help
-in the technological groundwork of civil society.
+in the technological groundwork of the civil society.
 
@@
 Today, current bleeding-edge version is 8.0.1257.
-I found no explicit stable, alpha or beta versions
-and found no scientific article on Vim 
-(this might be the first one).
+I found no explicit stable, alpha or beta versions.
+I found no scientific article on Vim 
+(this might be the first one), although there are
+books, software and third-party documentation on the web.
 
@@
 command-line mode.
+This section can be thought as a tutorial
+that focuses on the namespaces, i.e. sets of tokens
+that carry values or triggers procedures.
 
@@
 Go back to normal mode by pressing
-\texttt{<ESC>} for \texttt{<C-C}.
+\texttt{<ESC>} or \texttt{<C-C>}.
 You save the file by typing \texttt{:w<CR>},
-you exit vim by typing \texttt{:q<CR>}.
+and exit Vim by typing \texttt{:q<CR>}.
+You can save and quit with \tttt{:wq<CR>}
+or \tttt{:x<CR>} or \tttt{ZZ}.
 
@@
 You can find help for vim in various places.
-The standard resource is the Vim help resources.
+The standard resource is the Vim help files.
 They are accessed by typing \texttt{:h <anything><CR>}
@@
 Examples of such \texttt{<anything>} are:
-colour, navigation, :vs, vimtutor.
+color, navigation, :vs, vimtutor.
 Type 
@@
   This game is quite famous among Vim users.
-  \item There are official and semi-official Vim sites:
+  \item There are official and semi-official Vim sites e.g.:
   \url{www.vim.org}, \url{https://www.vi-improved.org} and
@@
   (e.g. how to make such a move) are asked and answered
-  in online forums (e.g. Quora, Stack Overflow, Stack Exchange, Reddit).
+  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
   One often finds these links through a search engine.
@@
 
-There are mailing lists and IRC channels listed in these
-sites for getting help on Vim usage and development.
-
 % any other commands? files?
-\subsection{Namespaces and state lists}
-Vim is a text editor ouroboros because
-text and writting alters text and writting.
+\subsection{Namespaces}\label{namespaces}
+Vim is a text editor ouroboros~\cite{ouroWiki} because
+text and writing alters text and writing.
 You have namespaces where tokens have scalar
 or complex values.
-In Vimscript, the colon commands (also Ex or command-line commands) are related through spaces, punctuations and keywords (see \texttt{:h script}.
 
-\subsubsection{Commands and mappinds}
+\subsubsection{Commands and mappings}
 There are commands, typing sequences which trigger automated actions,
@@
 	\item in Normal mode all keys are mapped to commands.
-		There redundancy and additional commands
+		There is redundancy and additional commands
 		using Ctrl and Shift keys.
@@
 		specially the z and g.
-	\item in the other modes, the sequences available for mappings are more obvious. One should look at \ttt{:h index} to know about all the mappings.
+    See Section~\ref{normal} and Appendix~\ref{notes}
+    for more insights into the commands available in the
+    normal mode.
+	\item In the other modes, the sequences available for mappings are more obvious and abundant.
+		One should look at \ttt{:h index} to know about all the standard mappings
+		and use \tttt{:map} to list the user-defined mappings.
 \end{itemize}
@@
 \ttt{C-\textbackslash} is always reserved for extensions,
-which make it a safe namespace to use (while there are no
+which makes it a safe namespace to use (while there are no
 such extensions).
@@
 modes, e.g. \tttt{:normal ?\^def },
-the \tttt{:execute} is the way to build commands in any mode:
+the \tttt{:execute} is a way to build commands in any mode,
+e.g.
 \tttt{:execute 'normal' tabn 'gt'}.
@@
 		More about internal variables in \tttt{:h internal-variables}.
-	\item The values of variables can be scalars, strings, lists, dictionaries, function references, etc (\tttt{:h eval}).
+	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
 \end{itemize}
@@
 and builting functions, refer to \ttt{:h vim-script}.
-Rudimentary classes are possible through dictionaries,
-but the language is otherwise overall quite powerful in
-dealing this text and commands.
+Classes are possible only in rudimentary forms, e.g. through dictionaries,
+but the language is otherwise overall quite powerful,
+specially in dealing with text and editor behavior, as expected
 
-\subsubsection{State lists}
+\subsubsection{State lists}\label{state}
 Vim keeps a number of lists which expresses the state of the editor.
@@
 \begin{itemize}
-	\item The entered commands are accessed through \tttt{:hist a}. the tokens a / s : can be used for specific types of commands, such
-		are search and colon commands.
-	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load it to the window with \tttt{:b <num or token in file name>}.
-	\item The windows open are listed is \tttt{:ls} with markers and in \tttt{:tabs}.
+	\item The entered commands are accessed through \tttt{:hist a}. 
+    The tokens a / s : can be used for specific types of commands, such
+		as search and colon commands.
+	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
+  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
 	\item Tabs list can be reached through \tttt{:tabs}.
@@
 	\item Jumps are available through \tttt{:jumps}.
-		One repositions the cursor at each jump position through \ttt{C-o} and \ttt{C-i}.
-	\item An undo list can be reached with \tttt{:changes}.
-	\item Quickfix and Location lists which are populated through \tttt{:make} and \tttt{:grep}.
-
-	\item Tags list
-	\item Argument list
+		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
+  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
+    They also keep track or your copy, edition and deletion and are promptly defined
+    by recording a typing sequence with the \tttt{q} normal command.
+	\item An undo list can be accessed with \tttt{:changes}.
+	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
+	\item The markers defined are listed with the \tttt{:marks} command.
+		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
+		Uppercase letters are cross buffer.
+	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
+		and variations, such as \tttt{:grep}.
+		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
+		to open the Quickfix window, where the lines of occurrence are in sequence
+		and one can \ttt{CR} one of them to have the cursor in the main window active at
+		the first character of the match.
+		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
+		to use the location-list window instead of the Quickfix, which
+		is very similar, but one per window instead of one per buffer.
+		More information in \tttt{:h quickfix}.
+	\item A tags list have to be made so one can use tags.
+		Most often one will generate the tags list using the exuberant ctags,
+		which supports dozens of languages.
+		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
+		and then using \ttt{C-]} to
+		go to the position of tag under cursor, and \tttt{:tabe tags} to open
+		the tags file.
+	\item The argument list holds a list of files to be edited or browsed. 
+		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
+	or using commands (e.g. \tttt{:ar ./*}).
+		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
+		one might perform actions on each file in argument list using \tttt{:argdo}.
+		All files in argument list is also in the buffers list.
+    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
 \end{itemize}
@@
 can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
-and the file has the commands to access various settings of Vim.
+and this script might be consulted because it has
+a collection of commands to access various settings of Vim.
 Another good list of commands to know about Vim's state is kept on
@@
 
-You can keep track of states though commands and persistent visual
+\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
+
+You can keep track of the editor state though commands and persistent visual
 cues, specially the tabs bar, the status line, and the line
 reserved for the command-line.
-For state persistense, one might keep an undo file for each file as
+For state persistence, one might keep an undo file for each file as
 in~\cite{vimrc}.
@@
 
-Because browsing the interface in Vim is so fast,
+Because browsing the interface in Vim is fast,
 and it favors copy and paste of text,
@@
 Now I toggle byobu/screen/tmux bar with \ttt{F5},
-status line and tabs bar with \ttt{localleader-TB} according to script~\cite{vimrc}.
-I am mostly using the cleanest setting, toggling the tabs bar sometimes. Numbering is always there. I rarely turn them off but keep the mappings \ttt{leader-nN} just in case.
+status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
+I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
+Numbering is always there.
+I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
 Instead of keeping the status bar, I use \ttt{C-g} to know about the
 file and \ttt{gC-g} to know more and rarily.
-It seems not possible to remove the statusbar between splits.
-After asking around and and experiments, I realized that it seems
+It seems not possible to remove the statusbar between horizontal splits.
+After asking around and experimenting, I realized that it seems
 reasonable to keep at least one line dividing the windows, so
@@
 My ideal of this feature would be to have a visual cue of the
-first and/or last line of the windows in the splits in the number
+first and/or last line of the windows in the lines-number
 column, or complete the spaces and empty chars with \$\$\$\$ or so.
 
-\subsection{Using Vim's modes}
+Autocommands are the standard way to define event-trigered routines in Vim.
+These are often related to particular file types, but are also often
+in defining the automated behavior of Vim.
+In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
+in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
+
+\subsection{Using Vim's modes}\label{modes}
 Vim has some basic and fully implemented modes of usage:
@@
   at the window.
-  A code goal of the normal mode it to allow fast
+  A core goal of the normal mode it to allow fast
   navigation of the document while allowing
@@
   The mode is also used for manupulating text
-  (e.g. copy, paste, delete, change case).
+  (e.g. copy, paste, delete, change case) and
+    changing to other modes.
   \item Insert mode: for inserting text.
@@
 the Terminal-Job mode.
-
 There are seven additional modes which are mostly subordinate 
@@
 \texttt{:h vim-mode}.
+Some of the modes are now further considered for
+an overview of the Vim usage possibilites.
 
-\subsubsection{Normal mode}
+\subsubsection{Normal mode}\label{normal}
 Sometimes also called navigation or command mode,
@@
     specified by the output of \texttt{:se iskeyword}.
-    To interate over space-separated tokens, use \texttt{W,B,E}).
-    To move to the end of last word, one might bumba use \texttt{be}
+    To iterate over space-separated tokens, use \texttt{W,B,E}).
+    To move to the end of last word, one might use \texttt{be}
     or \texttt{ge}.
-  \item \texttt{(fX,tX)} to move to or just before any X character.
-  \item \texttt{\},\{} for moving to the next blank line.
-  \item Search with / or ?, although these are in truth command-line 
+The \texttt{),(} commands iterate through sentences,
+		\tttt{\},\{} through text blocks separated by empty lines.
+  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
+	  \tttt{;} and \tttt{,} for next and previous found character.
+  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
   \item \texttt{CTRL+(o,i)} to move to an older or newer position in
@@
 as explained in \texttt{:h navigation}.
-E.g. one might find useful the \texttt{),(} commands
-for iterating through sentences.
 
@@
   \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
-  \item \texttt{x,X} to delete the character under or before the cursor  \item \texttt{~} to swap the case of a character.
-  \item \texttt{gu\{motion\},gU\{motion\},g~\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
+  \item \texttt{x,X} to delete the character under or before the cursor.
+  \item \texttt{$\sim$} to swap the case of a character.
+  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
 \end{itemize}
@@
 because they involve a transition to the insertion mode.
-A thorough consideration of the commands in the command mode
+A thorough consideration of the commands in the normal mode
 is found by executing \texttt{:h navigation} and 
@@
 Once in the insertion mode, the character keys
-input the characters at the cursor position.
+input the characters at the cursor position at the current buffer.
 One can exit insert mode by pressing \texttt{<Esc>},
 and Vim will be put in normal mode.
-Most used commands in insert mode include:
+Most useful commands in insert mode include:
 \begin{itemize}
@@
 	  This enter a secondary mode (see Section~\ref{modes})
-  \item \texttt{asdjn}
+  \item \ttt{C-R} to paste a register (a variable starting with
+    '@', defined, copied or recorded through a \tttt{q} command in
+    normal mode and as covered in Section~\ref{state}).
+  \item \ttt{C-T} to indent current line.
+  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
+    the line, the previous word.
+  \item \ttt{C-N,P} to find next, pervious keywords that match the prefix at hand..
+  \item \ttt{C-X} commands for scrolling the window with multiple
+    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
 \end{itemize}
@@
 \subsubsection{Command-line mode}
-% enter by typing :, ?, / in normal or visual modes.
+This mode is dedicated for writing colon, search and filter commands,
+entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
+Most useful commands in this mode include:
+\begin{itemize}
+  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
+    end of the line.
+  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
+    until the cursor.
+  \item \ttt{C-R} to paste a register as in insert mode.
+\end{itemize}
 
 \subsubsection{Terminal-Job mode}\label{terminal}
+This mode is reported to not have reached a stable usage design
+(see \tttt{:h terminal}.
+I find that it works exceptionaly well and have used it to run
+scripts in an IPython shell, compile latex files, and even open PDFs and images.
+Vim browsing of windows and text manipulation is well developed,
+so the terminal mode enables a very convenient integration,
+more traditionally achieved through Byobu/Tmux/Screen terminal
+multiplexers.
+Most useful commands in terminal-job mode include:
+\begin{itemize}
+  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the normal
+    and command-line modes.
+  \item \tttt{<C-W>\_"} to paste a register.
+  \item \tttt{i} for entering the terminal-job mode from the normal mode.
+\end{itemize}
 
-\subsection{Navigation + insertion}\label{navIn}
-Commands that bridge from Navigation to Insertion:
-csrCSR, iws, etc
-Default, line and block visual selection:
-  and AiIcCdD
+It is useful to define the same mappings for nativating splits and tabs
+for both normal and terminal-job modes, as in~\cite{vimrc}.
 
-\subsection{Netrw}
-Most important help files on Netrw
-No insert mode.
+\subsubsection{Normal $\rightarrow$ insertion}\label{navIn}
+Many commands bridge from Normal to Insertion modes:
+\tttt{csrCSR}, \tttt{iws}, etc.
+These make convenient the replacement of text and populates registers.
+The absence of a short command to insert one token is
+a known issue in Vim.
+Reasonable mappings to insert and append a token to and around
+other tokens are in~\cite{vimrc}.
+Vim commands are designed to couple operator and motion commands.
+There are many operator commands that take the editor from
+the normal mode to the insert mode, most of them
+favoring deletion or change, as detailed in \tttt{:h operator}.
+Motion commands are described in \tttt{:h motion}.
 
-Edit directory
-Sexplore
-Create and delete files and directories
-open files: on window, split, preview
-Change display of files
-Bookmarks
-Most important help files on Netrw
+\subsubsection{Netrw}\label{netrw}
+The standard interface of Vim for browsing file trees in Netrw.
+It starts when you open a directory, such as with \tttt{:e .<CR>}.
+It has solid support for editing remote files (such as over ssh or
+ftp) and handy e.g. mappings to open the files as splits and tabs (specially \tttt{pot}).
+Most useful commands in netrw include:
+\begin{itemize}
+  \item \tttt{d} and \tttt{\%} for creating directories and files.
+    \ttt{Delete} removes both.
+  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
+  \item \tttt{gb} and \tttt{uU} are used to load directories
+    while marked files might be copied, moved, edited, greped, tagged and migrated
+    to and from the argslist as in \tttt{:h netrw-mf}.
+\end{itemize}
+Last notes:
+there is no insert mode in netrw interface;
+the commands in \tttt{:h netrw-explore} are
+handy for opening the directory of the
+file being edited;
+further information in \tttt{:h netrw}.
 
-\subsection{Standard configuration files and directories and My .vim/vimrc}
+\subsubsection{Ex}
+This might be the least popular mode.
+One might use \tttt{q:,/,?} to have a window
+with colon or search command history to be edited normaly,
+and the chosen command can be run with \ttt{CR}.
+Vimscript was largerly based on the ex editor language~\cite{ex},
+and a more advanced user might use it for prototyping
+by defining mappings and settings and managing sessions and scripts
+propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
+In the default interface started with the \tttt{Q} command in normal mode,
+each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
+follow documentation from \tttt{:h Ex-mode} for further information.
+
+\subsection{Standard configuration files and directories and my .vim/vimrc}
+You can check the scripts Vim loads by using the debug script mentioned
+in Section~\ref{state}.
+By default, \tttt{~/.vimrc} and \tttt{~/.vim/vimrc} files are run by Vim at the beginning of the startup.
+One might edit the vimrc file with \tttt{:e \$MYVIMRC}
+and reload it with \tttt{:source \$MYVIMRC}, but
+I use the mappings in~\cite{vimrc} to encourage the constant enhancement of my settings.
+Any other file might be included to run at startup by
+adding a line \tttt{:source <afile.vim>} in vimrc.
+In fact, it is on vimrc that one usually specifies the plugins
+and plugin managers they use.
+Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
+or follow some patterns described on the next section to change the
+scripts and sequence of them to be loaded.
+
+\subsection{Plugins and packages}
+One can see the list of standard plugins
+with \tttt{:h standard-plugin-list} command.
+Any \tttt{plugins/**/*.vim} file inside a directory
+listed by \tttt{:se runtimepath} will be loaded
+(e.g. \tttt{.vim/plugin/something/ascript.vim}).
+There are various ways to automate the installation
+and enhnace the management of plugins.
+By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
+that downloads latest scripts from sourceforge as specified in \tttt{:h getscript-data},
+and the Vimball interface, which creates and loads a Vimball for a plugin
+with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
+that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
+The Vimball can be installed in a system by \tttt{:source <filename>.vba}
+or loading it at vim startup with \tttt{\$ vim <filename>.vba}.
+
+A Vim package is a directory that contain plugins.
+It should be located inside a \tttt{pack/} directory
+somewhere in the directories listed with \tttt{:se runtimepath}.
+The plugins found in \tttt{pack/<packName>/*/start/} are loaded
+at startup, the plugins found at
+\tttt{pack/<packName>/opt/**} are loaded with\\
+\tttt{:packdd <script\_or\_directory\_name>}.
+
+All directories vim looks for scripts are described
+in \tttt{:h vimfiles} and are basically set by
+\tttt{:se runtimepath} and conventions inside
+each path therein, such as to look for \tttt{vimrc} files
+and \tttt{plugin/} or \tttt{pack/} directories.
+Vim scripts can be loaded conditionaly,
+e.g. only if a function is used as in \tttt{:h autoload-functions}.
+Example of such are filetype plugins (enabled by in a \tttt{ftplugin/<filetype>.vim} file,
+e.g. inside a plugin directory).
+There is a number of plugin managers for Vim.
+Pathogen and Vundle seem to be the most popular,
+one because of its minimalism, the other because of
+advanced features, e.g. for searching and installing plugins with
+colon commands.
+
 \subsection{Spell and spelllang (en and pt\_br)}
-\subsection{Tabs, splits, buffers and namespaces}
-\& \% \$ and the following
-\subsection{Mappings and abbreviations}
+One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
+or \tttt{:se spelllang=pt\_br}
+and toggle spell checking with \tttt{:setl spell!}.
+These are used so often that one might use shortcuts as in~\cite{vimrc}.
+Currently, Vim will download the files for a specific language if
+not found in system.
+
+\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}
+In Vimscript,
+the colon commands (also Ex or command-line commands) are related through spaces,
+punctuations and keywords (see \texttt{:h script}).
+Scripting the Vim editor can also be accomplished using other languages,
+as well documented e.g. in \tttt{:h python}.
+Functions are defined through colon commands and are called
+inside colon commands e.g. \tttt{:call MyFunction()}
+or \tttt{:echo MyFunction(4)}.
+Notice that functions are not commands but might be binded to them
+through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
+Executing source files is very straightforward with \tttt{:so
+<filename>.vim},
+and one can always use the Ex mode for rapid scripting.
+
+At the same time, the \tttt{:term} and terminal-job mode make
+scripting other software more convenient, as output is
+proptly navigated and copied. The shell works well for generic use,
+such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
+See Section~\ref{terminal} for further directions about the terminal.
+
+\subsection{Color}\label{visual2}
+Newer versions of Vim support 24 bit true color (aka 16 million colors)
+in terminal Vim (not only in GVim).
+The terminal must support true color, and tests are available e.g. in
+\url{https://gist.github.com/XVilka/8346728}.
+Then Vim needs to be set to use true colors with
+\tttt{:set termguicolors}.
+If using 8 or 16 bit colors, Vim uses the color pallete from the
+terminal, if using true colors each color is defined directly.
+Settings for using true colors inside Byobu/Tmux involve tweakings
+and are available in~\cite{vimrc}.
+Good color schemes to use with true colors are Gruvbox~\cite{gruv}
+and Solarize~\cite{sol}.
+One might source syntax files at any time to change syntax, usually
+though linking tokens to syntax groups with
+\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
+and then relating the group to another group 
+\tttt{:highlight link <group\_name> <group\_name2>}
+or by specifying the group characteristics directly:
+\tttt{highlight <group\_name> guifg=\#ffffff}.
+If you change a syntax file, reloading a file with with \tttt{:e<CR>}
+updates the highlighting on the window with the corresponding filetype.
+A complete syntax highlighting support typically involves at least three files:
+\begin{itemize}
+  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
+    detected with e.g.
+    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
+  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
+    file type, such as: 
+    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
+    expandtab autoindent}.
+  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
+    tokens and highlighting groups; and highlighting group definitions.
+\end{itemize}
+
+This very scheme is implemented very straightforward in this
+plugin~\cite{tokipona}
+for highlighting text in the Toki-Pona language~\cite{tpLang}.
+Syntax highlighting plugins are as filetype plugins,
+but also have a \tttt{syntax/<filetype>.vim} file
+relating keywords to highlighting groups. 
+One might see every highlighting group, and their final visual results,
+with the command
+\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}.
+In my system, the \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
+directory, but \tttt{syntax/} files had to be moved to \tttt{$\sim$/.vim/syntax/}.
+
+\subsection{Fonts}
+The fonts are defined by your terminal or inside GVim.
+\ttt{C-+} and \ttt{C--} can be used to change font size.
+Some settings for fonts, such as boldface, might be set using
+the syntax highlighting facilities described in the previous section.
+
+\section{Conclusions and further work}\label{conc}
+This document seems reasonable as an overall reference of the Vim editor,
+at least for my usage and proficiency.
+Given the folkloric milestone of using Vim for 10 years,
+this article might serve as a benchmark for one to relate
+it's current use and understandings.
+As a pedagogical material, it seems to be unique in the emphasis
+on namespaces, understood as commands, variables, state-related lists, etc,
+especially in Section~\ref{namespaces},
+and the reference to the standard Vim documentation
+to achieve the DRY KISS style described in Section~\ref{intro}.
+
+\noindent {\bf Potential enhancements} to this document are:
+\begin{itemize}
+  \item The inclusion of facilities such as reading emails and connecting over ssh.
+    There is a working hack in~\cite{vimrc} for browsing over the WWW,
+    but such aspect of Vim usage might receive more attention
+    given that it is confortable to use the navigation and editing facilities
+    and the resulting integrated environment.
+  \item Updating of the information I can find about the issues discussed,
+    such as about status lines in Section~\ref{visual1}.
+  \item Include a discussion about Neovim.
+    I have never used it, but it seems to be reaching a considerable user base
+    and it might be feasable to give an account of Vim and Neovim
+    after some tests and researching the official and user base documentation.
+  \item Better cite documentation, plugins and Vim authors.
+    I prefered to keep the references inline through \tttt{:h} commands
+    and URLs, more in accordance with the style of Vim documentation,
+    but bibliographical items constitute a valuable asset for academic literature,
+    and some authors might find their work more pretiged if more
+    thoroughly cited.
+\end{itemize}
+
+\noindent {\bf Potential next steps} in using Vim:
+\begin{itemize}
+  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
+  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
+  \item Better integrate Python and Vimscript, especially for data visualization
+    and syntax highlighting management.
+    In accordance with the visualization issues described in
+    Sections~\ref{visual1} and~\ref{visual2}.
+  \item Make plugins for:
+    \begin{itemize}
+      \item listing all the mappings available in each mode and the typing combinations which are available
+        for new mappings.
+      \item Sessions, as described in Section~\ref{visual1}.
+      \item AA messages (shouts): to keep track, document and share of working sessions
+        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
+        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
+        relate AA sessions to other media, such as software repositories,
+        screencasts and images, interact with IRC channels and other social platforms.
+      \item Slick Vim: a collection of the settings, mappings and plugins I use.
+        Enhancements such as using \ttt{C-} commands also to browse tabs,
+        shortcuts to join a window into a tab,
+        and dummy minimal plugins as simplest, file type and syntax highlighting,
+        Some more elaborate tweaks should also be present, such
+        as breaking lines in sensible places for natural language texts
+        while respecting e.g. \tttt{:se textwidth}.
+      \item Dealing with .swo and .swp temporary files.
+        In summary, if the restored .swo file has the same content
+        and the file being opened,
+        the restoration phase can be omitted.
+        If the contents differ, Vim should open a tab with each file
+        in a vertical split and run \tttt{:windo :diffthis}.
+      \item Rendering images and equations.
+        These are useful for using Vim in presentations
+        or achieving a textual representation when it is mandatory,
+        such as to comply with the limitations of a platform (e.g. Vim editor). 
+        but also hold stylistic merits as ascii art is often
+        very appreciated.
+        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
+        and can directly render ascii charts from data using cues such are shape, position
+        and color..
+      \item Slide presentations. I've been using some automation for browsing
+        slides and opening figures and some Vim users asked for the settings and
+        commands. They are very elementary use of registers being executed over
+        arbitrary but consistent textual patterns.
+      \item Changing the color scheme an highlighting scheme incrementally
+        and selectively using the features described in Section~\ref{visual2}.
+    \end{itemize}
+\end{itemize}
+
+\subsection*{Acknowledgments}
+FAPESP, Vim developers and documentation maintainers,
+Vim user community.
+
+\appendix
+\section{Example of usage session}
+% open file, split, tab, find, replace char,
+% append to line, visual block, etc.
+\section{Key naming in Vim documentation}
+% CTRL, <C-X> <C-S-X>s 1250
+% Special chars..
+\section{My vimrc file and usage}
+In this Appendix is my vimrc file.
+Although it has comments, one should
+look for the options that (s)he does
+not understand, as a thorough explanation
+of the file is tedious and out of the scope
+of this document.
+
+Mostly, I toggle the status line with \texttt{\\\\B}
+and the tab line with \texttt{\\\\T}.
+Save and close windows with \texttt{\\w} and \texttt{\\q}.
+
+\section{Example of notations on mappings}\label{notes}
 \texttt{:h index} shows all the mappings.
+\subsection{Normal mode}
 Every letter and character in the keyboard is used.
@@
 
-\subsubsection{Insert mode}
+\subsection{Insert mode}
 All standard char keys are used for entering chars on text.
@@
 <C-J> and <C-M> are <CR>
-
-\subsection{Macros and registers for copy and paste}
-\subsection{History of commands}
-\subsection{Lists of Vim state variables}
-% of markers, jumps, registers, commands (hist)
-% :
-\subsection{Undo}
-\subsection{Scripting, Functions, Vimscript and Python}
-\subsection{Plugins}
-\subsubsection{Standard features}
-\subsubsection{Writting plugins}
-\subsubsection{Plugin systems: usage for using and writing plugins}
-\subsection{Colour}
-\subsubsection{Standard, 8 and 16 bits, and true color, Screen/Byobu}
-\subsubsection{Gruvebox, Solarize and other colourschemes}
-\subsection{Fonts}
-\subsubsection{Cools nd popular fonts}
-\subsubsection{How to set fonts in xterm, gnome-terminal and GVim.}
-\subsection{Verbosing, logs and possibilities of using it to study your own}
-\subsection{age (often used commands and typed sequences).}
-\subsection{Highlighting}
-\subsection{Bash and Vim commands}
-\subsection{Compiling, standard features and plugins}
-\subsection{Quickfix}
-\subsection{Persistence and vimrc}
-\subsubsection{Sessions}\label{sessions}
-% markers, undo, buffers, registers, bookmarked directories,
-% vimrc for settings
-\section{Issues and plugins}\label{issues}
-\subsection{Sessions}
-\subsection{AA messages}
-\subsection{Slick Vim}
-% tab line and status line and ruler
-
-\section{Final words and further work}
-
-Potential enhancements to this document are:
-\begin{itemize}
-  \item The inclusion of reading emails and connecting over ssh.
-\end{itemize}
-
-Potential next steps in using Vim:
-\begin{itemize}
-  \item Measure the performance text mining routines in Vim against those implemented in C or Python.
-  \item Make a functionalities to list all the mappings available in each mode and the typing combinations which are available.
-\end{itemize}
-
-\appendix
-\section{Example of usage session}
-% open file, split, tab, find, replace char,
-% append to line, visual block, etc.
-\section{Key naming in Vim documentation}
-% CTRL, <C-X> <C-S-X>s 1250
-% Special chars..
-\section{My vimrc file and usage}
-In this Appendix is my vimrc file.
-Although it has comments, one should
-look for the options that (s)he does
-not understand, as a thorough explanation
-of the file is tedious and out of the scope
-of this document.
-
-Mostly, I toggle the status line with \texttt{\\\\B}
-and the tab line with \texttt{\\\\T}.
-Save and close windows with \texttt{\\w} and \texttt{\\q}.
 
@@
 
+\section{My \tttt{:version}}
+
 % sessions, terminal for ipython3 and evince

### 2017-11-09 article/article.tex
finished first version of article

@@
 and the chosen command can be run with \ttt{CR}.
-Vimscript was largerly based on the ex editor language~\cite{ex},
+Vimscript was largerly based on the ex editor~\cite{ex},
 and a more advanced user might use it for prototyping
@@
 and are available in~\cite{vimrc}.
-Good color schemes to use with true colors are Gruvbox~\cite{gruv}
-and Solarize~\cite{sol}.
+Good color schemes to use with true colors are Gruvbox
+and Solarize.
 One might source syntax files at any time to change syntax, usually
@@
         and can directly render ascii charts from data using cues such are shape, position
-        and color..
+        and color.
+      \item Redirecting the commands that usually show the results in a 'more' interface,
+        which cannot be searched nor copied,
+        to a parsed and linked quickfix or location window.
+        The basic idea is to use \tttt{:redir} command to redirect the output
+        of such commands with \tttt{:se nomore}.
       \item Slide presentations. I've been using some automation for browsing
@@
 \subsection*{Acknowledgments}
-FAPESP, Vim developers and documentation maintainers,
+FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
 Vim user community.
@@
 \section{Example of usage session}
-% open file, split, tab, find, replace char,
-% append to line, visual block, etc.
-\section{Key naming in Vim documentation}
-% CTRL, <C-X> <C-S-X>s 1250
-% Special chars..
+I usually begin by opening a file or directory
+with \tttt{\$ vim <filename>}.
+The color scheme is alternated between
+blue and GruvBox with
+\tttt{:colo gruvbox} and \tttt{:colo blue}.
+I open an vertical split and then move
+the window to a new tab using
+\tttt{:vs} and \ttt{C-W\_T}.
+I then search tokens related to
+the enhancements I want to make or
+the knowledge I want to acquire.
+I go back to the previous tab with \tttt{gr}
+and make a global replace with
+\tttt{:\%s/<this>/<that/g}.
+On adding dots to sentences,
+I record in the \tttt{"q} register
+the sequence \tttt{jA.jj},
+using it as a macro 10 times by
+\tttt{10@q}.
+I move to the other tab with \tttt{gt}
+and open a terminal window with \tttt{:term}
+for compiling latex files.
+I start another terminal with \tttt{<C-W> :term}
+for opening the resulting PDFs with evince.
+If any new idea comes to mind and I have time,
+\tttt{\textbackslash\textbackslash s} opens 
+my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
+sources it.
+If there is e.g. code or notes in other projects I am working
+on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
+in Byobu/Tmux sessions and windows.
+
 \section{My vimrc file and usage}
@@
 
-Mostly, I toggle the status line with \texttt{\\\\B}
-and the tab line with \texttt{\\\\T}.
-Save and close windows with \texttt{\\w} and \texttt{\\q}.
+In using Vim with my vimrc file~\cite{vimrc},
+I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
+and the tab line with \texttt{\textbackslash\textbackslash T}.
+Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
+The mappings for transitioning through splits and tabs
+are also used constantly.
 
-\section{Example of notations on mappings}\label{notes}
-\texttt{:h index} shows all the mappings.
+\section{Example of notes on mappings}\label{notes}
+\texttt{:h index} shows all the default mappings
+while \tttt{:map} shows the user-defined mappings.
+By considering such information, one can make
+useful observations exemplified in this Appendix.
+
 \subsection{Normal mode}
 Every letter and character in the keyboard is used.
-In Normal mode: <TAB> is the same as <C-I>,
-<BS> and <C-H> is the same as h (<C-H> don't sowk here).
-<C-J> and <C-N> is the same as j.
-C-P is the same as k.
-Space is same as l
-CTRL-[ <Esc> are not used
-C-\ a-z reserved for extensions
-CTRL-\_ not used
+In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
+\ttt{BS} and \ttt{C-H} is the same as h.
+\ttt{C-J} and \ttt{C-N} is the same as j.
+\ttt{C-P} is the same as k.
+Space is same as l.
+\ttt{C-[} and \ttt{Esc} are not used
+\tttt{<C-\textbackslash> a-z} is reserved for extensions
+\ttt{C-\_} not used.
+\tttt{+} is the same as \ttt{CR} and they are both not very useful.
+Del is same as x.
 
-learn how to use filters ! and !!
-% .%#: registers
-\url{https://github.com/vim/vim/blob/master/runtime/doc/index.txt}.
-
-my usage of <C-(HJKL)> for jumping through windows.
-<C-L> is lost though, and require :redraw or :e
-
-+ same as <CR> which is not that useful.
-
-Many ][ combinations are not used, e.g.
-with abhjklfg
-\_ might be used as a more powerful \^, leaving it free for mappings.
+Many \tttt{][} combinations are not used, e.g.
+with \tttt{abhjklfg}.
+The \tttt{\_} command might be used as a more powerful \tttt{\^},
+leaving it free for mappings.
 
@@
 
-Del is same as x
+The \ttt{C-(HJKL)} commands are redundant,
+with the exception of \ttt{C-L} which redraws the screen,
+so it is a reasonable choice to use them to move focus
+of the editor to splits in the \tttt{hjkl} directions.
 
-text objects: redundancy between a/i(),{},[],<>
-
-many entries available through the g commands
-z for folds, spell checking and some movements (mainly when wrap is set).
+Many key combinations are available for new mappings through the \tttt{g}
+and \tttt{z}, and \tttt{v} commands.
+They have typical uses, e.g. \tttt{z} for folds,
+spell checking and some movements (mainly when wrap is set).
 
@@
 All standard char keys are used for entering chars on text.
-<C-G>jk can be achieved by <C-O>jk which is more powerful
+\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
 in moving through multiple lines.
-j-[ same as ESC
-		CTRL-\ a - z	reserved for extensions
-		CTRL-\ others	not used
-
-<C-J> and <C-M> are <CR>
-
-\section{Overview of my usage of Vim}
-% move to usage appendix TTM
-in some of the basic topics, such as in using numerous terminals 
-(it works pretty well!) and using sessions (Section~\ref{sessions}).
-These make Vim very convenient e.g. for users that edit multiple files,
-run scripts and compile programs and latex files.
-I used for years multiple terminal abs and windows,
-and sometimes a byobu/screen terminal multiplexer.
-Nowadays I am using standard Vim windows as terminals
-such as described in Section~\ref{terminals}.
+\ttt{C-[} is the same as \ttt{ESC}.
+\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+other combinations with \ttt{C-\textbackslash} are not used.
 
 \section{My \tttt{:version}}
+I should keep the output of \tttt{:version} executed on the system
+in which I wrote this document in this link:
+\url{}.
+It was compiled with this Makefile~\cite{makefile}
+in a 16.04 LTS Ubuntu Linux,
+and is tagged as 8.0, Included Patches 1-1173.
 
-% sessions, terminal for ipython3 and evince
-% alternative with byobu or many terminals
-% historical usage, other editors,
-% most often edited file types
-
+\begin{thebibliography}{99}
+\fontsize{11}{0}\selectfont
+\bibitem{anPh}
+	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
+	Available at \url{https://youtu.be/oeOKYc3-nbM}
+\bibitem{anPh2}
+	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
+\bibitem{ouroWiki}
+  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
+\bibitem{vimrc}
+	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
+\bibitem{ex}
+  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
+\bibitem{tokipona}
+	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
+\bibitem{tpLang}
+	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
+    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
+\bibitem{aa1}
+	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G.
+(2013).
+AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology.
+RESI.
+From \url{https://arxiv.org/abs/1604.08255}
+\bibitem{aa2}
+	Fabbri, R. (2017).
+The Algorithmic-Autoregulation (AA) Methodology and Software:
+a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
+\bibitem{makefile}
+	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
+\end{thebibliography}
 \end{document}

### 2017-11-10 article/article.tex
detalhe

@@
 and reload it with \tttt{:source \$MYVIMRC}, but
-I use the mappings in~\cite{vimrc} to encourage the constant enhancement of my settings.
+I use the mappings in~\cite{vimrc} to encourage a continuous enhancement of my settings.
 Any other file might be included to run at startup by

### 2017-11-10 article/article.tex
made an appendix on key notation

@@
 	\texttt{<#1>}}
-\newcommand{\tttt}[1] {
-	\texttt{#1}}
+\newcommand{\tttt}[1]{\texttt{#1}}
 
@@
 there is an anthropological component
-which is evdent in the use of the first person
+which is evident in the use of the first person
 in sentences.
@@
 that carry values or triggers procedures.
+One should see Appendix~\ref{keys}
+to understand the notation vim uses.
 
@@
 		Some keys expect a second key,
-		and have available possibilities,
+    and have combinations not used (available for new mappings),
 		specially the z and g.
@@
 	\item In the other modes, the sequences available for mappings are more obvious and abundant.
-		One should look at \ttt{:h index} to know about all the standard mappings
+		One should look at \tttt{:h index} to know about all the standard mappings
 		and use \tttt{:map} to list the user-defined mappings.
@@
 
-\ttt{C-\textbackslash} is always reserved for extensions,
+\ttt{C-\textbackslash} is often reserved for extensions,
 which makes it a safe namespace to use (while there are no
@@
 \begin{itemize}
-	\item Environment variables: names start with \$ an hold system
+	\item Environment variables: names start with \$ and hold system
 		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
-	\item Option variables: names start with a \& and are meant to control the behavior of the editor.
+	\item Option variables: names start with \& and are meant to control the behavior of the editor.
 		One might change a value through set or let, e.g.
@@
 For all the Vimscript capabilities, including loops, conditionals,
-and builting functions, refer to \ttt{:h vim-script}.
+and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
 Classes are possible only in rudimentary forms, e.g. through dictionaries,
 but the language is otherwise overall quite powerful,
-specially in dealing with text and editor behavior, as expected
+specially in dealing with text and editor behavior, as expected.
 
 \subsubsection{State lists}\label{state}
-Vim keeps a number of lists which expresses the state of the editor.
-These are examples of useful lists:
+Vim keeps a number of useful lists which expresses the state of the editor:
 \begin{itemize}
@@
     by recording a typing sequence with the \tttt{q} normal command.
+    Vim keeps only the last edition, in register \tttt{".}.
+    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
+    A hack to keep the latest deletions and copies in the standard register
+    might follow the same pattern, but use another \tttt{:h event} and
+    monitor register \tttt{""} (maybe also \tttt{"0}).
 	\item An undo list can be accessed with \tttt{:changes}.
@@
 		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
-		Uppercase letters are cross buffer.
+		Uppercase letters are cross buffers.
 	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
@@
 		one might perform actions on each file in argument list using \tttt{:argdo}.
-		All files in argument list is also in the buffers list.
+		All files in argument list are also in the buffers list.
     For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
@@
 it displays the tokens and related type of meanings when run
-inside e.g. a \tttt{.vim} or help file.
+inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
 
@@
 in a reasonable directory and makes it easier to remember and tweak
-then the standard commands to deal with sessions.
+the standard commands to deal with sessions.
 More information in \tttt{:h sessions}.
@@
 editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
-which I make available across terminal and normal modes.
+which I make available across terminal-job and normal modes.
 But I've been thinking on using \ttt{C-} commands also
@@
 
-A good strategy I find is to have selective visual cues of the state to make persistent or hide. A mapping to toggle each of them.
+A good strategy I find is to have selective visual cues of the state to make persistent or hide
+and a mapping to toggle each of them.
 Now I toggle byobu/screen/tmux bar with \ttt{F5},
@@
   \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
-  \item \texttt{(w,b,e)} to move to the next, previous and end-of-next word. There motions iterate over sequences of characters separated
+  \item \texttt{(w,b,e)} to move to the next, previous and next end-of-word.
+    There are motions to iterate over sequences of characters separated
     by special characters (e.g. punctuation and parenthesis) as
@@
     If X is lowercase, the mark is local to the buffer (the file),
-    if it is uppercase or numeric, it is global to the Vim session.
+    if it is uppercase or numeric, it is global to the Vim session
+    (cross buffers).
 \end{itemize}
@@
 \begin{itemize}
-  \item \texttt{Ctrl+o} to execute one and only command in normal mode.
+  \item \ttt{C-O} to execute one and only command in normal mode.
 	  This enter a secondary mode (see Section~\ref{modes})
@@
   \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
-    the line, the previous word.
-  \item \ttt{C-N,P} to find next, pervious keywords that match the prefix at hand..
+    the line or to the previous word.
+  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand..
   \item \ttt{C-X} commands for scrolling the window with multiple
@@
 This mode is reported to not have reached a stable usage design
-(see \tttt{:h terminal}.
+(see \tttt{:h terminal}).
 I find that it works exceptionaly well and have used it to run
@@
 These make convenient the replacement of text and populates registers.
-The absence of a short command to insert one token is
+The absence of a short command to insert one char is
 a known issue in Vim.
-Reasonable mappings to insert and append a token to and around
-other tokens are in~\cite{vimrc}.
+Reasonable mappings to insert and append a char to and around
+other char are in~\cite{vimrc}.
 Vim commands are designed to couple operator and motion commands.
@@
 \subsubsection{Netrw}\label{netrw}
-The standard interface of Vim for browsing file trees in Netrw.
+The standard interface of Vim for browsing file trees is Netrw.
 It starts when you open a directory, such as with \tttt{:e .<CR>}.
@@
   \item \tttt{gb} and \tttt{uU} are used to load directories
-    while marked files might be copied, moved, edited, greped, tagged and migrated
+    while marked files might be copied, moved, edited, grep-ed, tagged and migrated
     to and from the argslist as in \tttt{:h netrw-mf}.
@@
 the commands in \tttt{:h netrw-explore} are
-handy for opening the directory of the
+convenient for opening the directory of the
 file being edited;
@@
 \subsubsection{Ex}
-This might be the least popular mode.
-One might use \tttt{q:,/,?} to have a window
+One might use the normal commands \tttt{q(:,/,?)} to have a window
 with colon or search command history to be edited normaly,
@@
 in Section~\ref{state}.
-By default, \tttt{~/.vimrc} and \tttt{~/.vim/vimrc} files are run by Vim at the beginning of the startup.
+By default, \tttt{$\sim$/.vimrc} and \tttt{$\sim$/.vim/vimrc} files are run by Vim at the beginning of the startup.
 One might edit the vimrc file with \tttt{:e \$MYVIMRC}
@@
 with \tttt{:h standard-plugin-list} command.
-Any \tttt{plugins/**/*.vim} file inside a directory
+Any \tttt{plugin/**/*.vim} file inside a directory
 listed by \tttt{:se runtimepath} will be loaded
@@
 and \tttt{plugin/} or \tttt{pack/} directories.
-Vim scripts can be loaded conditionaly,
+Vim scripts can be loaded conditionally,
 e.g. only if a function is used as in \tttt{:h autoload-functions}.
@@
 
-\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}
+\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}\label{script}
 In Vimscript,
 the colon commands (also Ex or command-line commands) are related through spaces,
-punctuations and keywords (see \texttt{:h script}).
+punctuation and keywords (see \texttt{:h script}).
 Scripting the Vim editor can also be accomplished using other languages,
-as well documented e.g. in \tttt{:h python}.
+as well documented e.g. \tttt{:h python}.
 Functions are defined through colon commands and are called
@@
 or \tttt{:echo MyFunction(4)}.
-Notice that functions are not commands but might be binded to them
+Notice that functions are not commands but might be bind to them
 through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
@@
 scripting other software more convenient, as output is
-proptly navigated and copied. The shell works well for generic use,
+promptly navigated and copied. The shell works well for generic use,
 such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
@@
 Space is same as l.
-\ttt{C-[} and \ttt{Esc} are not used
-\tttt{<C-\textbackslash> a-z} is reserved for extensions
+\ttt{C-[} and \ttt{Esc} are not used,
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
 \ttt{C-\_} not used.
@@
 
+\section{Key notation and meaning for Vim}
+The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
+mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
+(or \tttt{Meta+x}).
+\ttt{A-X} is the same as \ttt{M-X} and refer to
+Alt or Meta keys.
+\ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x}  and \ttt{C-x} are the same,
+i.e. Vim does not distinguish lower and uppercase 
+letters in \ttt{C-} commands.
+The \ttt{M-x} and \ttt{M-X} (or \ttt{M-S-x} or \ttt{M-S-X})
+are different commands,
+and they are not used by Vim's builtin keys.
+Such \ttt{M-} key combinations are potentially conflicting
+with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
+but are otherwise a large and safe set of combinations
+for one to use.
+
+In summary,
+one might choose for mappings to overwrite defaults key combinations,
+use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
+normal commands, \ttt{C-} and \ttt{S-} commands, or any \ttt{M-} and
+\ttt{M-S-} command), or
+create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
+More information about key notation in \tttt{:h key-notation}.
+Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings
+in \tttt{:map}.
+
 \begin{thebibliography}{99}

### 2017-11-11 article/article.tex
started article

--- 2017-11-10 article/article.tex
@@
 \usepackage[utf8]{inputenc}
-\newcommand{\ttt}[1] {
-	\texttt{<#1>}}
-\newcommand{\tttt}[1] {
-	\texttt{#1}}
+\usepackage{tipa}
+% \usepackage{fontspec}
+% \usepackage[mathletters]{ucs}
+% \newcommand{\ttt}[1] {
+% 	\texttt{<#1>}}
+% \newcommand{\tttt}[1]{\texttt{#1}}
 
 \begin{document}
-\title{An anthropological account of the Vim editor:\\
-ouroboros, features and tweaks after 10 years of usage}
+\title{The Toki Pona Language:
+an overview and some hacks}
 % Anthropological computation
@@
 \begin{abstract}
-The Vim editor is very rich in capabilities
-and thus complex.
-This article is a description of the Vim text editor
-and a set of considerations on its usage and design.
-It is the result of more than ten years of experience
-with Vim for writing and editing various types of documents,
-e.g.:
-Python, C++, JavaScript, ChucK, programs;
-and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
-binary files.
-It is said that it takes about ten years to master (or start mastering)
-this text editor, and I find that other experienced users
-have a different view of Vim and that they use a different
-set of features.
-Therefore, this document exposes my insights in order
-to confront my usage with that of other Vim users.
-Another goal is
-to make available a reference document with which new users
-can grasp a sound overview by reading it and the discussions that
-it might generate.
-Also, it should be useful for users of any degree of experience,
-including me, as a compendium of commands, namespaces and tweaks.
-Upon feedback, and maturing of my Vim usage,
-this document might be enhanced or receive additional
-material.
+
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Vim is a very complex editor,
-considered by the Linux community to be
-matched only by Emacs.
-They both are the standard advanced text editors
-of the free software and open source communities
-and have been developed for decades.
-This document describes the Vim editor
-and proposes a set of enhancements of the user
-experience through simple tweaks.
-The editor has a very mature documentation
-and the contents herein presented is a
-report on the overall understandings I
-have of Vim after a little bit more
-of ten years using it.
-The purposes of this document are:
+Toki Pona is a minimalist conlang (constructed language)
+with only 126 words.
+% definition, historic, status (groups, speakers, etc).
+
+\subsection{Resources on the web}
+http://tokipona.net
+\subsection{Historical note}
+
+\subsection{Natural and constructed languages}
+% conlangs, natural langs, programming langs, Lojban as an
+% intermediary
+\section{Overview of the language}\label{basics}
+\subsection{Phonology}
+Words in Toki Pona are written using only 14 letters:
 \begin{itemize}
-  \item to help new users in grasping Vim essentials
-  and convenient practices.
-  \item To attain a sound overall description of the editor.
-  \item To record the view of the editor that
-  a user (me) has after $10$ years of usage.
-  \item To confront my usage with that of other experienced
-  users. This is helpful for me, but also for the other users
-  as they might benefit from this content and from discussions
-  that might arise from it.
-  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
-\end{itemize}
-
-Advanced users might just skim throughj
-Section~\ref{basics} and consider more carefully
-Section~\ref{namespaces}.
-The concluding remarks and proposed enhancements
-might also deserve some attention.
-
-I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
-I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
-Eventually, I edited database files and other types os files.
-With Vim, I mostly write software (web and scientific),
-music, poems and short stories.
-It is very useful because:
+  \item Vowels a (open), e (mid front), o (mid back), i (close front),
+    u (close back).
+  \item Consonants j, k, l, m, n, p, s, t, w:
     \begin{itemize}
-  \item it is meant to be a plain text (e.g. ascii, utf8) editor
-  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
-  \item It has a powerful architecture and set of commands.
-  \item It is highly configurable and most often the users
-  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
-\end{itemize}
-
-The standard capabilities of the editor
-should become clear in Section~\ref{basics}.
-Vim is constantly evolving and there are many plugins,
-some of them very popular for both general and specific
-users.
-Accordingly, there are many possible enhancements,
-and Section~\ref{conc} report the most prominent of them
-for me and potential workarounds made available as plugins.
-
-This document is written is a DRY KISS
-(Don't Repeat Yourself, Keep It Simple Stupid) style.
-Complex is to master the use of Vim
-and one finds sound
-references in help files and a nice vimrc.
-Therefore the following content should be kept
-as uncomplicated and original as possible.
-Also, because of Vim's complexity and entailed
-bond of this document to my usage,
-there is an anthropological component
-which is evdent in the use of the first person
-in sentences.
-This can be understood as anthropological computation
-or physics~\cite{anPh,anPh2} and observed to help
-in the technological groundwork of the civil society.
-
-% Vi and Vim
-\subsection{Historical note}
-Vim was first released publicly in 1991.
-It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
-Vim's development is coordinated (and performed) since the beginning
-mainly by Bram Moolenaar.
-Today, current bleeding-edge version is 8.0.1257.
-I found no explicit stable, alpha or beta versions.
-I found no scientific article on Vim 
-(this might be the first one), although there are
-books, software and third-party documentation on the web.
-
-\section{Basics}\label{basics}
-Vim's interface is text-based.
-In the GUI mode (gVim),
-there are convenient menus and toolbars
-but all functionalities are still available though
-the command line mode.
-Vimscript is the internal language of Vim,
-and is often used for scripting by users
-although other languages might be used 
-(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
-Each line of a Vimscript is a command on the
-command-line mode.
-This section can be thought as a tutorial
-that focuses on the namespaces, i.e. sets of tokens
-that carry values or triggers procedures.
-
-\subsection{The bare minimum}\label{minimum}
-You open a file with Vim by executing
-the command: \texttt{vim <filename>}.
-Inside Vim, you start in the normal
-mode, and might want to move around using
-\texttt{h-j-k-l} for left-down-up-right.
-To insert characters, move your
-cursor to the desired location an press i,
-which puts Vim in the insert mode.
-Go back to normal mode by pressing
-\texttt{<ESC>} or \texttt{<C-C>}.
-You save the file by typing \texttt{:w<CR>},
-and exit Vim by typing \texttt{:q<CR>}.
-You can save and quit with \tttt{:wq<CR>}
-or \tttt{:x<CR>} or \tttt{ZZ}.
-
-\subsection{Vim help}
-You can find help for vim in various places.
-The standard resource is the Vim help files.
-They are accessed by typing \texttt{:h <anything><CR>}
-in normal mode.
-Examples of such \texttt{<anything>} are:
-color, navigation, :vs, vimtutor.
-Type 
-\texttt{:h usr\_toc<CR>}
-to access the official User Manual,
-which is considerably lengthy and complex
-and is usually not read by users for a few years.
-In learning Vim, one
-might want to run the \texttt{vimtutor} command
-(outside Vim) to start the Vim Tutor.
-
-There are good resources on the Web for learning
-and tweaking Vim:
-\begin{itemize}
-  \item ``Vim Adventures'' is an online RPG game for practicing
-  and memorizing Vim commands while having fun.
-  This game is quite famous among Vim users.
-  \item There are official and semi-official Vim sites e.g.:
-  \url{www.vim.org}, \url{https://www.vi-improved.org} and
-  \url{http://vim.wikia.com/}.
-  \item Many hacks, understandings and general issues
-  (e.g. how to make such a move) are asked and answered
-  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
-  One often finds these links through a search engine.
-  \item You can find many videos about Vim.
-  One traditional site is \url{http://vimcasts.org},
-  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
-\end{itemize}
-
-% any other commands? files?
-\subsection{Namespaces}\label{namespaces}
-Vim is a text editor ouroboros~\cite{ouroWiki} because
-text and writing alters text and writing.
-You have namespaces where tokens have scalar
-or complex values.
-
-\subsubsection{Commands and mappings}
-There are commands, typing sequences which trigger automated actions,
-for each mode:
-\begin{itemize}
-	\item in Normal mode all keys are mapped to commands.
-		There is redundancy and additional commands
-		using Ctrl and Shift keys.
-		Some keys expect a second key,
-		and have available possibilities,
-		specially the z and g.
-    See Section~\ref{normal} and Appendix~\ref{notes}
-    for more insights into the commands available in the
-    normal mode.
-	\item In the other modes, the sequences available for mappings are more obvious and abundant.
-		One should look at \ttt{:h index} to know about all the standard mappings
-		and use \tttt{:map} to list the user-defined mappings.
-\end{itemize}
-
-\ttt{C-\textbackslash} is always reserved for extensions,
-which makes it a safe namespace to use (while there are no
-such extensions).
-
-A colon command can be written as a string
-and executed by the \tttt{:execute} colon command.
-E.g. \tttt{:execute 'vs afile.txt'}.
-As there are colon commands that execute commands in other
-modes, e.g. \tttt{:normal ?\^def },
-the \tttt{:execute} is a way to build commands in any mode,
-e.g.
-\tttt{:execute 'normal' tabn 'gt'}.
-
-\subsubsection{Variables}
-There are some types of variables in Vim:
-\begin{itemize}
-	\item Environment variables: names start with \$ an hold system
-		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
-	\item Option variables: names start with a \& and are meant to control the behavior of the editor.
-		One might change a value through set or let, e.g.
-		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
-	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
-	\item Internal variables are created with let and preferably have a prefix:
-	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
-		are local to the buffer, window, tab page, function, and
-		sourced Vim file, respectively.
- \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
-		\tttt{a:} is for function arguments.
-		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
-		More about internal variables in \tttt{:h internal-variables}.
-	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
-\end{itemize}
-You can echo any of such variables or use in expressions.
-Notice that you will only be able to echo a \tttt{b:} variable inside
-the buffer where it is defined.
-For all the Vimscript capabilities, including loops, conditionals,
-and builting functions, refer to \ttt{:h vim-script}.
-Classes are possible only in rudimentary forms, e.g. through dictionaries,
-but the language is otherwise overall quite powerful,
-specially in dealing with text and editor behavior, as expected
-
-\subsubsection{State lists}\label{state}
-Vim keeps a number of lists which expresses the state of the editor.
-These are examples of useful lists:
-\begin{itemize}
-	\item The entered commands are accessed through \tttt{:hist a}. 
-    The tokens a / s : can be used for specific types of commands, such
-		as search and colon commands.
-	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
-  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
-	\item Tabs list can be reached through \tttt{:tabs}.
-		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
-	\item Jumps are available through \tttt{:jumps}.
-		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
-  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
-    They also keep track or your copy, edition and deletion and are promptly defined
-    by recording a typing sequence with the \tttt{q} normal command.
-	\item An undo list can be accessed with \tttt{:changes}.
-	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
-	\item The markers defined are listed with the \tttt{:marks} command.
-		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
-		Uppercase letters are cross buffer.
-	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
-		and variations, such as \tttt{:grep}.
-		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
-		to open the Quickfix window, where the lines of occurrence are in sequence
-		and one can \ttt{CR} one of them to have the cursor in the main window active at
-		the first character of the match.
-		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
-		to use the location-list window instead of the Quickfix, which
-		is very similar, but one per window instead of one per buffer.
-		More information in \tttt{:h quickfix}.
-	\item A tags list have to be made so one can use tags.
-		Most often one will generate the tags list using the exuberant ctags,
-		which supports dozens of languages.
-		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
-		and then using \ttt{C-]} to
-		go to the position of tag under cursor, and \tttt{:tabe tags} to open
-		the tags file.
-	\item The argument list holds a list of files to be edited or browsed. 
-		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
-	or using commands (e.g. \tttt{:ar ./*}).
-		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
-		one might perform actions on each file in argument list using \tttt{:argdo}.
-		All files in argument list is also in the buffers list.
-    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
-\end{itemize}
-
-A file with information about the state of the editor
-can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
-and this script might be consulted because it has
-a collection of commands to access various settings of Vim.
-Another good list of commands to know about Vim's state is kept on
-\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
-I would specially highlight the \tttt{:syntax} command because
-it displays the tokens and related type of meanings when run
-inside e.g. a \tttt{.vim} or help file.
-
-\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
-
-You can keep track of the editor state though commands and persistent visual
-cues, specially the tabs bar, the status line, and the line
-reserved for the command-line.
-For state persistence, one might keep an undo file for each file as
-in~\cite{vimrc}.
-Sessions are easy to manage, enabling one to save and load the
-editor's state, with the opened windows, tabs, buffers, etc.
-I use the mappings in~\cite{vimrc} because they keep the sessions
-in a reasonable directory and makes it easier to remember and tweak
-then the standard commands to deal with sessions.
-More information in \tttt{:h sessions}.
-One might use \tttt{:h views} to keep the state of one window,
-but sessions keep all the states from all windows.
-This entails a strategy to deal with Vim that is similar to
-the use of Byobu/Tmux/Screen.
-The main limitation I found to this approach is that
-Vim is not keeping track of the terminals opened.
-If you open a terminal inside Vim with the \tttt{:term}
-command, you will save the session as usual, but when loading
-you get dummy empty windows for them and an error message.
-Screen and Tmux are the most popular terminal multiplexers.
-Byobu, built on top of them, has awesome keyboard shortcuts
-for managing sessions, windows and splits of terminals.
-Byoby/Screen/Tmux keep the state for future load.
-
-Because browsing the interface in Vim is fast,
-and it favors copy and paste of text,
-I tend to keeep a tab with some terminals:
-one with an IPython shell, another two for compiling latex
-and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
-One enters normal mode in terminal with \tttt{C-W N}
-and it is very corfortable to copy and browse the bash history of the
-terminal.
-I found the mappings on~\cite{vimrc} very helpful for directing
-editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
-which I make available across terminal and normal modes.
-But I've been thinking on using \ttt{C-} commands also
-for tabs (not only for splits).
-
-A good strategy I find is to have selective visual cues of the state to make persistent or hide. A mapping to toggle each of them.
-Now I toggle byobu/screen/tmux bar with \ttt{F5},
-status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
-I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
-Numbering is always there.
-I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
-Instead of keeping the status bar, I use \ttt{C-g} to know about the
-file and \ttt{gC-g} to know more and rarily.
-It seems not possible to remove the statusbar between horizontal splits.
-After asking around and experimenting, I realized that it seems
-reasonable to keep at least one line dividing the windows, so
-if it comes to it, I just \tttt{set statusline=-}.
-Unfortunatelly, as far as I could dig, one will need to enter
-Vim code to enable a horizontal split without losing a line.
-My ideal of this feature would be to have a visual cue of the
-first and/or last line of the windows in the lines-number
-column, or complete the spaces and empty chars with \$\$\$\$ or so.
-
-Autocommands are the standard way to define event-trigered routines in Vim.
-These are often related to particular file types, but are also often
-in defining the automated behavior of Vim.
-In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
-in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
-
-\subsection{Using Vim's modes}\label{modes}
-Vim has some basic and fully implemented modes of usage:
-\begin{itemize}
-  \item Normal mode: used for changing
-  the position of the cursor or the text displayed
-  at the window.
-  A core goal of the normal mode it to allow fast
-  navigation of the document while allowing
-  the typist to maintain its fingers on the home row
-  (i.e. on the center of the keyboard).
-  The mode is also used for manupulating text
-  (e.g. copy, paste, delete, change case) and
-    changing to other modes.
-  \item Insert mode: for inserting text.
-  \item Command-line mode: for entering Ex commands.
-  \item Ex mode: similar to command-line mode,
-  but more specialized for running various Ex commands.
-  \item Visual mode: for making, manipulating and navigating
-  selections of texts.
-  \item Select mode: similar to visual mode but
-  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
-\end{itemize}
-
-There is another basic mode, but it is not fully implemented:
-the Terminal-Job mode.
-There are seven additional modes which are mostly subordinate 
-to the basic modes and will be described when convenient.
-The manual page for Vim modes can be accessed by typing
-\texttt{:h vim-mode}.
-Some of the modes are now further considered for
-an overview of the Vim usage possibilites.
-
-\subsubsection{Normal mode}\label{normal}
-Sometimes also called navigation or command mode,
-the normal mode is most powerful for
-navigating, manipulating texts and changing to other modes.
-
-The simplest of these three is changing to other modes:
-type any of these letters to change to insert mode:
-\texttt{iIaAoOsScC}. More on the transition between
-normal and insert mode on Section~\ref{navIn}.
-Type any of these characters to change to command-line mode:
-\texttt{:/?}.
-Type \texttt{Q} to enter Ex mode.
-Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
-
-For most basic and naive navigation, one should check Section~\ref{minimum}.
-Most often, one uses:
-\begin{itemize}
-  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
-and whole page down and up, although these commands might
-be set to scroll a different number of lines.
-  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
-  \item \texttt{(w,b,e)} to move to the next, previous and end-of-next word. There motions iterate over sequences of characters separated
-    by special characters (e.g. punctuation and parenthesis) as
-    specified by the output of \texttt{:se iskeyword}.
-    To iterate over space-separated tokens, use \texttt{W,B,E}).
-    To move to the end of last word, one might use \texttt{be}
-    or \texttt{ge}.
-The \texttt{),(} commands iterate through sentences,
-		\tttt{\},\{} through text blocks separated by empty lines.
-  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
-	  \tttt{;} and \tttt{,} for next and previous found character.
-  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
-  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
-    jump list.
-  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
-    the first non-blank character of the line.
-    A mark is registered by the user in any cursor position
-    by typing \texttt{mX}, where X is any letter.
-    If X is lowercase, the mark is local to the buffer (the file),
-    if it is uppercase or numeric, it is global to the Vim session.
-\end{itemize}
-
-There are many more facilities to navigate Vim
-as explained in \texttt{:h navigation}.
-
-For changing the text, usual commands include:
-\begin{itemize}
-  \item \texttt{d\{motion\}} to delete the characters
-    involved in the motion command.
-  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
-  \item \texttt{x,X} to delete the character under or before the cursor.
-  \item \texttt{$\sim$} to swap the case of a character.
-  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
-\end{itemize}
-
-There are way more commands to change the texts.
-Some of them are discussed in Section~\ref{navIn}
-because they involve a transition to the insertion mode.
-A thorough consideration of the commands in the normal mode
-is found by executing \texttt{:h navigation} and 
-\texttt{:h change.txt}.
-
-\subsubsection{Insertion}
-Once in the insertion mode, the character keys
-input the characters at the cursor position at the current buffer.
-One can exit insert mode by pressing \texttt{<Esc>},
-and Vim will be put in normal mode.
-Most useful commands in insert mode include:
-\begin{itemize}
-  \item \texttt{Ctrl+o} to execute one and only command in normal mode.
-	  This enter a secondary mode (see Section~\ref{modes})
-  \item \ttt{C-R} to paste a register (a variable starting with
-    '@', defined, copied or recorded through a \tttt{q} command in
-    normal mode and as covered in Section~\ref{state}).
-  \item \ttt{C-T} to indent current line.
-  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
-    the line, the previous word.
-  \item \ttt{C-N,P} to find next, pervious keywords that match the prefix at hand..
-  \item \ttt{C-X} commands for scrolling the window with multiple
-    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
-\end{itemize}
-
-\subsubsection{Command-line mode}
-This mode is dedicated for writing colon, search and filter commands,
-entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
-Most useful commands in this mode include:
-\begin{itemize}
-  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
-    end of the line.
-  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
-    until the cursor.
-  \item \ttt{C-R} to paste a register as in insert mode.
-\end{itemize}
-
-\subsubsection{Terminal-Job mode}\label{terminal}
-This mode is reported to not have reached a stable usage design
-(see \tttt{:h terminal}.
-I find that it works exceptionaly well and have used it to run
-scripts in an IPython shell, compile latex files, and even open PDFs and images.
-Vim browsing of windows and text manipulation is well developed,
-so the terminal mode enables a very convenient integration,
-more traditionally achieved through Byobu/Tmux/Screen terminal
-multiplexers.
-Most useful commands in terminal-job mode include:
-\begin{itemize}
-  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the normal
-    and command-line modes.
-  \item \tttt{<C-W>\_"} to paste a register.
-  \item \tttt{i} for entering the terminal-job mode from the normal mode.
-\end{itemize}
-
-It is useful to define the same mappings for nativating splits and tabs
-for both normal and terminal-job modes, as in~\cite{vimrc}.
-
-\subsubsection{Normal $\rightarrow$ insertion}\label{navIn}
-Many commands bridge from Normal to Insertion modes:
-\tttt{csrCSR}, \tttt{iws}, etc.
-These make convenient the replacement of text and populates registers.
-The absence of a short command to insert one token is
-a known issue in Vim.
-Reasonable mappings to insert and append a token to and around
-other tokens are in~\cite{vimrc}.
-Vim commands are designed to couple operator and motion commands.
-There are many operator commands that take the editor from
-the normal mode to the insert mode, most of them
-favoring deletion or change, as detailed in \tttt{:h operator}.
-Motion commands are described in \tttt{:h motion}.
-
-\subsubsection{Netrw}\label{netrw}
-The standard interface of Vim for browsing file trees in Netrw.
-It starts when you open a directory, such as with \tttt{:e .<CR>}.
-It has solid support for editing remote files (such as over ssh or
-ftp) and handy e.g. mappings to open the files as splits and tabs (specially \tttt{pot}).
-Most useful commands in netrw include:
-\begin{itemize}
-  \item \tttt{d} and \tttt{\%} for creating directories and files.
-    \ttt{Delete} removes both.
-  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
-  \item \tttt{gb} and \tttt{uU} are used to load directories
-    while marked files might be copied, moved, edited, greped, tagged and migrated
-    to and from the argslist as in \tttt{:h netrw-mf}.
-\end{itemize}
-Last notes:
-there is no insert mode in netrw interface;
-the commands in \tttt{:h netrw-explore} are
-handy for opening the directory of the
-file being edited;
-further information in \tttt{:h netrw}.
-
-\subsubsection{Ex}
-This might be the least popular mode.
-One might use \tttt{q:,/,?} to have a window
-with colon or search command history to be edited normaly,
-and the chosen command can be run with \ttt{CR}.
-Vimscript was largerly based on the ex editor~\cite{ex},
-and a more advanced user might use it for prototyping
-by defining mappings and settings and managing sessions and scripts
-propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
-In the default interface started with the \tttt{Q} command in normal mode,
-each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
-follow documentation from \tttt{:h Ex-mode} for further information.
-
-\subsection{Standard configuration files and directories and my .vim/vimrc}
-You can check the scripts Vim loads by using the debug script mentioned
-in Section~\ref{state}.
-By default, \tttt{~/.vimrc} and \tttt{~/.vim/vimrc} files are run by Vim at the beginning of the startup.
-One might edit the vimrc file with \tttt{:e \$MYVIMRC}
-and reload it with \tttt{:source \$MYVIMRC}, but
-I use the mappings in~\cite{vimrc} to encourage a continuous enhancement of my settings.
-Any other file might be included to run at startup by
-adding a line \tttt{:source <afile.vim>} in vimrc.
-In fact, it is on vimrc that one usually specifies the plugins
-and plugin managers they use.
-Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
-or follow some patterns described on the next section to change the
-scripts and sequence of them to be loaded.
-
-\subsection{Plugins and packages}
-One can see the list of standard plugins
-with \tttt{:h standard-plugin-list} command.
-Any \tttt{plugins/**/*.vim} file inside a directory
-listed by \tttt{:se runtimepath} will be loaded
-(e.g. \tttt{.vim/plugin/something/ascript.vim}).
-There are various ways to automate the installation
-and enhnace the management of plugins.
-By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
-that downloads latest scripts from sourceforge as specified in \tttt{:h getscript-data},
-and the Vimball interface, which creates and loads a Vimball for a plugin
-with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
-that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
-The Vimball can be installed in a system by \tttt{:source <filename>.vba}
-or loading it at vim startup with \tttt{\$ vim <filename>.vba}.
-
-A Vim package is a directory that contain plugins.
-It should be located inside a \tttt{pack/} directory
-somewhere in the directories listed with \tttt{:se runtimepath}.
-The plugins found in \tttt{pack/<packName>/*/start/} are loaded
-at startup, the plugins found at
-\tttt{pack/<packName>/opt/**} are loaded with\\
-\tttt{:packdd <script\_or\_directory\_name>}.
-
-All directories vim looks for scripts are described
-in \tttt{:h vimfiles} and are basically set by
-\tttt{:se runtimepath} and conventions inside
-each path therein, such as to look for \tttt{vimrc} files
-and \tttt{plugin/} or \tttt{pack/} directories.
-Vim scripts can be loaded conditionaly,
-e.g. only if a function is used as in \tttt{:h autoload-functions}.
-Example of such are filetype plugins (enabled by in a \tttt{ftplugin/<filetype>.vim} file,
-e.g. inside a plugin directory).
-There is a number of plugin managers for Vim.
-Pathogen and Vundle seem to be the most popular,
-one because of its minimalism, the other because of
-advanced features, e.g. for searching and installing plugins with
-colon commands.
-
-\subsection{Spell and spelllang (en and pt\_br)}
-One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
-or \tttt{:se spelllang=pt\_br}
-and toggle spell checking with \tttt{:setl spell!}.
-These are used so often that one might use shortcuts as in~\cite{vimrc}.
-Currently, Vim will download the files for a specific language if
-not found in system.
-
-\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}
-In Vimscript,
-the colon commands (also Ex or command-line commands) are related through spaces,
-punctuations and keywords (see \texttt{:h script}).
-Scripting the Vim editor can also be accomplished using other languages,
-as well documented e.g. in \tttt{:h python}.
-Functions are defined through colon commands and are called
-inside colon commands e.g. \tttt{:call MyFunction()}
-or \tttt{:echo MyFunction(4)}.
-Notice that functions are not commands but might be binded to them
-through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
-Executing source files is very straightforward with \tttt{:so
-<filename>.vim},
-and one can always use the Ex mode for rapid scripting.
-
-At the same time, the \tttt{:term} and terminal-job mode make
-scripting other software more convenient, as output is
-proptly navigated and copied. The shell works well for generic use,
-such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
-See Section~\ref{terminal} for further directions about the terminal.
-
-\subsection{Color}\label{visual2}
-Newer versions of Vim support 24 bit true color (aka 16 million colors)
-in terminal Vim (not only in GVim).
-The terminal must support true color, and tests are available e.g. in
-\url{https://gist.github.com/XVilka/8346728}.
-Then Vim needs to be set to use true colors with
-\tttt{:set termguicolors}.
-If using 8 or 16 bit colors, Vim uses the color pallete from the
-terminal, if using true colors each color is defined directly.
-Settings for using true colors inside Byobu/Tmux involve tweakings
-and are available in~\cite{vimrc}.
-Good color schemes to use with true colors are Gruvbox
-and Solarize.
-One might source syntax files at any time to change syntax, usually
-though linking tokens to syntax groups with
-\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
-and then relating the group to another group 
-\tttt{:highlight link <group\_name> <group\_name2>}
-or by specifying the group characteristics directly:
-\tttt{highlight <group\_name> guifg=\#ffffff}.
-If you change a syntax file, reloading a file with with \tttt{:e<CR>}
-updates the highlighting on the window with the corresponding filetype.
-A complete syntax highlighting support typically involves at least three files:
-\begin{itemize}
-  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
-    detected with e.g.
-    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
-  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
-    file type, such as: 
-    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
-    expandtab autoindent}.
-  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
-    tokens and highlighting groups; and highlighting group definitions.
-\end{itemize}
-
-This very scheme is implemented very straightforward in this
-plugin~\cite{tokipona}
-for highlighting text in the Toki-Pona language~\cite{tpLang}.
-Syntax highlighting plugins are as filetype plugins,
-but also have a \tttt{syntax/<filetype>.vim} file
-relating keywords to highlighting groups. 
-One might see every highlighting group, and their final visual results,
-with the command
-\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}.
-In my system, the \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
-directory, but \tttt{syntax/} files had to be moved to \tttt{$\sim$/.vim/syntax/}.
-
-\subsection{Fonts}
-The fonts are defined by your terminal or inside GVim.
-\ttt{C-+} and \ttt{C--} can be used to change font size.
-Some settings for fonts, such as boldface, might be set using
-the syntax highlighting facilities described in the previous section.
-
-\section{Conclusions and further work}\label{conc}
-This document seems reasonable as an overall reference of the Vim editor,
-at least for my usage and proficiency.
-Given the folkloric milestone of using Vim for 10 years,
-this article might serve as a benchmark for one to relate
-it's current use and understandings.
-As a pedagogical material, it seems to be unique in the emphasis
-on namespaces, understood as commands, variables, state-related lists, etc,
-especially in Section~\ref{namespaces},
-and the reference to the standard Vim documentation
-to achieve the DRY KISS style described in Section~\ref{intro}.
-
-\noindent {\bf Potential enhancements} to this document are:
-\begin{itemize}
-  \item The inclusion of facilities such as reading emails and connecting over ssh.
-    There is a working hack in~\cite{vimrc} for browsing over the WWW,
-    but such aspect of Vim usage might receive more attention
-    given that it is confortable to use the navigation and editing facilities
-    and the resulting integrated environment.
-  \item Updating of the information I can find about the issues discussed,
-    such as about status lines in Section~\ref{visual1}.
-  \item Include a discussion about Neovim.
-    I have never used it, but it seems to be reaching a considerable user base
-    and it might be feasable to give an account of Vim and Neovim
-    after some tests and researching the official and user base documentation.
-  \item Better cite documentation, plugins and Vim authors.
-    I prefered to keep the references inline through \tttt{:h} commands
-    and URLs, more in accordance with the style of Vim documentation,
-    but bibliographical items constitute a valuable asset for academic literature,
-    and some authors might find their work more pretiged if more
-    thoroughly cited.
-\end{itemize}
-
-\noindent {\bf Potential next steps} in using Vim:
-\begin{itemize}
-  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
-  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
-  \item Better integrate Python and Vimscript, especially for data visualization
-    and syntax highlighting management.
-    In accordance with the visualization issues described in
-    Sections~\ref{visual1} and~\ref{visual2}.
-  \item Make plugins for:
-    \begin{itemize}
-      \item listing all the mappings available in each mode and the typing combinations which are available
-        for new mappings.
-      \item Sessions, as described in Section~\ref{visual1}.
-      \item AA messages (shouts): to keep track, document and share of working sessions
-        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
-        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
-        relate AA sessions to other media, such as software repositories,
-        screencasts and images, interact with IRC channels and other social platforms.
-      \item Slick Vim: a collection of the settings, mappings and plugins I use.
-        Enhancements such as using \ttt{C-} commands also to browse tabs,
-        shortcuts to join a window into a tab,
-        and dummy minimal plugins as simplest, file type and syntax highlighting,
-        Some more elaborate tweaks should also be present, such
-        as breaking lines in sensible places for natural language texts
-        while respecting e.g. \tttt{:se textwidth}.
-      \item Dealing with .swo and .swp temporary files.
-        In summary, if the restored .swo file has the same content
-        and the file being opened,
-        the restoration phase can be omitted.
-        If the contents differ, Vim should open a tab with each file
-        in a vertical split and run \tttt{:windo :diffthis}.
-      \item Rendering images and equations.
-        These are useful for using Vim in presentations
-        or achieving a textual representation when it is mandatory,
-        such as to comply with the limitations of a platform (e.g. Vim editor). 
-        but also hold stylistic merits as ascii art is often
-        very appreciated.
-        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
-        and can directly render ascii charts from data using cues such are shape, position
-        and color.
-      \item Redirecting the commands that usually show the results in a 'more' interface,
-        which cannot be searched nor copied,
-        to a parsed and linked quickfix or location window.
-        The basic idea is to use \tttt{:redir} command to redirect the output
-        of such commands with \tttt{:se nomore}.
-      \item Slide presentations. I've been using some automation for browsing
-        slides and opening figures and some Vim users asked for the settings and
-        commands. They are very elementary use of registers being executed over
-        arbitrary but consistent textual patterns.
-      \item Changing the color scheme an highlighting scheme incrementally
-        and selectively using the features described in Section~\ref{visual2}.
+      \item Nasal: m (labial), n (coronal).
+      \item Plosive: k (dorsal), p (labial), t (coronal).
+      \item Fricative: s (coronal).
+      \item Approximant: j (dorsal), l (coronal), w (labial).
     \end{itemize}
@@
 
+There are standard guidelines for pronunciation,
+but the language allows for considerable allophonic
+variation.
+For example, /p t k s l/ might be pronounced
+[p t k s l] or [b d g z \textipa{\!R}].
+
+Syllables are of the form (C)V(N):
+an optional consonant, a vowel and an optional nasal consonant.
+Non word-initial syllables must follow the pattern CV(N).
+The following sequences are forbidden: ji, wu, wo, ti, mn, nm, mm,
+nn.
+
+\subsection{Syntax}
+As in other natural languages, 
+colloquial Toki Pona might have
+incomplete sentences and deviate from the
+norm.
+The basic structure of sentences are in the form:
+<subject> li <predicate> e <object>.
+The li might be repeated to associate more than
+one predicate to the subject.
+The particle li is omitted if the subject is a simple mi (I or us)
+or sina (you). A discussion about problems with this rule
+and how I deal with them is in Appendix~\ref{mytoki}.
+
+The e might be repeated to associate more than
+one object to a predicate.
+Sentences might be related though la,
+'sentence' la 'sentence', where the second sentence is
+the main sentence, and the first sentence is a condition
+to the first.
+Multiple la-s are not described in literature,
+but I assume that one might assume the last sentence
+being a conditional to the next,
+except in cases where the context strongly suggests
+otherwise.
+
+Noun and verb phrases are built with the non-particle words.
+The first word is the noun and phrase and subsequent words
+qualify the noun or phrase.
+The pi particle might be used to separate sequences of words
+to be evaluated before the relations yield by pi:
+As pi is often ill understood and used,
+the following structures might be handy for newbies and as a
+reference:
+\begin{itemize}
+  \item No pi, `word word word':
+word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
+  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
+      word $\leftarrow$ (qualifies 1)  word ].
+  \item Two pi-s: `word pi word word word pi word word':
+    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
+or:
+    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
+\end{itemize}
+
+Notes on the usage of pi:
+\begin{itemize}
+  \item In a sequence of words, without pi, the second word qualifies
+    the first, the third word qualifies the phrase yield by the first
+    two words, the fourth word qualifies the noun yield by the first
+    three words and so on.
+  \item It is redundant to use pi before the last word in a noun or
+    verb phrase if there is no other pi, reason why it is most often
+    omitted.
+    Its use in this case is regarded as wrong~\cite{lipu,pije},
+    but, as one might notice, it does not introduce any ambiguity.
+  \item The book by jan Pije~\cite{kama} describes another use for pi:
+    after li to mean possession, e.g. `soweli li pi sina' (your pet).
+    This employment of pi might be regarded as correct, but are promptly written
+    as a noun phrase (e.g. `soweli sina') and is not mentioned
+    by the official book~\cite{tpLabg}.
+\end{itemize}
+
+All the words except the structural particles (li, e, la, pi)
+are usable in noun and verb phrases.
+Notice that the phrase expresses a noun in a noun phrase (subject or
+object) or a verb (in the predicate).
+
+At this point, the only missing syntax rule is related to
+the prepositions: kepeken, lon, sama, tan, tawa.
+They might appear at the end of noun phrases,
+should be followed by  another noun phrase,
+and require no particle. E.g.
+`toki tan jan Pije li pana e sona tawa mi'.
+
+Other particles are: a or kin, o, taso, anu, en, 
+nanpa, seme. mu?,
+
+Vocatives should have an o after the noun phase.
+
+\subsection{Further notes}
+The only synonyms on Toki Pona are:
+a and kin; lukin and oko;
+sin or kamako;
+ale or ali.
+
+toki e ni, wile e ni, etc.
+'e ni' can be omitted and : used by itself.
+
+Names are by default transliterated,
+
+% my view
+% considerations about the official book and jan Pije
+% pictorial forms
+\subsection{Main references for the language}
+\begin{itemize}
+  \item The official book is ``Toki Pona: The Language of Good''
+    and is authored by Sonja Lang, the creator of the language.
+  \item The book ``o kama sona e toki pona!'', from jan Pije,
+    is the other main reference for the language~\cite{kama}.
+  \item \cite{Wikipesija}.
+\end{itemize}
+
+\section{Analysis and hacks for the language}\label{hacks}
+% number of possible syllables
+% number of possible words of size X (1, 2, 3, 4....).
+% syntax highlighting, change colors on the fly
+% analysis of the language
+% synthesis of sentences and texts
+% hacks from other ppl
+\subsection{Hacks from other people}
+The \url{tokipona.net} has a number of tools,
+just as to transliterate names into Toki Pona phonemes,
+and search in corpus.
+
+\section{Conclusions and further work}\label{conc}
+\begin{itemize}
+  \item Relate Toki Pona to Wordnet: should one Toki Pona word
+    be related to more then one synset of the English language?
+  \item Understand how the corpus is gathered in \url{tokipona.net}.
+  \item Know about previously existing words that were used for Toki Pona
+    (e.g. suno and suwi might come from sun and sweet),
+    and about the reasons that lead Sonja (and maybe other people)
+    to choose the 14 letters and the syllable structure.
+    This might require a dedicated communication with the
+    speaker community and the documentation authors.
+  \item Corpus-based analysis.
+  \item Publication of original texts and translations.
+\end{itemize}
 \subsection*{Acknowledgments}
@@
 \appendix
-\section{Example of usage session}
-I usually begin by opening a file or directory
-with \tttt{\$ vim <filename>}.
-The color scheme is alternated between
-blue and GruvBox with
-\tttt{:colo gruvbox} and \tttt{:colo blue}.
-I open an vertical split and then move
-the window to a new tab using
-\tttt{:vs} and \ttt{C-W\_T}.
-I then search tokens related to
-the enhancements I want to make or
-the knowledge I want to acquire.
-I go back to the previous tab with \tttt{gr}
-and make a global replace with
-\tttt{:\%s/<this>/<that/g}.
-On adding dots to sentences,
-I record in the \tttt{"q} register
-the sequence \tttt{jA.jj},
-using it as a macro 10 times by
-\tttt{10@q}.
-I move to the other tab with \tttt{gt}
-and open a terminal window with \tttt{:term}
-for compiling latex files.
-I start another terminal with \tttt{<C-W> :term}
-for opening the resulting PDFs with evince.
-If any new idea comes to mind and I have time,
-\tttt{\textbackslash\textbackslash s} opens 
-my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
-sources it.
-If there is e.g. code or notes in other projects I am working
-on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
-in Byobu/Tmux sessions and windows.
+\section{My usage of Toki Pona}\label{mytoki}
+I use the standard sounds, but often use [z] for s.
+I often translate texts to Toki Pona (e.g. biblical excepts)
+and create new texts as poems and short stories.
+Most of them are in~\cite{tokisona}.
+I omit the li particle after subjects sina and mi,
+in accordance with the norm,
+but sometimes I use them when there are many predicates.
+E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
+In such cases, the first li is sometimes omitted.
+Also, sometimes I use li before mi and sina where I find
+that there is unwanted ambiguity, e.g.
+sina moku pona e jan 
+(might be sina li moku pona e jan or sina moku li pona e jan).
 
-\section{My vimrc file and usage}
-In this Appendix is my vimrc file.
-Although it has comments, one should
-look for the options that (s)he does
-not understand, as a thorough explanation
-of the file is tedious and out of the scope
-of this document.
-
-In using Vim with my vimrc file~\cite{vimrc},
-I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
-and the tab line with \texttt{\textbackslash\textbackslash T}.
-Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
-The mappings for transitioning through splits and tabs
-are also used constantly.
-
-\section{Example of notes on mappings}\label{notes}
-\texttt{:h index} shows all the default mappings
-while \tttt{:map} shows the user-defined mappings.
-By considering such information, one can make
-useful observations exemplified in this Appendix.
-
-\subsection{Normal mode}
-Every letter and character in the keyboard is used.
-In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
-\ttt{BS} and \ttt{C-H} is the same as h.
-\ttt{C-J} and \ttt{C-N} is the same as j.
-\ttt{C-P} is the same as k.
-Space is same as l.
-\ttt{C-[} and \ttt{Esc} are not used
-\tttt{<C-\textbackslash> a-z} is reserved for extensions
-\ttt{C-\_} not used.
-\tttt{+} is the same as \ttt{CR} and they are both not very useful.
-Del is same as x.
-
-Many \tttt{][} combinations are not used, e.g.
-with \tttt{abhjklfg}.
-The \tttt{\_} command might be used as a more powerful \tttt{\^},
-leaving it free for mappings.
-
-Directions, home, end, page up and down, insert, all have mappings
-in more centraly located keys.
-unbounded: g,z,[]
-
-The \ttt{C-(HJKL)} commands are redundant,
-with the exception of \ttt{C-L} which redraws the screen,
-so it is a reasonable choice to use them to move focus
-of the editor to splits in the \tttt{hjkl} directions.
-
-Many key combinations are available for new mappings through the \tttt{g}
-and \tttt{z}, and \tttt{v} commands.
-They have typical uses, e.g. \tttt{z} for folds,
-spell checking and some movements (mainly when wrap is set).
-
-\subsection{Insert mode}
-All standard char keys are used for entering chars on text.
-\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
-in moving through multiple lines.
-\ttt{C-[} is the same as \ttt{ESC}.
-\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
-\tttt{<C-\textbackslash> a-z} is reserved for extensions,
-other combinations with \ttt{C-\textbackslash} are not used.
-
-\section{My \tttt{:version}}
-I should keep the output of \tttt{:version} executed on the system
-in which I wrote this document in this link:
-\url{}.
-It was compiled with this Makefile~\cite{makefile}
-in a 16.04 LTS Ubuntu Linux,
-and is tagged as 8.0, Included Patches 1-1173.
+Names are by default transliterated,
+but I advocate that, as in other languages,
+names might be used as they are in the
+correspondent mother tongue.
+E.g. the name Erdös is used in
+English and Portuguese although the standard
+alphabet does not contain ö in such languages.
+I also tend to legitimate the use of English (or German) words
+in Toki Pona texts if it is the case,
+as happens often in scientific writing
+(kernel is a German word used in English,
+webpage is an English word used in Portuguese).
 
@@
 a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
-\bibitem{makefile}
-	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
+\bibitem{kama}
+  Knight, B. (2017) o kama sona e toki pona! Available at:
+    \url{http://tokipona.net/tp/janpije/okamasona.php}
+\bibitem{Wikipesija}
+  Toki Pona community (2017). Wikipesija. Available at:
+    \url{http://tokipona.wikia.com}
 \end{thebibliography}

### 2017-11-11 article/article.tex
first draft almost done

@@
 
-Other particles are: a or kin, o, taso, anu, en, 
-nanpa, seme. mu?,
+Other particles are:
+\begin{itemize}
+  \item a or kin, emphasis.
+  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
+  \item taso, means however or 'only' if adjective.
+  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
+    For repetition of verbs, repeat li.
+    For object nouns, repeat e.
+    If the noun is complementing a preposition (tawa, lon),
+    one might repeat the preposition.
+    As Toki Pona is a recent language, and is able to cope with
+    variation due to its simplicity, I would advocate for
+    using en and anu wherever there is no ambiguity.
+  \item nanpa, denotes numbering.
+  \item seme, for questions, used next to the thing being asked for.
+    'Why?' might be expressed as
+    'seme la sina pana e moku lon sewi'.
+  \item mu, for animal noises. For me it is not a particle, as in the
+    official dictionary, but a noun.
+    I also like to use it as a verb:
+    'mi pakala e luka. mu mute.
+\end{itemize}
 
-Vocatives should have an o after the noun phase.
+The vocabulary specifies morphosyntactic classes:
+nouns, adjectives, verbs, pre-verbs, adverbs, particles, prepositions, and numbers.
+I find that they might help the user and newcomer, but
+it might also suggest a deviation from what I understand and read:
+the words might be used indistinctly to be the nouns
+(subjects, the predicate when there is no object, objects,
+and preposition complements),
+the adjectives (anything that does not start the noun phrase or follows a pi),
+verbs
+(follows mi or sina or li or a preposition),
+adverbs (follows the verb).
+The pre-verbs (wile, ken, awen, kama, lukin, sona),
+might follow a verb, but might also be understood
+as the verb qualified with the next word,
+which carries a very similar if not identical meaning.
+The pre-verbs are all also defined as other morphosyntactic classes,
+such as adjective, noun, verb.
+The only exception is wile, which is only a pre-verb.
+
+Thus, the classes given in the dictionary dictate little
+in practice:
+jan kala li lape lon ni.
+Where kala, lape and ni are in this phrase
+as adjective, verb and noun,
+and are in the dictionary as noun,
+adjective and adjective.
+
+As far as I can see, one should regard
+the particles li, e, pi, and la
+and punctuation.
+The other tokens of the vocabulary
+might be used in any of the remaining positions.
+Detection:
+\begin{itemize}
+  \item Noun: the first word in a noun phrase.
+    After an 'e' and after a pi,
+    The first word in the sentence if
+    sentence does not start by the verb.
+    Might be in the position of the verb
+    if the sentence has no object.
+  \item Adjective: second word on after an e and after
+    Second word on in the subject phrase if present.
+  \item Verb: after a li, mi or sina.
+    If there is no object, the verb position
+    is often a noun.
+  \item After a preposition,
+    there can be a noun phrase, a verb phrase
+    or nothing.
+  \item Notice that there is ambiguity in the structure
+    introduced by the omission of li after mi and sina.
+    Also, when there is no object, a noun or a verb
+    or an adjective might be in the verb position
+    if there is no object.
+    The prepositional complement is also not defined.
+    (mi moku tawa pali, tawa tomo).
+    So, these are sources of syntactic ambiguities
+    in Toki Pona.
+    They might be solved or minimized by using the semantics
+    of the words.
+    One preliminary effort in this direction might be
+    using the classes in the official dictionary to resolve
+    ambiguities whenever possible.
+    This solution is not optimal in correct POS tagging,
+    and does not solve all possible ambiguities
+    (there are words classified as nouns and adjectives,
+    adjectives and verbs, nouns and verbs, particle and verb).
+
+    Another source of ambiguity is the pre-verbs as described
+    in the literature~\cite{tpLang,pije}.
+    But I find it reasonable to understand them as verbs.
+
+\end{itemize}
+
+Also, the prepositional complement might
+be a noun or a verb.
+I could not come up with a sentence
+where it would be understood as an adjective.
 
@@
 
-toki e ni, wile e ni, etc.
-'e ni' can be omitted and : used by itself.
+In formations such as
+toki e ni:, wile e ni:, tan ni: etc.
+'(e) ni' can be omitted and : used alone.
 
 Names are by default transliterated,
+but might not be, as described in Section~\ref{mytoki}.
 
@@
 \section{Analysis and hacks for the language}\label{hacks}
+
+\section{Statistics of the vocabulary}
+In~\cite{pije} there is statistics about Toki Pona corpus.
+This section focuses on the statistics of the vocabulary
+and syntactic rules:
+the letters, phonemes, word sizes,
+possible combinations for words and sentences.
+The statistics are in Appendix~\cite{listings},
+and the next paragraph is an overview.
+Python scripts were used to obtain the measurements
+and are available at~\cite{tokipona}.
+
+As described in Section~\cite{basics},
+there is only 14 letters,
+and phonemes respect a few rules.
+There are 120 different words in the official vocabulary,
+4 of them having synonyms.
+A total of 124 tokens.
+Not counting proper nouns (names)
+and punctuation.
+They include X nouns, Y verbs, Z adjectives,
+W prepositions, Y particles, K pre-verbs.
+Of which one might distinguish only between
+the particles and the other J words that
+might be used anywhere a particle will not.
+
+Most often vowels,
+consonants, consonant vs vowel.
+Most often phonemes in general an specific positions..
+Possible phonemes given the rules.
+Comparison of the occurring against the possible.
+
+Syntactic structures:
+Possibilities on noun, verb and prepositional phrases
+and in sentences.
+Counting along the number of words.
+
+\section{Synthesis of text}
+The same package~\cite{tokipona}
+has capabilities for synthesizing text.
+Noun, verb and prepositional phrases,
+sentences.
+It also aims at making larger scale texts
+by keeping a record of the used words and structures (context)
+and using stylistic outlines for poems and short narratives.
+
 % number of possible syllables
 % number of possible words of size X (1, 2, 3, 4....).
-% syntax highlighting, change colors on the fly
-% analysis of the language
-% synthesis of sentences and texts
-% hacks from other ppl
+
+\section{Syntax highlighting}
+The same package~\cite{tokipona}
+has a Vim syntax highlighting plugin
+for Toki Pona.
+Instructions for installing and using
+the syntax highlighting is at~\cite{tokipona}.
+
+Basically, it distinguishes the words among the morphosyntactic
+classes according to the official dictionary.
+As a word often belongs to more than one class,
+The precedence of them might be set by the user.
+Also, some classes might be further refined or joined,
+such as by distinguishing only particles and the rest,
+or maybe particles and prepositions and the rest.
+The colors are also promptly changed according to~\cite{vimArt}
+and exemplified in the package documentation.
+
+Currently, the Python package synthesizes the
+syntax file.
+The user has control of class precedence and
+merging.
+The choice of precise coloring schemes
+might involve hacking the colorscheme being
+used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
+and Vim's highlighting schemes as described in~\cite{vimArt}.
+In summary,
+the usage of the package and plugin might be performed
+through the following actions:
+\begin{itemize}
+  \item Installation of the plugin.
+  \item Tweak of the syntax file by hand.
+  \item Running the Python script to generate a new syntax file
+    according to other settings.
+  \item Write a file inside Vim using Toki Pona and save
+    the file with the .tokipona extension.
+    Reload the highlighting scheme whenever you
+    change the syntax file by hand or through the Python script.
+  \item Access the used highlighted groups with :syntax,
+    Access all the highlighting groups with :so \$VIMRUNTIME/syntax/hitest.vim.
+    Change the coloring of a set of terms by associating
+    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g. Visual):
+    :highlight link tokiponaADJECTIVE Visual.
+    The plugin comes with the :TokiStation command, which opens a window with
+    the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
+    Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files. Another with Readme, and PDF documentation. Another with an IPython.
+    Imported tokipona. It resets completely upon command :TokiClose, closing all created windows.
+\end{itemize}
+
 \subsection{Hacks from other people}
@@
   \item Publication of original texts and translations.
+  \item Make an article written in Toki Pona.
+    I wrote Section~\ref{finalToki}, and believe that
+    a summary in both English and Toki Pona for
+    facilitating the acquisition of context,
+    a reasonable article on some scientific topics is
+    possible.
+    I first conceived something around complexity, statistics, physics, or computer science.
+    But one possibility is to write about linguistics, philosophy, literature
+    or psychology with the partners I write in English.
+    I can start a draft,
+    they might learn the language in a few hours (with or without my help),
+    to contribute, and we can write a short paper.
+  \item Enhance the synthesis of text to yield better contextualized text
+    and stylistic traces, such as for poems and short stories.
+    Give the user the ability to choose the sentences (generate randomly according to
+    previously written or given text,
+    some rules input by the user, the package and the language guidelines,
+    outputs to the screen and asks to keep and discard).
+  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
+    It uses only syntactic cues to choose POS tags,
+    which might be overcome by using n-grams and
+    further techniques from Natural Language Processing~\cite{POS}.
 \end{itemize}
+
 \subsection*{Acknowledgments}
@@
 webpage is an English word used in Portuguese).
+
+Proposed notations for numbers seem numerous.
+I tend to think that one might indicate is two numbers
+are multiplying (pi) or are in different scales
+(such as in decimal or binary notations).
+For example, I take luka two to mean 52.
+Or it might not be taken to such level of strictidness,
+for 52 is a reasonable notation for a simple language.
+mi jo e jan sama nanpa 12.
+Or even:
+ona li lon e soweli 27.
+
+Remove the 'li'.
+I've been avoiding e ni: and using only :.
+'ve been omitting the subject if it is the
+same as in the last sentence.
+I've been avoiding also the li sometimes,
+and starting only a predicate + object phrase,
+or a whole phrase altogether.
+
+\input{pos}
+
+\section{Final words in Toki Pona}
+toki li nasin e lawa.
+li nasin tawa (pi) toki insa.
+
+% jan li ken toki li ken pali pona e toki pona.
+toki pona li pona e nasin tan ni:
+ona li jo e sitelen mute lili.
+ona li pona.
+pona kepeken weka 'p' li ona, a.
+
+o taso la toki pona li kalama li lukin pona.
+sitelen en nasin li open e sitelen suli\cite{tpLang,pije,fb1,fb2,tokisona,Wikipesija}.
+li sona. li nasin e toki e sona e lawa e lon.
+
+toki ni li wawa tawa jan mute nasin en toki.
+wawa tawa toki pi jan sona.
+pi ilo nanpa en nanpa nasin.
+taso tawa toki sona a.
+sitelen sona, sitelen musi.
+
+ilo lon linha~\ref{hacks} li pana e sitelen
+e sona tan sitelen,
+en nasin.
+linha~\ref{basics} li pana e nasin pi toki pona.
+e sitelen tawa kama sona.
+
+mi wile pali e sitelen lon toki pona.
+sitelen lon nasin, sitelen sona,
+sitelen musi.
+ante e toki pona la
+la ona li nasa.
+taso nasa li pona mute.
+li pona mute tawa lawa,
+tawa kama sona, tawa sitelen e toki.
+ante la toki pona o.
+toki e toki pona tawa sina.
+kama sona e toki pona sina.
+
+o pona tawa jan pi toki pona.
+jan Sonja, Birns-Sprage, Kipo, Pije,
+Siwejo, Malija, jan kulupu mute.
 

### 2017-11-11 article/pos.tex
first draft almost done

\begin{table*}[h!]
\begin{center}
\caption{"POS tags incident and chosen.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the text highlighting Plugin, for example,
            a token has to have an established tag to have
            a defined color.
            On the Chosen column, the tokens were regarded only once
            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}
\begin{tabular}{| l | c | c |}\hline
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\\hline
ADJECTIVE & 40  & 35 \\\hline
VERB & 18  & 13 \\\hline
PARTICLE & 12  & 12 \\\hline
PRE & 5  & 5 \\\hline
PREPOSITION & 5  & 5 \\\hline
NUMBER & 4  & 1 \\\hline
total & 142  & 120 \\\hline
\end{tabular}\end{center}
\end{table*}

### 2017-11-11 tokipona/makeVimSyntax.py

@@ -1,0 +1,179 @@
+# import ezodf
+import pandas as pd
+import numpy as n
+import os
+# from ODSReader import *
+
+# ods = ezodf.newdoc(doctype='ods',filename='../data/dictionary.ods')
+
+# ezodf.config.set_table_expand_strategy('all')
+
+# spreadsheet = ezodf.opendoc('../data/dictionary.ods')
+
+
+os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+xl_file = pd.ExcelFile('../data/dictionary.xlsx')
+dfs = {sheet_name: xl_file.parse(sheet_name)
+       for sheet_name in xl_file.sheet_names}
+df = dfs['Sheet1']
+table = n.vstack((df.keys(), df.values))
+
+all = set()
+for row in table:
+    classes = " ".join(i for i in row[1].split() if i.isupper())
+    print(row[0], classes)
+    all.add(classes)
+all2 = set()
+
+colors = {'ADJECTIVE': 'Comment', 'NOUN': 'Constant', 'NUMBER': 'Identifier',
+          'PARTICLE': 'Statement', 'PRE': 'PreProc',
+          'PREPOSITION': 'Type', 'VERB': 'Special'}
+words = {}
+for color in colors:
+    words[color] = set()
+words_all = words.copy()
+
+synonyms = 0
+precedence = 'first'
+precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
+allp = list(colors.keys())
+precedence += [i for i in allp if i not in precedence]
+
+class_count = colors.copy()
+for i in class_count: class_count[i] = 0
+class_count_ = colors.copy()
+for i in class_count_: class_count_[i] = 0
+
+word_classes = {}
+for row in table:
+    # Better choose the class
+    classes = [i for i in row[1].split() if i.isupper() and i not in ("I,","O!")]
+    if precedence=="first":
+        class_ = classes[0]
+    else:
+        class__ = classes
+        not_chosen = 1
+        for pos in precedence:
+            if pos in class__ and not_chosen:
+                class_ = pos
+                class_count[pos] += 1
+                class_count_[pos] += 1
+                not_chosen = 0
+            elif pos in class__:
+                class_count[pos] += 1
+
+    print(row[0], class_, colors[class_])
+    all2.add(class_)
+    words_ = row[0].split()
[…]

### 2017-11-11 tokipona/renderWords.py

@@ -1,64 +1,0 @@
-# import ezodf
-import pandas as pd
-import numpy as n
-import os
-# from ODSReader import *
-
-# ods = ezodf.newdoc(doctype='ods',filename='../data/dictionary.ods')
-
-# ezodf.config.set_table_expand_strategy('all')
-
-# spreadsheet = ezodf.opendoc('../data/dictionary.ods')
-
-
-os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
-xl_file = pd.ExcelFile('../data/dictionary.xlsx')
-dfs = {sheet_name: xl_file.parse(sheet_name)
-       for sheet_name in xl_file.sheet_names}
-df = dfs['Sheet1']
-table = n.vstack((df.keys(), df.values))
-
-all = set()
-for row in table:
-    classes = " ".join(i for i in row[1].split() if i.isupper())
-    print(row[0], classes)
-    all.add(classes)
-all2 = set()
-
-colors = {'ADJECTIVE': 'Comment', 'NOUN': 'Constant', 'NUMBER': 'Identifier',
-          'PARTICLE': 'Statement', 'PRE': 'PreProc',
-          'PREPOSITION': 'Type', 'VERB': 'Special'}
-words = {}
-for color in colors:
-    words[color] = set()
-
-for row in table:
-    class_ = [i for i in row[1].split() if i.isupper()][0]
-    print(row[0], class_, colors[class_])
-    all2.add(class_)
-    words_ = row[0].split()
-    for word in words_:
-        if word != 'or':
-            words[class_].add(word)
-lines = []
-for class_ in words:
-    words_ = ' '.join(words[class_])
-    group = colors[class_]
-    lines.append('\nsyntax keyword '+'tokipona'+class_+' '+words_)
-    lines.append('highlight link '+'tokipona'+class_+' '+group)
-
-coloring = '\n'.join(lines)
-header = '''" Vim syntax file
-" Language:	toki pona
-" Maintainer:	Renato Fabbri <fabbri@usp.br>
-" Last Change:	2016 Apr 30
-" Remark:	toki pona is a conlang, not a programming language
-
-if exists("b:current_syntax")
-    finish
-endif
-
-syntax clear
-syntax case ignore\n\n'''
-with open('../syntax/tokipona.vim','w') as f:
-    f.write(header+coloring)

### 2017-11-11 article/article.tex
ideas for enhancements

--- 2017-11-10 article/article.tex
@@
     thoroughly cited.
+  \item An analysis of my usage, e.g. according to \url{http://www.drbunsen.org/vim-croquet/}.
+    This undertake might benefit from data from many users,
+    which favors the potential plugin for usage analytics described in the next bullet list.
 \end{itemize}
@@
         and selectively using the features described in Section~\ref{visual2}.
+      \item For bringing back all the splits after an \tttt{:only}.
+        Also for bringing back the tabs after a \tttt{:tabo}.
+      \item Run Python excepts from file in an IPython shell. 
+        The buffer number of the terminal window should be stored,
+        and than any selected lines should be run on that instance.
+        Mappings should make available all movements to catch the excepts,
+        execution of registers, remain on script or in the IPython shell.
+        Also, terminal mode can use some improvements, such as \ttt{C-O} for
+        one normal mode command.
+      \item For keeping track of the usage and making analysis
+        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}. Usage analytics.
     \end{itemize}

### 2017-11-12 article/article.tex
ideas for enhancements

@@
 follow documentation from \tttt{:h Ex-mode} for further information.
+I might be loosing something, but for most
+hacking I use the command-line window accessed with \tttt{q:}.
+I find it comfortable to browse the history in a normal-like mode,
+and edit them also using the insert mode as usual.
+The mapping have to be defined as autocommands as in~\cite{vimrc}.
 
@@
         for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}. Usage analytics.
+      \item Colorschemes for true color (gui).
+        The standard colors (e.g. blue, elflord) loose some of the distinctions,
+        e.g. SpellBad tokens are not highlighted on these colorschemes
+        it you have a functioning \tttt{se termguicolors} (standard GUI mode
+        and available in terminals since recent versions of Vim 8, see Section~\ref{visual2}). In~\cite{vimrc} are some lines that make such colorschemes after over
+        \tttt{:colorscheme blue}.
+        This design of color schemes over the standard color schemes might be
+        a very simple and effective way to make new color schemes.
     \end{itemize}

### 2017-11-19 article/article.tex
andado com artigo

--- 2017-11-11 article/article.tex
@@
 \usepackage{tipa}
+\newcommand{\ttt}[1] {
+	\texttt{<#1>}}
+\newcommand{\tttt}[1]{\texttt{#1}}
 % \usepackage{fontspec}
@@
 \begin{document}
-\title{The Toki Pona Language:
-an overview and some hacks}
+\title{Basic concepts and tools for the Toki Pona minimalist language:
+Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
 % Anthropological computation
@@
 \begin{abstract}
-
+  A minimalist constructed language (conlang)
+  is useful for experiments and is comfortable for making tools.
+  The Toki Pona conlang is minimalist both in the vocabulary
+  (only 14 letters, X phonemes, and 120 words + 4 synonyms)
+  and in the $\approx10$ syntax rules.
+  The language is useful for being a used and somewhat established
+  minimalist conlang with at least hundreds of speakers.
+  In this article, we describe current concepts and resources
+  for Toki Pona,
+  and make available Python scripted routines for
+  analysis and synthesis of texts and the specification of syntax highlighting schemes.
+  We focus on the analysis of the basic vocabulary,
+  as corpus analyses were found in~\cite{corpus}.
+  The synthesis is based on the basic sentence template,
+  relates to context by keeping track of used words,
+  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
+  and number of sentences, words and letters (e.g. for paragraphs).
+  Syntax highlighting is performed in 24bit true colors, currently only
+  with the Vim text editor.
+  Word colors reflect morphosyntactic classes given in the official dictionary
+  and different solutions to necessary and arbitrary choices are described and implemented.
+  In summary, this text holds potentially novel conceptualizations about,
+  and tools and results in analyzing, synthesizing or syntax highlighting the
+  Toki Pona language.
+  % subj (li), obj (e), prepositions, qualification (word concatenation), 
+  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
 \end{abstract}
+{\bf keywords:} Natural Language Processing, Textual synthesis, Syntax highlighting, Wordnet, Toki Pona
 
@@
 Toki Pona is a minimalist conlang (constructed language)
-with only 126 words.
+with only 124 words (120 without the synonyms).
+Therefore, the concepts are usually very general and different,
+and, without context, the words are rarely related through
+meronymy and hyponymy.
+Such a linguistic setting is desired because of the simplicity
+which entails easier e.g. learning and tool making.
+Another reason why the minimalistic language design is compelling
+is the study and harnessing of the strong and weak forms of
+the Sapir-Whorf hypothesis (linguistic relativity),
+i.e. that language influences or dictates ones thought
+and world experience.
+Accordingly, one uses a conlang as a thinking tool (or platform)
+or to make experiments about the influence language has on
+the thoughts of the `speaker' (also writer and reader).
+Toki Pona is often described as a tool to meditate,
+to simplify the thinking processes, and as a way to modify
+the mood and impressions about the world.~\cite{tp1,tp2,tp3,tp4}
+In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
+describes that she has seen the language been used successfuly
+in the context of management, creation of texts, legal texts,
+etc.
+
+In this article, we present a conceptual overview of the language,
+which is fit both to the newcomer and to the expert.
+We also summarize
+texts and software gadgets
+for Toki Pona that
+were developed by us and by other Toki Pona users.
+Most importantly, the conceptual overview is considerably different
+from what we found in the literature~\cite{pu,pije,memrise},
+with emphasis on simplicity and flexibility,
+and the software routines yield analyses, syntheses
+and highlighting Toki Pona what was thought to be complementary
+to the software currently available.
+
+Next subsections hold a description of the general resources available,
+a historical note, and some words about natural and constructed languages.
+Section~\ref{basic} describes Toki Pona's phonology and syntax.
+Section~\ref{hacks} presents the software routines we made available
+and immediate results, such as listings and statistics of words,
+poems and short stories, and coloring schemes.
+Conclusions and further work are in Section~\ref{conc}.
+Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
+with thoughts of rule breaking and potentially new conlangs.
+Appendix~\ref{ftp} holds final words in Toki Pona.
+
 % definition, historic, status (groups, speakers, etc).
 
-\subsection{Resources on the web}
-http://tokipona.net
+\subsection{Resources on Toki Pona}
+One might organize current resources for the Toki Pona
+language in: references and learning material, corpus,
+websites, interaction groups (here users talk a post texts and comments), and software gadgets.
+The main references of the language are:
+  the official book ``Toki Pona: The Language of Good'',
+    authored by Sonja Lang, the creator of the language;
+the online book ``o kama sona e toki pona!'',
+from jan Pije~\cite{kama}.
+For a more comprehensive view of the resources available
+for the user, we suggest following the links 
+from~\cite{wikiToki}.
+Section~\cite{third} holds a summary of what we found in software
+gadgets related to Toki Pona.
+
 \subsection{Historical note}
+Toki Pona was developed as an internal and personal language
+by Sonja Lang~\cite{interview}.
+It was released as a draft in 2001 and in 2007 some documents
+reported it to have a few hundred speakers.
+The English official book~\cite{lipu} was released only in 2014.
+In 2016, a version of the official book was released in French.
+Nowadays, one finds a number of texts about Toki Pona and written
+in it, and uses of Toki Pona for artificial intelligence~\cite{roila}
+and software tools~\cite{tptoll1,tptool2,tptool3},
+in social platforms such as Facebook groups~\cite{fbtp,fbtpt},
+microblogging~\cite{tokilili}, Telegram~\cite{telegram} and IRC~\cite{irc}.
 
-\subsection{Natural and constructed languages}
+\subsection{Natural and constructed and artificial languages}
+A language one uses (or might use) to communicate by speaking
+and writing is called `natural language'.
+A `constructed language' is a natural language built by someone or
+a group, such as Esperanto, Toki Pona, and Lojban.
+An artificial language is a language yield by artificial agents,
+such as in AI routines, and might be considered
+within `cultural evolution' studies.
+Formal languages are defined by tokens and rules to
+operate them, they span from computer programming languages
+to math and formal models for natural languages.
+Lojban is a constructed language with the purpose of being
+an artificial language because it is more axiomatic.
+
+Constructed languages are in some traditions called artificial
+languages, but creators most often prefer to use the term `standardized'
+or 'planned' language with the argument that the conlang
+is rooted on natural languages, and 'artificial' is misleading~\cite{conlanWikip}.
+The preferred terms seem to be planned or constructed languages
+or conlang.
+The construction of languages is called glossopoeia.
+Toki Pona is a conlang which might be classified
+as engineered for experiments, meditation and philosophy;
+and suitable for auxiliary international language and
+as an artistic language. 
+% taxonomy, vocabulary, ontology, terminology,
+
 % conlangs, natural langs, programming langs, Lojban as an
 % intermediary
+% non-verbal languages, musical corporal language, 
+% non-verbal languages confront Sapir-Whorf hypothesis.
+
 \section{Overview of the language}\label{basics}
+This section describes very succinctly the formation of
+words and sentences in the Toki Pona language.
+It should enable a newcomer to grasp the essentials
+of the language and the experienced to
+acquire new insights.
+Furthermore, it is a solid reference of the syntax and phonological
+rules.
+
 \subsection{Phonology}
@@
 \subsection{Syntax}
+\subsubsection{Fundamental notions}
 As in other natural languages, 
@@
 norm.
-The basic structure of sentences are in the form:
-<subject> li <predicate> e <object>.
+The basic structure of sentences is:
+`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
 The li might be repeated to associate more than
@@
 The particle li is omitted if the subject is a simple mi (I or us)
-or sina (you). A discussion about problems with this rule
+or sina (you). A discussion about issues (problems)
+yield by this rule
 and how I deal with them is in Appendix~\ref{mytoki}.
@@
 
-Noun and verb phrases are built with the non-particle words.
-The first word is the noun and phrase and subsequent words
-qualify the noun or phrase.
+Noun and verb phrases are (usually) built with the non-particle words.
+The first word is the noun or verb and subsequent words
+qualify the noun or verb.
 The pi particle might be used to separate sequences of words
-to be evaluated before the relations yield by pi:
+to be evaluated before the relation yield by pi.
 As pi is often ill understood and used,
@@
     Its use in this case is regarded as wrong~\cite{lipu,pije},
-    but, as one might notice, it does not introduce any ambiguity.
+    but, as one might notice, it does not add (much) information
+    through syntax (order of qualifies is conserved).
+    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
+    Also used in~\cite{akesiWawa}.
   \item The book by jan Pije~\cite{kama} describes another use for pi:
@@
 All the words except the structural particles (li, e, la, pi)
-are usable in noun and verb phrases.
+are usable in any position in noun and verb phrases,
+(i.e. 116 words).
 Notice that the phrase expresses a noun in a noun phrase (subject or
 object) or a verb (in the predicate).
+And that the first word of the phrase is the noun or verb,
+and that subsequent words are adjectives or adverbs.
+The pi and other particles restarts the phrase to
+first word being a noun.
 
@@
 the prepositions: kepeken, lon, sama, tan, tawa.
-They might appear at the end of noun phrases,
-should be followed by  another noun phrase,
+They might appear at the end of phrases,
+should be followed by another phrase,
 and require no particle. E.g.
-`toki tan jan Pije li pana e sona tawa mi'.
+`toki *tan* jan Pije li pana e sona *tawa* mi'.
+Because prepositions can be used as adjectives
+or nouns or verbs, adverbs, etc, and what follows
+a preposition complement might also be of any morphosyntactic class,
+they are core sources of ambiguity.
+three:
+* preposition
+* absence of li on sina and mi
+* Incomplete sentences
+(ona li) tawa lon kon e ilo pi suli mute
 
-Other particles are:
+\subsubsection{Particles}
+Beyond the structural particles (li, e, la, pi) presented
+in last section, other particles are:
+
 \begin{itemize}
@@
   \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
-  \item taso, means however or 'only' if adjective.
+  \item taso, means however as sentence or 'only' if adjective.
   \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
@@
     using en and anu wherever there is no ambiguity.
+    E.g.: 'mi tawa anu moku lon tenpo lili?'
+    In the official documentation that is not described and
+    might be regarded as wrong in strict canonical.
   \item nanpa, denotes numbering.
@@
     I also like to use it as a verb:
-    'mi pakala e luka. mu mute.
+    'mi pakala e luka. mu mute.'
 \end{itemize}
@@
 might be used in any of the remaining positions.
-Detection:
+
+\subsubsection{Convention for recognizing POS tags by speaker and machine}
+
 \begin{itemize}
@@
 
-\subsection{Further notes}
+\subsubsection{Further notes}
 The only synonyms on Toki Pona are:
@@
 % pictorial forms
-\subsection{Main references for the language}
-\begin{itemize}
-  \item The official book is ``Toki Pona: The Language of Good''
-    and is authored by Sonja Lang, the creator of the language.
-  \item The book ``o kama sona e toki pona!'', from jan Pije,
-    is the other main reference for the language~\cite{kama}.
-  \item \cite{Wikipesija}.
-\end{itemize}
+\section{Software for analysis and synthesis and for syntax highlighting}\label{hacks}
+In this section,
+we describe software, natural language, audiovisual and numeric results
+on four directions:
+1)
+the analysis of the Toki Pona language through statistics
+obtained by processing the dictionary;
+2)
+the synthesis of sentences, poems and short stories
+in Toki Pona;
+3)
+syntax highlighting for Toki Pona,
+including fine-tuning and theoretical
+considerations;
+4)
+initial considerations about the WordNet of
+Toki Pona words.
+Corpus analysis was already found in~\cite{}.
 
-\section{Analysis and hacks for the language}\label{hacks}
-
-\section{Statistics of the vocabulary}
+\subsection{Statistics of the vocabulary}
 In~\cite{pije} there is statistics about Toki Pona corpus.
@@
 
-\section{Synthesis of text}
+\subsection{Synthesis of text}
+number of elisoes
+
 The same package~\cite{tokipona}
@@
 
-\section{Syntax highlighting}
+\subsection{Syntax highlighting}
 The same package~\cite{tokipona}
@@
 
+The tokipona.vim syntax file has an association of toki pona
+words with highlighting groups.
+It also holds association of these groups to other groups
+to use their color settings.
+Therefore, one is able to use various color schemes
+with syntax-highlighted Toki Pona texts.
+A Vim user might run \tttt{:colorscheme blue},
+\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox}
+\tttt{:colorscheme elflord} to see the same text colored
+with different colorschemes.
+To interfere directly on the colors chosen,
+\tttt{:highlight Normal guifg=\#00000 guibg=\#0000ff}
+will change the standard foreground (text) color to pure black
+and background to pure blue.
+See \tttt{:h gui-colors} and \tttt{:h highlight}
+for the way you might edit colors directly.
+Useful commands are given as commented lines
+in the end of the syntax file.
+
+\subsubsection{Advanced syntax highlighting considerations}
+Standard guidelines for syntax highlighting
+depend heavily on cultural and use factors
+(https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies),
+although there are projects such as Solarized~\cite{solazired}
+(http://ethanschoonover.com/solarized)
+which try to give general solutions for some contexts.
+Standards guidelines for syntax highlighting were not found,
+and might be driven from data visualization considerations.
+Most evident examples of such guidelines are:
+\begin{itemize}
+  \item The use of blue and other high frequency colors for
+  background.
+  This is extended to other high frequency (small wavelength)
+  colors such as violet.
+  \item The avoidance of colors of blue and other high frequency
+  colors in small objects, such as letters and words.
+  \item Explore simplicity and elegance through the use of discrete
+  coloring schemes. Most of the tokens should be preferably of
+  the same color.
+  One might use a power-law to mimic natural occurring phenomena
+  and perception. Or an exponential to map parameters to perceptual
+  stimuli.
+  \item You might always wish to maximize contrast,
+  although taste and less consative combinations might
+  also dictate choices.
+  \item Stipulate axes of parameters with which to set colors.
+  Wavelength or frequency is an obvious axis given the considerations
+  above.
+  Other parameters are hue, transparency etc.
+
+  two types of colorschemes: those with dark background,
+  which are more commode at first; and those with a light background,
+  which are usually impressive (and even annoying) at first,
+  but the eye gets used to it and it keeps you more stimulated.
+  The blue color seems to be specially related to physiological
+  stimulation of the body by color~\cite{blue,blue}.
+  \item Consider that the user might stay many hours (and we often do)
+  at a text editor, and that the colors and formats involved are
+  prone to entail a considerable effect in the body and mental
+  activity.
+  \item Syntax sonification: consider mapping the textual structures
+  to sound (e.g. counting of tokens or context as inside a function or
+  class or over or around a variable of conditional or loop.
+  E.g. through MASSA.
+  \item Mode for tuning syntax highlighting with keyboard shortcuts
+  are used to tune individual colors from background and foreground
+  colors and groups.
+  \item How would these techniques be used for coloring Toki Pona?
+  Coloring of keywords that reflects the dictionary classification
+  of words are not in a power law distribution.
+  Maybe better described by a half-normal or by
+  overlapping normal distributions over the three peaks ($\sim35-58,
+  12-18, 1-5 occurrences$).
+  \item Coloring for poems: division by syllables, words, POS tag,
+    total number of syllables or letters of words.
+\end{itemize}
+
+\url{http://pygments.org/docs/}
+
+\subsection{Wordnet Toki Pona}\label{wn}
+exception: jan (people)
+is a hyponym of soweli (mammal),
+kili is hyponym of kasi,
+(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
+But otherwise such relations hold:
+e.g. luka (as hand) is e meronym of sijelo (body).
+
+For a first achievement of a Toki Pona Wordnet,
+each of the Toki Pona words in the dictionary
+were related to Wordnet synsets
+through the English lemmas.
+All synsets related to the lemma with the POS tag
+given in the dictionary are added.
+Another immediate possibility would be to
+relate each Toki Pona word to the synsets
+related to all the words bounded by a semicolon.
+Another possibility, which I find more consistent
+but also more controversial,
+is to choose all synsets related to the Wnglish
+words given that the Toki Pona words can be used
+in many roles.
+WordNet only contains nouns, adjectives, verbs, and adverbs.
+Thus, particles and prepositions were not considered.
+Numbers were considered adjectives.
+Words presented as adjectives in the dictionary
+were considered adjectives and adverbs.
+
+The total number of different synsets is 2493 in a very conservative
+counting: discarding prepositions and particles,
+only considering the synsets which have the POS tag
+given in the dictionary.
+If all the synsets (that have any POS tag)
+related to a lemma are considered, 
+then Toki Pona vocabulary reaches 3961 synsets.
+The prepositions are particularly missing in these
+countings because e.g. tawa or lon are used also
+as ordinary words.
+If synsets for them are considered those of the English
+lemmas with any POS tag, the vocabulary reaches 4052
+synsets.
+The total number of synsets in Wordnet is X.
+Total number of synsets counting neighbors.
+Depth of TP synsets.
+Using such features to devise other conlangs.
+
 \subsection{Hacks from other people}
@@
     be related to more then one synset of the English language?
+    An immediate implementation would be linking each Toki Pona word
+    to the English words in the official vocabulary.
+    A Toki Pona synset would be related to all the synsets of each English
+    words regarded as lemmas (sequence of characters or sound).
+  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
+    of words and text.
+    Enhancements of such translators would use POS tagging and templates
+    in English and Toki Pona and/or corpus and further statistical methods.
+    An efficiency of Toki Pona can be studied in terms of the coverage of
+    the Wordnet cloud.
+  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
+    (conlanging), and insights derived from Toki Pona might be used for that end.
   \item Understand how the corpus is gathered in \url{tokipona.net}.
@@
     I can start a draft,
-    they might learn the language in a few hours (with or without my help),
+    qhey might learn the language in a few hours (with or without my help),
     to contribute, and we can write a short paper.
@@
     further techniques from Natural Language Processing~\cite{POS}.
+  \item We are particularly interested in software tools,
+    and one possibility is to use a conlang to describe software
+    routines.
+    In this sense, it seems profitable
+    to understand what `non-ambiguous' means in Lojban,
+    and in what sense one is able to compile and parse Lojban.
+    This study will probably yield considerations about
+    parsing Toki Pona files.
+  \item Conlang idealized for describing data, programs,
+    scientific writings, creative writing
+    (exploring the thinking process).
+    Different modes that might be specified by a section header or tag.
 \end{itemize}
@@
 \subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
-Vim user community.
+FAPESP (project 2017/05838-3);
+Vim, Python, and Toki Pona authors, developers and literature maintainers;
+user and speaker communities.
 
@@
 \section{My usage of Toki Pona}\label{mytoki}
+
 I use the standard sounds, but often use [z] for s.
@@
 
-\input{pos}
+pali moku is addiction.
+noka and ampa; luka, suno, sike and lawa;
+I tend to group these things, as pali and pana.
+These clusters of meaning hint me that an even simpler
+language is possible and maybe more optimized to the
+core goals of Toki Pona.
+Also, ambiguities introduced by omission of li
+and absence of a token to denote preposition,
+suggest to me the possibility of a syntax that is uniquely
+parsed.
+
+I am tending to think on a language with the same syntax of toki pona:
+<sentence> la <sentence = subject + predicate + object + preposition complement>
+but always using the syntax separators,
+repeating them when the function in performed in the smallest scale:
+'(toki pi toki pona) pi jan sona' (language in toki pona and of intellectuals) vs
+'toki pi (toki pona pi pi jan sona)' (language in toki pona of intelectuals).
+The keyword for preposition complement is the catch.
+An 'a' seems good, but it is taken in Toki Pona.
+Maybe use it as the preposition particle and use ha or he
+instead of Toki Pona a.
+In such a setting,
+one might enable pi li e a (concept-qualifier, subject-predicate,
+predicate-object, concept-predicate)
+in any of the terms of the template:
+<sentence> la <sentence = subject + predicate + object + preposition complement>
+Between sentences one would use:
+la pi, la li, la e, la a, and la la as normally.
+This is a very fractal conlang proposal.
+Maybe have a way to distinct between noun, qualifiers (adjectives/adverbs) and verb terms,
+and accept any of them for subject, predicate, object, predicate,
+slots. Or assume a part of speech for template slot or for first word in slot.
+
+Because toki pona's vocabulary is so small,
+one might make variations of it for languages
+in specific.
+For example,
+suno might be sun in an English variant,
+and sol in a Portuguese variant.
+In Appendix~\ref{enPo} are an English and Portuguese
+proposition of such variants.
+A naive version might be obtained through
+choosing the first word in the description of
+an official vocabulary, as they exist in English and French.
+
+mute, mute ike,
+something can be good, numbered,
+or numbered in a bad way, too much,
+as in 'o awen, sina pali mute, lape lili'.
+
+% \input{pos}
 
@@
     \url{http://tokipona.wikia.com}
+\bibitem{conlanWikip}
+  Constructed language.
+(2017, November 1).
+In Wikipedia, The Free Encyclopedia.
+Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
 \end{thebibliography}

### 2017-11-19 article/article.tex
adiantada escrita e scripts de analise

@@
 [p t k s l] or [b d g z \textipa{\!R}].
+Especially for poetry, one might consider
+j and w to be compound vowels
+(e.g. j is 'i' and w is 'u'),
+or as consonants.
 
@@
 Non word-initial syllables must follow the pattern CV(N).
-The following sequences are forbidden: ji, wu, wo, ti, mn, nm, mm,
-nn.
+The following sequences are forbidden: ji, wu, wo, ti, nm, nn.
 
@@
 \subsection{Statistics of the vocabulary}
-In~\cite{pije} there is statistics about Toki Pona corpus.
+In~\cite{pije} are statistics about Toki Pona corpus.
 This section focuses on the statistics of the vocabulary
@@
 and punctuation.
-They include X nouns, Y verbs, Z adjectives,
-W prepositions, Y particles, K pre-verbs.
-Of which one might distinguish only between
-the particles and the other J words that
-might be used anywhere a particle will not.
+Table~\ref{tab:pos} shows the number of words
+related to each POS tag specified in the dictionary.
 
-Most often vowels,
-consonants, consonant vs vowel.
-Most often phonemes in general an specific positions..
-Possible phonemes given the rules.
-Comparison of the occurring against the possible.
+\input{pos}
 
-Syntactic structures:
+Vowel frequencies are as shown in Table~\ref{tab:vowels}
+for starting, end and internal position.
+Consonants are shown in Table~\ref{tab:cons}
+for meaningful values.
+Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
+The high incidences of 'l' and 'a' favor
+the naturalness of the language prosody
+as it resembles lallation.
+
+Most often phonemes are in
+Table~\ref{tab:phon}.
+A toolbox for obtaining such statistics and sets
+of words is in~\cite{tokipona}.
+
+Given the rules given in Section~\ref{phonology}
+(of 15 letters, 5 vowels, (C)V(N) phonemes,
+forbidden ji, wu, wo, ti, nn, nm),
+X words are possible with 1 syllable,
+Y with 2 syllables, Z with 3 syllables
+and W with 4 syllables.
+These are shown in Table~\ref{tab:pw}
+with fraction of the space actually used in Toki Pona.
+
+Possible syntactic structures are quantified
+in Table~\ref{tab:syn} for
 Possibilities on noun, verb and prepositional phrases
 and in sentences.
-Counting along the number of words.
+(Counting along the number of words.)
+
+% number of possible syllables
+% number of possible words of size X (1, 2, 3, 4....).
 
 \subsection{Synthesis of text}
-number of elisoes
+Such counting exercise is also useful
+for (semi-)automate writing through scripting.
+The syntax organizes the words in larger structures.
+The rhymes are restricted,
+the ending is as a tonality in music.
+
+There are some specific tasks for achieving texts,
+such as finding the number of syllables considering the elisoes,
+or handling interaction of the writer with the script
+to choose sentences or verses or stanzas.
 
@@
 
-% number of possible syllables
-% number of possible words of size X (1, 2, 3, 4....).
+Figure~\ref{fig:syn} holds some synthesized texts
+with different coloring schemes for syntax highlighting,
+as presented in the next section.
 

### 2017-11-19 article/pos.tex
adiantada escrita e scripts de analise

@@
 NOUN & 58  & 49 \\\hline
-ADJECTIVE & 40  & 35 \\\hline
-VERB & 18  & 13 \\\hline
+ADJECTIVE & 40  & 37 \\\hline
+VERB & 15  & 15 \\\hline
 PARTICLE & 12  & 12 \\\hline
-PRE & 5  & 5 \\\hline
 PREPOSITION & 5  & 5 \\\hline
 NUMBER & 4  & 1 \\\hline
-total & 142  & 120 \\\hline
+PRE & 0  & 0 \\\hline
+total & 134  & 120 \\\hline
 \end{tabular}\end{center}

### 2017-11-19 tokipona/makeStatistics.py

@@ -1,0 +1,287 @@
+# import ezodf
+import pandas as pd
+import numpy as n
+import os
+# from ODSReader import *
+
+# ods = ezodf.newdoc(doctype='ods',filename='../data/dictionary.ods')
+
+# ezodf.config.set_table_expand_strategy('all')
+
+# spreadsheet = ezodf.opendoc('../data/dictionary.ods')
+
+
+os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+xl_file = pd.ExcelFile('../data/dictionary.xlsx')
+dfs = {sheet_name: xl_file.parse(sheet_name)
+       for sheet_name in xl_file.sheet_names}
+df = dfs['Sheet1']
+table = n.vstack((df.keys(), df.values))
+
+all = set()
+for row in table:
+    classes = " ".join(i for i in row[1].split() if i.isupper())
+    print(row[0], classes)
+    all.add(classes)
+all2 = set()
+
+colors = {'ADJECTIVE': 'Comment', 'NOUN': 'Constant', 'NUMBER': 'Identifier',
+          'PARTICLE': 'Statement', 'PRE': 'PreProc',
+          'PREPOSITION': 'Type', 'VERB': 'Special'}
+words = {}
+for color in colors:
+    words[color] = set()
+words_all = words.copy()
+
+synonyms = 0
+precedence = 'first'
+precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
+allp = list(colors.keys())
+precedence += [i for i in allp if i not in precedence]
+
+class_count = colors.copy()
+for i in class_count: class_count[i] = 0
+class_count_ = colors.copy()
+for i in class_count_: class_count_[i] = 0
+
+word_classes = {}
+for row in table:
+    # Better choose the class
+    classes = [i for i in row[1].split() if i.isupper() and i not in ("I,","O!")]
+    if precedence=="first":
+        class_ = classes[0]
+    else:
+        class__ = classes
+        not_chosen = 1
+        for pos in precedence:
+            if pos in class__ and not_chosen:
+                class_ = pos
+                class_count[pos] += 1
+                class_count_[pos] += 1
+                not_chosen = 0
+            elif pos in class__:
+                class_count[pos] += 1
+
+    print(row[0], class_, colors[class_])
+    all2.add(class_)
+    words_ = row[0].split()
+    if "PRE-VERB" in classes:
[…]

### 2017-11-19 tokipona/makeWordnet.py

@@ -1,0 +1,87 @@
+# import ezodf
+import pandas as pd
+import numpy as n
+import os
+
+# to make tp wn, I need to know, for each word,
+# the English lemmas and pos of each lemma.
+# words[word] = [(lemma1, pos1), (lemma2, pos2),...]
+
+os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+xl_file = pd.ExcelFile('../data/dictionary.xlsx')
+dfs = {sheet_name: xl_file.parse(sheet_name)
+       for sheet_name in xl_file.sheet_names}
+df = dfs['Sheet1']
+table = n.vstack((df.keys(), df.values))
+
+words = {}
+
+for row in table:
+    all_pos = row[1].split("\n")
+    lemma_pos = []
+    for pos in all_pos:
+        class_ = [i for i in pos.split() if i.isupper() and i not in ("I,","O!")][0]
+        en_words = [i for i in pos.replace(" or ", ' ').split() if i.islower() or i in ("I,","O!")]
+        for enw in en_words:
+            w = enw.replace(";", "").replace(',', '')
+            lemma_pos.append( (w, class_) )
+
+    words_ = row[0].split()
+    if 'or' in words_:
+        words_.remove("or")
+    for w in words_:
+        words[w] = lemma_pos
+
+pos = {'NOUN': 'n',
+        'VERB': 'v',
+        'PRE-VERB': 'v', # pre-verbs are also verbs
+        'ADJECTIVE': 'asr', # adjectives are also adverbs and satellite adjectives
+        'NUMBER': 'asr'
+        }
+
+import nltk as k
+tp_wordnet = {}
+tp_wordnet_ = {}
+tp_wordnet__ = {}
+for word in words:
+    # if word in ('li', 'e', 'la', 'pi'):
+    #     continue
+    synsets = []
+    synsets_ = []
+    synsets__ = []
+    for enw in words[word]:
+        class_ = enw[1]
+        if class_ in ('PREPOSITION'):
+            w = enw[0]
+            ss__ = k.wordnet.wordnet.synsets(w)
+            synsets__ += ss__
+        if class_ in ('PARTICLE', 'PREPOSITION'):
+            continue
+        w = enw[0]
+        c = pos[class_]
+        ss = k.wordnet.wordnet.synsets(w, pos=c)
+        synsets += ss
[…]

### 2017-11-24 article/article.tex
andando com as estatisticas

@@
 % pictorial forms
-\section{Software for analysis and synthesis and for syntax highlighting}\label{hacks}
+\section{Software for analysis, synthesis, and syntax highlighting}\label{hacks}
 In this section,
@@
 As described in Section~\cite{basics},
-there is only 14 letters,
+there are only 14 letters,
 and phonemes respect a few rules.
@@
 4 of them having synonyms.
-A total of 124 tokens.
-Not counting proper nouns (names)
+A total of 124 tokens,
+not counting proper nouns (names)
 and punctuation.
@@
 \input{pos}
+
+From the official 124 words, 
+26 of them (20.97\%)
+have only one syllable,
+85 (68.55\%) have two syllables,
+and 13 (10.48\%) have three syllables.
+No official words has four or more syllables.
+Of all the 235 syllables of the official words,
+li, la, ka, na, and pa are the most often (with 5.53\%,
+4.26\%, 3.83\%,  3.83\%, 3.83\%).
+On the first syllable of each word,
+a, o, pi, ka, and la are the most often
+(with 6.45\%, 4.03\%, 4.03\%, 3.23\%, 3.23\%).
+On the last syllable of each word,
+li, lo, na, la, and ma are the most often
+(with 8.06\%, 4.84\%, 4.84\%, 4.03\%, 4.03\%).
+Middle-word syllables only occur 13 times,
+and are all different, with the exception of `la',
+which occurs twice.
+For the frequencies of all syllables in every position,
+run \tttt{makeStatistics.py} in~\cite{tokipona}.
 
@@
 as it resembles lallation.
+
+\input{vowels}
 

### 2017-11-24 article/vowels.tex
andando com as estatisticas

\begin{table*}[h!]
\begin{center}
\caption{"Frequency of vowels in Toki Pona.}
\begin{tabular}{| l | c |}\hline
Vowel & freq \\\hline
a & 33.19 \\\hline
e & 17.45 \\\hline
i & 23.40 \\\hline
o & 15.32 \\\hline
u & 10.64 \\\hline
\end{tabular}\end{center}
\end{table*}

### 2017-11-24 tokipona/makeStatistics.py

@@ -185,11 +185,11 @@ consonants = 'jklmnpstw'
 vowels_ = ''.join(v for v in tpwords_ if v in vowels)
 consonants_ = ''.join(v for v in tpwords_ if v in consonants)
 
-fracv = (tpwords_.count(i)/len(tpwords_) for i in vowels)
-fracv_ = (vowels_.count(i)/len(vowels_) for i in vowels)
+fracv =  [100*tpwords_.count(i)/len(tpwords_) for i in vowels]
+fracv_ = [100*vowels_.count(i)/len(vowels_) for i in vowels]
 
-fracc = (tpwords_.count(i)/len(tpwords_) for i in consonants)
-fracc_ = (consonants_.count(i)/len(consonants_) for i in consonants)
+fracc =  [100*tpwords_.count(i)/len(tpwords_) for i in consonants]
+fracc_ = [100*consonants_.count(i)/len(consonants_) for i in consonants]
 # overall, only vowels, only consonants, begin, end, middle
 
 firstl = ''.join(i[0] for i in tpwords)
@@ -216,9 +216,6 @@ all_data = []
 for ls in letter_sets:
     data = statMe(ls)
     all_data += data
-
-labels = vowels + consonants
-labels = [i for i in labels]
 
 def firstSyllable(token):
     if len(token) <= 2:
@@ -268,7 +265,9 @@ def histSyl(syllables_):
     syllables__.sort(key=len)
 
     freq_syl = [100*syllables_.count(i)/len(syllables_) for i in syllables__]
-    return list(zip(syllables__, freq_syl))
+    aa = list(zip(syllables__, freq_syl))
+    bb = sorted(aa, key=lambda t: -t[1])
+    return bb
 
 syls = [syllables_, syllables_0, syllables_1, syllables_i]
 hsyls = [histSyl(i) for i in syls]
@@ -279,6 +278,21 @@ hlsyl_ = [lsyl.count(i) for i in (1, 2, 3)]
 hlsyl__ = [[w for w in tpwords if len(getSyllables(w)) == i] for i in (1, 2, 3)]
 
 
+def v(syl):
+    for i in syl:
+        if i in 'aeiou':
+            return i
+labels = vowels + consonants
+labels = [i for i in labels]
+fracv0_ = [v(i) for i in syllables_0]
+fracv1_ = [v(i) for i in syllables_1]
+p.mediaRendering.tables.writeTex(
+        p.mediaRendering.tables.encapsulateTable(
[…]

### 2017-12-01 article/article.tex
encaminhada tabela de frequencia de letras

@@
 acquire new insights.
-Furthermore, it is a solid reference of the syntax and phonological
+Furthermore, it is a solid reference of the phonological and syntax
 rules.
@@
       \item Nasal: m (labial), n (coronal).
-      \item Plosive: k (dorsal), p (labial), t (coronal).
+      \item Plosive: p (labial), t (coronal), k (dorsal).
       \item Fricative: s (coronal).
-      \item Approximant: j (dorsal), l (coronal), w (labial).
+      \item Approximant: w (labial), l (coronal), j (dorsal).
     \end{itemize}
@@
 Especially for poetry, one might consider
-j and w to be compound vowels
-(e.g. j is 'i' and w is 'u'),
-or as consonants.
+j and w to be vowels
+(e.g. j as 'i' and w as 'u').
 
@@
 
-Vowel frequencies are as shown in Table~\ref{tab:vowels}
+Vowel frequencies are as shown in Table~\ref{freqLet}
 for starting, end and internal position.
@@
 
-\subsection{Syntax highlighting}
+\subsection{Syntax highlighting}\label{shigh}
 The same package~\cite{tokipona}
@@
 \begin{itemize}
+  \item Make a syntax highlighting that colors the tokens with respect
+    to the syntatic position and function, and not in a fixed manner
+    as is implemented and exposed in Section~\ref{shigh}.
   \item Relate Toki Pona to Wordnet: should one Toki Pona word

### 2017-12-01 tokipona/makeStatistics.py

@@ -48,6 +48,8 @@ word_classes = {}
 for row in table:
     # Better choose the class
     classes = [i for i in row[1].split() if i.isupper() and i not in ("I,","O!")]
+    classes = ["PRE" if "PRE-VERB" in i else i for i in classes]
+    print(classes)
     if precedence=="first":
         class_ = classes[0]
     else:
@@ -286,13 +288,39 @@ labels = vowels + consonants
 labels = [i for i in labels]
 fracv0_ = [v(i) for i in syllables_0]
 fracv1_ = [v(i) for i in syllables_1]
+l0 = ['freq', 'v', 'c', 'freq\\_I', 'v', 'c', 'freq\\_L', 'v', 'c', 'freq\\_M', 'v', 'c']
+l_ = [[i[j] for i in all_data] for j in range(14)]
+ll = [l0]+l_
 p.mediaRendering.tables.writeTex(
         p.mediaRendering.tables.encapsulateTable(
             p.mediaRendering.tables.makeTabular(
-                ["Vowel"]+[i for i in vowels], 
-                [[i] for i in ["freq"]+fracv_], True),
-            """Frequency of vowels in Toki Pona."""),
+                ["letter"]+[i for i in vowels]+[i for i in consonants], 
+                ll,
+                True),
+            """Frequency of letters in Toki Pona.
+            freq, freq\\_I, freq\\_L and freq\\_M are
+            the frequencies of the letters in any, initial, last and middle
+            positions.
+            The columns 'v' and 'c' that follow them are frequencies
+            considering only vowels and consonants.
+            The most frequent vowel is 'a' in any position,
+            although it is more salient among words starting with a vowel
+            and among the last letter of the words.
+            For starting, ending and middle positions, the second most frequent
+            vowel varies.
+            Among the consonants, 'n' is the most frequent because it is
+            the only consonant allowed in the last position and because
+            almost 20\\% of the words end with 'n'.
+            On the initial position, 's' is the most frequent consonant,
+            while in middle position 'l' is the most frequent consonant.
+            Many other conclusions can be drawn from this table and are
+            useful e.g. for exploring sonorities in poems.""",
+            'freqLet'),
         "../article/vowels.tex")
+p.mediaRendering.tables.fontSize("../article/vowels.tex", write=1)
+p.mediaRendering.tables.doubleLines("../article/vowels.tex",
[…]

### 2017-12-01 article/article.tex
andando com stats

@@
 
-\subsection{Statistics of the vocabulary}
+\subsection{Statistics of the vocabulary}\label{sec:stat}
 In~\cite{pije} are statistics about Toki Pona corpus.
@@
 
-Vowel frequencies are as shown in Table~\ref{freqLet}
+Vowel and consonant frequencies are as shown in Table~\ref{freqLet}
 for starting, end and internal position.
-Consonants are shown in Table~\ref{tab:cons}
-for meaningful values.
-Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
-The high incidences of 'l' and 'a' favor
+Comparison of vowels against consonant is in Table~\ref{tab:vcon}.
+The high incidences e.g. of 'l' and 'a' favor
 the naturalness of the language prosody
-as it resembles lallation.
+as it resembles child speech.
 
@@
 
-Most often phonemes are in
-Table~\ref{tab:phon}.
-A toolbox for obtaining such statistics and sets
-of words is in~\cite{tokipona}.
+26, 85 and 13 words have one, two and three syllables
+(20.96, 68.54 and 10.48\% of all 124 tokens).
+No official word have more than three syllables
+(although proper names may be of any size).
+In all the 124 tokens, there are 235 syllables (68 different).
+Most often syllables are in
+Table~\ref{syl}.
+A complete list of the syllables and their frequencies (in different
+positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
+script.
+A list of all words, grouped by their size in syllables,
+and ordered alphabetically and by the number of letters,
+is in \tttt{hlsyl\_\_}.
+
+\input{syls}
 
 Given the rules given in Section~\ref{phonology}
-(of 15 letters, 5 vowels, (C)V(N) phonemes,
+(of 14 letters, 5 vowels, (C)V(N) phonemes,
 forbidden ji, wu, wo, ti, nn, nm),
-X words are possible with 1 syllable,
-Y with 2 syllables, Z with 3 syllables
-and W with 4 syllables.
+46 words are possible with 1 syllable,
+1886 with 2 syllables, 77326 with 3 syllables.
+No middle syllable end with the nasal consonant 'n'.
 These are shown in Table~\ref{tab:pw}
@@
 (Counting along the number of words.)
+
+A toolbox for obtaining such statistics and sets
+of words is in~\cite{tokipona}.
 

### 2017-12-01 tokipona/makeStatistics.py

@@ -219,6 +219,43 @@ for ls in letter_sets:
     data = statMe(ls)
     all_data += data
 
+l0 = ['freq', 'v', 'c', 'freq\\_I', 'v', 'c', 'freq\\_L', 'v', 'c', 'freq\\_M', 'v', 'c']
+l_ = [[i[j] for i in all_data] for j in range(14)]
+ll = [l0]+l_
+p.mediaRendering.tables.writeTex(
+        p.mediaRendering.tables.encapsulateTable(
+            p.mediaRendering.tables.makeTabular(
+                ["letter"]+[i for i in vowels]+[i for i in consonants], 
+                ll,
+                True),
+            """Frequency of letters in Toki Pona.
+            freq, freq\\_I, freq\\_L and freq\\_M are
+            the frequencies of the letters in any, initial, last and middle
+            positions.
+            The columns 'v' and 'c' that follow them are frequencies
+            considering only vowels and consonants.
+            The most frequent vowel is 'a' in any position,
+            although it is more salient among words starting with a vowel
+            and among the last letter of the words.
+            For starting, ending and middle positions, the second most frequent
+            vowel varies.
+            Among the consonants, 'n' is the most frequent because it is
+            the only consonant allowed in the last position and because
+            almost 20\\% of the words end with 'n'.
+            On the initial position, 's' is the most frequent consonant,
+            while in middle position 'l' is the most frequent consonant.
+            Many other conclusions can be drawn from this table and are
+            useful e.g. for exploring sonorities in poems.""",
+            'freqLet'),
+        "../article/vowels.tex")
+p.mediaRendering.tables.fontSize("../article/vowels.tex", write=1)
+p.mediaRendering.tables.doubleLines("../article/vowels.tex",
+        hlines=[], vlines=[], hlines_=[0,2,3,4,5,7,8,10,11,13,14,15],
+        vlines_=[0,2,3,5,6,8,9,11,12,13])
+
+##########################
+# Syllables
+
 def firstSyllable(token):
     if len(token) <= 2:
         sy = token
@@ -257,18 +294,19 @@ def getSyllables(token):
 
 syllables = [getSyllables(i) for i in tpwords]
 syllables_ = [j for i in syllables for j in i]
+
+syllables__ = list(set(syllables_))
+syllables__.sort()
+syllables__.sort(key=len)
+
[…]

### 2017-12-01 article/article.tex
andando com stats

@@
 forbidden ji, wu, wo, ti, nn, nm),
-46 words are possible with 1 syllable,
-1886 with 2 syllables, 77326 with 3 syllables.
+96 words are possible with 1 syllable,
+ 96, 8256, 710016
+8256 with 2 syllables, 710016 with 3 syllables.
 No middle syllable end with the nasal consonant 'n'.

### 2017-12-01 tokipona/makeStatistics.py

@@ -362,20 +362,20 @@ p.mediaRendering.tables.doubleLines("../article/syls.tex",
 syllables_possible = set()
 for c in consonants:
     for v in vowels:
-        syllables_possible.append(c+v)
-        syllables_possible.append(c+v+'n')
+        syllables_possible.add(c+v)
+        syllables_possible.add(c+v+'n')
 forbiden = ['ji', 'wu', 'wo', 'ti', 'nn', 'nm']
 for i in forbiden[:-2]:
     syllables_possible.remove(i)
 
-syllables_start = [i for i in vowels] + [i+'n' for i in vowels] + syllables_possible
+syllables_start = set([i for i in vowels]) | set([i+'n' for i in vowels]) | syllables_possible
 
 words1 = syllables_start
-words2 = []
+words2 = set()
 for i in words1:
     for j in syllables_possible:
-        words2.append(i+j)
-words2_ = words2[:]
+        words2.add(i+j)
+words2_ = words2.copy()
 for i in forbiden:
     for j in words2:
         if i in j:
@@ -383,11 +383,11 @@ for i in forbiden:
             if j in words2_:
                 words2_.remove(j)
 
-words3 = []
+words3 = set()
 for i in words2:
     for j in syllables_possible:
-        words3.append(i+j)
-words3_ = words3[:]
+        words3.add(i+j)
+words3_ = words3.copy()
 for i in forbiden:
     for j in words3:
         if i in j:

### 2017-12-03 article/article.tex
adiantada sintese de textos

--- 2017-12-01 article/article.tex
@@
 
-\subsection{Syntax}
+\subsection{Syntax}\label{syntax}
 \subsubsection{Fundamental notions}
@@
 
+The Python scripts (or the toolbox)
+for obtaining all the results in this sections,
+and more, is publicly available in~\cite{tokipona},
+to facilitate both the inspection of the results and the
+generation of derivatives by other interested parties.
+
 \subsection{Statistics of the vocabulary}\label{sec:stat}
@@
 and the next paragraph is an overview.
-Python scripts were used to obtain the measurements
-and are available at~\cite{tokipona}.
+The script \tttt{makeStatistics.py}
+was used to obtain the all the measurements and tables
+discussed herein,
+and hold some other measurements and sets of words
+which were regarded as less suitable for this exposition.
 
@@
 For the frequencies of all syllables in every position,
-run \tttt{makeStatistics.py} in~\cite{tokipona}.
+run the script \tttt{makeStatistics.py} in~\cite{tokipona}.
 
@@
 forbidden ji, wu, wo, ti, nn, nm),
-46 words are possible with 1 syllable,
-1886 with 2 syllables, 77326 with 3 syllables.
+96 words are possible with 1 syllable,
+8256 with 2 syllables, 710016 with 3 syllables.
 No middle syllable end with the nasal consonant 'n'.
-These are shown in Table~\ref{tab:pw}
-with fraction of the space actually used in Toki Pona.
-
-Possible syntactic structures are quantified
-in Table~\ref{tab:syn} for
-Possibilities on noun, verb and prepositional phrases
-and in sentences.
-(Counting along the number of words.)
-
-A toolbox for obtaining such statistics and sets
-of words is in~\cite{tokipona}.
-
-% number of possible syllables
-% number of possible words of size X (1, 2, 3, 4....).
-
-\subsection{Synthesis of text}
-Such counting exercise is also useful
-for (semi-)automate writing through scripting.
+
+Possible syntactic structures are unlimited.
+For a noun or verb phrase, one might quantify the possibilities
+by considering all words but the particles that are nothing else
+(i.e. li, e, la, pi, a, o, anu, en, seme, mu).
+This yields $120-10=110 words$.
+As discussed in Section~\ref{syntax},
+I advocate that, although such words are classified in the dictionary,
+they might be used indistinctly as nouns, adjectives,
+verbs and adverbs.
+Thus, one have 110 possibilities of noun and verb phrases with
+one word, $110^2=12100$ possibilities with two words,
+$110^3=1331000$ with three words and so on.
+Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
+(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
+and assume that the sentence has at most one prepositional phrase,
+one might quantify the possibilities using the formula:
+$\delta = 110^n\times 110^v\times 10^o \times 5\times 110^p$,
+where $5$ stands for the possible prepositions.
+To account for the particles, one possibility is
+to assume for simplicity that one might use none or one particle
+(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
+at each phrase, yielding:
+$\delta = 110^n\times 110^v\times 110^o \times 5\times 110^p\times 9^4$.
+For example, assume $n=v=p=o=1$, then
+$\delta=110^1\times 110^1\times 110^1 \times 5\times 110^1\times 9^4=
+4.80298^{12}$
+i.e. almost 5 trillion possible
+sentences with single word phrases (i.e. no adjectives or adverbs),
+while allowing only one particle per phrase
+and only one predicate phrase 
+(e.g. 'ona li moku e soweli lon supa').
+The general case is implemented as a function
+in the script \tttt{makeStatistics.py}~\cite{tokipona}
+
+\subsection{Synthesis of text}\label{synth}
+Such counting exercises is also useful
+for (semi-)automated writing through scripting.
 The syntax organizes the words in larger structures.
-The rhymes are restricted,
-the ending is as a tonality in music.
-
+The rhymes are restricted and sonorities are
+restricted by the small vocabulary and simple syntax.
 There are some specific tasks for achieving texts,
-such as finding the number of syllables considering the elisoes,
+such as finding the number of syllables considering the elisions,
 or handling interaction of the writer with the script
@@
 The same package~\cite{tokipona}
-has capabilities for synthesizing text.
+has capabilities for synthesizing Toki Pona text.
 Noun, verb and prepositional phrases,
@@
 as presented in the next section.
+The textual synthesis described here,
+and implemented in \tttt{makeTexts.py}
+might be enhanced as described in Section~\ref{conc}.
 
@@
 \begin{itemize}
+  \item Further develop the text synthesis facilities described in
+    Section~\ref{synth} might be enhanced by further employing e.g.
+    rhythm, meter, rhyme and form techniques~\cite{wikipPoetry}.
   \item Make a syntax highlighting that colors the tokens with respect
@@
 Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
+\bibitem{wikiPoetry}
+Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
+    Retrieved 19:46, December 3, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
 \end{thebibliography}

### 2017-12-03 tokipona/makeStatistics.py

@@ -400,3 +400,14 @@ nwords = [len(i) for i in (words1, words2, words3)]
 # largest in letters
 # walk through levenshtein distance
 # 
+##########################################
+# Possible sentences
+
+def possibleSentences(n, v, o, p=[1], nparticles=1):
+    p_ = 1
+    for pp in p:
+        p_ *= 5 * 110**pp
+
+    delta = 110**n * 110**v * 110**o * p_ * 10^3
+    return delta 
+

### 2017-12-03 tokipona/makeTexts.py

@@ -1,0 +1,157 @@
+import pandas as pd
+import numpy as n
+import os
+# from makeStatistics import getSyllables
+
+vowels = 'aeiou'
+consonants = 'jklmnpstw'
+
+os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+xl_file = pd.ExcelFile('../data/dictionary.xlsx')
+dfs = {sheet_name: xl_file.parse(sheet_name)
+       for sheet_name in xl_file.sheet_names}
+df = dfs['Sheet1']
+table = n.vstack((df.keys(), df.values))
+
+all = set()
+for row in table:
+    classes = " ".join(i for i in row[1].split() if i.isupper())
+    print(row[0], classes)
+    all.add(classes)
+all2 = set()
+
+colors = {'ADJECTIVE': 'Comment', 'NOUN': 'Constant', 'NUMBER': 'Identifier',
+          'PARTICLE': 'Statement', 'PRE': 'PreProc',
+          'PREPOSITION': 'Type', 'VERB': 'Special'}
+words = {}
+for color in colors:
+    words[color] = set()
+words_all = words.copy()
+
+synonyms = 0
+precedence = 'first'
+precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
+allp = list(colors.keys())
+precedence += [i for i in allp if i not in precedence]
+
+class_count = colors.copy()
+for i in class_count: class_count[i] = 0
+class_count_ = colors.copy()
+for i in class_count_: class_count_[i] = 0
+
+word_classes = {}
+for row in table:
+    # Better choose the class
+    classes = [i for i in row[1].split() if i.isupper() and i not in ("I,","O!")]
+    classes = ["PRE" if "PRE-VERB" in i else i for i in classes]
+    print(classes)
+    if precedence=="first":
+        class_ = classes[0]
+    else:
+        class__ = classes
+        not_chosen = 1
+        for pos in precedence:
+            if pos in class__ and not_chosen:
+                class_ = pos
+                class_count[pos] += 1
+                class_count_[pos] += 1
+                not_chosen = 0
+            elif pos in class__:
+                class_count[pos] += 1
+
+    print(row[0], class_, colors[class_])
+    all2.add(class_)
+    words_ = row[0].split()
+    if "PRE-VERB" in classes:
+        classes[classes.index("PRE-VERB")] = "PRE"
[…]

### 2017-12-03 article/article.tex
adiantada a sintese de textos

@@
 by considering all words but the particles that are nothing else
-(i.e. li, e, la, pi, a, o, anu, en, seme, mu).
-This yields $120-10=110 words$.
+(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
+and the presotions that are nothing else (i.e. kepeken, lon, tan).
+This yields $120-13=107 words$.
 As discussed in Section~\ref{syntax},
@@
 verbs and adverbs.
-Thus, one have 110 possibilities of noun and verb phrases with
-one word, $110^2=12100$ possibilities with two words,
-$110^3=1331000$ with three words and so on.
+Thus, one have 107 possibilities of noun and verb phrases with
+one word, $107^2=11449$ possibilities with two words,
+$107^3=1225043$ with three words and so on.
 Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
@@
 one might quantify the possibilities using the formula:
-$\delta = 110^n\times 110^v\times 10^o \times 5\times 110^p$,
+$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
 where $5$ stands for the possible prepositions.
@@
 at each phrase, yielding:
-$\delta = 110^n\times 110^v\times 110^o \times 5\times 110^p\times 9^4$.
-For example, assume $n=v=p=o=1$, then
-$\delta=110^1\times 110^1\times 110^1 \times 5\times 110^1\times 9^4=
-4.80298^{12}$
-i.e. almost 5 trillion possible
+$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
+For example, assume $n=v=o=p=1$, then
+$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
+ 4.300066\times 10^+12$
+i.e. more then 4 trillion possible
 sentences with single word phrases (i.e. no adjectives or adverbs),

### 2017-12-03 tokipona/makeTexts.py

@@ -114,7 +114,8 @@ def getSyllables(token):
         syl.append(sy)
     return syl
 
-def createPhrase(nsy=0):
+forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
+def createPhrase(nsy=0, vowel=False):
     """
     Create a Toki Pona phrase.
 
@@ -126,11 +127,16 @@ def createPhrase(nsy=0):
     nsy : integer
         The number of syllables of the phrase.
         If nsy=0, the phrase will have a random number of syllables
+    vowel : boolean
 
     Returns
     -------
     ph : string
-        A phrase of words in Toki Pona.
+        A phrase in Toki Pona.
+
+    See Also
+    --------
+    createSentence : creates a sentence in Toki Pona.
 
     Notes
     -----
@@ -150,8 +156,55 @@ def createPhrase(nsy=0):
         w = n.random.choice(plain_words)
         s = getSyllables(w)
         l = len(s)
+        if w[0] in vowels and vowel:  # elision
+            l -= 1
         if nsy_+l > nsy:
+            continue
+        if w in forbiden:
+            print(w)
             continue
         nsy_ += l
         phrase += w + ' '
+        if w[-1] in vowels:
+            vowel = 1
+        else:
+            vowel = 0
     return phrase[:-1]
+
+
+def createSentence(nsy=0, prep=1, n=0, v=0, o=0, p=0):
+    """
+    Create a Toki Pona sentence.
+
+    Such phrase will always have a subject, predicate
+    and object. Preposition is optional.
+
+    Parameters
+    ----------
+    nsy : integer
+        The number of syllables of the sentence.
+        If nsy=0, the phrase will have a random number of syllables
+
+    Returns
+    -------
+    se : string
+        A sentence in Toki Pona.
+
+    See Also
+    --------
+    createPhrase : creates a phrase in Toki Pona.
+
+    Notes
+    -----
+    See createPhrase to understand how the phrases are created.
+
+    References
+    ----------
+    .. [1] Fabbri, Renato, et al. "Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts." arXiv preprint arXiv:abs/XXX.XXXX (2017)
+
+    """
+    if nsy == 0:
+        nsy = n.random.randint(5,20)
+    nsy_ = 0
+    phrase = ''
+    while nsy_ < nsy:

### 2017-12-04 article/article.tex
updating vimrc and article

--- 2017-12-01 article/article.tex
@@
 \usepackage[utf8]{inputenc}
-\usepackage{tipa}
 \newcommand{\ttt}[1] {
@@
 \newcommand{\tttt}[1]{\texttt{#1}}
-% \usepackage{fontspec}
-% \usepackage[mathletters]{ucs}
-% \newcommand{\ttt}[1] {
-% 	\texttt{<#1>}}
-% \newcommand{\tttt}[1]{\texttt{#1}}
 
 \begin{document}
-\title{Basic concepts and tools for the Toki Pona minimalist language:
-Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
+\title{An anthropological account of the Vim editor:\\
+ouroboros, features and tweaks after 10 years of usage}
 % Anthropological computation
@@
 \begin{abstract}
-  A minimalist constructed language (conlang)
-  is useful for experiments and is comfortable for making tools.
-  The Toki Pona conlang is minimalist both in the vocabulary
-  (only 14 letters, X phonemes, and 120 words + 4 synonyms)
-  and in the $\approx10$ syntax rules.
-  The language is useful for being a used and somewhat established
-  minimalist conlang with at least hundreds of speakers.
-  In this article, we describe current concepts and resources
-  for Toki Pona,
-  and make available Python scripted routines for
-  analysis and synthesis of texts and the specification of syntax highlighting schemes.
-  We focus on the analysis of the basic vocabulary,
-  as corpus analyses were found in~\cite{corpus}.
-  The synthesis is based on the basic sentence template,
-  relates to context by keeping track of used words,
-  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
-  and number of sentences, words and letters (e.g. for paragraphs).
-  Syntax highlighting is performed in 24bit true colors, currently only
-  with the Vim text editor.
-  Word colors reflect morphosyntactic classes given in the official dictionary
-  and different solutions to necessary and arbitrary choices are described and implemented.
-  In summary, this text holds potentially novel conceptualizations about,
-  and tools and results in analyzing, synthesizing or syntax highlighting the
-  Toki Pona language.
-  % subj (li), obj (e), prepositions, qualification (word concatenation), 
-  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
+The Vim editor is very rich in capabilities
+and thus complex.
+This article is a description of the Vim text editor
+and a set of considerations on its usage and design.
+It is the result of more than ten years of experience
+with Vim for writing and editing various types of documents,
+e.g.:
+Python, C++, JavaScript, ChucK, programs;
+and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
+binary files.
+It is said that it takes about ten years to master (or start mastering)
+this text editor, and I find that other experienced users
+have a different view of Vim and that they use a different
+set of features.
+Therefore, this document exposes my insights in order
+to confront my usage with that of other Vim users.
+Another goal is
+to make available a reference document with which new users
+can grasp a sound overview by reading it and the discussions that
+it might generate.
+Also, it should be useful for users of any degree of experience,
+including me, as a compendium of commands, namespaces and tweaks.
+Upon feedback, and maturing of my Vim usage,
+this document might be enhanced or receive additional
+material.
 \end{abstract}
-{\bf keywords:} Natural Language Processing, Textual synthesis, Syntax highlighting, Wordnet, Toki Pona
 
 \section{Introduction}\label{intro}
-Toki Pona is a minimalist conlang (constructed language)
-with only 124 words (120 without the synonyms).
-Therefore, the concepts are usually very general and different,
-and, without context, the words are rarely related through
-meronymy and hyponymy.
-Such a linguistic setting is desired because of the simplicity
-which entails easier e.g. learning and tool making.
-Another reason why the minimalistic language design is compelling
-is the study and harnessing of the strong and weak forms of
-the Sapir-Whorf hypothesis (linguistic relativity),
-i.e. that language influences or dictates ones thought
-and world experience.
-Accordingly, one uses a conlang as a thinking tool (or platform)
-or to make experiments about the influence language has on
-the thoughts of the `speaker' (also writer and reader).
-Toki Pona is often described as a tool to meditate,
-to simplify the thinking processes, and as a way to modify
-the mood and impressions about the world.~\cite{tp1,tp2,tp3,tp4}
-In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
-describes that she has seen the language been used successfuly
-in the context of management, creation of texts, legal texts,
-etc.
-
-In this article, we present a conceptual overview of the language,
-which is fit both to the newcomer and to the expert.
-We also summarize
-texts and software gadgets
-for Toki Pona that
-were developed by us and by other Toki Pona users.
-Most importantly, the conceptual overview is considerably different
-from what we found in the literature~\cite{pu,pije,memrise},
-with emphasis on simplicity and flexibility,
-and the software routines yield analyses, syntheses
-and highlighting Toki Pona what was thought to be complementary
-to the software currently available.
-
-Next subsections hold a description of the general resources available,
-a historical note, and some words about natural and constructed languages.
-Section~\ref{basic} describes Toki Pona's phonology and syntax.
-Section~\ref{hacks} presents the software routines we made available
-and immediate results, such as listings and statistics of words,
-poems and short stories, and coloring schemes.
-Conclusions and further work are in Section~\ref{conc}.
-Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
-with thoughts of rule breaking and potentially new conlangs.
-Appendix~\ref{ftp} holds final words in Toki Pona.
-
-% definition, historic, status (groups, speakers, etc).
-
-\subsection{Resources on Toki Pona}
-One might organize current resources for the Toki Pona
-language in: references and learning material, corpus,
-websites, interaction groups (here users talk a post texts and comments), and software gadgets.
-The main references of the language are:
-  the official book ``Toki Pona: The Language of Good'',
-    authored by Sonja Lang, the creator of the language;
-the online book ``o kama sona e toki pona!'',
-from jan Pije~\cite{kama}.
-For a more comprehensive view of the resources available
-for the user, we suggest following the links 
-from~\cite{wikiToki}.
-Section~\cite{third} holds a summary of what we found in software
-gadgets related to Toki Pona.
-
+Vim is a very complex editor,
+considered by the Linux community to be
+matched only by Emacs.
+They both are the standard advanced text editors
+of the free software and open source communities
+and have been developed for decades.
+This document describes the Vim editor
+and proposes a set of enhancements of the user
+experience through simple tweaks.
+The editor has a very mature documentation
+and the contents herein presented is a
+report on the overall understandings I
+have of Vim after a little bit more
+of ten years using it.
+The purposes of this document are:
+\begin{itemize}
+  \item to help new users in grasping Vim essentials
+  and convenient practices.
+  \item To attain a sound overall description of the editor.
+  \item To record the view of the editor that
+  a user (me) has after $10$ years of usage.
+  \item To confront my usage with that of other experienced
+  users. This is helpful for me, but also for the other users
+  as they might benefit from this content and from discussions
+  that might arise from it.
+  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
+\end{itemize}
+
+Advanced users might just skim throughj
+Section~\ref{basics} and consider more carefully
+Section~\ref{namespaces}.
+The concluding remarks and proposed enhancements
+might also deserve some attention.
+
+I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
+I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
+Eventually, I edited database files and other types os files.
+With Vim, I mostly write software (web and scientific),
+music, poems and short stories.
+It is very useful because:
+\begin{itemize}
+  \item it is meant to be a plain text (e.g. ascii, utf8) editor
+  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
+  \item It has a powerful architecture and set of commands.
+  \item It is highly configurable and most often the users
+  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
+\end{itemize}
+
+The standard capabilities of the editor
+should become clear in Section~\ref{basics}.
+Vim is constantly evolving and there are many plugins,
+some of them very popular for both general and specific
+users.
+Accordingly, there are many possible enhancements,
+and Section~\ref{conc} report the most prominent of them
+for me and potential workarounds made available as plugins.
+
+This document is written is a DRY KISS
+(Don't Repeat Yourself, Keep It Simple Stupid) style.
+Complex is to master the use of Vim
+and one finds sound
+references in help files and a nice vimrc.
+Therefore the following content should be kept
+as uncomplicated and original as possible.
+Also, because of Vim's complexity and entailed
+bond of this document to my usage,
+there is an anthropological component
+which is evident in the use of the first person
+in sentences.
+This can be understood as anthropological computation
+or physics~\cite{anPh,anPh2} and observed to help
+in the technological groundwork of the civil society.
+
+% Vi and Vim
 \subsection{Historical note}
-Toki Pona was developed as an internal and personal language
-by Sonja Lang~\cite{interview}.
-It was released as a draft in 2001 and in 2007 some documents
-reported it to have a few hundred speakers.
-The English official book~\cite{lipu} was released only in 2014.
-In 2016, a version of the official book was released in French.
-Nowadays, one finds a number of texts about Toki Pona and written
-in it, and uses of Toki Pona for artificial intelligence~\cite{roila}
-and software tools~\cite{tptoll1,tptool2,tptool3},
-in social platforms such as Facebook groups~\cite{fbtp,fbtpt},
-microblogging~\cite{tokilili}, Telegram~\cite{telegram} and IRC~\cite{irc}.
-
-\subsection{Natural and constructed and artificial languages}
-A language one uses (or might use) to communicate by speaking
-and writing is called `natural language'.
-A `constructed language' is a natural language built by someone or
-a group, such as Esperanto, Toki Pona, and Lojban.
-An artificial language is a language yield by artificial agents,
-such as in AI routines, and might be considered
-within `cultural evolution' studies.
-Formal languages are defined by tokens and rules to
-operate them, they span from computer programming languages
-to math and formal models for natural languages.
-Lojban is a constructed language with the purpose of being
-an artificial language because it is more axiomatic.
-
-Constructed languages are in some traditions called artificial
-languages, but creators most often prefer to use the term `standardized'
-or 'planned' language with the argument that the conlang
-is rooted on natural languages, and 'artificial' is misleading~\cite{conlanWikip}.
-The preferred terms seem to be planned or constructed languages
-or conlang.
-The construction of languages is called glossopoeia.
-Toki Pona is a conlang which might be classified
-as engineered for experiments, meditation and philosophy;
-and suitable for auxiliary international language and
-as an artistic language. 
-% taxonomy, vocabulary, ontology, terminology,
-
-% conlangs, natural langs, programming langs, Lojban as an
-% intermediary
-% non-verbal languages, musical corporal language, 
-% non-verbal languages confront Sapir-Whorf hypothesis.
-
-\section{Overview of the language}\label{basics}
-This section describes very succinctly the formation of
-words and sentences in the Toki Pona language.
-It should enable a newcomer to grasp the essentials
-of the language and the experienced to
-acquire new insights.
-Furthermore, it is a solid reference of the phonological and syntax
-rules.
-
-\subsection{Phonology}
-Words in Toki Pona are written using only 14 letters:
-\begin{itemize}
-  \item Vowels a (open), e (mid front), o (mid back), i (close front),
-    u (close back).
-  \item Consonants j, k, l, m, n, p, s, t, w:
-    \begin{itemize}
-      \item Nasal: m (labial), n (coronal).
-      \item Plosive: p (labial), t (coronal), k (dorsal).
-      \item Fricative: s (coronal).
-      \item Approximant: w (labial), l (coronal), j (dorsal).
-    \end{itemize}
-\end{itemize}
-
-There are standard guidelines for pronunciation,
-but the language allows for considerable allophonic
-variation.
-For example, /p t k s l/ might be pronounced
-[p t k s l] or [b d g z \textipa{\!R}].
-Especially for poetry, one might consider
-j and w to be vowels
-(e.g. j as 'i' and w as 'u').
-
-Syllables are of the form (C)V(N):
-an optional consonant, a vowel and an optional nasal consonant.
-Non word-initial syllables must follow the pattern CV(N).
-The following sequences are forbidden: ji, wu, wo, ti, nm, nn.
-
-\subsection{Syntax}
-\subsubsection{Fundamental notions}
-As in other natural languages, 
-colloquial Toki Pona might have
-incomplete sentences and deviate from the
-norm.
-The basic structure of sentences is:
-`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
-The li might be repeated to associate more than
-one predicate to the subject.
-The particle li is omitted if the subject is a simple mi (I or us)
-or sina (you). A discussion about issues (problems)
-yield by this rule
-and how I deal with them is in Appendix~\ref{mytoki}.
-
-The e might be repeated to associate more than
-one object to a predicate.
-Sentences might be related though la,
-'sentence' la 'sentence', where the second sentence is
-the main sentence, and the first sentence is a condition
-to the first.
-Multiple la-s are not described in literature,
-but I assume that one might assume the last sentence
-being a conditional to the next,
-except in cases where the context strongly suggests
-otherwise.
-
-Noun and verb phrases are (usually) built with the non-particle words.
-The first word is the noun or verb and subsequent words
-qualify the noun or verb.
-The pi particle might be used to separate sequences of words
-to be evaluated before the relation yield by pi.
-As pi is often ill understood and used,
-the following structures might be handy for newbies and as a
-reference:
-\begin{itemize}
-  \item No pi, `word word word':
-word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
-  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
-      word $\leftarrow$ (qualifies 1)  word ].
-  \item Two pi-s: `word pi word word word pi word word':
-    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
-or:
-    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
-\end{itemize}
-
-Notes on the usage of pi:
-\begin{itemize}
-  \item In a sequence of words, without pi, the second word qualifies
-    the first, the third word qualifies the phrase yield by the first
-    two words, the fourth word qualifies the noun yield by the first
-    three words and so on.
-  \item It is redundant to use pi before the last word in a noun or
-    verb phrase if there is no other pi, reason why it is most often
-    omitted.
-    Its use in this case is regarded as wrong~\cite{lipu,pije},
-    but, as one might notice, it does not add (much) information
-    through syntax (order of qualifies is conserved).
-    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
-    Also used in~\cite{akesiWawa}.
-  \item The book by jan Pije~\cite{kama} describes another use for pi:
-    after li to mean possession, e.g. `soweli li pi sina' (your pet).
-    This employment of pi might be regarded as correct, but are promptly written
-    as a noun phrase (e.g. `soweli sina') and is not mentioned
-    by the official book~\cite{tpLabg}.
-\end{itemize}
-
-All the words except the structural particles (li, e, la, pi)
-are usable in any position in noun and verb phrases,
-(i.e. 116 words).
-Notice that the phrase expresses a noun in a noun phrase (subject or
-object) or a verb (in the predicate).
-And that the first word of the phrase is the noun or verb,
-and that subsequent words are adjectives or adverbs.
-The pi and other particles restarts the phrase to
-first word being a noun.
-
-At this point, the only missing syntax rule is related to
-the prepositions: kepeken, lon, sama, tan, tawa.
-They might appear at the end of phrases,
-should be followed by another phrase,
-and require no particle. E.g.
-`toki *tan* jan Pije li pana e sona *tawa* mi'.
-Because prepositions can be used as adjectives
-or nouns or verbs, adverbs, etc, and what follows
-a preposition complement might also be of any morphosyntactic class,
-they are core sources of ambiguity.
-three:
-* preposition
-* absence of li on sina and mi
-* Incomplete sentences
-(ona li) tawa lon kon e ilo pi suli mute
-
-\subsubsection{Particles}
-Beyond the structural particles (li, e, la, pi) presented
-in last section, other particles are:
-
-\begin{itemize}
-  \item a or kin, emphasis.
-  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
-  \item taso, means however as sentence or 'only' if adjective.
-  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
-    For repetition of verbs, repeat li.
-    For object nouns, repeat e.
-    If the noun is complementing a preposition (tawa, lon),
-    one might repeat the preposition.
-    As Toki Pona is a recent language, and is able to cope with
-    variation due to its simplicity, I would advocate for
-    using en and anu wherever there is no ambiguity.
-    E.g.: 'mi tawa anu moku lon tenpo lili?'
-    In the official documentation that is not described and
-    might be regarded as wrong in strict canonical.
-  \item nanpa, denotes numbering.
-  \item seme, for questions, used next to the thing being asked for.
-    'Why?' might be expressed as
-    'seme la sina pana e moku lon sewi'.
-  \item mu, for animal noises. For me it is not a particle, as in the
-    official dictionary, but a noun.
-    I also like to use it as a verb:
-    'mi pakala e luka. mu mute.'
-\end{itemize}
-
-The vocabulary specifies morphosyntactic classes:
-nouns, adjectives, verbs, pre-verbs, adverbs, particles, prepositions, and numbers.
-I find that they might help the user and newcomer, but
-it might also suggest a deviation from what I understand and read:
-the words might be used indistinctly to be the nouns
-(subjects, the predicate when there is no object, objects,
-and preposition complements),
-the adjectives (anything that does not start the noun phrase or follows a pi),
-verbs
-(follows mi or sina or li or a preposition),
-adverbs (follows the verb).
-The pre-verbs (wile, ken, awen, kama, lukin, sona),
-might follow a verb, but might also be understood
-as the verb qualified with the next word,
-which carries a very similar if not identical meaning.
-The pre-verbs are all also defined as other morphosyntactic classes,
-such as adjective, noun, verb.
-The only exception is wile, which is only a pre-verb.
-
-Thus, the classes given in the dictionary dictate little
-in practice:
-jan kala li lape lon ni.
-Where kala, lape and ni are in this phrase
-as adjective, verb and noun,
-and are in the dictionary as noun,
-adjective and adjective.
-
-As far as I can see, one should regard
-the particles li, e, pi, and la
-and punctuation.
-The other tokens of the vocabulary
-might be used in any of the remaining positions.
-
-\subsubsection{Convention for recognizing POS tags by speaker and machine}
-
-\begin{itemize}
-  \item Noun: the first word in a noun phrase.
-    After an 'e' and after a pi,
-    The first word in the sentence if
-    sentence does not start by the verb.
-    Might be in the position of the verb
-    if the sentence has no object.
-  \item Adjective: second word on after an e and after
-    Second word on in the subject phrase if present.
-  \item Verb: after a li, mi or sina.
-    If there is no object, the verb position
-    is often a noun.
-  \item After a preposition,
-    there can be a noun phrase, a verb phrase
-    or nothing.
-  \item Notice that there is ambiguity in the structure
-    introduced by the omission of li after mi and sina.
-    Also, when there is no object, a noun or a verb
-    or an adjective might be in the verb position
-    if there is no object.
-    The prepositional complement is also not defined.
-    (mi moku tawa pali, tawa tomo).
-    So, these are sources of syntactic ambiguities
-    in Toki Pona.
-    They might be solved or minimized by using the semantics
-    of the words.
-    One preliminary effort in this direction might be
-    using the classes in the official dictionary to resolve
-    ambiguities whenever possible.
-    This solution is not optimal in correct POS tagging,
-    and does not solve all possible ambiguities
-    (there are words classified as nouns and adjectives,
-    adjectives and verbs, nouns and verbs, particle and verb).
-
-    Another source of ambiguity is the pre-verbs as described
-    in the literature~\cite{tpLang,pije}.
-    But I find it reasonable to understand them as verbs.
-
-\end{itemize}
-
-Also, the prepositional complement might
-be a noun or a verb.
-I could not come up with a sentence
-where it would be understood as an adjective.
-
-\subsubsection{Further notes}
-The only synonyms on Toki Pona are:
-a and kin; lukin and oko;
-sin or kamako;
-ale or ali.
-
-In formations such as
-toki e ni:, wile e ni:, tan ni: etc.
-'(e) ni' can be omitted and : used alone.
-
-Names are by default transliterated,
-but might not be, as described in Section~\ref{mytoki}.
-
-% my view
-% considerations about the official book and jan Pije
-% pictorial forms
-\section{Software for analysis, synthesis, and syntax highlighting}\label{hacks}
-In this section,
-we describe software, natural language, audiovisual and numeric results
-on four directions:
-1)
-the analysis of the Toki Pona language through statistics
-obtained by processing the dictionary;
-2)
-the synthesis of sentences, poems and short stories
-in Toki Pona;
-3)
-syntax highlighting for Toki Pona,
-including fine-tuning and theoretical
-considerations;
-4)
-initial considerations about the WordNet of
-Toki Pona words.
-Corpus analysis was already found in~\cite{}.
-
-\subsection{Statistics of the vocabulary}\label{sec:stat}
-In~\cite{pije} are statistics about Toki Pona corpus.
-This section focuses on the statistics of the vocabulary
-and syntactic rules:
-the letters, phonemes, word sizes,
-possible combinations for words and sentences.
-The statistics are in Appendix~\cite{listings},
-and the next paragraph is an overview.
-Python scripts were used to obtain the measurements
-and are available at~\cite{tokipona}.
-
-As described in Section~\cite{basics},
-there are only 14 letters,
-and phonemes respect a few rules.
-There are 120 different words in the official vocabulary,
-4 of them having synonyms.
-A total of 124 tokens,
-not counting proper nouns (names)
-and punctuation.
-Table~\ref{tab:pos} shows the number of words
-related to each POS tag specified in the dictionary.
-
-\input{pos}
-
-From the official 124 words, 
-26 of them (20.97\%)
-have only one syllable,
-85 (68.55\%) have two syllables,
-and 13 (10.48\%) have three syllables.
-No official words has four or more syllables.
-Of all the 235 syllables of the official words,
-li, la, ka, na, and pa are the most often (with 5.53\%,
-4.26\%, 3.83\%,  3.83\%, 3.83\%).
-On the first syllable of each word,
-a, o, pi, ka, and la are the most often
-(with 6.45\%, 4.03\%, 4.03\%, 3.23\%, 3.23\%).
-On the last syllable of each word,
-li, lo, na, la, and ma are the most often
-(with 8.06\%, 4.84\%, 4.84\%, 4.03\%, 4.03\%).
-Middle-word syllables only occur 13 times,
-and are all different, with the exception of `la',
-which occurs twice.
-For the frequencies of all syllables in every position,
-run \tttt{makeStatistics.py} in~\cite{tokipona}.
-
-Vowel and consonant frequencies are as shown in Table~\ref{freqLet}
-for starting, end and internal position.
-Comparison of vowels against consonant is in Table~\ref{tab:vcon}.
-The high incidences e.g. of 'l' and 'a' favor
-the naturalness of the language prosody
-as it resembles child speech.
-
-\input{vowels}
-
-26, 85 and 13 words have one, two and three syllables
-(20.96, 68.54 and 10.48\% of all 124 tokens).
-No official word have more than three syllables
-(although proper names may be of any size).
-In all the 124 tokens, there are 235 syllables (68 different).
-Most often syllables are in
-Table~\ref{syl}.
-A complete list of the syllables and their frequencies (in different
-positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
-script.
-A list of all words, grouped by their size in syllables,
-and ordered alphabetically and by the number of letters,
-is in \tttt{hlsyl\_\_}.
-
-\input{syls}
-
-Given the rules given in Section~\ref{phonology}
-(of 14 letters, 5 vowels, (C)V(N) phonemes,
-forbidden ji, wu, wo, ti, nn, nm),
-96 words are possible with 1 syllable,
- 96, 8256, 710016
-8256 with 2 syllables, 710016 with 3 syllables.
-No middle syllable end with the nasal consonant 'n'.
-These are shown in Table~\ref{tab:pw}
-with fraction of the space actually used in Toki Pona.
-
-Possible syntactic structures are quantified
-in Table~\ref{tab:syn} for
-Possibilities on noun, verb and prepositional phrases
-and in sentences.
-(Counting along the number of words.)
-
-A toolbox for obtaining such statistics and sets
-of words is in~\cite{tokipona}.
-
-% number of possible syllables
-% number of possible words of size X (1, 2, 3, 4....).
-
-\subsection{Synthesis of text}
-Such counting exercise is also useful
-for (semi-)automate writing through scripting.
-The syntax organizes the words in larger structures.
-The rhymes are restricted,
-the ending is as a tonality in music.
-
-There are some specific tasks for achieving texts,
-such as finding the number of syllables considering the elisoes,
-or handling interaction of the writer with the script
-to choose sentences or verses or stanzas.
-
-The same package~\cite{tokipona}
-has capabilities for synthesizing text.
-Noun, verb and prepositional phrases,
-sentences.
-It also aims at making larger scale texts
-by keeping a record of the used words and structures (context)
-and using stylistic outlines for poems and short narratives.
-
-Figure~\ref{fig:syn} holds some synthesized texts
-with different coloring schemes for syntax highlighting,
-as presented in the next section.
-
-\subsection{Syntax highlighting}\label{shigh}
-The same package~\cite{tokipona}
-has a Vim syntax highlighting plugin
-for Toki Pona.
-Instructions for installing and using
-the syntax highlighting is at~\cite{tokipona}.
-
-Basically, it distinguishes the words among the morphosyntactic
-classes according to the official dictionary.
-As a word often belongs to more than one class,
-The precedence of them might be set by the user.
-Also, some classes might be further refined or joined,
-such as by distinguishing only particles and the rest,
-or maybe particles and prepositions and the rest.
-The colors are also promptly changed according to~\cite{vimArt}
-and exemplified in the package documentation.
-
-Currently, the Python package synthesizes the
-syntax file.
-The user has control of class precedence and
-merging.
-The choice of precise coloring schemes
-might involve hacking the colorscheme being
-used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
-and Vim's highlighting schemes as described in~\cite{vimArt}.
+Vim was first released publicly in 1991.
+It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
+Vim's development is coordinated (and performed) since the beginning
+mainly by Bram Moolenaar.
+Today, current bleeding-edge version is 8.0.1257.
+I found no explicit stable, alpha or beta versions.
+I found no scientific article on Vim 
+(this might be the first one), although there are
+books, software and third-party documentation on the web.
+
+\section{Basics}\label{basics}
+Vim's interface is text-based.
+In the GUI mode (gVim),
+there are convenient menus and toolbars
+but all functionalities are still available though
+the command line mode.
+Vimscript is the internal language of Vim,
+and is often used for scripting by users
+although other languages might be used 
+(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
+Each line of a Vimscript is a command on the
+command-line mode.
+This section can be thought as a tutorial
+that focuses on the namespaces, i.e. sets of tokens
+that carry values or triggers procedures.
+One should see Appendix~\ref{keys}
+to understand the notation vim uses.
+
+\subsection{The bare minimum}\label{minimum}
+You open a file with Vim by executing
+the command: \texttt{vim <filename>}.
+Inside Vim, you start in the normal
+mode, and might want to move around using
+\texttt{h-j-k-l} for left-down-up-right.
+To insert characters, move your
+cursor to the desired location an press i,
+which puts Vim in the insert mode.
+Go back to normal mode by pressing
+\texttt{<ESC>} or \texttt{<C-C>}.
+You save the file by typing \texttt{:w<CR>},
+and exit Vim by typing \texttt{:q<CR>}.
+You can save and quit with \tttt{:wq<CR>}
+or \tttt{:x<CR>} or \tttt{ZZ}.
+
+\subsection{Vim help}
+You can find help for vim in various places.
+The standard resource is the Vim help files.
+They are accessed by typing \texttt{:h <anything><CR>}
+in normal mode.
+Examples of such \texttt{<anything>} are:
+color, navigation, :vs, vimtutor.
+Type 
+\texttt{:h usr\_toc<CR>}
+to access the official User Manual,
+which is considerably lengthy and complex
+and is usually not read by users for a few years.
+In learning Vim, one
+might want to run the \texttt{vimtutor} command
+(outside Vim) to start the Vim Tutor.
+
+There are good resources on the Web for learning
+and tweaking Vim:
+\begin{itemize}
+  \item ``Vim Adventures'' is an online RPG game for practicing
+  and memorizing Vim commands while having fun.
+  This game is quite famous among Vim users.
+  \item There are official and semi-official Vim sites e.g.:
+  \url{www.vim.org}, \url{https://www.vi-improved.org} and
+  \url{http://vim.wikia.com/}.
+  \item Many hacks, understandings and general issues
+  (e.g. how to make such a move) are asked and answered
+  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
+  One often finds these links through a search engine.
+  \item You can find many videos about Vim.
+  One traditional site is \url{http://vimcasts.org},
+  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+\end{itemize}
+
+% any other commands? files?
+\subsection{Namespaces}\label{namespaces}
+Vim is a text editor ouroboros~\cite{ouroWiki} because
+text and writing alters text and writing.
+You have namespaces where tokens have scalar
+or complex values.
+
+\subsubsection{Commands and mappings}
+There are commands, typing sequences which trigger automated actions,
+for each mode:
+\begin{itemize}
+	\item in Normal mode all keys are mapped to commands.
+		There is redundancy and additional commands
+		using Ctrl and Shift keys.
+		Some keys expect a second key,
+    and have combinations not used (available for new mappings),
+		specially the z and g.
+    See Section~\ref{normal} and Appendix~\ref{notes}
+    for more insights into the commands available in the
+    normal mode.
+	\item In the other modes, the sequences available for mappings are more obvious and abundant.
+		One should look at \tttt{:h index} to know about all the standard mappings
+		and use \tttt{:map} to list the user-defined mappings.
+\end{itemize}
+
+\ttt{C-\textbackslash} is often reserved for extensions,
+which makes it a safe namespace to use (while there are no
+such extensions).
+
+A colon command can be written as a string
+and executed by the \tttt{:execute} colon command.
+E.g. \tttt{:execute 'vs afile.txt'}.
+As there are colon commands that execute commands in other
+modes, e.g. \tttt{:normal ?\^def },
+the \tttt{:execute} is a way to build commands in any mode,
+e.g.
+\tttt{:execute 'normal' tabn 'gt'}.
+
+\subsubsection{Variables}
+There are some types of variables in Vim:
+\begin{itemize}
+	\item Environment variables: names start with \$ and hold system
+		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
+	\item Option variables: names start with \& and are meant to control the behavior of the editor.
+		One might change a value through set or let, e.g.
+		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
+	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
+	\item Internal variables are created with let and preferably have a prefix:
+	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
+		are local to the buffer, window, tab page, function, and
+		sourced Vim file, respectively.
+ \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
+		\tttt{a:} is for function arguments.
+		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
+		More about internal variables in \tttt{:h internal-variables}.
+	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
+\end{itemize}
+You can echo any of such variables or use in expressions.
+Notice that you will only be able to echo a \tttt{b:} variable inside
+the buffer where it is defined.
+For all the Vimscript capabilities, including loops, conditionals,
+and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
+Classes are possible only in rudimentary forms, e.g. through dictionaries,
+but the language is otherwise overall quite powerful,
+specially in dealing with text and editor behavior, as expected.
+
+\subsubsection{State lists}\label{state}
+Vim keeps a number of useful lists which expresses the state of the editor:
+\begin{itemize}
+	\item The entered commands are accessed through \tttt{:hist a}. 
+    The tokens a / s : can be used for specific types of commands, such
+		as search and colon commands.
+	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
+  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
+	\item Tabs list can be reached through \tttt{:tabs}.
+		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
+	\item Jumps are available through \tttt{:jumps}.
+		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
+  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
+    They also keep track or your copy, edition and deletion and are promptly defined
+    by recording a typing sequence with the \tttt{q} normal command.
+    Vim keeps only the last edition, in register \tttt{".}.
+    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
+    A hack to keep the latest deletions and copies in the standard register
+    might follow the same pattern, but use another \tttt{:h event} and
+    monitor register \tttt{""} (maybe also \tttt{"0}).
+	\item An undo list can be accessed with \tttt{:changes}.
+	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
+	\item The markers defined are listed with the \tttt{:marks} command.
+		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
+		Uppercase letters are cross buffers.
+	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
+		and variations, such as \tttt{:grep}.
+		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
+		to open the Quickfix window, where the lines of occurrence are in sequence
+		and one can \ttt{CR} one of them to have the cursor in the main window active at
+		the first character of the match.
+		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
+		to use the location-list window instead of the Quickfix, which
+		is very similar, but one per window instead of one per buffer.
+		More information in \tttt{:h quickfix}.
+	\item A tags list have to be made so one can use tags.
+		Most often one will generate the tags list using the exuberant ctags,
+		which supports dozens of languages.
+		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
+		and then using \ttt{C-]} to
+		go to the position of tag under cursor, and \tttt{:tabe tags} to open
+		the tags file.
+	\item The argument list holds a list of files to be edited or browsed. 
+		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
+	or using commands (e.g. \tttt{:ar ./*}).
+		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
+		one might perform actions on each file in argument list using \tttt{:argdo}.
+		All files in argument list are also in the buffers list.
+    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
+  \item All the autocommands might be accessed through \tttt{:au}.
+    They are event-triggered actions Vim performs.
+\end{itemize}
+
+A file with information about the state of the editor
+can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
+and this script might be consulted because it has
+a collection of commands to access various settings of Vim.
+Another good list of commands to know about Vim's state is kept on
+\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
+I would specially highlight the \tttt{:syntax} command because
+it displays the tokens and related type of meanings when run
+inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
+
+\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
+
+You can keep track of the editor state though commands and persistent visual
+cues, specially the tabs bar, the status line, and the line
+reserved for the command-line.
+For state persistence, one might keep an undo file for each file as
+in~\cite{vimrc}.
+Sessions are easy to manage, enabling one to save and load the
+editor's state, with the opened windows, tabs, buffers, etc.
+I use the mappings in~\cite{vimrc} because they keep the sessions
+in a reasonable directory and makes it easier to remember and tweak
+the standard commands to deal with sessions.
+More information in \tttt{:h sessions}.
+One might use \tttt{:h views} to keep the state of one window,
+but sessions keep all the states from all windows.
+This entails a strategy to deal with Vim that is similar to
+the use of Byobu/Tmux/Screen.
+The main limitation I found to this approach is that
+Vim is not keeping track of the terminals opened.
+If you open a terminal inside Vim with the \tttt{:term}
+command, you will save the session as usual, but when loading
+you get dummy empty windows for them and an error message.
+Screen and Tmux are the most popular terminal multiplexers.
+Byobu, built on top of them, has awesome keyboard shortcuts
+for managing sessions, windows and splits of terminals.
+Byoby/Screen/Tmux keep the state for future load.
+
+Because browsing the interface in Vim is fast,
+and it favors copy and paste of text,
+I tend to keeep a tab with some terminals:
+one with an IPython shell, another two for compiling latex
+and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
+One enters normal mode in terminal with \tttt{C-W N}
+and it is very corfortable to copy and browse the bash history of the
+terminal.
+I found the mappings on~\cite{vimrc} very helpful for directing
+editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
+which I make available across terminal-job and normal modes.
+But I've been thinking on using \ttt{C-} commands also
+for tabs (not only for splits).
+
+A good strategy I find is to have selective visual cues of the state to make persistent or hide
+and a mapping to toggle each of them.
+Now I toggle byobu/screen/tmux bar with \ttt{F5},
+status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
+I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
+Numbering is always there.
+I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
+Instead of keeping the status bar, I use \ttt{C-g} to know about the
+file and \ttt{gC-g} to know more and rarily.
+It seems not possible to remove the statusbar between horizontal splits.
+After asking around and experimenting, I realized that it seems
+reasonable to keep at least one line dividing the windows, so
+if it comes to it, I just \tttt{set statusline=-}.
+Unfortunatelly, as far as I could dig, one will need to enter
+Vim code to enable a horizontal split without losing a line.
+My ideal of this feature would be to have a visual cue of the
+first and/or last line of the windows in the lines-number
+column, or complete the spaces and empty chars with \$\$\$\$ or so.
+
+Autocommands are the standard way to define event-trigered routines in Vim.
+These are often related to particular file types, but are also often
+in defining the automated behavior of Vim.
+In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
+in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
+
+\subsection{Using Vim's modes}\label{modes}
+Vim has some basic and fully implemented modes of usage:
+\begin{itemize}
+  \item Normal mode: used for changing
+  the position of the cursor or the text displayed
+  at the window.
+  A core goal of the normal mode it to allow fast
+  navigation of the document while allowing
+  the typist to maintain its fingers on the home row
+  (i.e. on the center of the keyboard).
+  The mode is also used for manupulating text
+  (e.g. copy, paste, delete, change case) and
+    changing to other modes.
+  \item Insert mode: for inserting text.
+  \item Command-line mode: for entering Ex commands.
+  \item Ex mode: similar to command-line mode,
+  but more specialized for running various Ex commands.
+  \item Visual mode: for making, manipulating and navigating
+  selections of texts.
+  \item Select mode: similar to visual mode but
+  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
+\end{itemize}
+
+There is another basic mode, but it is not fully implemented:
+the Terminal-Job mode.
+There are seven additional modes which are mostly subordinate 
+to the basic modes and will be described when convenient.
+The manual page for Vim modes can be accessed by typing
+\texttt{:h vim-mode}.
+Some of the modes are now further considered for
+an overview of the Vim usage possibilites.
+
+\subsubsection{Normal mode}\label{normal}
+Sometimes also called navigation or command mode,
+the normal mode is most powerful for
+navigating, manipulating texts and changing to other modes.
+
+The simplest of these three is changing to other modes:
+type any of these letters to change to insert mode:
+\texttt{iIaAoOsScC}. More on the transition between
+normal and insert mode on Section~\ref{navIn}.
+Type any of these characters to change to command-line mode:
+\texttt{:/?}.
+Type \texttt{Q} to enter Ex mode.
+Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
+
+For most basic and naive navigation, one should check Section~\ref{minimum}.
+Most often, one uses:
+\begin{itemize}
+  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
+and whole page down and up, although these commands might
+be set to scroll a different number of lines.
+  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
+  \item \texttt{(w,b,e)} to move to the next, previous and next end-of-word.
+    There are motions to iterate over sequences of characters separated
+    by special characters (e.g. punctuation and parenthesis) as
+    specified by the output of \texttt{:se iskeyword}.
+    To iterate over space-separated tokens, use \texttt{W,B,E}).
+    To move to the end of last word, one might use \texttt{be}
+    or \texttt{ge}.
+The \texttt{),(} commands iterate through sentences,
+		\tttt{\},\{} through text blocks separated by empty lines.
+  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
+	  \tttt{;} and \tttt{,} for next and previous found character.
+  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
+  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
+    jump list.
+  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
+    the first non-blank character of the line.
+    A mark is registered by the user in any cursor position
+    by typing \texttt{mX}, where X is any letter.
+    If X is lowercase, the mark is local to the buffer (the file),
+    if it is uppercase or numeric, it is global to the Vim session
+    (cross buffers).
+\end{itemize}
+
+There are many more facilities to navigate Vim
+as explained in \texttt{:h navigation}.
+
+For changing the text, usual commands include:
+\begin{itemize}
+  \item \texttt{d\{motion\}} to delete the characters
+    involved in the motion command.
+  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
+  \item \texttt{x,X} to delete the character under or before the cursor.
+  \item \texttt{$\sim$} to swap the case of a character.
+  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
+\end{itemize}
+
+There are way more commands to change the texts.
+Some of them are discussed in Section~\ref{navIn}
+because they involve a transition to the insertion mode.
+A thorough consideration of the commands in the normal mode
+is found by executing \texttt{:h navigation} and 
+\texttt{:h change.txt}.
+
+\subsubsection{Insertion}
+Once in the insertion mode, the character keys
+input the characters at the cursor position at the current buffer.
+One can exit insert mode by pressing \texttt{<Esc>},
+and Vim will be put in normal mode.
+Most useful commands in insert mode include:
+\begin{itemize}
+  \item \ttt{C-O} to execute one and only command in normal mode.
+	  This enter a secondary mode (see Section~\ref{modes})
+  \item \ttt{C-R} to paste a register (a variable starting with
+    '@', defined, copied or recorded through a \tttt{q} command in
+    normal mode and as covered in Section~\ref{state}).
+  \item \ttt{C-T} to indent current line.
+  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
+    the line or to the previous word.
+  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand..
+  \item \ttt{C-X} commands for scrolling the window with multiple
+    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
+\end{itemize}
+
+\subsubsection{Command-line mode}
+This mode is dedicated for writing colon, search and filter commands,
+entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
+Most useful commands in this mode include:
+\begin{itemize}
+  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
+    end of the line.
+  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
+    until the cursor.
+  \item \ttt{C-R} to paste a register as in insert mode.
+\end{itemize}
+
+\subsubsection{Terminal-Job mode}\label{terminal}
+This mode is reported to not have reached a stable usage design
+(see \tttt{:h terminal}).
+I find that it works exceptionaly well and have used it to run
+scripts in an IPython shell, compile latex files, and even open PDFs and images.
+Vim browsing of windows and text manipulation is well developed,
+so the terminal mode enables a very convenient integration,
+more traditionally achieved through Byobu/Tmux/Screen terminal
+multiplexers.
+Most useful commands in terminal-job mode include:
+\begin{itemize}
+  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the normal
+    and command-line modes.
+  \item \tttt{<C-W>\_"} to paste a register.
+  \item \tttt{i} for entering the terminal-job mode from the normal mode.
+\end{itemize}
+
+It is useful to define the same mappings for nativating splits and tabs
+for both normal and terminal-job modes, as in~\cite{vimrc}.
+
+\subsubsection{Normal $\rightarrow$ insertion}\label{navIn}
+Many commands bridge from Normal to Insertion modes:
+\tttt{csrCSR}, \tttt{iws}, etc.
+These make convenient the replacement of text and populates registers.
+The absence of a short command to insert one char is
+a known issue in Vim.
+Reasonable mappings to insert and append a char to and around
+other char are in~\cite{vimrc}.
+Vim commands are designed to couple operator and motion commands.
+There are many operator commands that take the editor from
+the normal mode to the insert mode, most of them
+favoring deletion or change, as detailed in \tttt{:h operator}.
+Motion commands are described in \tttt{:h motion}.
+
+\subsubsection{Netrw}\label{netrw}
+The standard interface of Vim for browsing file trees is Netrw.
+It starts when you open a directory, such as with \tttt{:e .<CR>}.
+It has solid support for editing remote files (such as over ssh or
+ftp) and handy e.g. mappings to open the files as splits and tabs (specially \tttt{pot}).
+Most useful commands in netrw include:
+\begin{itemize}
+  \item \tttt{d} and \tttt{\%} for creating directories and files.
+    \ttt{Delete} removes both.
+  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
+  \item \tttt{gb} and \tttt{uU} are used to load directories
+    while marked files might be copied, moved, edited, grep-ed, tagged and migrated
+    to and from the argslist as in \tttt{:h netrw-mf}.
+\end{itemize}
+Last notes:
+there is no insert mode in netrw interface;
+the commands in \tttt{:h netrw-explore} are
+convenient for opening the directory of the
+file being edited;
+further information in \tttt{:h netrw}.
+
+\subsubsection{Ex}
+One might use the normal commands \tttt{q(:,/,?)} to have a window
+with colon or search command history to be edited normaly,
+and the chosen command can be run with \ttt{CR}.
+Vimscript was largerly based on the ex editor~\cite{ex},
+and a more advanced user might use it for prototyping
+by defining mappings and settings and managing sessions and scripts
+propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
+In the default interface started with the \tttt{Q} command in normal mode,
+each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
+follow documentation from \tttt{:h Ex-mode} for further information.
+I might be loosing something, but for most
+hacking I use the command-line window accessed with \tttt{q:}.
+I find it comfortable to browse the history in a normal-like mode,
+and edit them also using the insert mode as usual.
+The mapping have to be defined as autocommands as in~\cite{vimrc}.
+
+\subsection{Standard configuration files and directories and my .vim/vimrc}
+You can check the scripts Vim loads by using the debug script mentioned
+in Section~\ref{state}.
+By default, \tttt{$\sim$/.vimrc} and \tttt{$\sim$/.vim/vimrc} files are run by Vim at the beginning of the startup.
+One might edit the vimrc file with \tttt{:e \$MYVIMRC}
+and reload it with \tttt{:source \$MYVIMRC}, but
+I use the mappings in~\cite{vimrc} to encourage a continuous enhancement of my settings.
+Any other file might be included to run at startup by
+adding a line \tttt{:source <afile.vim>} in vimrc.
+In fact, it is on vimrc that one usually specifies the plugins
+and plugin managers they use.
+Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
+or follow some patterns described on the next section to change the
+scripts and sequence of them to be loaded.
+
+\subsection{Plugins and packages}
+One can see the list of standard plugins
+with \tttt{:h standard-plugin-list} command.
+Any \tttt{plugin/**/*.vim} file inside a directory
+listed by \tttt{:se runtimepath} will be loaded
+(e.g. \tttt{.vim/plugin/something/ascript.vim}).
+There are various ways to automate the installation
+and enhnace the management of plugins.
+By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
+that downloads latest scripts from sourceforge as specified in \tttt{:h getscript-data},
+and the Vimball interface, which creates and loads a Vimball for a plugin
+with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
+that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
+The Vimball can be installed in a system by \tttt{:source <filename>.vba}
+or loading it at vim startup with \tttt{\$ vim <filename>.vba}.
+
+A Vim package is a directory that contain plugins.
+It should be located inside a \tttt{pack/} directory
+somewhere in the directories listed with \tttt{:se runtimepath}.
+The plugins found in \tttt{pack/<packName>/*/start/} are loaded
+at startup, the plugins found at
+\tttt{pack/<packName>/opt/**} are loaded with\\
+\tttt{:packdd <script\_or\_directory\_name>}.
+
+All directories vim looks for scripts are described
+in \tttt{:h vimfiles} and are basically set by
+\tttt{:se runtimepath} and conventions inside
+each path therein, such as to look for \tttt{vimrc} files
+and \tttt{plugin/} or \tttt{pack/} directories.
+Vim scripts can be loaded conditionally,
+e.g. only if a function is used as in \tttt{:h autoload-functions}.
+Example of such are filetype plugins (enabled by in a \tttt{ftplugin/<filetype>.vim} file,
+e.g. inside a plugin directory).
+There is a number of plugin managers for Vim.
+Pathogen and Vundle seem to be the most popular,
+one because of its minimalism, the other because of
+advanced features, e.g. for searching and installing plugins with
+colon commands.
+
+\subsection{Spell and spelllang (en and pt\_br)}
+One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
+or \tttt{:se spelllang=pt\_br}
+and toggle spell checking with \tttt{:setl spell!}.
+These are used so often that one might use shortcuts as in~\cite{vimrc}.
+Currently, Vim will download the files for a specific language if
+not found in system.
+
+\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}\label{script}
+In Vimscript,
+the colon commands (also Ex or command-line commands) are related through spaces,
+punctuation and keywords (see \texttt{:h script}).
+Scripting the Vim editor can also be accomplished using other languages,
+as well documented e.g. \tttt{:h python}.
+Functions are defined through colon commands and are called
+inside colon commands e.g. \tttt{:call MyFunction()}
+or \tttt{:echo MyFunction(4)}.
+Notice that functions are not commands but might be bind to them
+through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
+Executing source files is very straightforward with \tttt{:so
+<filename>.vim},
+and one can always use the Ex mode for rapid scripting.
+
+At the same time, the \tttt{:term} and terminal-job mode make
+scripting other software more convenient, as output is
+promptly navigated and copied. The shell works well for generic use,
+such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
+See Section~\ref{terminal} for further directions about the terminal.
+
+\subsection{Color}\label{visual2}
+Newer versions of Vim support 24 bit true color (aka 16 million colors)
+in terminal Vim (not only in GVim).
+The terminal must support true color, and tests are available e.g. in
+\url{https://gist.github.com/XVilka/8346728}.
+Then Vim needs to be set to use true colors with
+\tttt{:set termguicolors}.
+If using 8 or 16 bit colors, Vim uses the color pallete from the
+terminal, if using true colors each color is defined directly.
+Settings for using true colors inside Byobu/Tmux involve tweakings
+and are available in~\cite{vimrc}.
+Good color schemes to use with true colors are Gruvbox
+and Solarize.
+One might source syntax files at any time to change syntax, usually
+though linking tokens to syntax groups with
+\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
+and then relating the group to another group 
+\tttt{:highlight link <group\_name> <group\_name2>}
+or by specifying the group characteristics directly:
+\tttt{highlight <group\_name> guifg=\#ffffff}.
+If you change a syntax file, reloading a file with with \tttt{:e<CR>}
+updates the highlighting on the window with the corresponding filetype.
+A complete syntax highlighting support typically involves at least three files:
+\begin{itemize}
+  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
+    detected with e.g.
+    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
+  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
+    file type, such as: 
+    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
+    expandtab autoindent}.
+  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
+    tokens and highlighting groups; and highlighting group definitions.
+\end{itemize}
+
+This very scheme is implemented very straightforward in this
+plugin~\cite{tokipona}
+for highlighting text in the Toki-Pona language~\cite{tpLang}.
+Syntax highlighting plugins are as filetype plugins,
+but also have a \tttt{syntax/<filetype>.vim} file
+relating keywords to highlighting groups. 
+One might see every highlighting group, and their final visual results,
+with the command
+\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}.
+In my system, the \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
+directory, but \tttt{syntax/} files had to be moved to \tttt{$\sim$/.vim/syntax/}.
+
+\subsection{Fonts}
+The fonts are defined by your terminal or inside GVim.
+\ttt{C-+} and \ttt{C--} can be used to change font size.
+Some settings for fonts, such as boldface, might be set using
+the syntax highlighting facilities described in the previous section.
+
+\section{Conclusions and further work}\label{conc}
+This document seems reasonable as an overall reference of the Vim editor,
+at least for my usage and proficiency.
+Given the folkloric milestone of using Vim for 10 years,
+this article might serve as a benchmark for one to relate
+it's current use and understandings.
+As a pedagogical material, it seems to be unique in the emphasis
+on namespaces, understood as commands, variables, state-related lists, etc,
+especially in Section~\ref{namespaces},
+and the reference to the standard Vim documentation
+to achieve the DRY KISS style described in Section~\ref{intro}.
+
+\noindent {\bf Potential enhancements} to this document are:
+\begin{itemize}
+  \item The inclusion of facilities such as reading emails and connecting over ssh.
+    There is a working hack in~\cite{vimrc} for browsing over the WWW,
+    but such aspect of Vim usage might receive more attention
+    given that it is confortable to use the navigation and editing facilities
+    and the resulting integrated environment.
+  \item Updating of the information I can find about the issues discussed,
+    such as about status lines in Section~\ref{visual1}.
+  \item Include a discussion about Neovim.
+    I have never used it, but it seems to be reaching a considerable user base
+    and it might be feasable to give an account of Vim and Neovim
+    after some tests and researching the official and user base documentation.
+  \item Better cite documentation, plugins and Vim authors.
+    I prefered to keep the references inline through \tttt{:h} commands
+    and URLs, more in accordance with the style of Vim documentation,
+    but bibliographical items constitute a valuable asset for academic literature,
+    and some authors might find their work more pretiged if more
+    thoroughly cited.
+  \item An analysis of my usage, e.g. according to \url{http://www.drbunsen.org/vim-croquet/}.
+    This undertake might benefit from data from many users,
+    which favors the potential plugin for usage analytics described in the next bullet list.
+\end{itemize}
+
+\noindent {\bf Potential next steps} in using Vim:
+\begin{itemize}
+  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
+  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
+  \item Better integrate Python and Vimscript, especially for data visualization
+    and syntax highlighting management.
+    In accordance with the visualization issues described in
+    Sections~\ref{visual1} and~\ref{visual2}.
+  \item Make plugins for:
+    \begin{itemize}
+      \item listing all the mappings available in each mode and the typing combinations which are available
+        for new mappings.
+      \item Sessions, as described in Section~\ref{visual1}.
+      \item AA messages (shouts): to keep track, document and share of working sessions
+        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
+        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
+        relate AA sessions to other media, such as software repositories,
+        screencasts and images, interact with IRC channels and other social platforms.
+      \item Slick Vim: a collection of the settings, mappings and plugins I use.
+        Enhancements such as using \ttt{C-} commands also to browse tabs,
+        shortcuts to join a window into a tab,
+        and dummy minimal plugins as simplest, file type and syntax highlighting,
+        Some more elaborate tweaks should also be present, such
+        as breaking lines in sensible places for natural language texts
+        while respecting e.g. \tttt{:se textwidth}.
+      \item Dealing with .swo and .swp temporary files.
+        In summary, if the restored .swo file has the same content
+        and the file being opened,
+        the restoration phase can be omitted.
+        If the contents differ, Vim should open a tab with each file
+        in a vertical split and run \tttt{:windo :diffthis}.
+      \item Rendering images and equations.
+        These are useful for using Vim in presentations
+        or achieving a textual representation when it is mandatory,
+        such as to comply with the limitations of a platform (e.g. Vim editor). 
+        but also hold stylistic merits as ascii art is often
+        very appreciated.
+        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
+        and can directly render ascii charts from data using cues such are shape, position
+        and color.
+      \item Redirecting the commands that usually show the results in a 'more' interface,
+        which cannot be searched nor copied,
+        to a parsed and linked quickfix or location window.
+        The basic idea is to use \tttt{:redir} command to redirect the output
+        of such commands with \tttt{:se nomore}.
+      \item Slide presentations. I've been using some automation for browsing
+        slides and opening figures and some Vim users asked for the settings and
+        commands. They are very elementary use of registers being executed over
+        arbitrary but consistent textual patterns.
+      \item Changing the color scheme an highlighting scheme incrementally
+        and selectively using the features described in Section~\ref{visual2}.
+      \item For bringing back all the splits after an \tttt{:only}.
+        Also for bringing back the tabs after a \tttt{:tabo}.
+      \item Run Python excepts from file in an IPython shell. 
+        The buffer number of the terminal window should be stored,
+        and than any selected lines should be run on that instance.
+        Mappings should make available all movements to catch the excepts,
+        execution of registers, remain on script or in the IPython shell.
+        Also, terminal mode can use some improvements, such as \ttt{C-O} for
+        one normal mode command.
+      \item For keeping track of the usage and making analysis
+        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}. Usage analytics.
+      \item Colorschemes for true color (gui).
+        The standard colors (e.g. blue, elflord) loose some of the distinctions,
+        e.g. SpellBad tokens are not highlighted on these colorschemes
+        it you have a functioning \tttt{se termguicolors} (standard GUI mode
+        and available in terminals since recent versions of Vim 8, see Section~\ref{visual2}). In~\cite{vimrc} are some lines that make such colorschemes after over
+        \tttt{:colorscheme blue}.
+        This design of color schemes over the standard color schemes might be
+        a very simple and effective way to make new color schemes.
+      \item Opening the editor status listings such as in Section~\ref{}
+        in a quickfix or location window where they might be searched, edited
+        and used as wanted.
+        These windows open in a more-like interface that is very restricted in what
+        they allow (e.g. no search, no persistence).
+    \end{itemize}
+\end{itemize}
+
+\subsection*{Acknowledgments}
+FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
+Vim user community.
+
+\appendix
+\section{Example of usage session}
+I usually begin by opening a file or directory
+with \tttt{\$ vim <filename>}.
+The color scheme is alternated between
+blue and GruvBox with
+\tttt{:colo gruvbox} and \tttt{:colo blue}.
+I open an vertical split and then move
+the window to a new tab using
+\tttt{:vs} and \ttt{C-W\_T}.
+I then search tokens related to
+the enhancements I want to make or
+the knowledge I want to acquire.
+I go back to the previous tab with \tttt{gr}
+and make a global replace with
+\tttt{:\%s/<this>/<that/g}.
+On adding dots to sentences,
+I record in the \tttt{"q} register
+the sequence \tttt{jA.jj},
+using it as a macro 10 times by
+\tttt{10@q}.
+I move to the other tab with \tttt{gt}
+and open a terminal window with \tttt{:term}
+for compiling latex files.
+I start another terminal with \tttt{<C-W> :term}
+for opening the resulting PDFs with evince.
+If any new idea comes to mind and I have time,
+\tttt{\textbackslash\textbackslash s} opens 
+my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
+sources it.
+If there is e.g. code or notes in other projects I am working
+on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
+in Byobu/Tmux sessions and windows.
+
+\section{My vimrc file and usage}
+In this Appendix is my vimrc file.
+Although it has comments, one should
+look for the options that (s)he does
+not understand, as a thorough explanation
+of the file is tedious and out of the scope
+of this document.
+
+In using Vim with my vimrc file~\cite{vimrc},
+I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
+and the tab line with \texttt{\textbackslash\textbackslash T}.
+Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
+The mappings for transitioning through splits and tabs
+are also used constantly.
+
+\section{Example of notes on mappings}\label{notes}
+\texttt{:h index} shows all the default mappings
+while \tttt{:map} shows the user-defined mappings.
+By considering such information, one can make
+useful observations exemplified in this Appendix.
+
+\subsection{Normal mode}
+Every letter and character in the keyboard is used.
+In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
+\ttt{BS} and \ttt{C-H} is the same as h.
+\ttt{C-J} and \ttt{C-N} is the same as j.
+\ttt{C-P} is the same as k.
+Space is same as l.
+\ttt{C-[} and \ttt{Esc} are not used,
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+\ttt{C-\_} not used.
+\tttt{+} is the same as \ttt{CR} and they are both not very useful.
+Del is same as x.
+
+Many \tttt{][} combinations are not used, e.g.
+with \tttt{abhjklfg}.
+The \tttt{\_} command might be used as a more powerful \tttt{\^},
+leaving it free for mappings.
+
+Directions, home, end, page up and down, insert, all have mappings
+in more centraly located keys.
+unbounded: g,z,[]
+
+The \ttt{C-(HJKL)} commands are redundant,
+with the exception of \ttt{C-L} which redraws the screen,
+so it is a reasonable choice to use them to move focus
+of the editor to splits in the \tttt{hjkl} directions.
+
+Many key combinations are available for new mappings through the \tttt{g}
+and \tttt{z}, and \tttt{v} commands.
+They have typical uses, e.g. \tttt{z} for folds,
+spell checking and some movements (mainly when wrap is set).
+
+\subsection{Insert mode}
+All standard char keys are used for entering chars on text.
+\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
+in moving through multiple lines.
+\ttt{C-[} is the same as \ttt{ESC}.
+\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+other combinations with \ttt{C-\textbackslash} are not used.
+
+\section{My \tttt{:version}}
+I should keep the output of \tttt{:version} executed on the system
+in which I wrote this document in this link:
+\url{}.
+It was compiled with this Makefile~\cite{makefile}
+in a 16.04 LTS Ubuntu Linux,
+and is tagged as 8.0, Included Patches 1-1173.
+
+\section{Key notation and meaning for Vim}
+The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
+mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
+(or \tttt{Meta+x}).
+\ttt{A-X} is the same as \ttt{M-X} and refer to
+Alt or Meta keys.
+\ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x}  and \ttt{C-x} are the same,
+i.e. Vim does not distinguish lower and uppercase 
+letters in \ttt{C-} commands.
+The \ttt{M-x} and \ttt{M-X} (or \ttt{M-S-x} or \ttt{M-S-X})
+are different commands,
+and they are not used by Vim's builtin keys.
+Such \ttt{M-} key combinations are potentially conflicting
+with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
+but are otherwise a large and safe set of combinations
+for one to use.
+
 In summary,
-the usage of the package and plugin might be performed
-through the following actions:
-\begin{itemize}
-  \item Installation of the plugin.
-  \item Tweak of the syntax file by hand.
-  \item Running the Python script to generate a new syntax file
-    according to other settings.
-  \item Write a file inside Vim using Toki Pona and save
-    the file with the .tokipona extension.
-    Reload the highlighting scheme whenever you
-    change the syntax file by hand or through the Python script.
-  \item Access the used highlighted groups with :syntax,
-    Access all the highlighting groups with :so \$VIMRUNTIME/syntax/hitest.vim.
-    Change the coloring of a set of terms by associating
-    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g. Visual):
-    :highlight link tokiponaADJECTIVE Visual.
-    The plugin comes with the :TokiStation command, which opens a window with
-    the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
-    Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files. Another with Readme, and PDF documentation. Another with an IPython.
-    Imported tokipona. It resets completely upon command :TokiClose, closing all created windows.
-\end{itemize}
-
-The tokipona.vim syntax file has an association of toki pona
-words with highlighting groups.
-It also holds association of these groups to other groups
-to use their color settings.
-Therefore, one is able to use various color schemes
-with syntax-highlighted Toki Pona texts.
-A Vim user might run \tttt{:colorscheme blue},
-\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox}
-\tttt{:colorscheme elflord} to see the same text colored
-with different colorschemes.
-To interfere directly on the colors chosen,
-\tttt{:highlight Normal guifg=\#00000 guibg=\#0000ff}
-will change the standard foreground (text) color to pure black
-and background to pure blue.
-See \tttt{:h gui-colors} and \tttt{:h highlight}
-for the way you might edit colors directly.
-Useful commands are given as commented lines
-in the end of the syntax file.
-
-\subsubsection{Advanced syntax highlighting considerations}
-Standard guidelines for syntax highlighting
-depend heavily on cultural and use factors
-(https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies),
-although there are projects such as Solarized~\cite{solazired}
-(http://ethanschoonover.com/solarized)
-which try to give general solutions for some contexts.
-Standards guidelines for syntax highlighting were not found,
-and might be driven from data visualization considerations.
-Most evident examples of such guidelines are:
-\begin{itemize}
-  \item The use of blue and other high frequency colors for
-  background.
-  This is extended to other high frequency (small wavelength)
-  colors such as violet.
-  \item The avoidance of colors of blue and other high frequency
-  colors in small objects, such as letters and words.
-  \item Explore simplicity and elegance through the use of discrete
-  coloring schemes. Most of the tokens should be preferably of
-  the same color.
-  One might use a power-law to mimic natural occurring phenomena
-  and perception. Or an exponential to map parameters to perceptual
-  stimuli.
-  \item You might always wish to maximize contrast,
-  although taste and less consative combinations might
-  also dictate choices.
-  \item Stipulate axes of parameters with which to set colors.
-  Wavelength or frequency is an obvious axis given the considerations
-  above.
-  Other parameters are hue, transparency etc.
-
-  two types of colorschemes: those with dark background,
-  which are more commode at first; and those with a light background,
-  which are usually impressive (and even annoying) at first,
-  but the eye gets used to it and it keeps you more stimulated.
-  The blue color seems to be specially related to physiological
-  stimulation of the body by color~\cite{blue,blue}.
-  \item Consider that the user might stay many hours (and we often do)
-  at a text editor, and that the colors and formats involved are
-  prone to entail a considerable effect in the body and mental
-  activity.
-  \item Syntax sonification: consider mapping the textual structures
-  to sound (e.g. counting of tokens or context as inside a function or
-  class or over or around a variable of conditional or loop.
-  E.g. through MASSA.
-  \item Mode for tuning syntax highlighting with keyboard shortcuts
-  are used to tune individual colors from background and foreground
-  colors and groups.
-  \item How would these techniques be used for coloring Toki Pona?
-  Coloring of keywords that reflects the dictionary classification
-  of words are not in a power law distribution.
-  Maybe better described by a half-normal or by
-  overlapping normal distributions over the three peaks ($\sim35-58,
-  12-18, 1-5 occurrences$).
-  \item Coloring for poems: division by syllables, words, POS tag,
-    total number of syllables or letters of words.
-\end{itemize}
-
-\url{http://pygments.org/docs/}
-
-\subsection{Wordnet Toki Pona}\label{wn}
-exception: jan (people)
-is a hyponym of soweli (mammal),
-kili is hyponym of kasi,
-(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
-But otherwise such relations hold:
-e.g. luka (as hand) is e meronym of sijelo (body).
-
-For a first achievement of a Toki Pona Wordnet,
-each of the Toki Pona words in the dictionary
-were related to Wordnet synsets
-through the English lemmas.
-All synsets related to the lemma with the POS tag
-given in the dictionary are added.
-Another immediate possibility would be to
-relate each Toki Pona word to the synsets
-related to all the words bounded by a semicolon.
-Another possibility, which I find more consistent
-but also more controversial,
-is to choose all synsets related to the Wnglish
-words given that the Toki Pona words can be used
-in many roles.
-WordNet only contains nouns, adjectives, verbs, and adverbs.
-Thus, particles and prepositions were not considered.
-Numbers were considered adjectives.
-Words presented as adjectives in the dictionary
-were considered adjectives and adverbs.
-
-The total number of different synsets is 2493 in a very conservative
-counting: discarding prepositions and particles,
-only considering the synsets which have the POS tag
-given in the dictionary.
-If all the synsets (that have any POS tag)
-related to a lemma are considered, 
-then Toki Pona vocabulary reaches 3961 synsets.
-The prepositions are particularly missing in these
-countings because e.g. tawa or lon are used also
-as ordinary words.
-If synsets for them are considered those of the English
-lemmas with any POS tag, the vocabulary reaches 4052
-synsets.
-The total number of synsets in Wordnet is X.
-Total number of synsets counting neighbors.
-Depth of TP synsets.
-Using such features to devise other conlangs.
-
-\subsection{Hacks from other people}
-The \url{tokipona.net} has a number of tools,
-just as to transliterate names into Toki Pona phonemes,
-and search in corpus.
-
-\section{Conclusions and further work}\label{conc}
-\begin{itemize}
-  \item Make a syntax highlighting that colors the tokens with respect
-    to the syntatic position and function, and not in a fixed manner
-    as is implemented and exposed in Section~\ref{shigh}.
-  \item Relate Toki Pona to Wordnet: should one Toki Pona word
-    be related to more then one synset of the English language?
-    An immediate implementation would be linking each Toki Pona word
-    to the English words in the official vocabulary.
-    A Toki Pona synset would be related to all the synsets of each English
-    words regarded as lemmas (sequence of characters or sound).
-  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
-    of words and text.
-    Enhancements of such translators would use POS tagging and templates
-    in English and Toki Pona and/or corpus and further statistical methods.
-    An efficiency of Toki Pona can be studied in terms of the coverage of
-    the Wordnet cloud.
-  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
-    (conlanging), and insights derived from Toki Pona might be used for that end.
-  \item Understand how the corpus is gathered in \url{tokipona.net}.
-  \item Know about previously existing words that were used for Toki Pona
-    (e.g. suno and suwi might come from sun and sweet),
-    and about the reasons that lead Sonja (and maybe other people)
-    to choose the 14 letters and the syllable structure.
-    This might require a dedicated communication with the
-    speaker community and the documentation authors.
-  \item Corpus-based analysis.
-  \item Publication of original texts and translations.
-  \item Make an article written in Toki Pona.
-    I wrote Section~\ref{finalToki}, and believe that
-    a summary in both English and Toki Pona for
-    facilitating the acquisition of context,
-    a reasonable article on some scientific topics is
-    possible.
-    I first conceived something around complexity, statistics, physics, or computer science.
-    But one possibility is to write about linguistics, philosophy, literature
-    or psychology with the partners I write in English.
-    I can start a draft,
-    qhey might learn the language in a few hours (with or without my help),
-    to contribute, and we can write a short paper.
-  \item Enhance the synthesis of text to yield better contextualized text
-    and stylistic traces, such as for poems and short stories.
-    Give the user the ability to choose the sentences (generate randomly according to
-    previously written or given text,
-    some rules input by the user, the package and the language guidelines,
-    outputs to the screen and asks to keep and discard).
-  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
-    It uses only syntactic cues to choose POS tags,
-    which might be overcome by using n-grams and
-    further techniques from Natural Language Processing~\cite{POS}.
-  \item We are particularly interested in software tools,
-    and one possibility is to use a conlang to describe software
-    routines.
-    In this sense, it seems profitable
-    to understand what `non-ambiguous' means in Lojban,
-    and in what sense one is able to compile and parse Lojban.
-    This study will probably yield considerations about
-    parsing Toki Pona files.
-  \item Conlang idealized for describing data, programs,
-    scientific writings, creative writing
-    (exploring the thinking process).
-    Different modes that might be specified by a section header or tag.
-\end{itemize}
-
-\subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3);
-Vim, Python, and Toki Pona authors, developers and literature maintainers;
-user and speaker communities.
-
-\appendix
-\section{My usage of Toki Pona}\label{mytoki}
-
-I use the standard sounds, but often use [z] for s.
-I often translate texts to Toki Pona (e.g. biblical excepts)
-and create new texts as poems and short stories.
-Most of them are in~\cite{tokisona}.
-I omit the li particle after subjects sina and mi,
-in accordance with the norm,
-but sometimes I use them when there are many predicates.
-E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
-In such cases, the first li is sometimes omitted.
-Also, sometimes I use li before mi and sina where I find
-that there is unwanted ambiguity, e.g.
-sina moku pona e jan 
-(might be sina li moku pona e jan or sina moku li pona e jan).
-
-Names are by default transliterated,
-but I advocate that, as in other languages,
-names might be used as they are in the
-correspondent mother tongue.
-E.g. the name Erdös is used in
-English and Portuguese although the standard
-alphabet does not contain ö in such languages.
-I also tend to legitimate the use of English (or German) words
-in Toki Pona texts if it is the case,
-as happens often in scientific writing
-(kernel is a German word used in English,
-webpage is an English word used in Portuguese).
-
-Proposed notations for numbers seem numerous.
-I tend to think that one might indicate is two numbers
-are multiplying (pi) or are in different scales
-(such as in decimal or binary notations).
-For example, I take luka two to mean 52.
-Or it might not be taken to such level of strictidness,
-for 52 is a reasonable notation for a simple language.
-mi jo e jan sama nanpa 12.
-Or even:
-ona li lon e soweli 27.
-
-Remove the 'li'.
-I've been avoiding e ni: and using only :.
-'ve been omitting the subject if it is the
-same as in the last sentence.
-I've been avoiding also the li sometimes,
-and starting only a predicate + object phrase,
-or a whole phrase altogether.
-
-pali moku is addiction.
-noka and ampa; luka, suno, sike and lawa;
-I tend to group these things, as pali and pana.
-These clusters of meaning hint me that an even simpler
-language is possible and maybe more optimized to the
-core goals of Toki Pona.
-Also, ambiguities introduced by omission of li
-and absence of a token to denote preposition,
-suggest to me the possibility of a syntax that is uniquely
-parsed.
-
-I am tending to think on a language with the same syntax of toki pona:
-<sentence> la <sentence = subject + predicate + object + preposition complement>
-but always using the syntax separators,
-repeating them when the function in performed in the smallest scale:
-'(toki pi toki pona) pi jan sona' (language in toki pona and of intellectuals) vs
-'toki pi (toki pona pi pi jan sona)' (language in toki pona of intelectuals).
-The keyword for preposition complement is the catch.
-An 'a' seems good, but it is taken in Toki Pona.
-Maybe use it as the preposition particle and use ha or he
-instead of Toki Pona a.
-In such a setting,
-one might enable pi li e a (concept-qualifier, subject-predicate,
-predicate-object, concept-predicate)
-in any of the terms of the template:
-<sentence> la <sentence = subject + predicate + object + preposition complement>
-Between sentences one would use:
-la pi, la li, la e, la a, and la la as normally.
-This is a very fractal conlang proposal.
-Maybe have a way to distinct between noun, qualifiers (adjectives/adverbs) and verb terms,
-and accept any of them for subject, predicate, object, predicate,
-slots. Or assume a part of speech for template slot or for first word in slot.
-
-Because toki pona's vocabulary is so small,
-one might make variations of it for languages
-in specific.
-For example,
-suno might be sun in an English variant,
-and sol in a Portuguese variant.
-In Appendix~\ref{enPo} are an English and Portuguese
-proposition of such variants.
-A naive version might be obtained through
-choosing the first word in the description of
-an official vocabulary, as they exist in English and French.
-
-mute, mute ike,
-something can be good, numbered,
-or numbered in a bad way, too much,
-as in 'o awen, sina pali mute, lape lili'.
-
-% \input{pos}
-
-\section{Final words in Toki Pona}
-toki li nasin e lawa.
-li nasin tawa (pi) toki insa.
-
-% jan li ken toki li ken pali pona e toki pona.
-toki pona li pona e nasin tan ni:
-ona li jo e sitelen mute lili.
-ona li pona.
-pona kepeken weka 'p' li ona, a.
-
-o taso la toki pona li kalama li lukin pona.
-sitelen en nasin li open e sitelen suli\cite{tpLang,pije,fb1,fb2,tokisona,Wikipesija}.
-li sona. li nasin e toki e sona e lawa e lon.
-
-toki ni li wawa tawa jan mute nasin en toki.
-wawa tawa toki pi jan sona.
-pi ilo nanpa en nanpa nasin.
-taso tawa toki sona a.
-sitelen sona, sitelen musi.
-
-ilo lon linha~\ref{hacks} li pana e sitelen
-e sona tan sitelen,
-en nasin.
-linha~\ref{basics} li pana e nasin pi toki pona.
-e sitelen tawa kama sona.
-
-mi wile pali e sitelen lon toki pona.
-sitelen lon nasin, sitelen sona,
-sitelen musi.
-ante e toki pona la
-la ona li nasa.
-taso nasa li pona mute.
-li pona mute tawa lawa,
-tawa kama sona, tawa sitelen e toki.
-ante la toki pona o.
-toki e toki pona tawa sina.
-kama sona e toki pona sina.
-
-o pona tawa jan pi toki pona.
-jan Sonja, Birns-Sprage, Kipo, Pije,
-Siwejo, Malija, jan kulupu mute.
+one might choose for mappings to overwrite defaults key combinations,
+use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
+normal commands, \ttt{C-} and \ttt{S-} commands, or any \ttt{M-} and
+\ttt{M-S-} command), or
+create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
+More information about key notation in \tttt{:h key-notation}.
+Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings
+in \tttt{:map}.
 
@@
 a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
-\bibitem{kama}
-  Knight, B. (2017) o kama sona e toki pona! Available at:
-    \url{http://tokipona.net/tp/janpije/okamasona.php}
-\bibitem{Wikipesija}
-  Toki Pona community (2017). Wikipesija. Available at:
-    \url{http://tokipona.wikia.com}
-\bibitem{conlanWikip}
-  Constructed language.
-(2017, November 1).
-In Wikipedia, The Free Encyclopedia.
-Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
+\bibitem{makefile}
+	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
 \end{thebibliography}

### 2017-12-04 tokipona/makeTexts.py
advanced in text synthesis

@@ -76,6 +76,7 @@ for row in table:
 
 # use variables words and word_classes to synthesize texts
 plain_words = list(word_classes.keys())
+prepositions = list(words['PREPOSITION'])
 
 # functions repeated from makeStatistics, join them into a utils.py or similar
 def firstSyllable(token):
@@ -115,7 +116,10 @@ def getSyllables(token):
     return syl
 
 forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
-def createPhrase(nsy=0, vowel=False):
+for f in forbiden:
+    plain_words.remove(f)
+
+def createPhrase(nsy=0, nwords=0, vowel=False):
     """
     Create a Toki Pona phrase.
 
@@ -127,6 +131,9 @@ def createPhrase(nsy=0, vowel=False):
     nsy : integer
         The number of syllables of the phrase.
         If nsy=0, the phrase will have a random number of syllables
+    nwords : integer
+        The number of words in the phrase.
+        If nwords != 0, nsy is ignored.
     vowel : boolean
 
     Returns
@@ -145,45 +152,58 @@ def createPhrase(nsy=0, vowel=False):
 
     References
     ----------
-    .. [1] Fabbri, Renato, et al. "Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts." arXiv preprint arXiv:abs/XXX.XXXX (2017)
+    .. [1] Fabbri, Renato, "Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts." arXiv preprint arXiv:abs/XXX.XXXX (2017)
 
     """
-    if nsy == 0:
-        nsy = n.random.randint(1,11)
-    nsy_ = 0
     phrase = ''
-    while nsy_ < nsy:
-        w = n.random.choice(plain_words)
-        s = getSyllables(w)
-        l = len(s)
-        if w[0] in vowels and vowel:  # elision
-            l -= 1
-        if nsy_+l > nsy:
-            continue
-        if w in forbiden:
-            print(w)
-            continue
-        nsy_ += l
-        phrase += w + ' '
-        if w[-1] in vowels:
-            vowel = 1
-        else:
-            vowel = 0
+    if nwords:
+        for i in range(nwords):
+            w = n.random.choice(plain_words)
+            phrase += w + ' '
+    else:
+        if nsy == 0:
[…]

### 2017-12-06 article/article.tex
advanced in text synthesis
arrumada sintese de poemas

--- 2017-12-03 article/article.tex
@@
 \subsection{Synthesis of text}\label{synth}
-Such counting exercises is also useful
-for (semi-)automated writing through scripting.
+Such counting exercises are also useful
+for facilitating (semi-)automated writing through scripting.
 The syntax organizes the words in larger structures.
@@
 has capabilities for synthesizing Toki Pona text.
-Noun, verb and prepositional phrases,
+On its most basic level, the routines yield
+noun, verb and prepositional phrases,
 sentences.
-It also aims at making larger scale texts
+They also aim at making larger scale texts
 by keeping a record of the used words and structures (context)
@@
 might be enhanced as described in Section~\ref{conc}.
+
+Automatic and randomic synthesis of texts in Toki Pona are
+useful because of the reduced vocabulary: ideas often make
+sense in unexpected way, and yield a way to 
+explore the semantic field possible by Toki Pona.
+Each words is related to a broad semantic field,
+thus the randomized selection of words are more likely
+to make sense.
+One might object that the resulting texts are unusual 
+and even consider that they often hold insubstantial or unsound meaning.
+We advocate that these unexpected formations are desirable
+for exploring the possibilities of the language and of thought,
+and for artistic endeavors.
+Also, to obtain texts which one finds usual or satisfactory
+in more strict or personal terms, such a person might just write
+them normally.
+
+% phrase
+% sentence
+% poem
+% text
+% each with a different color scheme
 
@@
 in the end of the syntax file.
+
+% :TOhtml command
 

### 2017-12-06 tokipona/makeTexts.py

@@ -115,9 +115,29 @@ def getSyllables(token):
         syl.append(sy)
     return syl
 
+def countSyllables(token):
+    t = token.split(' ')
+    nsy = 0
+    vowel = 0
+    for i in t:
+        nsy += len(getSyllables(i))
+        if i[0] in vowels and vowel:
+            nsy -= 1
+        if i[-1] in vowels:
+            vowel = 1
+        else:
+            vowel = 0
+    return nsy
+
 forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
 for f in forbiden:
     plain_words.remove(f)
+
+words_nsy = [
+        [i for i in plain_words if len(getSyllables(i)) == 1],
+        [i for i in plain_words if len(getSyllables(i)) == 2],
+        [i for i in plain_words if len(getSyllables(i)) == 3]
+        ]
 
 def createPhrase(nsy=0, nwords=0, vowel=False):
     """
@@ -162,7 +182,7 @@ def createPhrase(nsy=0, nwords=0, vowel=False):
             phrase += w + ' '
     else:
         if nsy == 0:
-            nsy = n.random.randint(1,11)
+            nsy = n.random.randint(1,7)
         nsy_ = 0
         while nsy_ < nsy:
             w = n.random.choice(plain_words)
@@ -178,9 +198,32 @@ def createPhrase(nsy=0, nwords=0, vowel=False):
                 vowel = 1
             else:
                 vowel = 0
-    return phrase[:-1]
+    p = phrase[:-1]
+    p_ = p.split()
+    nsy_ = sum([len(getSyllables(i)) for i in p_])
+    if len(p_) > 2 and nsy_ > 4:
+        if n.random.random() < .2:  # use pi in 1/5 of sentences
+            print('\n', p, countSyllables(p))
+            where = n.random.randint(len(p_)-2)
+            p_.insert(where+1, 'pi')
+            if p_[where+2][0] not in vowels or (p_[where+2][0] in vowels and p_[where][-1] in vowels):
+                # choose a words with two or three syllables
+                # and make them one syllable shorter
+                p__ = ' '.join(p_)
+                while countSyllables(p__) != countSyllables(p):
+                    sy__ = n.array([len(getSyllables(i)) for i in p_])
+                    ok = (sy__ > 1).nonzero()[0]
+                    chosen = n.random.choice(ok)
[…]

### 2017-12-06 tokipona/makeTexts.py
arrumada sintese de sentencas

@@ -281,7 +281,7 @@ def createSentence(n=0, v=0, o=0, p=0,prep=True):
     if prep:
         if not p:
             p = np.random.randint(1, 5)
-        sentence += ' ' + createPhrase(nwords=p)
+        sentence += ' ' + np.random.choice(prepositions) + ' ' + createPhrase(nwords=p)
     return sentence
 
 def createPoem(nsyl=10, lstan=4, nstan=6, rhyme=True):
@@ -356,7 +356,7 @@ def createParagraph(nsentences=0, words=[]):
         nsentences = n.random.randint(3,10)
     sentences = []
     for i in range(nsentences):
-        sentences.append(createSentence(prep=n.random.randint(0,2)))
+        sentences.append(createSentence(1,1,1,1,prep=n.random.randint(0,2)))
 
     w = set()
     s1 = []
@@ -375,18 +375,11 @@ def createParagraph(nsentences=0, words=[]):
     w_ = w2.difference(w)
     while s2 and w_:
         ss = createSentence(prep=n.random.randint(0,2))
-        for w in w_:
+        w__ = w_.copy()
+        for w in w__:
             if w in ss:
                 s1.append(ss)
                 s2.pop()
                 w_.remove(w)
     paragraph = ". ".join(s1+s2)
     return paragraph
-
-
-            
-
-
-
-
-

### 2017-12-07 article/article.tex
added some synthesized texts to article

@@
 % \newcommand{\tttt}[1]{\texttt{#1}}
+\usepackage{subcaption}
 
@@
 
-Automatic and randomic synthesis of texts in Toki Pona are
+\begin{figure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/phrase_}
+          \caption{1a}
+            \label{fig:sfig1}
+  \end{subfigure}%
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/sentence_}
+          \caption{1b}
+            \label{fig:sfig2}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/paragraph_}
+          \caption{1b}
+            \label{fig:sfig2}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/poem_}
+          \caption{1b}
+            \label{fig:sfig2}
+  \end{subfigure}
+  \caption{plots of....}
+  \label{fig:fig}
+\end{figure}
+
+Automatic and randomized synthesis of texts in Toki Pona are
 useful because of the reduced vocabulary: ideas often make
@@
 % each with a different color scheme
+
+% nanpa followed by numbers
+% pu, taso, kin/namako
+% particles
+% rhymes 
+% modes in which the words are only used as specified in the dictionary
 

### 2017-12-07 tokipona/makeTexts.py

@@ -2,6 +2,7 @@ import pandas as pd
 import numpy as n
 import os
 # from makeStatistics import getSyllables
+ri = n.random.randint
 
 vowels = 'aeiou'
 consonants = 'jklmnpstw'
@@ -356,7 +357,7 @@ def createParagraph(nsentences=0, words=[]):
         nsentences = n.random.randint(3,10)
     sentences = []
     for i in range(nsentences):
-        sentences.append(createSentence(1,1,1,1,prep=n.random.randint(0,2)))
+        sentences.append(createSentence(ri(1,4),ri(1,4),ri(1,4),ri(1,4),prep=n.random.randint(0,2)))
 
     w = set()
     s1 = []

### 2017-12-08 article/article.tex
revisada parte de syntax highlighting

@@
 
-\subsection{Phonology}
+\subsection{Phonology}\label{phonology}
 Words in Toki Pona are written using only 14 letters:
@@
 
+\begin{figure}     \center
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/phrase_}\vspace{-0.15cm}
+          \caption{Synthesized phrase.}
+            \label{fig:sfig1}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.9\textwidth}
+    \centering
+        \includegraphics[width=1.2\linewidth]{figs/sentence_}\vspace{-0.15cm}
+          \caption{Synthesized sentence.}
+            \label{fig:sfig2}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.8\textwidth}
+    \centering
+        \includegraphics[width=1.2\linewidth]{figs/paragraph_}\vspace{-0.15cm}
+          \caption{Synthesized paragraph.}
+            \label{fig:sfig2}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.7\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/poem_}\vspace{-0.15cm}
+          \caption{Synthesized poem.}
+            \label{fig:sfig2}
+  \end{subfigure}
+  \caption{Toki Pona texts synthesized by the scripts
+  in~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
+  and (d) a poem.
+  The unusual formations are convenient for exploring semantic
+  possibilities.
+  A user might obtain many of such structures to select those
+  he/she finds fit for the intended employment.
+  For more information, see Section~\ref{synth}.
+  The words are colored in accordance to the considerations
+  of Section~\ref{shigh}.}
+  \label{fig:syn}
+\end{figure}
+
 The same package~\cite{tokipona}
@@
 might be enhanced as described in Section~\ref{conc}.
-
-\begin{figure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/phrase_}
-          \caption{1a}
-            \label{fig:sfig1}
-  \end{subfigure}%
-  \begin{subfigure}{.5\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/sentence_}
-          \caption{1b}
-            \label{fig:sfig2}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/paragraph_}
-          \caption{1b}
-            \label{fig:sfig2}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/poem_}
-          \caption{1b}
-            \label{fig:sfig2}
-  \end{subfigure}
-  \caption{plots of....}
-  \label{fig:fig}
-\end{figure}
 
@@
 
+\clearpage
 \subsection{Syntax highlighting}\label{shigh}
 The same package~\cite{tokipona}
-has a Vim syntax highlighting plugin
-for Toki Pona.
-Instructions for installing and using
-the syntax highlighting is at~\cite{tokipona}.
+has a Vim~\cite{vim} syntax highlighting plugin
+for Toki Pona,
+routines to arrange the coloring schemes
+(i.e. to yield the text files with the specifications for which tokens
+to relate to which coloring groups),
+and instructions for installing and using
+the resulting syntax highlighting for Toki Pona texts~\cite{tokipona}.
 
 Basically, it distinguishes the words among the morphosyntactic
-classes according to the official dictionary.
-As a word often belongs to more than one class,
-The precedence of them might be set by the user.
-Also, some classes might be further refined or joined,
-such as by distinguishing only particles and the rest,
-or maybe particles and prepositions and the rest.
+classes according to the official dictionary as given in
+Table~\ref{foobar}.
+A word often belongs to more than one class,
+thus the precedence of them might be set by the user.
+Also, some classes might be further refined (e.g. words beginning with
+vowels) or joined (e.g. distinguishing only particles and the rest,
+or particles and prepositions and the rest).
 The colors are also promptly changed according to~\cite{vimArt}
@@
 Currently, the Python package synthesizes the
-syntax file.
+syntax file through the\\ \tttt{makeVimSyntax.py} script.
 The user has control of class precedence and
-merging.
+merging and further details through tweaking such routine.
 The choice of precise coloring schemes
-might involve hacking the colorscheme being
+involves fine tuning the color scheme being
 used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
@@
   \item Installation of the plugin.
-  \item Tweak of the syntax file by hand.
-  \item Running the Python script to generate a new syntax file
+  \item Tweaking of the syntax file by hand.
+  \item Running the \tttt{makeVimSyntax.py}
+    Python script to generate a new syntax file
     according to other settings.
-  \item Write a file inside Vim using Toki Pona and save
-    the file with the .tokipona extension.
-    Reload the highlighting scheme whenever you
+  \item Write a file names with the \tttt{.tokipona} or \tttt{.tp} extension
+    (inside Vim) using Toki Pona words.
+    Reload the highlighting scheme using \tttt{:e} whenever you
     change the syntax file by hand or through the Python script.
   \item Access the used highlighted groups with :syntax,
-    Access all the highlighting groups with :so \$VIMRUNTIME/syntax/hitest.vim.
+    Access all the highlighting groups with \tttt{:so
+    \$VIMRUNTIME/syntax/hitest.vim} or \tttt{:hi}.
     Change the coloring of a set of terms by associating
-    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g. Visual):
-    :highlight link tokiponaADJECTIVE Visual.
-    The plugin comes with the :TokiStation command, which opens a window with
-    the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
-    Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files. Another with Readme, and PDF documentation. Another with an IPython.
-    Imported tokipona. It resets completely upon command :TokiClose, closing all created windows.
+    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g.
+    Visual) such as in \tttt{:highlight link tokiponaADJECTIVE Visual}.
 \end{itemize}
@@
 with syntax-highlighted Toki Pona texts.
-A Vim user might run \tttt{:colorscheme blue},
-\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox}
+A Vim user might run e.g. \tttt{:colorscheme blue},
+\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox} or
 \tttt{:colorscheme elflord} to see the same text colored
-with different colorschemes.
+with different color schemes (association of colors to sets of tokens).
 To interfere directly on the colors chosen,
@@
 in the end of the syntax file.
-
-% :TOhtml command
+One might obtain an HTML file with the colored Toki Pona text
+by using the \tttt{:TOhtml} command.
 
@@
 depend heavily on cultural and use factors
-(https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies),
-although there are projects such as Solarized~\cite{solazired}
-(http://ethanschoonover.com/solarized)
-which try to give general solutions for some contexts.
-Standards guidelines for syntax highlighting were not found,
-and might be driven from data visualization considerations.
-Most evident examples of such guidelines are:
+and have scarce scientific studies~\cite{stack}.
+There are informed projects such as Solarized~\cite{solazired},
+which present solutions for some contexts.
+Strikingly, standard guidelines for syntax highlighting were not found.
+Therefore, we considered current data visualization
+theory~\cite{dv1,dv2,dv3,dv4} to glimpse at the potential guidelines:
 \begin{itemize}
@@
   colors such as violet.
-  \item The avoidance of colors of blue and other high frequency
-  colors in small objects, such as letters and words.
+  \item The avoidance of blue and other high frequency
+  colors to fill small objects, such as letters and words.
   \item Explore simplicity and elegance through the use of discrete
@@
   the same color.
-  One might use a power-law to mimic natural occurring phenomena
-  and perception. Or an exponential to map parameters to perceptual
-  stimuli.
-  \item You might always wish to maximize contrast,
-  although taste and less consative combinations might
-  also dictate choices.
+  One might use a power-law to mimic natural occurring phenomena.
+
+  Unfortunately, the dictionary classification of words are not in a
+    power-law distribution and might be better described by a half-normal or by
+overlapping normal distributions over the three peaks ($\sim35-58,
+12-18, 1-5 occurrences$).
+  \item Physical stimuli might be related to perceptual stimuli both 
+    through a power-law or an exponential law (respectively known
+    as Steven's law and Weber-Fechner's law).
+  \item One usually wishes to maximize contrast,
+  although taste and less wearing combinations might
+  also dictate the coloring choices.
   \item Stipulate axes of parameters with which to set colors.
@@
   Other parameters are hue, transparency etc.
-
-  two types of colorschemes: those with dark background,
-  which are more commode at first; and those with a light background,
+  \item One might consider two types of color schemes: those with a dark background,
+  which are more comfortable at first; and those with a light background,
   which are usually impressive (and even annoying) at first,
   but the eye gets used to it and it keeps you more stimulated.
-  The blue color seems to be specially related to physiological
-  stimulation of the body by color~\cite{blue,blue}.
+\item The blue color is specially related to physiological
+  stimulation of the body~\cite{blue,blue2},
+    and programmers often report that a blue background keeps them
+    awake and more concentrated.
+    I found no evidence of this in the scientific literature,
+    and one might simply take my word that programmers say so
+    since I now have programmed daily for more than 10 years.
   \item Consider that the user might stay many hours (and we often do)
@@
   prone to entail a considerable effect in the body and mental
-  activity.
+  activity and thus in the quality of life and work of a programmer.
   \item Syntax sonification: consider mapping the textual structures
-  to sound (e.g. counting of tokens or context as inside a function or
-  class or over or around a variable of conditional or loop.
-  E.g. through MASSA.
-  \item Mode for tuning syntax highlighting with keyboard shortcuts
-  are used to tune individual colors from background and foreground
-  colors and groups.
-  \item How would these techniques be used for coloring Toki Pona?
-  Coloring of keywords that reflects the dictionary classification
-  of words are not in a power law distribution.
-  Maybe better described by a half-normal or by
-  overlapping normal distributions over the three peaks ($\sim35-58,
-  12-18, 1-5 occurrences$).
-  \item Coloring for poems: division by syllables, words, POS tag,
-    total number of syllables or letters of words.
+  to sound (i.e. parametrize the synthesis of sounds e.g. 
+    by the counting of specific tokens inside a script, a function or
+    class or around a variable of conditional or
+    loop)~\cite{mass}.
+  \item Ideally, one should have facilities for tuning 
+    the syntax highlighting (e.g. through keyboard shortcuts)
+  as envisioned in~\cite{vimArt}.
 \end{itemize}
 
-\url{http://pygments.org/docs/}
+On the consideration of writing and of Toki Pona,
+the considerable irrelevance of the morphsyntactic classes as
+described in Section~\ref{} suggest that a appropriate coloring should
+consider the syntax to identify the nouns, verbs, adjectives, adverbs,
+prepositions and particles.
+Also, the coloring for poems might be more appealing if considered
+e.g. the counting of syllables or repeated vowels and ending
+syllables.
+Insights might be obtained through the observation of the
+choices made and advocated in software packages such as text editors
+(e.g. Vim, Emacs, Sublime, etc.),
+software dedicated to syntax highlighting (e.g.
+pigments \footnote{\url{http://pygments.org/docs/}} and 
+highlight.js \footnote{\url{https://highlightjs.org/}},
+other software (e.g. Linux terminals such as Xterm and gnome-terminal).
+Finally, the careful choice of fonts is known to have
+a relevant impact in comfort and productivity~\cite{fonts},
+e.g. sans-sheriff fonts are more promptly read and yield a cleaner
+text than a fot with sheriff.
 
@@
     Different modes that might be specified by a section header or tag.
+  \item The plugin comes with the :TokiStation command, which opens a window with the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
+Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files.
+Another with Readme, and PDF documentation.
+Another with an IPython.
+Imported tokipona.
+It resets completely upon command :TokiClose, closing all created windows.
 \end{itemize}
@@
     \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
+\bibitem{stack}
+  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
+    Software Engineering in Stack Exchange. Available at
+    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
+\bibitem{solarized}
+  Schoonover, E. Solarized color scheme. Available at
+  \url{http://ethanschoonover.com/solarized}
 \end{thebibliography}

### 2017-12-08 tokipona/makeTexts.py

@@ -346,6 +346,7 @@ def createParagraph(nsentences=0, words=[]):
 
     Might not have all words if nsentences < len(words).
     Discards Toki Pona words not recognized.
+
     """
     ww = []
     for w in words:

### 2017-12-13 article/article.tex
terminado primeiro rascunho

@@
 
+% reference to figure of the texts.
+
 \subsubsection{Advanced syntax highlighting considerations}
@@
 e.g. sans-sheriff fonts are more promptly read and yield a cleaner
-text than a fot with sheriff.
+text than a for with sheriff.
 
 \subsection{Wordnet Toki Pona}\label{wn}
-exception: jan (people)
-is a hyponym of soweli (mammal),
-kili is hyponym of kasi,
-(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
-But otherwise such relations hold:
-e.g. luka (as hand) is e meronym of sijelo (body).
-
-For a first achievement of a Toki Pona Wordnet,
+For the achievement of a first Toki Pona Wordnet,
 each of the Toki Pona words in the dictionary
@@
 relate each Toki Pona word to the synsets
-related to all the words bounded by a semicolon.
-Another possibility, which I find more consistent
+simultaneously associated to all the English
+words bounded by a semicolon in the dictionary.
+Yet another possibility, which I find more consistent
 but also more controversial,
-is to choose all synsets related to the Wnglish
+is to choose all synsets related to the English
 words given that the Toki Pona words can be used
 in many roles.
-WordNet only contains nouns, adjectives, verbs, and adverbs.
+Wordnet only contains nouns, adjectives, verbs, and adverbs.
 Thus, particles and prepositions were not considered.
@@
 Words presented as adjectives in the dictionary
-were considered adjectives and adverbs.
-
-The total number of different synsets is 2493 in a very conservative
+were considered adjectives and adverbs.~\cite{wordnet}
+
+The total number of different synsets is $2493$ in a very conservative
 counting: discarding prepositions and particles,
-only considering the synsets which have the POS tag
+and only considering the synsets which have the POS tag
 given in the dictionary.
@@
 related to a lemma are considered, 
-then Toki Pona vocabulary reaches 3961 synsets.
-The prepositions are particularly missing in these
+then Toki Pona vocabulary reaches $3961$ synsets.
+The prepositions are particularly missing in such
 countings because e.g. tawa or lon are used also
@@
 If synsets for them are considered those of the English
-lemmas with any POS tag, the vocabulary reaches 4052
+lemmas with any POS tag, the vocabulary reaches $4052$
 synsets.
-The total number of synsets in Wordnet is X.
-Total number of synsets counting neighbors.
-Depth of TP synsets.
-Using such features to devise other conlangs.
-
-\subsection{Hacks from other people}
-The \url{tokipona.net} has a number of tools,
-just as to transliterate names into Toki Pona phonemes,
-and search in corpus.
+The total number of synsets in Wordnet is $117659$.
+Total number of synsets counting neighbors is Y.
+Depth of TP synsets is .
+We envision the usage of such features to guide creation
+of other conlangs.
+
+The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
+makes available such a tentative Wordnet in the simplest form:
+the Toki Pona words are keys in a dictionary that returns the
+corresponding synsets.
+Three of such dictionaries are implemented:
+one where all synsets related to the lemma is
+returned; another as such but excluding the prepositions;
+one that returns only the synsets
+registered in Wordnet with the same POS tag as the lemma
+is in the official dictionary.
+
+It is worth mentioning that Toki Pona words are very general
+and each of them might mean many things.
+Thus, the words are not very easily related e.g. by hyponymy
+or meronymy.
+Exceptions: jan (people) is a hyponym of soweli (mammal);
+kili is an hyponym of kasi;
+walo, pimeja, jelo, loje, laso
+are hyponyms of kule;
+luka (as hand) is a meronym of sijelo (as body).
+
+\subsection{Hacks and resources from other people}
+A number of tools and other resources for dealing with Toki Pona
+were developed by the community.
+The most complete list is supposedly~\cite{gdoc} and includes
+videos, musical pieces, artistic texts, reference documents (e.g.
+a Toki Pona - Esperanto dictionary), journalistic articles,
+and software tools.
 

### 2017-12-13 tokipona/makeWordnet.py

@@ -23,7 +23,9 @@ for row in table:
         class_ = [i for i in pos.split() if i.isupper() and i not in ("I,","O!")][0]
         en_words = [i for i in pos.replace(" or ", ' ').split() if i.islower() or i in ("I,","O!")]
         for enw in en_words:
-            w = enw.replace(";", "").replace(',', '')
+            if enw in ('to','a','of','and','in','take','on','form','at','for'):
+                continue
+            w = enw.replace(";", "").replace(',', '').replace('(', '').replace(')', '').replace('?', '')
             lemma_pos.append( (w, class_) )
 
     words_ = row[0].split()
@@ -36,7 +38,7 @@ pos = {'NOUN': 'n',
         'VERB': 'v',
         'PRE-VERB': 'v', # pre-verbs are also verbs
         'ADJECTIVE': 'asr', # adjectives are also adverbs and satellite adjectives
-        'NUMBER': 'asr'
+        'NUMBER': 'asr' # numbers are qualifiers
         }
 
 import nltk as k

### 2017-12-15 article/article.tex
terminada revisao

--- 2017-12-08 article/article.tex
@@
   A minimalist constructed language (conlang)
-  is useful for experiments and is comfortable for making tools.
-  The Toki Pona conlang is minimalist both in the vocabulary
-  (only 14 letters, X phonemes, and 120 words + 4 synonyms)
+  is useful for experiments and comfortable for making tools.
+  The Toki Pona (TP) conlang is minimalist both in the vocabulary
+  (with only 14 letters and 124 words)
   and in the $\approx10$ syntax rules.
   The language is useful for being a used and somewhat established
-  minimalist conlang with at least hundreds of speakers.
+  minimalist conlang with at least hundreds of fluent speakers.
   In this article, we describe current concepts and resources
-  for Toki Pona,
+  for TP,
   and make available Python scripted routines for
-  analysis and synthesis of texts and the specification of syntax highlighting schemes.
+  the analysis of the language,
+  the synthesis of texts, the specification of syntax highlighting
+  schemes, and the achievement of a preliminary TP
+  Wordnet~\cite{wordnet}.
   We focus on the analysis of the basic vocabulary,
   as corpus analyses were found in~\cite{corpus}.
-  The synthesis is based on the basic sentence template,
+  The synthesis is based on sentence templates,
   relates to context by keeping track of used words,
@@
   and number of sentences, words and letters (e.g. for paragraphs).
-  Syntax highlighting is performed in 24bit true colors, currently only
-  with the Vim text editor.
-  Word colors reflect morphosyntactic classes given in the official dictionary
-  and different solutions to necessary and arbitrary choices are described and implemented.
+  Syntax highlighting 
+  reflects morphosyntactic classes given in the official dictionary
+  and different solutions are described and implemented
+  using 24bit true colors in the well-established Vim text editor~\cite{vim}.
+  The tentative TP Wordnet is made available in three forms that
+  reflect the choices of the synsets related to each word.
   In summary, this text holds potentially novel conceptualizations about,
   and tools and results in analyzing, synthesizing or syntax highlighting the
-  Toki Pona language.
+  TP language.
   % subj (li), obj (e), prepositions, qualification (word concatenation), 
@@
 \end{abstract}
-{\bf keywords:} Natural Language Processing, Textual synthesis, Syntax highlighting, Wordnet, Toki Pona
+{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona
 
 \section{Introduction}\label{intro}
-Toki Pona is a minimalist conlang (constructed language)
+Toki Pona (TP) is a minimalist conlang (constructed language)
 with only 124 words (120 without the synonyms).
@@
 which entails easier e.g. learning and tool making.
-Another reason why the minimalistic language design is compelling
+Another reason why the minimalist language design is compelling
 is the study and harnessing of the strong and weak forms of
 the Sapir-Whorf hypothesis (linguistic relativity),
-i.e. that language influences or dictates ones thought
-and world experience.
+i.e. that language influences or dictates thought
+and world experience~\cite{sapWho}.
 Accordingly, one uses a conlang as a thinking tool (or platform)
@@
 to simplify the thinking processes, and as a way to modify
-the mood and impressions about the world.~\cite{tp1,tp2,tp3,tp4}
+the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
 In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
-describes that she has seen the language been used successfuly
+describes that she has seen the language been used successfully
 in the context of management, creation of texts, legal texts,
@@
 In this article, we present a conceptual overview of the language,
-which is fit both to the newcomer and to the expert.
-We also summarize
-texts and software gadgets
-for Toki Pona that
-were developed by us and by other Toki Pona users.
-Most importantly, the conceptual overview is considerably different
-from what we found in the literature~\cite{pu,pije,memrise},
-with emphasis on simplicity and flexibility,
-and the software routines yield analyses, syntheses
-and highlighting Toki Pona what was thought to be complementary
-to the software currently available.
+which is fit both to the newcomer and to the expert
+for being considerably different
+from what we found in the literature~\cite{tpLang,kama,memrise},
+with emphasis on simplicity and flexibility.
+We also present novel software routines for analysis, synthesis,
+syntax highlighting, and the achievement of a tentative TP
+Wordnet~\cite{wordnet}.
 
@@
 a historical note, and some words about natural and constructed languages.
-Section~\ref{basic} describes Toki Pona's phonology and syntax.
+Section~\ref{basics} describes TP's phonology and syntax.
 Section~\ref{hacks} presents the software routines we made available
 and immediate results, such as listings and statistics of words,
-poems and short stories, and coloring schemes.
+poems and short stories, coloring schemes, and preliminary TP Wordnet
+versions.
 Conclusions and further work are in Section~\ref{conc}.
 Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
-with thoughts of rule breaking and potentially new conlangs.
+with thoughts about rule breaking and potentially new conlangs.
 Appendix~\ref{ftp} holds final words in Toki Pona.
@@
 language in: references and learning material, corpus,
-websites, interaction groups (here users talk a post texts and comments), and software gadgets.
+websites, interaction groups
+(where users talk a post texts and comments), and software gadgets.
 The main references of the language are:
-  the official book ``Toki Pona: The Language of Good'',
+  the official book ``Toki Pona: The Language of Good''~\cite{tpLang},
     authored by Sonja Lang, the creator of the language;
-the online book ``o kama sona e toki pona!'',
+and the online book ``o kama sona e toki pona!'',
 from jan Pije~\cite{kama}.
+A number of tools and other resources for dealing with Toki Pona
+were developed by the community.
+The most complete list is supposedly~\cite{gdoc} and includes
+videos, musical pieces, artistic texts, reference documents (e.g.
+a Toki Pona - Esperanto dictionary), journalistic articles,
+and software tools.
 For a more comprehensive view of the resources available
 for the user, we suggest following the links 
-from~\cite{wikiToki}.
-Section~\cite{third} holds a summary of what we found in software
-gadgets related to Toki Pona.
+in~\cite{gdoc,wikiToki}.
 
 \subsection{Historical note}
-Toki Pona was developed as an internal and personal language
+TP was developed as an internal and personal language
 by Sonja Lang~\cite{interview}.
@@
 reported it to have a few hundred speakers.
-The English official book~\cite{lipu} was released only in 2014.
+The English official book~\cite{tpLang} was released only in 2014.
 In 2016, a version of the official book was released in French.
-Nowadays, one finds a number of texts about Toki Pona and written
-in it, and uses of Toki Pona for artificial intelligence~\cite{roila}
-and software tools~\cite{tptoll1,tptool2,tptool3},
-in social platforms such as Facebook groups~\cite{fbtp,fbtpt},
-microblogging~\cite{tokilili}, Telegram~\cite{telegram} and IRC~\cite{irc}.
+Nowadays, one finds a number of texts about TP and written
+in it
+(e.g. in social platforms such as Facebook groups,
+microblogging, Telegram and
+IRC), and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.
 
@@
 and writing is called `natural language'.
-A `constructed language' is a natural language built by someone or
+A `constructed language' (also planned or invented language) is a natural language built by someone or
 a group, such as Esperanto, Toki Pona, and Lojban.
 An artificial language is a language yield by artificial agents,
-such as in AI routines, and might be considered
-within `cultural evolution' studies.
+such as by AI routines, or by humans in controlled experiments,
+and are considered e.g. within `cultural evolution' studies.
 Formal languages are defined by tokens and rules to
@@
 to math and formal models for natural languages.
-Lojban is a constructed language with the purpose of being
-an artificial language because it is more axiomatic.
 
 Constructed languages are in some traditions called artificial
-languages, but creators most often prefer to use the term `standardized'
+languages, but creators most often prefer to use the term
+`standardized', 'constructed'
 or 'planned' language with the argument that the conlang
-is rooted on natural languages, and 'artificial' is misleading~\cite{conlanWikip}.
-The preferred terms seem to be planned or constructed languages
-or conlang.
+is rooted on natural languages,
+and 'artificial' is misleading.
+The preferred terms seem to be planned or constructed language
+or simply conlang.
 The construction of languages is called glossopoeia.
-Toki Pona is a conlang which might be classified
+TP is a conlang which might be classified
 as engineered for experiments, meditation and philosophy;
-and suitable for auxiliary international language and
-as an artistic language. 
+and suitable for use as an auxiliary international language and
+as an artistic language.~\cite{conlanWikip,artlanWikip}
 % taxonomy, vocabulary, ontology, terminology,
@@
 % non-verbal languages, musical corporal language, 
-% non-verbal languages confront Sapir-Whorf hypothesis.
+% non-verbal languages confront the strong form of the Sapir-Whorf hypothesis.
 
@@
 Furthermore, it is a solid reference of the phonological and syntax
-rules.
+rules, and enables one to understand and modify the software presented
+in Section~\ref{hacks}.
 
@@
 
-Syllables are of the form (C)V(N):
-an optional consonant, a vowel and an optional nasal consonant.
+Syllables are of the form (C)V(n):
+an optional consonant, a vowel and an optional coronal nasal consonant
+(n).
 Non word-initial syllables must follow the pattern CV(N).
-The following sequences are forbidden: ji, wu, wo, ti, nm, nn.
+These sequences are forbidden: ji, wu, wo, ti, nm, nn.
 
@@
 As in other natural languages, 
-colloquial Toki Pona might have
+colloquial TP might have
 incomplete sentences and deviate from the
@@
 The particle li is omitted if the subject is a simple mi (I or us)
-or sina (you). A discussion about issues (problems)
-yield by this rule
+or sina (you). A discussion about issues (potential problems)
+yield by this last rule
 and how I deal with them is in Appendix~\ref{mytoki}.
-
 The e might be repeated to associate more than
 one object to a predicate.
-Sentences might be related though la,
+Sentences might be related through la,
 'sentence' la 'sentence', where the second sentence is
 the main sentence, and the first sentence is a condition
-to the first.
+to the second.
 Multiple la-s are not described in literature,
-but I assume that one might assume the last sentence
-being a conditional to the next,
+but one might assume that the last sentence
+is a conditional to the next,
 except in cases where the context strongly suggests
@@
 The first word is the noun or verb and subsequent words
-qualify the noun or verb.
+qualify the noun or verb (i.e. are adjectives and adverbs).
 The pi particle might be used to separate sequences of words
@@
 
-Notes on the usage of pi:
+\noindent Further notes on the usage of pi:
 \begin{itemize}
@@
     the first, the third word qualifies the phrase yield by the first
-    two words, the fourth word qualifies the noun yield by the first
+    two words, the fourth word qualifies the phrase yield by the first
     three words and so on.
   \item It is redundant to use pi before the last word in a noun or
-    verb phrase if there is no other pi, reason why it is most often
+    verb phrase, reason why it is most often
     omitted.
-    Its use in this case is regarded as wrong~\cite{lipu,pije},
+    Its use in this case is considered an error~\cite{tpLang,kama},
     but, as one might notice, it does not add (much) information
-    through syntax (order of qualifies is conserved).
+    through syntax because the order of qualifications is conserved.
     It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
@@
     after li to mean possession, e.g. `soweli li pi sina' (your pet).
-    This employment of pi might be regarded as correct, but are promptly written
+    This employment of pi might be regarded as correct, but is promptly written
     as a noun phrase (e.g. `soweli sina') and is not mentioned
-    by the official book~\cite{tpLabg}.
+    by the official book~\cite{tpLang}.
 \end{itemize}
 
-All the words except the structural particles (li, e, la, pi)
+All the words except the particles (li, e, la, pi, a, o, anu, en, seme, mu)
+and the words classified solely as
+prepositions (kepeken, lon, tan)
 are usable in any position in noun and verb phrases,
-(i.e. 116 words).
-Notice that the phrase expresses a noun in a noun phrase (subject or
-object) or a verb (in the predicate).
+(i.e. 107 words discarding synonyms).
+Notice that the phrase expresses a noun (in a subject or
+object) or a verb (in a predicate).
 And that the first word of the phrase is the noun or verb,
 and that subsequent words are adjectives or adverbs.
-The pi and other particles restarts the phrase to
-first word being a noun.
+The pi precedes a noun or a verb to be qualified and then qualify
+the phrase that comes before it.
 
@@
 the prepositions: kepeken, lon, sama, tan, tawa.
-They might appear at the end of phrases,
+They might appear at the end of a phrase,
 should be followed by another phrase,
@@
 Because prepositions can be used as adjectives
-or nouns or verbs, adverbs, etc, and what follows
-a preposition complement might also be of any morphosyntactic class,
-they are core sources of ambiguity.
-three:
-* preposition
-* absence of li on sina and mi
-* Incomplete sentences
-(ona li) tawa lon kon e ilo pi suli mute
-
-\subsubsection{Particles}
+or nouns or verbs or adverbs, etc, and might follow
+anu noun or verb phrase,
+they are one of the three main sources of ambiguity:
+prepositions; absence of li after sina and mi; incomplete sentences.
+E.g. '(mi [li]) pana tawa kon e ilo pi suli mute'.
+
+\subsubsection{Particles}\label{sec:par}
 Beyond the structural particles (li, e, la, pi) presented
@@
   \item taso, means however as sentence or 'only' if adjective.
-  \item anu, en: 'or' and 'and'. Used for nouns in nous phrases.
-    For repetition of verbs, repeat li.
+  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
+    For verbs, repeat li.
     For object nouns, repeat e.
-    If the noun is complementing a preposition (tawa, lon),
+    If the noun is complementing a preposition (e.g. tawa, lon),
     one might repeat the preposition.
-    As Toki Pona is a recent language, and is able to cope with
+    As TP is a recent language, and is particularly able to cope with
     variation due to its simplicity, I would advocate for
     using en and anu wherever there is no ambiguity.
-    E.g.: 'mi tawa anu moku lon tenpo lili?'
+    E.g.: 'mi tawa mute anu moku lon tenpo lili'
     In the official documentation that is not described and
-    might be regarded as wrong in strict canonical.
+    might be regarded as wrong in strict canonical TP.
   \item nanpa, denotes numbering.
   \item seme, for questions, used next to the thing being asked for.
-    'Why?' might be expressed as
-    'seme la sina pana e moku lon sewi'.
+    'tan seme la sina pana e sike?'.
   \item mu, for animal noises. For me it is not a particle, as in the
@@
     I also like to use it as a verb:
-    'mi pakala e luka. mu mute.'
+    'mi pakala e luka. mi mu mute.'
 \end{itemize}
 
-The vocabulary specifies morphosyntactic classes:
-nouns, adjectives, verbs, pre-verbs, adverbs, particles, prepositions, and numbers.
-I find that they might help the user and newcomer, but
+The vocabulary specifies these morphosyntactic classes:
+nouns, adjectives, verbs, pre-verbs, adverbs, prepositions, particles, and numbers.
+I find that it might help the speaker, specially the newcomer, but
 it might also suggest a deviation from what I understand and read:
-the words might be used indistinctly to be the nouns
-(subjects, the predicate when there is no object, objects,
-and preposition complements),
-the adjectives (anything that does not start the noun phrase or follows a pi),
-verbs
-(follows mi or sina or li or a preposition),
-adverbs (follows the verb).
+the words might be used indistinctly to be a noun
+(the first word of: a subject, a predicate when there is no object, an object,
+or a prepositional complement),
+an adjective (anything that follows an noun and is not a particle or a
+preposition),
+a verb (follows mi or sina or li),
+or an adverbs (follows a verb).
 The pre-verbs (wile, ken, awen, kama, lukin, sona),
-might follow a verb, but might also be understood
-as the verb qualified with the next word,
-which carries a very similar if not identical meaning.
-The pre-verbs are all also defined as other morphosyntactic classes,
-such as adjective, noun, verb.
-The only exception is wile, which is only a pre-verb.
+might be followed by a verb, but might also be understood
+as a verb qualified by the next word,
+which carries a very similar if not identical meaning
+('wile moku' might be understood both as a pre-verb followed
+by a verb and as a verb followed by an adverb).
+The pre-verb words are all also specified in the official dictionary
+to have other morphosyntactic classes,
+such as adjective, noun, and verb.
+The only exception is wile, which is specified to be only a pre-verb.
 
@@
 
-As far as I can see, one should regard
-the particles li, e, pi, and la
-and punctuation.
-The other tokens of the vocabulary
-might be used in any of the remaining positions.
-
-\subsubsection{Convention for recognizing POS tags by speaker and machine}
-
+\subsubsection{Recognizing POS tags by speaker and machine}
+One might perform a POS tagging by following these rules:
 \begin{itemize}
   \item Noun: the first word in a noun phrase.
-    After an 'e' and after a pi,
-    The first word in the sentence if
-    sentence does not start by the verb.
-    Might be in the position of the verb
+    Starting a sentence if it is complete,
+    after an 'e', after a pi, and after a preposition.
+    The predicate is often a noun
     if the sentence has no object.
   \item Adjective: second word on after an e and after
-    Second word on in the subject phrase if present.
+    second word on in the subject phrase if present.
   \item Verb: after a li, mi or sina.
-    If there is no object, the verb position
-    is often a noun.
+    If there is no object, the predicate
+    might be a noun or a verb.
   \item After a preposition,
-    there can be a noun phrase, a verb phrase
-    or nothing.
+    there is a noun phrase.
   \item Notice that there is ambiguity in the structure
@@
     Also, when there is no object, a noun or a verb
-    or an adjective might be in the verb position
-    if there is no object.
-    The prepositional complement is also not defined.
-    (mi moku tawa pali, tawa tomo).
-    So, these are sources of syntactic ambiguities
-    in Toki Pona.
-    They might be solved or minimized by using the semantics
+    or an adjective might be in the verb position.
+    % The prepositional complement is also not defined.
+    % (mi moku tawa pali, tawa tomo). % TTM
+    These are sources of syntactic ambiguities
+    in TP.
+    They might be resolved or minimized by using the semantics
     of the words.
-    One preliminary effort in this direction might be
+    An initial effort in this direction might be
     using the classes in the official dictionary to resolve
     ambiguities whenever possible.
-    This solution is not optimal in correct POS tagging,
+    This solution is not optimal for correct POS tagging,
     and does not solve all possible ambiguities
-    (there are words classified as nouns and adjectives,
-    adjectives and verbs, nouns and verbs, particle and verb).
-
+    (there are words classified as noun and adjective,
+    adjective and verb, noun and verb, particle and verb).
     Another source of ambiguity is the pre-verbs as described
-    in the literature~\cite{tpLang,pije}.
-    But I find it reasonable to understand them as verbs.
-
+    in the literature~\cite{tpLang,kama}.
+    But I find it reasonable to understand them as verbs,
+    as described in the end of Section~\ref{sec:par}.
+  \item If there is no object, the predicate might be parsed
+as being a hybrid of noun, verb and adjective.
+  \item The prepositional phrase might also be taken to be a hybrid
+    of noun, verb and adjective.
 \end{itemize}
 
-Also, the prepositional complement might
-be a noun or a verb.
-I could not come up with a sentence
-where it would be understood as an adjective.
-
-\subsubsection{Further notes}
+% Also, the prepositional complement might
+% be a noun or a verb.
+% I could not come up with a sentence
+% where it would be understood as an adjective. TTM
+
+\subsubsection{Additional notes}
+John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
+states that there are not noun, verb or prepositional
+phrases in TP, but only phrases in a structure
+yield by particles and content words~\cite{janKipo}.
 The only synonyms on Toki Pona are:
-a and kin; lukin and oko;
-sin or kamako;
+a and kin; lukin and oko; sin or kamako;
 ale or ali.
-
 In formations such as
-toki e ni:, wile e ni:, tan ni: etc.
-'(e) ni' can be omitted and : used alone.
-
+toki e ni:, wile e ni:, tan ni: etc.,
+'(e) ni' might be omitted and : used alone.
 Names are by default transliterated,
-but might not be, as described in Section~\ref{mytoki}.
+but might not be.
+These and other issues are further described in Appendix~\ref{mytoki}.
 
 % my view
+Many other considerations might be made about the language,
+such as additional (deprecated and new) words,
+situations where li is used and avoided,
+canonical and alternative employment of pi,
+counting systems,
+associations of TP word sequences to words in English,
+usage of images to represent TP words and sentences,
+and best practices in writing TP texts or for translation.
+For these and other matters, visit~\cite{tpLang,kama,gdoc,tp4}.
+
 % considerations about the official book and jan Pije
@@
 In this section,
-we describe software, natural language, audiovisual and numeric results
-on four directions:
+we describe software, statistic, natural language processing,
+and visualization results:
 1)
 the analysis of the Toki Pona language through statistics
-obtained by processing the dictionary;
+of the vocabulary obtained by processing the official dictionary;
 2)
-the synthesis of sentences, poems and short stories
+the synthesis of sentences, poems and paragraphs (e.g. short stories)
 in Toki Pona;
@@
 4)
-initial considerations about the WordNet of
+an initial Wordnet~\cite{wordnet} of
 Toki Pona words.
-Corpus analysis was already found in~\cite{}.
-
+Corpus analysis was already found in~\cite{corpus}.
 The Python scripts (or the toolbox)
@@
 generation of derivatives by other interested parties.
+Such scripts were made very friendly for one to understand and alter
+at will.
 
 \subsection{Statistics of the vocabulary}\label{sec:stat}
-In~\cite{pije} are statistics about Toki Pona corpus.
+In~\cite{corpus} are statistics about a Toki Pona corpus.
 This section focuses on the statistics of the vocabulary
-and syntactic rules:
+and of the syntactic rules:
 the letters, phonemes, word sizes,
 possible combinations for words and sentences.
-The statistics are in Appendix~\cite{listings},
-and the next paragraph is an overview.
+% The statistics are in Appendix~\cite{listings},
+% and the next paragraph is an overview. TTM
 The script \tttt{makeStatistics.py}
@@
 discussed herein,
-and hold some other measurements and sets of words
-which were regarded as less suitable for this exposition.
-
-As described in Section~\cite{basics},
+and hold further measurements and sets of words
+which were regarded as less suitable for this exposition,
+but is useful for a deeper investigation.
+
+As described in Section~\ref{basics},
 there are only 14 letters,
-and phonemes respect a few rules.
+and phonemes are also very restricted.
 There are 120 different words in the official vocabulary,
 4 of them having synonyms.
-A total of 124 tokens,
+A total of 124 tokens (or lemmas),
 not counting proper nouns (names)
 and punctuation.
-Table~\ref{tab:pos} shows the number of words
-related to each POS tag specified in the dictionary.
+Table~\ref{foobar} shows the number of words
+related to each POS tag (or morphosyntactic class)
+specified in the dictionary.
 
@@
 and 13 (10.48\%) have three syllables.
-No official words has four or more syllables.
-Of all the 235 syllables of the official words,
-li, la, ka, na, and pa are the most often (with 5.53\%,
-4.26\%, 3.83\%,  3.83\%, 3.83\%).
-On the first syllable of each word,
-a, o, pi, ka, and la are the most often
-(with 6.45\%, 4.03\%, 4.03\%, 3.23\%, 3.23\%).
-On the last syllable of each word,
-li, lo, na, la, and ma are the most often
-(with 8.06\%, 4.84\%, 4.84\%, 4.03\%, 4.03\%).
+No official word has four or more syllables.
+In all the 124 tokens, there are 235 syllables (68 different),
+and Table~\ref{tab:freqsy} exhibits the 10 most frequent
+syllables in the first, last, middle and all positions.
 Middle-word syllables only occur 13 times,
-and are all different, with the exception of `la',
+and are all different, with the exception of 'la'
 which occurs twice.
-For the frequencies of all syllables in every position,
-run the script \tttt{makeStatistics.py} in~\cite{tokipona}.
-
-Vowel and consonant frequencies are as shown in Table~\ref{freqLet}
-for starting, end and internal position.
-Comparison of vowels against consonant is in Table~\ref{tab:vcon}.
-The high incidences e.g. of 'l' and 'a' favor
-the naturalness of the language prosody
-as it resembles child speech.
-
-\input{vowels}
-
-26, 85 and 13 words have one, two and three syllables
-(20.96, 68.54 and 10.48\% of all 124 tokens).
-No official word have more than three syllables
-(although proper names may be of any size).
-In all the 124 tokens, there are 235 syllables (68 different).
-Most often syllables are in
-Table~\ref{syl}.
 A complete list of the syllables and their frequencies (in different
@@
 and ordered alphabetically and by the number of letters,
-is in \tttt{hlsyl\_\_}.
+is in the \tttt{hlsyl\_\_} variable.
 
@@
 
+Vowel and consonant frequencies,
+and comparisons of vowels against consonants,
+are as shown in Table~\ref{freqLet}
+for starting, end and internal position.
+The limited number of consonants favors
+the naturalness of the language prosody
+as it resembles babbling.
+
+\input{vowels}
+
 Given the rules given in Section~\ref{phonology}
-(of 14 letters, 5 vowels, (C)V(N) phonemes,
+(of 14 letters, 5 vowels, (C)V(n) phonemes,
 forbidden ji, wu, wo, ti, nn, nm),
@@
 8256 with 2 syllables, 710016 with 3 syllables.
-No middle syllable end with the nasal consonant 'n'.
-
-Possible syntactic structures are unlimited.
+In the incident words of the official dictionary,
+no middle syllable end with the nasal consonant 'n'.
+
+Possible sentence structures are unlimited.
 For a noun or verb phrase, one might quantify the possibilities
-by considering all words but the particles that are nothing else
+by considering all words but the particles that are not classified
+also as something else
 (i.e. li, e, la, pi, a, o, anu, en, seme, mu)
-and the presotions that are nothing else (i.e. kepeken, lon, tan).
+and the prepositions that are not classified also as something
+else (i.e. kepeken, lon, tan).
 This yields $120-13=107 words$.
@@
 verbs and adverbs.
-Thus, one have 107 possibilities of noun and verb phrases with
-one word, $107^2=11449$ possibilities with two words,
-$107^3=1225043$ with three words and so on.
+Thus, one has 107 possibilities of noun and verb phrases with
+one word, $107^2=11,449$ possibilities with two words,
+$107^3=1,225,043$ with three words and so on.
 Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
@@
 $\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
- 4.300066\times 10^+12$
+ 4.300066\times 10^+12$,
 i.e. more then 4 trillion possible
@@
 The general case is implemented as a function
-in the script \tttt{makeStatistics.py}~\cite{tokipona}
+in the script \tttt{makeStatistics.py}~\cite{tokipona}.
 
@@
 The syntax organizes the words in larger structures.
-The rhymes are restricted and sonorities are
-restricted by the small vocabulary and simple syntax.
+The rhymes are very restricted and sonorities are
+bounded by the small vocabulary and simple syntax.
 There are some specific tasks for achieving texts,
@@
 to choose sentences or verses or stanzas.
+The same package~\cite{tokipona}
+has capabilities for synthesizing TP text.
+On its most basic level, these routines yield
+noun, verb and prepositional phrases,
+and sentences.
+They also aim at making larger scale texts
+by assuring the use of specific words (to entail context),
+e.g. to assist the creation of short narratives,
+and by employing stylistic outlines for poems.
 
@@
   \end{subfigure}
-  \caption{Toki Pona texts synthesized by the scripts
+  \caption{Toki Pona texts synthesized by the \tttt{makeTexts.py} script
   in~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
   and (d) a poem.
-  The unusual formations are convenient for exploring semantic
+  The unusual word combinations are convenient for exploring semantic
   possibilities.
-  A user might obtain many of such structures to select those
-  he/she finds fit for the intended employment.
+  One might obtain many of such texts to select the excerpts
+  he/she finds fit for the intended use.
   For more information, see Section~\ref{synth}.
   The words are colored in accordance to the considerations
-  of Section~\ref{shigh}.}
+  in Section~\ref{shigh}.}
   \label{fig:syn}
@@
 
-The same package~\cite{tokipona}
-has capabilities for synthesizing Toki Pona text.
-On its most basic level, the routines yield
-noun, verb and prepositional phrases,
-sentences.
-They also aim at making larger scale texts
-by keeping a record of the used words and structures (context)
-and using stylistic outlines for poems and short narratives.
-
 Figure~\ref{fig:syn} holds some synthesized texts
-with different coloring schemes for syntax highlighting,
+with different color schemes for the syntax highlighting,
 as presented in the next section.
 The textual synthesis described here,
-and implemented in \tttt{makeTexts.py}
-might be enhanced as described in Section~\ref{conc}.
-
-Automatic and randomized synthesis of texts in Toki Pona are
-useful because of the reduced vocabulary: ideas often make
-sense in unexpected way, and yield a way to 
-explore the semantic field possible by Toki Pona.
-Each words is related to a broad semantic field,
-thus the randomized selection of words are more likely
-to make sense.
+and implemented in \tttt{makeTexts.py},
+might be enhanced in countless ways,
+e.g. as described in Section~\ref{conc}.
+
+Automatic and randomized synthesis of texts in TP is particularly
+useful because of the reduced vocabulary where each word is related
+to a broad semantic field.
+Ideas often make sense in unexpected ways, and thus
+the synthesis yields a procedure to 
+explore the semantic possibilities within TP.
 One might object that the resulting texts are unusual 
 and even consider that they often hold insubstantial or unsound meaning.
-We advocate that these unexpected formations are desirable
+I advocate that these unexpected formations are desirable
 for exploring the possibilities of the language and of thought,
@@
 
-\clearpage
+% \clearpage
 \subsection{Syntax highlighting}\label{shigh}
@@
 the resulting syntax highlighting for Toki Pona texts~\cite{tokipona}.
-
-Basically, it distinguishes the words among the morphosyntactic
+An online syntax highlighting gadget is found at~\cite{tpNetSH},
+and the solution here described presents a number of enhancements:
+it is capable of coloring all morphosyntactic classes;
+it might be fine-tuned in the colors and their relations to sets of
+tokens; it is designed to be used within Vim
+and might be exported as HTML; the highlighting scheme is promptly
+rendered by a script; the resulting script might be used with
+many color schemes.
+
+Basically, the resulting syntax highlighting
+distinguishes the words among the morphosyntactic
 classes according to the official dictionary as given in
@@
 or particles and prepositions and the rest).
-The colors are also promptly changed according to~\cite{vimArt}
+The colors are also promptly changed according to~\cite{vim}
 and exemplified in the package documentation.
@@
 used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
-and Vim's highlighting schemes as described in~\cite{vimArt}.
+and Vim's highlighting schemes as described in~\cite{vim}.
 In summary,
@@
   \item Running the \tttt{makeVimSyntax.py}
-    Python script to generate a new syntax file
+    Python script to generate a new \tttt{tokipona.vim} syntax file
     according to other settings.
-  \item Write a file names with the \tttt{.tokipona} or \tttt{.tp} extension
+  \item Write a file with the \tttt{.tokipona} or \tttt{.tp} extension
     (inside Vim) using Toki Pona words.
@@
     change the syntax file by hand or through the Python script.
-  \item Access the used highlighted groups with :syntax,
+  \item Access the used highlighted groups with \tttt{:syntax}.
     Access all the highlighting groups with \tttt{:so
@@
 
-The tokipona.vim syntax file has an association of toki pona
+The syntax file has an association of TP
 words with highlighting groups.
-It also holds association of these groups to other groups
+It also holds associations of these groups to other groups
 to use their color settings.
 Therefore, one is able to use various color schemes
-with syntax-highlighted Toki Pona texts.
+in syntax-highlighted TP texts, such as in Figure~\ref{fig:syn}.
 A Vim user might run e.g. \tttt{:colorscheme blue},
@@
 Useful commands are given as commented lines
-in the end of the syntax file.
-One might obtain an HTML file with the colored Toki Pona text
+in the end of the syntax file. % TTM
+One might obtain an HTML file with the colored TP text
 by using the \tttt{:TOhtml} command.
+
+% reference to figure of the texts.
 
@@
 \begin{itemize}
-  \item The use of blue and other high frequency colors for
-  background.
-  This is extended to other high frequency (small wavelength)
-  colors such as violet.
+  \item The use of blue and other high frequency colors (such as
+    violet) for the background.
   \item The avoidance of blue and other high frequency
@@
   the same color.
-  One might use a power-law to mimic natural occurring phenomena.
-
-  Unfortunately, the dictionary classification of words are not in a
+    One might use only a small number of colors to achieve a clean
+    visualization of texts.
+  Also, a power-law distribution of tokens among colors might be
+    well-suited to mimic natural occurring phenomena.
+    (Unfortunately, the TP dictionary classification of words are not in a
     power-law distribution and might be better described by a half-normal or by
-overlapping normal distributions over the three peaks ($\sim35-58,
+overlapping normal distributions over the three peaks: $\sim35-58,
 12-18, 1-5 occurrences$).
@@
     as Steven's law and Weber-Fechner's law).
+    This might be useful e.g. for coloring sets of tokens in
+    a way that their similarity is considered.
   \item One usually wishes to maximize contrast,
@@
   above.
-  Other parameters are hue, transparency etc.
+  Other evident parameters are hue, transparency etc.
   \item One might consider two types of color schemes: those with a dark background,
@@
     and programmers often report that a blue background keeps them
-    awake and more concentrated.
-    I found no evidence of this in the scientific literature,
-    and one might simply take my word that programmers say so
-    since I now have programmed daily for more than 10 years.
-  \item Consider that the user might stay many hours (and we often do)
-  at a text editor, and that the colors and formats involved are
+    awake and more concentrated (I also notice such effect,
+    and I have programmed daily for more than 10 years).
+  \item Colors have been associated to enhancements in specific tasks,
+    e.g. blue and red are respectively associated to enhanced
+    creativity and detail-oriented tasks in~\cite{blue}.
+  \item It is important to consider that the user might 
+    stay many hours
+  at a text editor (and we very often do),
+    and that the colors and formats involved are
   prone to entail a considerable effect in the body and mental
-  activity and thus in the quality of life and work of a programmer.
-  \item Syntax sonification: consider mapping the textual structures
-  to sound (i.e. parametrize the synthesis of sounds e.g. 
-    by the counting of specific tokens inside a script, a function or
-    class or around a variable of conditional or
-    loop)~\cite{mass}.
+    activity and thus in the quality of life and work of a writer
+    (e.g. a programmer).
   \item Ideally, one should have facilities for tuning 
     the syntax highlighting (e.g. through keyboard shortcuts)
-  as envisioned in~\cite{vimArt}.
+  as envisioned in~\cite{vim}.
 \end{itemize}
 
-On the consideration of writing and of Toki Pona,
-the considerable irrelevance of the morphsyntactic classes as
-described in Section~\ref{} suggest that a appropriate coloring should
-consider the syntax to identify the nouns, verbs, adjectives, adverbs,
+The considerable irrelevance of the morphosyntactic classes as
+described in Section~\ref{syntax} suggests that an appropriate coloring
+of words should
+either distinguish only between particles, prepositions and the rest,
+or consider the syntax to identify the nouns, verbs, adjectives, adverbs,
 prepositions and particles.
 Also, the coloring for poems might be more appealing if considered
-e.g. the counting of syllables or repeated vowels and ending
-syllables.
+e.g. the counting of syllables, the repeated letters or syllables,
+and the ending syllables.
 Insights might be obtained through the observation of the
-choices made and advocated in software packages such as text editors
-(e.g. Vim, Emacs, Sublime, etc.),
-software dedicated to syntax highlighting (e.g.
-pigments \footnote{\url{http://pygments.org/docs/}} and 
-highlight.js \footnote{\url{https://highlightjs.org/}},
+choices made and advocated in software such as text editors
+(e.g. Vim, Emacs, Sublime),
+packages dedicated to syntax highlighting (e.g.
+pigments\footnote{\url{http://pygments.org/docs/}} and 
+highlight.js\footnote{\url{https://highlightjs.org/}},
+and
 other software (e.g. Linux terminals such as Xterm and gnome-terminal).
@@
 e.g. sans-sheriff fonts are more promptly read and yield a cleaner
-text than a fot with sheriff.
-
-\subsection{Wordnet Toki Pona}\label{wn}
-exception: jan (people)
-is a hyponym of soweli (mammal),
-kili is hyponym of kasi,
-(walo, pimeja, jelo, loje, laso) are hyponyms of kule.
-But otherwise such relations hold:
-e.g. luka (as hand) is e meronym of sijelo (body).
-
-For a first achievement of a Toki Pona Wordnet,
-each of the Toki Pona words in the dictionary
-were related to Wordnet synsets
+text than a serif font, and the same applies to a monospaced font
+which is more likely to yield a cleaner text, at least for programming.
+
+One might also 
+consider the mapping the textual structures
+to sound~\cite{mass} (i.e. parametrize the synthesis of sounds e.g. 
+by the counting of specific tokens inside a script, a function or
+class or around a variable of conditional or
+loop), which might be called ``syntax sonification''.
+
+\subsection{Toki Pona Wordnet}\label{wn}
+For the achievement of a first Toki Pona Wordnet,
+each of the TP words in the dictionary
+was related to Wordnet synsets~\cite{wordnet}
 through the English lemmas.
-All synsets related to the lemma with the POS tag
-given in the dictionary are added.
-Another immediate possibility would be to
-relate each Toki Pona word to the synsets
-related to all the words bounded by a semicolon.
-Another possibility, which I find more consistent
-but also more controversial,
-is to choose all synsets related to the Wnglish
-words given that the Toki Pona words can be used
-in many roles.
-WordNet only contains nouns, adjectives, verbs, and adverbs.
+The most canonical (i.e. Princeton) Wordnet only contains nouns,
+adjectives, verbs, and adverbs.
 Thus, particles and prepositions were not considered.
@@
 Words presented as adjectives in the dictionary
-were considered adjectives and adverbs.
-
-The total number of different synsets is 2493 in a very conservative
-counting: discarding prepositions and particles,
-only considering the synsets which have the POS tag
-given in the dictionary.
-If all the synsets (that have any POS tag)
-related to a lemma are considered, 
-then Toki Pona vocabulary reaches 3961 synsets.
-The prepositions are particularly missing in these
-countings because e.g. tawa or lon are used also
-as ordinary words.
-If synsets for them are considered those of the English
-lemmas with any POS tag, the vocabulary reaches 4052
-synsets.
-The total number of synsets in Wordnet is X.
-Total number of synsets counting neighbors.
-Depth of TP synsets.
-Using such features to devise other conlangs.
-
-\subsection{Hacks from other people}
-The \url{tokipona.net} has a number of tools,
-just as to transliterate names into Toki Pona phonemes,
-and search in corpus.
+were considered both as adjectives and adverbs.~\cite{wordnet}
+
+The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
+makes available such tentative TP Wordnets in the simplest form:
+the TP words are keys in a dictionary that returns the
+corresponding synsets.
+Three of such dictionaries are implemented:
+one where all synsets related to the lemma is
+returned (the version most consistent
+with the expositions in Section~\ref{basics}),
+with 4,052 synsets;
+another as such but excluding the prepositions,
+with 3,961 synsets;
+a last one that returns only the synsets
+registered in Wordnet with the same POS tag as the lemma
+is in the official dictionary,
+with 2,493 synsets. 
+Another immediate possibility, not implemented,
+would be to relate each Toki Pona word to the synsets
+simultaneously associated to all the English
+words bounded by a semicolon in the dictionary.
+Such collection of synsets, and its relation to the whole
+Wordnet
+(the total number of synsets in Wordnet is $117,659$),
+might guide creation of other conlangs
+(e.g. one might seek to use lemmas related to synsets that
+are very far apart, or that has the most complete neighborhood
+possible).
+% Total number of synsets counting neighbors is Y. TTM
+% Depth of TP synsets is .
+
+It is worth mentioning that TP words are very general
+and each of them might mean many things.
+Thus, the words are not very easily related e.g. by hyponymy
+or meronymy.
+Exceptions: jan (people) is a hyponym of soweli (mammal);
+kili is an hyponym of kasi;
+walo, pimeja, jelo, loje, laso
+are hyponyms of kule.
+There is at least one caveat:
+if a particular meaning of each word is chosen,
+then there might be many other of such relations,
+e.g. 
+luka (as hand) is a meronym of sijelo (as body).
 
 \section{Conclusions and further work}\label{conc}
+This document presented a potentially novel description of Toki Pona (TP)
+language in Section~\ref{basics}, and innovative and useful software routines for dealing with
+the language in Section~\ref{hacks}.
+As a minimalist conlang, TP is very convenient for cognitive
+experiments
+through linguistic semantics and
+for devising tools for analysis, creation and visualization.
+Other conlangs, and even non-planned languages might benefit from
+TP and the content presented here in many ways, e.g. one might
+synthesize TP texts and translate them as convenient,
+adapt the routines to obtain statistics of the vocabulary,
+or take advantage of the language description to devise new conlangs
+or stilistic guidelines for any language.
+
+Possible next steps are numerous:
 \begin{itemize}
@@
     Section~\ref{synth} might be enhanced by further employing e.g.
-    rhythm, meter, rhyme and form techniques~\cite{wikipPoetry}.
+    rhythm, meter, rhyme and form techniques~\cite{wikiPoetry}.
   \item Make a syntax highlighting that colors the tokens with respect
@@
     as is implemented and exposed in Section~\ref{shigh}.
-  \item Relate Toki Pona to Wordnet: should one Toki Pona word
-    be related to more then one synset of the English language?
-    An immediate implementation would be linking each Toki Pona word
-    to the English words in the official vocabulary.
-    A Toki Pona synset would be related to all the synsets of each English
-    words regarded as lemmas (sequence of characters or sound).
-  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
-    of words and text.
-    Enhancements of such translators would use POS tagging and templates
-    in English and Toki Pona and/or corpus and further statistical methods.
-    An efficiency of Toki Pona can be studied in terms of the coverage of
-    the Wordnet cloud.
-  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
-    (conlanging), and insights derived from Toki Pona might be used for that end.
-  \item Understand how the corpus is gathered in \url{tokipona.net}.
-  \item Know about previously existing words that were used for Toki Pona
+    This might be implemented by using n-grams and
+    further techniques from Natural Language Processing
+  \item Enhance and study the TP Wordnet described in Section~\ref{wn}:
+    implement a NLTK-like Wordnet interface; check if each synset
+    is correctly associated to the TP words; seek further synsets
+    that were not found by the English lemmas in the official
+    dictionary; implement the most restricted version described in
+    Section~\ref{wn}, which relates each TP word only to the synsets
+    that are related to all the English words not separated by a
+    semicolon; find out the neighborhood of the synsets of the TP
+    Wordnet in the English Wordnet; explore the possibilities for
+    devising new conlangs by means of criteria related to Wordnet.
+%   \item Translators.
+%    This Toki Pona Wordnet would be useful e.g. for naive translators
+%    of words and text.
+%    Enhancements of such translators would use POS tagging and templates
+%    in English and Toki Pona and/or corpus and further statistical methods.
+%    An efficiency of Toki Pona can be studied in terms of the coverage of
+%    the Wordnet cloud.
+  \item Conlanging: use Wordnet for making new conlangs;
+    use insights derived from TP.
+    Maybe create conlangs for specific uses:
+    describing data, programs,
+    scientific writings, creative writing
+    (exploring the thinking process).
+    A conlang might have different modes that might 
+    be specified by a section header or tag. 
+  \item Understand how the corpus was gathered in \url{tokipona.net}.
+  \item Know about previously existing words that were used for TP
     (e.g. suno and suwi might come from sun and sweet),
-    and about the reasons that lead Sonja (and maybe other people)
+    and about the reasons that led Sonja (and maybe other people)
     to choose the 14 letters and the syllable structure.
     This might require a dedicated communication with the
-    speaker community and the documentation authors.
+    speaker community and the documentation keepers.
   \item Corpus-based analysis.
-  \item Publication of original texts and translations.
-  \item Make an article written in Toki Pona.
-    I wrote Section~\ref{finalToki}, and believe that
-    a summary in both English and Toki Pona for
+  \item Publication of original TP texts and translations.
+  \item Make an article written in TP
+    (extending Section~\ref{finalToki} which might be the
+    first section of a scientific document written in TP).
+    The outline might be:
+    a summary in both English and Toki Pona, for
     facilitating the acquisition of context,
-    a reasonable article on some scientific topics is
-    possible.
+    and a article in a canonical structure
+    (introduction and related work, materials and methods,
+    results and discussion, conclusions and further work),
+    with an Appendix explaining the content and context in English.
     I first conceived something around complexity, statistics, physics, or computer science.
@@
     I can start a draft,
-    qhey might learn the language in a few hours (with or without my help),
+    they might learn the language in a few hours or weeks
+    (with or without my help),
     to contribute, and we can write a short paper.
   \item Enhance the synthesis of text to yield better contextualized text
-    and stylistic traces, such as for poems and short stories.
-    Give the user the ability to choose the sentences (generate randomly according to
+    and stylistic traces.
+    Give the user the ability to choose the sentences:
+    generate a sentence randomly according to
     previously written or given text,
-    some rules input by the user, the package and the language guidelines,
-    outputs to the screen and asks to keep and discard).
-  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
-    It uses only syntactic cues to choose POS tags,
-    which might be overcome by using n-grams and
-    further techniques from Natural Language Processing~\cite{POS}.
+    to some rules input by the user,
+    to the package and the language guidelines.
+    Should output to the screen and asks to keep and discard.
   \item We are particularly interested in software tools,
@@
     and in what sense one is able to compile and parse Lojban.
-    This study will probably yield considerations about
-    parsing Toki Pona files.
-  \item Conlang idealized for describing data, programs,
-    scientific writings, creative writing
-    (exploring the thinking process).
-    Different modes that might be specified by a section header or tag.
-  \item The plugin comes with the :TokiStation command, which opens a window with the files: ijositelen.tokipona, makevimSyntax.py, Highlight Test (created by hitest.vim above), syntax/tokipona.vim.
-Another tab with shells: python to run over the makeVimSyntax.py and make new syntax/tokipona.vim files.
-Another with Readme, and PDF documentation.
-Another with an IPython.
-Imported tokipona.
-It resets completely upon command :TokiClose, closing all created windows.
+    Such study will probably entail considerations about
+    parsing TP.
+%  \item The plugin comes with the :TokiStation command, which opens a
+    %  window with the files: ijositelen.tokipona, makevimSyntax.py,
+    %  Highlight Test (created by hitest.vim above),
+    %  syntax/tokipona.vim.  Another tab with shells: python to run
+    %  over the makeVimSyntax.py and make new syntax/tokipona.vim
+    %  files. Another with Readme, and PDF documentation. Another with
+    %  an IPython. Imported tokipona. It resets completely upon
+    %  command :TokiClose, closing all created windows. TTM
 \end{itemize}
@@
 FAPESP (project 2017/05838-3);
-Vim, Python, and Toki Pona authors, developers and literature maintainers;
-user and speaker communities.
+Toki Pona, Python and Vim authors, developers and literature maintainers;
+Toki Pona user and speaker communities. 
+I thank Mario Alzate for introducing me to Toki Pona,
+and Silverio Guazzelli Donatti for the stimulus that resulted
+from the usage of Toki Pona for chatting.
 
@@
 \section{My usage of Toki Pona}\label{mytoki}
-
 I use the standard sounds, but often use [z] for s.
@@
 
-\section{Final words in Toki Pona}
+\section{Final words in Toki Pona}\label{ftp}
 toki li nasin e lawa.
@@
 \fontsize{11}{0}\selectfont
-\bibitem{anPh}
-	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
-	Available at \url{https://youtu.be/oeOKYc3-nbM}
-\bibitem{anPh2}
-	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
-\bibitem{ouroWiki}
-  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
-\bibitem{vimrc}
-	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
-\bibitem{ex}
-  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
+\bibitem{wordnet}
+  Fellbaum, C. (1998). WordNet. John Wiley \& Sons, Inc..
+\bibitem{corpus}
+    Henry, J. lipu pi toki pona pi jan Jakopo. Available at         
+    \url{http://jimhenry.conlang.org/conlang/tokipona/tokipona.htm} 
+\bibitem{vim}
+    Fabbri, R. An anthropological account of the Vim editor:
+features and tweaks after 10 years of usage. Available at         
+    \url{https://github.com/ttm/vim/raw/master/article/article.pdf} 
+\bibitem{sapWho}
+  O'Neill, S. P. (2015). Sapir–Whorf Hypothesis. The International
+    Encyclopedia of Language and Social Interaction.
+\bibitem{gdoc}
+  Toki Pona Community (2017). toki pona central hub preparation.
+    Available at
+    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
 \bibitem{tokipona}
@@
     ISBN-10: 0978292308, ISBN-13: 978-0978292300.
-\bibitem{aa1}
-	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G.
-(2013).
-AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology.
-RESI.
-From \url{https://arxiv.org/abs/1604.08255}
-\bibitem{aa2}
-	Fabbri, R. (2017).
-The Algorithmic-Autoregulation (AA) Methodology and Software:
-a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
+\bibitem{tp3}
+  Blahuš, M. (2011). Toki pona–eine minimalistische Plansprache.
+    Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
 \bibitem{kama}
@@
     Software Engineering in Stack Exchange. Available at
-    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
+    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
 \bibitem{solarized}
@@
   \url{http://ethanschoonover.com/solarized}
+\bibitem{tp4}
+  Schneider, S. (2017). nasin toki pi toki pona. Available at
+  \url{https://github.com/stefichjo/toki-pona}
+\bibitem{interview}
+  Lang, S. and Broholm, K. AFP 20 – Sonja Lang: Toki Pona, Conlanging
+    and the meaning of life (Sonja Lang interview). Available at
+    \url{https://actualfluency.com/afp-20-sonja-lang-toki-pona-conlanging-meaning-life/}
+\bibitem{memrise}
+  Toki Pona Community. (2017). Various Memrise courses on Toki Pona.
+    \url{https://www.memrise.com/courses/english/?q=toki+pona}
+\bibitem{wikiToki}
+  Toki Pona. (2017, December 11). In Wikipedia, The Free Encyclopedia.
+    Retrieved 22:48, December 14, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
+\bibitem{artlanWikip}
+  Artificial language. (2017, November 26). In Wikipedia, The Free
+  Encyclopedia. Retrieved 23:00, December 14, 2017, from
+  \url{https://en.wikipedia.org/w/index.php?title=Artificial_language&oldid=812127746}
+\bibitem{akesiWawa}
+  Fabbri, R. (2017, November 6). o awen, akesi wawa.
+  In the group toki pona taso. Available at
+  \url{https://www.facebook.com/groups/tokiponataso/permalink/1895942817391318/}
+\bibitem{janKipo}
+  Clifford, J. (2017, December 15). Considerations about Toki Pona
+  phrases.
+  In the Facebook group 'toki pona'. Available at
+  \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
+\bibitem{tpNetSH}
+  Knight, B. (2017). Online gadget for syntax highlighting Toki Pona
+  texts. . Available at
+  \url{http://tokipona.net/tp/DisplayText.aspx}
+\bibitem{blue}
+  Mehta, R., \& Zhu, R. J. (2009). Blue or red? Exploring the effect of
+  color on cognitive task performances. Science, 323(5918), 1226-1229.
+\bibitem{blue2}
+  Viola, A. U., James, L. M., Schlangen, L. J., \& Dijk, D. J. (2008).
+  Blue-enriched white light in the workplace improves self-reported
+  alertness, performance and sleep quality. Scandinavian journal of
+  work, environment \& health, 297-306.
+\bibitem{fonts}
+  Spolsky, A. J. (2008). User interface design for programmers.
+  Apress.
+\bibitem{mass}
+	Fabbri, R., Vieira, V.
+, Pessotti, A.
+C.
+S., Corrêa, D.
+C., Oliveira Jr, O.
+N.
+O.
+Musical elements in the discrete-time representation of sound (2017).
+arXiv preprint arXiv:1412.6853.
+Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
 \end{thebibliography}

### 2017-12-15 article/pos.tex
terminada revisao

@@
 \begin{center}
-\caption{"POS tags incident and chosen.
+\caption{POS tags incident and chosen.
             The official dictionary often relates tokens
             to more than one POS tag.
-            For the text highlighting Plugin, for example,
+            For the current version of the text highlighting plugin described in
+            Section~\ref{shigh}, for example,
             a token has to have an established tag to have
             a defined color.
-            On the Chosen column, the tokens were regarded only once
-            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}
+            On the 'Chosen' column, the tokens were regarded only once
+            by choosing the class in the dictionary respecting the
+            precedence order: PRE, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
 \begin{tabular}{| l | c | c |}\hline
@@
 NOUN & 58  & 49 \\\hline
-ADJECTIVE & 40  & 37 \\\hline
-VERB & 15  & 15 \\\hline
+ADJECTIVE & 40  & 34 \\\hline
+VERB & 15  & 13 \\\hline
 PARTICLE & 12  & 12 \\\hline
+PRE & 6  & 6 \\\hline
 PREPOSITION & 5  & 5 \\\hline
 NUMBER & 4  & 1 \\\hline
-PRE & 0  & 0 \\\hline
-total & 134  & 120 \\\hline
+total & 140  & 120 \\\hline
 \end{tabular}\end{center}

### 2017-12-15 article/syls.tex
terminada revisao

\begin{table*}[h!]
\begin{center}
\caption{Frequency of syllables in Toki Pona
            considering all 235 syllables of the 124 tokens,
            only the first or last syllables or only the middle
            syllable.
            In parenthesis are the count and percentage of the
            corresponding syllable. For more information and a
            complete list of syllables, see Section~\ref{sec:stat}.
            }\label{tab:freqsy}
\begin{tabular}{  l | c   c   c   c  }
rank & all  & first  & last  & middle \\\hline
1 & li (13, 5.53\%)  & a (8, 6.45\%)  & li (10, 8.06\%)  & la (2, 15.38\%) \\
2 & la (10, 4.26\%)  & o (5, 4.03\%)  & lo (6, 4.84\%)  & je (1, 7.69\%) \\
3 & ka (9, 3.83\%)  & pi (5, 4.03\%)  & na (6, 4.84\%)  & ka (1, 7.69\%) \\
4 & na (9, 3.83\%)  & ka (4, 3.23\%)  & la (5, 4.03\%)  & ke (1, 7.69\%) \\
5 & pa (9, 3.83\%)  & la (4, 3.23\%)  & ma (5, 4.03\%)  & li (1, 7.69\%) \\
6 & a (8, 3.40\%)  & pa (4, 3.23\%)  & pa (5, 4.03\%)  & lu (1, 7.69\%) \\
7 & ma (8, 3.40\%)  & se (4, 3.23\%)  & ka (4, 3.23\%)  & ma (1, 7.69\%) \\
8 & si (8, 3.40\%)  & si (4, 3.23\%)  & sa (4, 3.23\%)  & me (1, 7.69\%) \\
9 & lo (7, 2.98\%)  & su (4, 3.23\%)  & si (4, 3.23\%)  & pe (1, 7.69\%) \\
10 & pi (6, 2.55\%)  & i (3, 2.42\%)  & te (4, 3.23\%)  & ta (1, 7.69\%) \\
\end{tabular}\end{center}
\end{table*}

### 2017-12-15 article/vowels.tex
terminada revisao

@@
-\begin{table*}[h!]
+\begin{table*}[h]
+\scriptsize
 \begin{center}
-\caption{"Frequency of vowels in Toki Pona.}
-\begin{tabular}{| l | c |}\hline
-Vowel & freq \\\hline
-a & 33.19 \\\hline
-e & 17.45 \\\hline
-i & 23.40 \\\hline
-o & 15.32 \\\hline
-u & 10.64 \\\hline
+\caption{Frequency of letters in Toki Pona.
+            freq, freq\_I, freq\_L and freq\_M are
+            the frequencies of the letters in any, initial, last and middle
+            positions.
+            The columns 'v' and 'c' that follow them are frequencies
+            considering only vowels and consonants.
+            The most frequent vowel is 'a' in any position,
+            although it is more salient among words starting with a vowel
+            and among the last letter of the words.
+            For starting, ending and middle positions, the second most frequent
+            vowel varies.
+            Among the consonants, 'n' is the most frequent because it is
+            the only consonant allowed in the last position and because
+            almost 20\% of the words end with 'n'.
+            On the initial position, 's' is the most frequent consonant,
+            while in middle position 'l' is the most frequent consonant.
+            Many other conclusions can be drawn from this table and are
+            useful e.g. for exploring sonorities in poems.}\label{freqLet}
+\hspace*{-1.3cm}
+\begin{tabular}{  l | c   c   c | c   c   c | c   c   c | c   c   c  }
+letter & freq  & v  & c  & freq\_I  & v  & c  & freq\_L  & v  & c  & freq\_M  & v  & c \\\hline
+a & 16.35  & 33.19  & -  & 8.06  & 40.00  & -  & 29.03  & 35.64  & -  & 14.22  & 29.46  & - \\
+e & 8.60  & 17.45  & -  & 2.42  & 12.00  & -  & 11.29  & 13.86  & -  & 10.78  & 22.32  & - \\
+i & 11.53  & 23.40  & -  & 3.23  & 16.00  & -  & 20.97  & 25.74  & -  & 10.78  & 22.32  & - \\
+o & 7.55  & 15.32  & -  & 4.03  & 20.00  & -  & 14.52  & 17.82  & -  & 6.03  & 12.50  & - \\
+u & 5.24  & 10.64  & -  & 2.42  & 12.00  & -  & 5.65  & 6.93  & -  & 6.47  & 13.39  & - \\\hline
+j & 2.10  & -  & 4.13  & 3.23  & -  & 4.04  & 0.00  & -  & 0.00  & 2.59  & -  & 5.00 \\
+k & 6.29  & -  & 12.40  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 6.90  & -  & 13.33 \\
+l & 9.22  & -  & 18.18  & 12.10  & -  & 15.15  & 0.00  & -  & 0.00  & 12.50  & -  & 24.17 \\\hline
+m & 4.61  & -  & 9.09  & 10.48  & -  & 13.13  & 0.00  & -  & 0.00  & 3.88  & -  & 7.50 \\
+n & 10.48  & -  & 20.66  & 6.45  & -  & 8.08  & 18.55  & -  & 100.00  & 8.19  & -  & 15.83 \\
+p & 5.66  & -  & 11.16  & 11.29  & -  & 14.14  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\\hline
+s & 6.29  & -  & 12.40  & 13.71  & -  & 17.17  & 0.00  & -  & 0.00  & 5.60  & -  & 10.83 \\
+t & 3.14  & -  & 6.20  & 6.45  & -  & 8.08  & 0.00  & -  & 0.00  & 3.02  & -  & 5.83 \\
+w & 2.94  & -  & 5.79  & 4.84  & -  & 6.06  & 0.00  & -  & 0.00  & 3.45  & -  & 6.67 \\
 \end{tabular}\end{center}

### 2017-12-15 article/article.tex
terminada revisao dos apendices

@@
     Should output to the screen and asks to keep and discard.
+  \item Make TP variations where each TP word is replaced by a word in
+    a language with a large speaker population. 
+For example,
+suno might be sun in an English variant,
+and sol in a Portuguese variant.
+A naive version might be obtained through
+choosing the first word in the description of
+an official vocabulary, as they already exist in English and French.
   \item We are particularly interested in software tools,
@@
 and create new texts as poems and short stories.
-Most of them are in~\cite{tokisona}.
+Most of them are in~\cite{tokisona}
+and/or published in the TP Facebook groups.
 I omit the li particle after subjects sina and mi,
@@
 In such cases, the first li is sometimes omitted.
-Also, sometimes I use li before mi and sina where I find
+Also, sometimes I use li after mi and sina where I find
 that there is unwanted ambiguity, e.g.
-sina moku pona e jan 
-(might be sina li moku pona e jan or sina moku li pona e jan).
+sina moku pona 
+which 
+might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.
 
@@
 I also tend to legitimate the use of English (or German) words
-in Toki Pona texts if it is the case,
+in TP texts if it is the case,
 as happens often in scientific writing
 (kernel is a German word used in English,
-webpage is an English word used in Portuguese).
+webpage and software are English words used in Portuguese).
 
 Proposed notations for numbers seem numerous.
-I tend to think that one might indicate is two numbers
-are multiplying (pi) or are in different scales
+I tend to think that one might indicate if two numbers
+are being multiplied (pi) or are in different scales
 (such as in decimal or binary notations).
-For example, I take luka two to mean 52.
-Or it might not be taken to such level of strictidness,
-for 52 is a reasonable notation for a simple language.
-mi jo e jan sama nanpa 12.
-Or even:
-ona li lon e soweli 27.
-
-Remove the 'li'.
+For example, luka (pi) two might mean 52.
+Most importantly,
+52 is a reasonable notation for a simple language.
+E.g.
+'mi jo e jan sama nanpa 12',
+or
+'ona li lon e soweli 27'.
+
+Avoiding needless words...
 I've been avoiding e ni: and using only :.
-'ve been omitting the subject if it is the
+I've been omitting the subject (sometimes also the li)
+if it is the
 same as in the last sentence.
-I've been avoiding also the li sometimes,
-and starting only a predicate + object phrase,
-or a whole phrase altogether.
-
-pali moku is addiction.
-noka and ampa; luka, suno, sike and lawa;
-I tend to group these things, as pali and pana.
+I tend to group these words:
+noka and anpa; luka, suno, sike and lawa; pali and pana.
 These clusters of meaning hint me that an even simpler
 language is possible and maybe more optimized to the
-core goals of Toki Pona.
+core goals of TP.
+
 Also, ambiguities introduced by omission of li
 and absence of a token to denote preposition,
-suggest to me the possibility of a syntax that is uniquely
-parsed.
-
-I am tending to think on a language with the same syntax of toki pona:
-<sentence> la <sentence = subject + predicate + object + preposition complement>
-but always using the syntax separators,
-repeating them when the function in performed in the smallest scale:
-'(toki pi toki pona) pi jan sona' (language in toki pona and of intellectuals) vs
-'toki pi (toki pona pi pi jan sona)' (language in toki pona of intelectuals).
-The keyword for preposition complement is the catch.
-An 'a' seems good, but it is taken in Toki Pona.
-Maybe use it as the preposition particle and use ha or he
-instead of Toki Pona a.
+suggest to me the possibility of a syntax that is always uniquely
+parsed (or at least less ambiguous).
+
+I have been inclined to conceive a language with the same syntax of TP:
+<sentence> la <sentence = subject + predicate + object (each term with
+a possible preposition complement)>
+but always using particles to bound the sentence sections,
+repeating them when the function is performed in the smallest scale:
+'toki pi toki pona pi jan sona' meaning
+language in toki pona and of intellectuals vs
+'toki pi toki pona pi pi jan sona'
+meaning language in [toki pona of intelectuals].
+About the keyword for initiating a preposition complement 
+might be 'a', but it is already taken in TP.
+Maybe use 'a' before a preposition and use 'ha' or 'he'
+instead of TP a.
 In such a setting,
 one might enable pi li e a (concept-qualifier, subject-predicate,
-predicate-object, concept-predicate)
-in any of the terms of the template:
-<sentence> la <sentence = subject + predicate + object + preposition complement>
+predicate-object, concept-preposition)
+in any slot of the template:
+<sentence> la <sentence = subject + predicate + object (each term with
+a possible preposition complement)>
 Between sentences one would use:
-la pi, la li, la e, la a, and la la as normally.
+la pi, la li, la e, la a, and la la.
 This is a very fractal conlang proposal.
-Maybe have a way to distinct between noun, qualifiers (adjectives/adverbs) and verb terms,
-and accept any of them for subject, predicate, object, predicate,
-slots. Or assume a part of speech for template slot or for first word in slot.
-
-Because toki pona's vocabulary is so small,
-one might make variations of it for languages
-in specific.
-For example,
-suno might be sun in an English variant,
-and sol in a Portuguese variant.
-In Appendix~\ref{enPo} are an English and Portuguese
-proposition of such variants.
-A naive version might be obtained through
-choosing the first word in the description of
-an official vocabulary, as they exist in English and French.
-
-mute, mute ike,
-something can be good, numbered,
-or numbered in a bad way, too much,
-as in 'o awen, sina pali mute, lape lili'.
-
-% \input{pos}
+Maybe also have a way to discern between a noun,
+a qualifier (adjectives/adverbs) and a verb,
+and accept any of them for the subject, predicate, object, preposition,
+slots. Or assume a part of speech as default for a slot or for the
+first word in a slot.
 
@@
 toki pona li pona e nasin tan ni:
-ona li jo e sitelen mute lili.
+ona li jo e nimi nanpa lili.
 ona li pona.
@@
 o taso la toki pona li kalama li lukin pona.
-sitelen en nasin li open e sitelen suli\cite{tpLang,pije,fb1,fb2,tokisona,Wikipesija}.
+sitelen en nasin li open e sitelen
+suli\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
 li sona. li nasin e toki e sona e lawa e lon.
@@
 
-ilo lon linha~\ref{hacks} li pana e sitelen
+ilo lon Poki~\ref{hacks} li pana e sitelen
 e sona tan sitelen,
 en nasin.
-linha~\ref{basics} li pana e nasin pi toki pona.
+Poki~\ref{basics} li pana e nasin pi toki pona.
 e sitelen tawa kama sona.
@@
 sitelen musi.
-ante e toki pona la
+ante e toki pona
 la ona li nasa.
@@
 o pona tawa jan pi toki pona.
-jan Sonja, Birns-Sprage, Kipo, Pije,
-Siwejo, Malija, jan kulupu mute.
+tawa jan Sonja, Birns-Sprage, Kipo, Pije,
+Siwejo, Malija, Tepan, jan kulupu mute.
 
@@
 Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
+\bibitem{tokisona}
+  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
+  Available at \url{http://tokisona.github.io/}
 \end{thebibliography}

### 2017-12-15 article/article.tex
finalizada primeira versao revisada do artigo para mandar p arxiv

@@
 and best practices in writing TP texts or for translation.
-For these and other matters, visit~\cite{tpLang,kama,gdoc,tp4}.
+For these and other matters, visit~\cite{tpLang,kama,tp4,gdoc}.
 
@@
 and have scarce scientific studies~\cite{stack}.
-There are informed projects such as Solarized~\cite{solazired},
+There are informed projects such as Solarized~\cite{solarized},
 which present solutions for some contexts.
@@
   \item Make an article written in TP
-    (extending Section~\ref{finalToki} which might be the
+    (extending Section~\ref{ftp} which might be the
     first section of a scientific document written in TP).
@@
     Encyclopedia of Language and Social Interaction.
-\bibitem{gdoc}
-  Toki Pona Community (2017). toki pona central hub preparation.
-    Available at
-    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
-\bibitem{tokipona}
-	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
 \bibitem{tpLang}
@@
     ISBN-10: 0978292308, ISBN-13: 978-0978292300.
+\bibitem{kama}
+  Knight, B. (2017) o kama sona e toki pona! Available at:
+    \url{http://tokipona.net/tp/janpije/okamasona.php}
 \bibitem{tp3}
@@
     Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
-\bibitem{kama}
-  Knight, B. (2017) o kama sona e toki pona! Available at:
-    \url{http://tokipona.net/tp/janpije/okamasona.php}
-\bibitem{Wikipesija}
-  Toki Pona community (2017). Wikipesija. Available at:
-    \url{http://tokipona.wikia.com}
-\bibitem{conlanWikip}
-  Constructed language.
-(2017, November 1).
-In Wikipedia, The Free Encyclopedia.
-Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
-\bibitem{wikiPoetry}
-Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
-    Retrieved 19:46, December 3, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
-\bibitem{stack}
-  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
-    Software Engineering in Stack Exchange. Available at
-    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
-\bibitem{solarized}
-  Schoonover, E. Solarized color scheme. Available at
-  \url{http://ethanschoonover.com/solarized}
 \bibitem{tp4}
@@
     \url{https://www.memrise.com/courses/english/?q=toki+pona}
+\bibitem{gdoc}
+  Toki Pona Community (2017). toki pona central hub preparation.
+    Available at
+    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
 \bibitem{wikiToki}
@@
     \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
+\bibitem{conlanWikip}
+  Constructed language.
+(2017, November 1).
+In Wikipedia, The Free Encyclopedia.
+Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
 \bibitem{artlanWikip}
@@
   \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
+\bibitem{tokipona}
+	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
 \bibitem{tpNetSH}
@@
   \url{http://tokipona.net/tp/DisplayText.aspx}
+\bibitem{stack}
+  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
+    Software Engineering in Stack Exchange. Available at
+    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
+\bibitem{solarized}
+  Schoonover, E. Solarized color scheme. Available at
+  \url{http://ethanschoonover.com/solarized}
+\bibitem{dv1}
+	Munzner, T. (2014). Visualization analysis and design. CRC press.
+\bibitem{dv2}
+  Telea, A. C. (2014). Data visualization: principles and practice.
+  CRC Press.
+\bibitem{dv3}
+  Ward, M. O., Grinstein, G., \& Keim, D. (2010). Interactive data
+  visualization: foundations, techniques, and applications. CRC Press.
+\bibitem{dv4}
+  Ware, C. (2012). Information visualization: perception for design.
+  Elsevier.
 \bibitem{blue}
@@
 Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
+\bibitem{wikiPoetry}
+Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
+    Retrieved 19:46, December 3, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
 \bibitem{tokisona}
@@
   Available at \url{http://tokisona.github.io/}
+\bibitem{Wikipesija}
+  Toki Pona community (2017). Wikipesija. Available at:
+    \url{http://tokipona.wikia.com}
 \end{thebibliography}

### 2017-12-15 article/article.tex
general update

--- 2017-12-15 article/article.tex
@@
 \newcommand{\tttt}[1]{\texttt{#1}}
-% \usepackage{fontspec}
-% \usepackage[mathletters]{ucs}
-% \newcommand{\ttt}[1] {
-% 	\texttt{<#1>}}
-% \newcommand{\tttt}[1]{\texttt{#1}}
-\usepackage{subcaption}
 
 \begin{document}
-\title{Basic concepts and tools for the Toki Pona minimalist language:
-Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
+\title{An anthropological account of the Vim editor:\\
+features and tweaks after 10 years of usage}
 % Anthropological computation
@@
 \begin{abstract}
-  A minimalist constructed language (conlang)
-  is useful for experiments and comfortable for making tools.
-  The Toki Pona (TP) conlang is minimalist both in the vocabulary
-  (with only 14 letters and 124 words)
-  and in the $\approx10$ syntax rules.
-  The language is useful for being a used and somewhat established
-  minimalist conlang with at least hundreds of fluent speakers.
-  In this article, we describe current concepts and resources
-  for TP,
-  and make available Python scripted routines for
-  the analysis of the language,
-  the synthesis of texts, the specification of syntax highlighting
-  schemes, and the achievement of a preliminary TP
-  Wordnet~\cite{wordnet}.
-  We focus on the analysis of the basic vocabulary,
-  as corpus analyses were found in~\cite{corpus}.
-  The synthesis is based on sentence templates,
-  relates to context by keeping track of used words,
-  and renders larger texts by using a fixed number of phonemes (e.g. for poems)
-  and number of sentences, words and letters (e.g. for paragraphs).
-  Syntax highlighting 
-  reflects morphosyntactic classes given in the official dictionary
-  and different solutions are described and implemented
-  using 24bit true colors in the well-established Vim text editor~\cite{vim}.
-  The tentative TP Wordnet is made available in three forms that
-  reflect the choices of the synsets related to each word.
-  In summary, this text holds potentially novel conceptualizations about,
-  and tools and results in analyzing, synthesizing or syntax highlighting the
-  TP language.
-  % subj (li), obj (e), prepositions, qualification (word concatenation), 
-  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
+The Vim editor is very rich in capabilities
+and thus complex.
+This article is a description of the Vim text editor
+and a set of considerations on its usage and design.
+It is the result of more than ten years of experience
+with Vim for writing and editing various types of documents,
+e.g.:
+Python, C++, JavaScript, ChucK, programs;
+and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
+binary files.
+It is said that it takes about ten years to master (or start mastering)
+this text editor, and I find that other experienced users
+have a different view of Vim and that they use a different
+set of features.
+Therefore, this document exposes my insights in order
+to confront my usage with that of other Vim users.
+Another goal is
+to make available a reference document with which new users
+can grasp a sound overview by reading it and the discussions that
+it might generate.
+Also, it should be useful for users of any degree of experience,
+including me, as a compendium of commands, namespaces and tweaks.
+Upon feedback, and maturing of my Vim usage,
+this document might be enhanced or receive additional
+material.
 \end{abstract}
-{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona
 
 \section{Introduction}\label{intro}
-Toki Pona (TP) is a minimalist conlang (constructed language)
-with only 124 words (120 without the synonyms).
-Therefore, the concepts are usually very general and different,
-and, without context, the words are rarely related through
-meronymy and hyponymy.
-Such a linguistic setting is desired because of the simplicity
-which entails easier e.g. learning and tool making.
-Another reason why the minimalist language design is compelling
-is the study and harnessing of the strong and weak forms of
-the Sapir-Whorf hypothesis (linguistic relativity),
-i.e. that language influences or dictates thought
-and world experience~\cite{sapWho}.
-Accordingly, one uses a conlang as a thinking tool (or platform)
-or to make experiments about the influence language has on
-the thoughts of the `speaker' (also writer and reader).
-Toki Pona is often described as a tool to meditate,
-to simplify the thinking processes, and as a way to modify
-the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
-In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
-describes that she has seen the language been used successfully
-in the context of management, creation of texts, legal texts,
-etc.
-
-In this article, we present a conceptual overview of the language,
-which is fit both to the newcomer and to the expert
-for being considerably different
-from what we found in the literature~\cite{tpLang,kama,memrise},
-with emphasis on simplicity and flexibility.
-We also present novel software routines for analysis, synthesis,
-syntax highlighting, and the achievement of a tentative TP
-Wordnet~\cite{wordnet}.
-
-Next subsections hold a description of the general resources available,
-a historical note, and some words about natural and constructed languages.
-Section~\ref{basics} describes TP's phonology and syntax.
-Section~\ref{hacks} presents the software routines we made available
-and immediate results, such as listings and statistics of words,
-poems and short stories, coloring schemes, and preliminary TP Wordnet
-versions.
-Conclusions and further work are in Section~\ref{conc}.
-Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
-with thoughts about rule breaking and potentially new conlangs.
-Appendix~\ref{ftp} holds final words in Toki Pona.
-
-% definition, historic, status (groups, speakers, etc).
-
-\subsection{Resources on Toki Pona}
-One might organize current resources for the Toki Pona
-language in: references and learning material, corpus,
-websites, interaction groups
-(where users talk a post texts and comments), and software gadgets.
-The main references of the language are:
-  the official book ``Toki Pona: The Language of Good''~\cite{tpLang},
-    authored by Sonja Lang, the creator of the language;
-and the online book ``o kama sona e toki pona!'',
-from jan Pije~\cite{kama}.
-A number of tools and other resources for dealing with Toki Pona
-were developed by the community.
-The most complete list is supposedly~\cite{gdoc} and includes
-videos, musical pieces, artistic texts, reference documents (e.g.
-a Toki Pona - Esperanto dictionary), journalistic articles,
-and software tools.
-For a more comprehensive view of the resources available
-for the user, we suggest following the links 
-in~\cite{gdoc,wikiToki}.
-
+Vim is a very complex editor,
+considered by the Linux community to be
+matched only by Emacs.
+They both are the standard advanced text editors
+of the free software and open source communities
+and have been developed for decades.
+This document describes the Vim editor
+and proposes a set of enhancements of the user
+experience through simple tweaks.
+The editor has a very mature documentation
+and the contents herein presented is a
+report on the overall understandings I
+have of Vim after a little bit more
+of ten years using it.
+The purposes of this document are:
+\begin{itemize}
+  \item to help new users in grasping Vim essentials
+  and convenient practices.
+  \item To attain a sound overall description of the editor.
+  \item To record the view of the editor that
+  a user (me) has after $10$ years of usage.
+  \item To confront my usage with that of other experienced
+  users. This is helpful for me, but also for the other users
+  as they might benefit from this content and from discussions
+  that might arise from it.
+  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
+\end{itemize}
+
+Advanced users might just skim throughj
+Section~\ref{basics} and consider more carefully
+Section~\ref{namespaces}.
+The concluding remarks and proposed enhancements
+might also deserve some attention.
+
+I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
+I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
+Eventually, I edited database files and other types os files.
+With Vim, I mostly write software (web and scientific),
+music, poems and short stories.
+It is very useful because:
+\begin{itemize}
+  \item it is meant to be a plain text (e.g. ascii, utf8) editor
+  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
+  \item It has a powerful architecture and set of commands.
+  \item It is highly configurable and most often the users
+  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
+\end{itemize}
+
+The standard capabilities of the editor
+should become clear in Section~\ref{basics}.
+Vim is constantly evolving and there are many plugins,
+some of them very popular for both general and specific
+users.
+Accordingly, there are many possible enhancements,
+and Section~\ref{conc} report the most prominent of them
+for me and potential workarounds made available as plugins.
+
+This document is written is a DRY KISS
+(Don't Repeat Yourself, Keep It Simple Stupid) style.
+Complex is to master the use of Vim
+and one finds sound
+references in help files and a nice vimrc.
+Therefore the following content should be kept
+as uncomplicated and original as possible.
+Also, because of Vim's complexity and entailed
+bond of this document to my usage,
+there is an anthropological component
+which is evident in the use of the first person
+in sentences.
+This can be understood as anthropological computation
+or physics~\cite{anPh,anPh2} and observed to help
+in the technological groundwork of the civil society.
+
+% Vi and Vim
 \subsection{Historical note}
-TP was developed as an internal and personal language
-by Sonja Lang~\cite{interview}.
-It was released as a draft in 2001 and in 2007 some documents
-reported it to have a few hundred speakers.
-The English official book~\cite{tpLang} was released only in 2014.
-In 2016, a version of the official book was released in French.
-Nowadays, one finds a number of texts about TP and written
-in it
-(e.g. in social platforms such as Facebook groups,
-microblogging, Telegram and
-IRC), and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.
-
-\subsection{Natural and constructed and artificial languages}
-A language one uses (or might use) to communicate by speaking
-and writing is called `natural language'.
-A `constructed language' (also planned or invented language) is a natural language built by someone or
-a group, such as Esperanto, Toki Pona, and Lojban.
-An artificial language is a language yield by artificial agents,
-such as by AI routines, or by humans in controlled experiments,
-and are considered e.g. within `cultural evolution' studies.
-Formal languages are defined by tokens and rules to
-operate them, they span from computer programming languages
-to math and formal models for natural languages.
-
-Constructed languages are in some traditions called artificial
-languages, but creators most often prefer to use the term
-`standardized', 'constructed'
-or 'planned' language with the argument that the conlang
-is rooted on natural languages,
-and 'artificial' is misleading.
-The preferred terms seem to be planned or constructed language
-or simply conlang.
-The construction of languages is called glossopoeia.
-TP is a conlang which might be classified
-as engineered for experiments, meditation and philosophy;
-and suitable for use as an auxiliary international language and
-as an artistic language.~\cite{conlanWikip,artlanWikip}
-% taxonomy, vocabulary, ontology, terminology,
-
-% conlangs, natural langs, programming langs, Lojban as an
-% intermediary
-% non-verbal languages, musical corporal language, 
-% non-verbal languages confront the strong form of the Sapir-Whorf hypothesis.
-
-\section{Overview of the language}\label{basics}
-This section describes very succinctly the formation of
-words and sentences in the Toki Pona language.
-It should enable a newcomer to grasp the essentials
-of the language and the experienced to
-acquire new insights.
-Furthermore, it is a solid reference of the phonological and syntax
-rules, and enables one to understand and modify the software presented
-in Section~\ref{hacks}.
-
-\subsection{Phonology}\label{phonology}
-Words in Toki Pona are written using only 14 letters:
-\begin{itemize}
-  \item Vowels a (open), e (mid front), o (mid back), i (close front),
-    u (close back).
-  \item Consonants j, k, l, m, n, p, s, t, w:
-    \begin{itemize}
-      \item Nasal: m (labial), n (coronal).
-      \item Plosive: p (labial), t (coronal), k (dorsal).
-      \item Fricative: s (coronal).
-      \item Approximant: w (labial), l (coronal), j (dorsal).
-    \end{itemize}
-\end{itemize}
-
-There are standard guidelines for pronunciation,
-but the language allows for considerable allophonic
-variation.
-For example, /p t k s l/ might be pronounced
-[p t k s l] or [b d g z \textipa{\!R}].
-Especially for poetry, one might consider
-j and w to be vowels
-(e.g. j as 'i' and w as 'u').
-
-Syllables are of the form (C)V(n):
-an optional consonant, a vowel and an optional coronal nasal consonant
-(n).
-Non word-initial syllables must follow the pattern CV(N).
-These sequences are forbidden: ji, wu, wo, ti, nm, nn.
-
-\subsection{Syntax}\label{syntax}
-\subsubsection{Fundamental notions}
-As in other natural languages, 
-colloquial TP might have
-incomplete sentences and deviate from the
-norm.
-The basic structure of sentences is:
-`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
-The li might be repeated to associate more than
-one predicate to the subject.
-The particle li is omitted if the subject is a simple mi (I or us)
-or sina (you). A discussion about issues (potential problems)
-yield by this last rule
-and how I deal with them is in Appendix~\ref{mytoki}.
-The e might be repeated to associate more than
-one object to a predicate.
-Sentences might be related through la,
-'sentence' la 'sentence', where the second sentence is
-the main sentence, and the first sentence is a condition
-to the second.
-Multiple la-s are not described in literature,
-but one might assume that the last sentence
-is a conditional to the next,
-except in cases where the context strongly suggests
-otherwise.
-
-Noun and verb phrases are (usually) built with the non-particle words.
-The first word is the noun or verb and subsequent words
-qualify the noun or verb (i.e. are adjectives and adverbs).
-The pi particle might be used to separate sequences of words
-to be evaluated before the relation yield by pi.
-As pi is often ill understood and used,
-the following structures might be handy for newbies and as a
-reference:
-\begin{itemize}
-  \item No pi, `word word word':
-word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
-  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
-      word $\leftarrow$ (qualifies 1)  word ].
-  \item Two pi-s: `word pi word word word pi word word':
-    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
-or:
-    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
-\end{itemize}
-
-\noindent Further notes on the usage of pi:
-\begin{itemize}
-  \item In a sequence of words, without pi, the second word qualifies
-    the first, the third word qualifies the phrase yield by the first
-    two words, the fourth word qualifies the phrase yield by the first
-    three words and so on.
-  \item It is redundant to use pi before the last word in a noun or
-    verb phrase, reason why it is most often
-    omitted.
-    Its use in this case is considered an error~\cite{tpLang,kama},
-    but, as one might notice, it does not add (much) information
-    through syntax because the order of qualifications is conserved.
-    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
-    Also used in~\cite{akesiWawa}.
-  \item The book by jan Pije~\cite{kama} describes another use for pi:
-    after li to mean possession, e.g. `soweli li pi sina' (your pet).
-    This employment of pi might be regarded as correct, but is promptly written
-    as a noun phrase (e.g. `soweli sina') and is not mentioned
-    by the official book~\cite{tpLang}.
-\end{itemize}
-
-All the words except the particles (li, e, la, pi, a, o, anu, en, seme, mu)
-and the words classified solely as
-prepositions (kepeken, lon, tan)
-are usable in any position in noun and verb phrases,
-(i.e. 107 words discarding synonyms).
-Notice that the phrase expresses a noun (in a subject or
-object) or a verb (in a predicate).
-And that the first word of the phrase is the noun or verb,
-and that subsequent words are adjectives or adverbs.
-The pi precedes a noun or a verb to be qualified and then qualify
-the phrase that comes before it.
-
-At this point, the only missing syntax rule is related to
-the prepositions: kepeken, lon, sama, tan, tawa.
-They might appear at the end of a phrase,
-should be followed by another phrase,
-and require no particle. E.g.
-`toki *tan* jan Pije li pana e sona *tawa* mi'.
-Because prepositions can be used as adjectives
-or nouns or verbs or adverbs, etc, and might follow
-anu noun or verb phrase,
-they are one of the three main sources of ambiguity:
-prepositions; absence of li after sina and mi; incomplete sentences.
-E.g. '(mi [li]) pana tawa kon e ilo pi suli mute'.
-
-\subsubsection{Particles}\label{sec:par}
-Beyond the structural particles (li, e, la, pi) presented
-in last section, other particles are:
-
-\begin{itemize}
-  \item a or kin, emphasis.
-  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
-  \item taso, means however as sentence or 'only' if adjective.
-  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
-    For verbs, repeat li.
-    For object nouns, repeat e.
-    If the noun is complementing a preposition (e.g. tawa, lon),
-    one might repeat the preposition.
-    As TP is a recent language, and is particularly able to cope with
-    variation due to its simplicity, I would advocate for
-    using en and anu wherever there is no ambiguity.
-    E.g.: 'mi tawa mute anu moku lon tenpo lili'
-    In the official documentation that is not described and
-    might be regarded as wrong in strict canonical TP.
-  \item nanpa, denotes numbering.
-  \item seme, for questions, used next to the thing being asked for.
-    'tan seme la sina pana e sike?'.
-  \item mu, for animal noises. For me it is not a particle, as in the
-    official dictionary, but a noun.
-    I also like to use it as a verb:
-    'mi pakala e luka. mi mu mute.'
-\end{itemize}
-
-The vocabulary specifies these morphosyntactic classes:
-nouns, adjectives, verbs, pre-verbs, adverbs, prepositions, particles, and numbers.
-I find that it might help the speaker, specially the newcomer, but
-it might also suggest a deviation from what I understand and read:
-the words might be used indistinctly to be a noun
-(the first word of: a subject, a predicate when there is no object, an object,
-or a prepositional complement),
-an adjective (anything that follows an noun and is not a particle or a
-preposition),
-a verb (follows mi or sina or li),
-or an adverbs (follows a verb).
-The pre-verbs (wile, ken, awen, kama, lukin, sona),
-might be followed by a verb, but might also be understood
-as a verb qualified by the next word,
-which carries a very similar if not identical meaning
-('wile moku' might be understood both as a pre-verb followed
-by a verb and as a verb followed by an adverb).
-The pre-verb words are all also specified in the official dictionary
-to have other morphosyntactic classes,
-such as adjective, noun, and verb.
-The only exception is wile, which is specified to be only a pre-verb.
-
-Thus, the classes given in the dictionary dictate little
-in practice:
-jan kala li lape lon ni.
-Where kala, lape and ni are in this phrase
-as adjective, verb and noun,
-and are in the dictionary as noun,
-adjective and adjective.
-
-\subsubsection{Recognizing POS tags by speaker and machine}
-One might perform a POS tagging by following these rules:
-\begin{itemize}
-  \item Noun: the first word in a noun phrase.
-    Starting a sentence if it is complete,
-    after an 'e', after a pi, and after a preposition.
-    The predicate is often a noun
-    if the sentence has no object.
-  \item Adjective: second word on after an e and after
-    second word on in the subject phrase if present.
-  \item Verb: after a li, mi or sina.
-    If there is no object, the predicate
-    might be a noun or a verb.
-  \item After a preposition,
-    there is a noun phrase.
-  \item Notice that there is ambiguity in the structure
-    introduced by the omission of li after mi and sina.
-    Also, when there is no object, a noun or a verb
-    or an adjective might be in the verb position.
-    % The prepositional complement is also not defined.
-    % (mi moku tawa pali, tawa tomo). % TTM
-    These are sources of syntactic ambiguities
-    in TP.
-    They might be resolved or minimized by using the semantics
-    of the words.
-    An initial effort in this direction might be
-    using the classes in the official dictionary to resolve
-    ambiguities whenever possible.
-    This solution is not optimal for correct POS tagging,
-    and does not solve all possible ambiguities
-    (there are words classified as noun and adjective,
-    adjective and verb, noun and verb, particle and verb).
-    Another source of ambiguity is the pre-verbs as described
-    in the literature~\cite{tpLang,kama}.
-    But I find it reasonable to understand them as verbs,
-    as described in the end of Section~\ref{sec:par}.
-  \item If there is no object, the predicate might be parsed
-as being a hybrid of noun, verb and adjective.
-  \item The prepositional phrase might also be taken to be a hybrid
-    of noun, verb and adjective.
-\end{itemize}
-
-% Also, the prepositional complement might
-% be a noun or a verb.
-% I could not come up with a sentence
-% where it would be understood as an adjective. TTM
-
-\subsubsection{Additional notes}
-John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
-states that there are not noun, verb or prepositional
-phrases in TP, but only phrases in a structure
-yield by particles and content words~\cite{janKipo}.
-The only synonyms on Toki Pona are:
-a and kin; lukin and oko; sin or kamako;
-ale or ali.
-In formations such as
-toki e ni:, wile e ni:, tan ni: etc.,
-'(e) ni' might be omitted and : used alone.
-Names are by default transliterated,
-but might not be.
-These and other issues are further described in Appendix~\ref{mytoki}.
-
-% my view
-Many other considerations might be made about the language,
-such as additional (deprecated and new) words,
-situations where li is used and avoided,
-canonical and alternative employment of pi,
-counting systems,
-associations of TP word sequences to words in English,
-usage of images to represent TP words and sentences,
-and best practices in writing TP texts or for translation.
-For these and other matters, visit~\cite{tpLang,kama,gdoc,tp4}.
-
-% considerations about the official book and jan Pije
-% pictorial forms
-\section{Software for analysis, synthesis, and syntax highlighting}\label{hacks}
-In this section,
-we describe software, statistic, natural language processing,
-and visualization results:
-1)
-the analysis of the Toki Pona language through statistics
-of the vocabulary obtained by processing the official dictionary;
-2)
-the synthesis of sentences, poems and paragraphs (e.g. short stories)
-in Toki Pona;
-3)
-syntax highlighting for Toki Pona,
-including fine-tuning and theoretical
-considerations;
-4)
-an initial Wordnet~\cite{wordnet} of
-Toki Pona words.
-Corpus analysis was already found in~\cite{corpus}.
-The Python scripts (or the toolbox)
-for obtaining all the results in this sections,
-and more, is publicly available in~\cite{tokipona},
-to facilitate both the inspection of the results and the
-generation of derivatives by other interested parties.
-Such scripts were made very friendly for one to understand and alter
-at will.
-
-\subsection{Statistics of the vocabulary}\label{sec:stat}
-In~\cite{corpus} are statistics about a Toki Pona corpus.
-This section focuses on the statistics of the vocabulary
-and of the syntactic rules:
-the letters, phonemes, word sizes,
-possible combinations for words and sentences.
-% The statistics are in Appendix~\cite{listings},
-% and the next paragraph is an overview. TTM
-The script \tttt{makeStatistics.py}
-was used to obtain the all the measurements and tables
-discussed herein,
-and hold further measurements and sets of words
-which were regarded as less suitable for this exposition,
-but is useful for a deeper investigation.
-
-As described in Section~\ref{basics},
-there are only 14 letters,
-and phonemes are also very restricted.
-There are 120 different words in the official vocabulary,
-4 of them having synonyms.
-A total of 124 tokens (or lemmas),
-not counting proper nouns (names)
-and punctuation.
-Table~\ref{foobar} shows the number of words
-related to each POS tag (or morphosyntactic class)
-specified in the dictionary.
-
-\input{pos}
-
-From the official 124 words, 
-26 of them (20.97\%)
-have only one syllable,
-85 (68.55\%) have two syllables,
-and 13 (10.48\%) have three syllables.
-No official word has four or more syllables.
-In all the 124 tokens, there are 235 syllables (68 different),
-and Table~\ref{tab:freqsy} exhibits the 10 most frequent
-syllables in the first, last, middle and all positions.
-Middle-word syllables only occur 13 times,
-and are all different, with the exception of 'la'
-which occurs twice.
-A complete list of the syllables and their frequencies (in different
-positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
-script.
-A list of all words, grouped by their size in syllables,
-and ordered alphabetically and by the number of letters,
-is in the \tttt{hlsyl\_\_} variable.
-
-\input{syls}
-
-Vowel and consonant frequencies,
-and comparisons of vowels against consonants,
-are as shown in Table~\ref{freqLet}
-for starting, end and internal position.
-The limited number of consonants favors
-the naturalness of the language prosody
-as it resembles babbling.
-
-\input{vowels}
-
-Given the rules given in Section~\ref{phonology}
-(of 14 letters, 5 vowels, (C)V(n) phonemes,
-forbidden ji, wu, wo, ti, nn, nm),
-96 words are possible with 1 syllable,
-8256 with 2 syllables, 710016 with 3 syllables.
-In the incident words of the official dictionary,
-no middle syllable end with the nasal consonant 'n'.
-
-Possible sentence structures are unlimited.
-For a noun or verb phrase, one might quantify the possibilities
-by considering all words but the particles that are not classified
-also as something else
-(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
-and the prepositions that are not classified also as something
-else (i.e. kepeken, lon, tan).
-This yields $120-13=107 words$.
-As discussed in Section~\ref{syntax},
-I advocate that, although such words are classified in the dictionary,
-they might be used indistinctly as nouns, adjectives,
-verbs and adverbs.
-Thus, one has 107 possibilities of noun and verb phrases with
-one word, $107^2=11,449$ possibilities with two words,
-$107^3=1,225,043$ with three words and so on.
-Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
-(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
-and assume that the sentence has at most one prepositional phrase,
-one might quantify the possibilities using the formula:
-$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
-where $5$ stands for the possible prepositions.
-To account for the particles, one possibility is
-to assume for simplicity that one might use none or one particle
-(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
-at each phrase, yielding:
-$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
-For example, assume $n=v=o=p=1$, then
-$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
- 4.300066\times 10^+12$,
-i.e. more then 4 trillion possible
-sentences with single word phrases (i.e. no adjectives or adverbs),
-while allowing only one particle per phrase
-and only one predicate phrase
-(e.g. 'ona li moku e soweli lon supa').
-The general case is implemented as a function
-in the script \tttt{makeStatistics.py}~\cite{tokipona}.
-
-\subsection{Synthesis of text}\label{synth}
-Such counting exercises are also useful
-for facilitating (semi-)automated writing through scripting.
-The syntax organizes the words in larger structures.
-The rhymes are very restricted and sonorities are
-bounded by the small vocabulary and simple syntax.
-There are some specific tasks for achieving texts,
-such as finding the number of syllables considering the elisions,
-or handling interaction of the writer with the script
-to choose sentences or verses or stanzas.
-The same package~\cite{tokipona}
-has capabilities for synthesizing TP text.
-On its most basic level, these routines yield
-noun, verb and prepositional phrases,
-and sentences.
-They also aim at making larger scale texts
-by assuring the use of specific words (to entail context),
-e.g. to assist the creation of short narratives,
-and by employing stylistic outlines for poems.
-
-\begin{figure}     \center
-  \begin{subfigure}{.5\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/phrase_}\vspace{-0.15cm}
-          \caption{Synthesized phrase.}
-            \label{fig:sfig1}
-  \end{subfigure}\\\vspace{0.3cm}
-  \begin{subfigure}{.9\textwidth}
-    \centering
-        \includegraphics[width=1.2\linewidth]{figs/sentence_}\vspace{-0.15cm}
-          \caption{Synthesized sentence.}
-            \label{fig:sfig2}
-  \end{subfigure}\\\vspace{0.3cm}
-  \begin{subfigure}{.8\textwidth}
-    \centering
-        \includegraphics[width=1.2\linewidth]{figs/paragraph_}\vspace{-0.15cm}
-          \caption{Synthesized paragraph.}
-            \label{fig:sfig2}
-  \end{subfigure}\\\vspace{0.3cm}
-  \begin{subfigure}{.7\textwidth}
-    \centering
-        \includegraphics[width=.8\linewidth]{figs/poem_}\vspace{-0.15cm}
-          \caption{Synthesized poem.}
-            \label{fig:sfig2}
-  \end{subfigure}
-  \caption{Toki Pona texts synthesized by the \tttt{makeTexts.py} script
-  in~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
-  and (d) a poem.
-  The unusual word combinations are convenient for exploring semantic
-  possibilities.
-  One might obtain many of such texts to select the excerpts
-  he/she finds fit for the intended use.
-  For more information, see Section~\ref{synth}.
-  The words are colored in accordance to the considerations
-  in Section~\ref{shigh}.}
-  \label{fig:syn}
-\end{figure}
-
-Figure~\ref{fig:syn} holds some synthesized texts
-with different color schemes for the syntax highlighting,
-as presented in the next section.
-The textual synthesis described here,
-and implemented in \tttt{makeTexts.py},
-might be enhanced in countless ways,
-e.g. as described in Section~\ref{conc}.
-
-Automatic and randomized synthesis of texts in TP is particularly
-useful because of the reduced vocabulary where each word is related
-to a broad semantic field.
-Ideas often make sense in unexpected ways, and thus
-the synthesis yields a procedure to 
-explore the semantic possibilities within TP.
-One might object that the resulting texts are unusual 
-and even consider that they often hold insubstantial or unsound meaning.
-I advocate that these unexpected formations are desirable
-for exploring the possibilities of the language and of thought,
-and for artistic endeavors.
-Also, to obtain texts which one finds usual or satisfactory
-in more strict or personal terms, such a person might just write
-them normally.
-
-% phrase
-% sentence
-% poem
-% text
-% each with a different color scheme
-
-% nanpa followed by numbers
-% pu, taso, kin/namako
-% particles
-% rhymes 
-% modes in which the words are only used as specified in the dictionary
-
-% \clearpage
-\subsection{Syntax highlighting}\label{shigh}
-The same package~\cite{tokipona}
-has a Vim~\cite{vim} syntax highlighting plugin
-for Toki Pona,
-routines to arrange the coloring schemes
-(i.e. to yield the text files with the specifications for which tokens
-to relate to which coloring groups),
-and instructions for installing and using
-the resulting syntax highlighting for Toki Pona texts~\cite{tokipona}.
-An online syntax highlighting gadget is found at~\cite{tpNetSH},
-and the solution here described presents a number of enhancements:
-it is capable of coloring all morphosyntactic classes;
-it might be fine-tuned in the colors and their relations to sets of
-tokens; it is designed to be used within Vim
-and might be exported as HTML; the highlighting scheme is promptly
-rendered by a script; the resulting script might be used with
-many color schemes.
-
-Basically, the resulting syntax highlighting
-distinguishes the words among the morphosyntactic
-classes according to the official dictionary as given in
-Table~\ref{foobar}.
-A word often belongs to more than one class,
-thus the precedence of them might be set by the user.
-Also, some classes might be further refined (e.g. words beginning with
-vowels) or joined (e.g. distinguishing only particles and the rest,
-or particles and prepositions and the rest).
-The colors are also promptly changed according to~\cite{vim}
-and exemplified in the package documentation.
-
-Currently, the Python package synthesizes the
-syntax file through the\\ \tttt{makeVimSyntax.py} script.
-The user has control of class precedence and
-merging and further details through tweaking such routine.
-The choice of precise coloring schemes
-involves fine tuning the color scheme being
-used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
-and Vim's highlighting schemes as described in~\cite{vim}.
+Vim was first released publicly in 1991.
+It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
+Vim's development is coordinated (and performed) since the beginning
+mainly by Bram Moolenaar.
+Today, current bleeding-edge version is 8.0.1257.
+I found no explicit stable, alpha or beta versions.
+I found no scientific article on Vim 
+(this might be the first one), although there are
+books, software and third-party documentation on the web.
+
+\section{Basics}\label{basics}
+Vim's interface is text-based.
+In the GUI mode (gVim),
+there are convenient menus and toolbars
+but all functionalities are still available though
+the command line mode.
+Vimscript is the internal language of Vim,
+and is often used for scripting by users
+although other languages might be used 
+(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
+Each line of a Vimscript is a command on the
+command-line mode.
+This section can be thought as a tutorial
+that focuses on the namespaces, i.e. sets of tokens
+that carry values or triggers procedures.
+One should see Appendix~\ref{keys}
+to understand the notation vim uses.
+
+\subsection{The bare minimum}\label{minimum}
+You open a file with Vim by executing
+the command: \texttt{vim <filename>}.
+Inside Vim, you start in the normal
+mode, and might want to move around using
+\texttt{h-j-k-l} for left-down-up-right.
+To insert characters, move your
+cursor to the desired location an press i,
+which puts Vim in the insert mode.
+Go back to normal mode by pressing
+\texttt{<ESC>} or \texttt{<C-C>}.
+You save the file by typing \texttt{:w<CR>},
+and exit Vim by typing \texttt{:q<CR>}.
+You can save and quit with \tttt{:wq<CR>}
+or \tttt{:x<CR>} or \tttt{ZZ}.
+
+\subsection{Vim help}
+You can find help for vim in various places.
+The standard resource is the Vim help files.
+They are accessed by typing \texttt{:h <anything><CR>}
+in normal mode.
+Examples of such \texttt{<anything>} are:
+color, navigation, :vs, vimtutor.
+Type 
+\texttt{:h usr\_toc<CR>}
+to access the official User Manual,
+which is considerably lengthy and complex
+and is usually not read by users for a few years.
+In learning Vim, one
+might want to run the \texttt{vimtutor} command
+(outside Vim) to start the Vim Tutor.
+
+There are good resources on the Web for learning
+and tweaking Vim:
+\begin{itemize}
+  \item ``Vim Adventures'' is an online RPG game for practicing
+  and memorizing Vim commands while having fun.
+  This game is quite famous among Vim users.
+  \item There are official and semi-official Vim sites e.g.:
+  \url{www.vim.org}, \url{https://www.vi-improved.org} and
+  \url{http://vim.wikia.com/}.
+  \item Many hacks, understandings and general issues
+  (e.g. how to make such a move) are asked and answered
+  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
+  One often finds these links through a search engine.
+  \item You can find many videos about Vim.
+  One traditional site is \url{http://vimcasts.org},
+  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+\end{itemize}
+
+% any other commands? files?
+\subsection{Namespaces}\label{namespaces}
+Vim is a text editor ouroboros~\cite{ouroWiki} because
+text and writing alters text and writing.
+You have namespaces where tokens have scalar
+or complex values.
+
+\subsubsection{Commands and mappings}
+There are commands, typing sequences which trigger automated actions,
+for each mode:
+\begin{itemize}
+	\item in Normal mode all keys are mapped to commands.
+		There is redundancy and additional commands
+		using Ctrl and Shift keys.
+		Some keys expect a second key,
+    and have combinations not used (available for new mappings),
+		specially the z and g.
+    See Section~\ref{normal} and Appendix~\ref{notes}
+    for more insights into the commands available in the
+    normal mode.
+	\item In the other modes, the sequences available for mappings are more obvious and abundant.
+		One should look at \tttt{:h index} to know about all the standard mappings
+		and use \tttt{:map} to list the user-defined mappings.
+\end{itemize}
+
+\ttt{C-\textbackslash} is often reserved for extensions,
+which makes it a safe namespace to use (while there are no
+such extensions).
+
+A colon command can be written as a string
+and executed by the \tttt{:execute} colon command.
+E.g. \tttt{:execute 'vs afile.txt'}.
+As there are colon commands that execute commands in other
+modes, e.g. \tttt{:normal ?\^def },
+the \tttt{:execute} is a way to build commands in any mode,
+e.g.
+\tttt{:execute 'normal' tabn 'gt'}.
+
+\subsubsection{Variables}
+There are some types of variables in Vim:
+\begin{itemize}
+	\item Environment variables: names start with \$ and hold system
+		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
+	\item Option variables: names start with \& and are meant to control the behavior of the editor.
+		One might change a value through set or let, e.g.
+		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
+	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
+	\item Internal variables are created with let and preferably have a prefix:
+	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
+		are local to the buffer, window, tab page, function, and
+		sourced Vim file, respectively.
+ \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
+		\tttt{a:} is for function arguments.
+		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
+		More about internal variables in \tttt{:h internal-variables}.
+	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
+\end{itemize}
+You can echo any of such variables or use in expressions.
+Notice that you will only be able to echo a \tttt{b:} variable inside
+the buffer where it is defined.
+For all the Vimscript capabilities, including loops, conditionals,
+and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
+Classes are possible only in rudimentary forms, e.g. through dictionaries,
+but the language is otherwise overall quite powerful,
+specially in dealing with text and editor behavior, as expected.
+
+\subsubsection{State lists}\label{state}
+Vim keeps a number of useful lists which expresses the state of the editor:
+\begin{itemize}
+	\item The entered commands are accessed through \tttt{:hist a}. 
+    The tokens a / s : can be used for specific types of commands, such
+		as search and colon commands.
+	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
+  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
+	\item Tabs list can be reached through \tttt{:tabs}.
+		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
+	\item Jumps are available through \tttt{:jumps}.
+		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
+  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
+    They also keep track or your copy, edition and deletion and are promptly defined
+    by recording a typing sequence with the \tttt{q} normal command.
+    Vim keeps only the last edition, in register \tttt{".}.
+    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
+    A hack to keep the latest deletions and copies in the standard register
+    might follow the same pattern, but use another \tttt{:h event} and
+    monitor register \tttt{""} (maybe also \tttt{"0}).
+	\item An undo list can be accessed with \tttt{:changes}.
+	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
+	\item The markers defined are listed with the \tttt{:marks} command.
+		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
+		Uppercase letters are cross buffers.
+	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
+		and variations, such as \tttt{:grep}.
+		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
+		to open the Quickfix window, where the lines of occurrence are in sequence
+		and one can \ttt{CR} one of them to have the cursor in the main window active at
+		the first character of the match.
+		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
+		to use the location-list window instead of the Quickfix, which
+		is very similar, but one per window instead of one per buffer.
+		More information in \tttt{:h quickfix}.
+	\item A tags list have to be made so one can use tags.
+		Most often one will generate the tags list using the exuberant ctags,
+		which supports dozens of languages.
+		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
+		and then using \ttt{C-]} to
+		go to the position of tag under cursor, and \tttt{:tabe tags} to open
+		the tags file.
+	\item The argument list holds a list of files to be edited or browsed. 
+		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
+	or using commands (e.g. \tttt{:ar ./*}).
+		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
+		one might perform actions on each file in argument list using \tttt{:argdo}.
+		All files in argument list are also in the buffers list.
+    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
+  \item All the autocommands might be accessed through \tttt{:au}.
+    They are event-triggered actions Vim performs.
+\end{itemize}
+
+A file with information about the state of the editor
+can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
+and this script might be consulted because it has
+a collection of commands to access various settings of Vim.
+Another good list of commands to know about Vim's state is kept on
+\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
+I would specially highlight the \tttt{:syntax} command because
+it displays the tokens and related type of meanings when run
+inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
+
+\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
+
+You can keep track of the editor state though commands and persistent visual
+cues, specially the tabs bar, the status line, and the line
+reserved for the command-line.
+For state persistence, one might keep an undo file for each file as
+in~\cite{vimrc}.
+Sessions are easy to manage, enabling one to save and load the
+editor's state, with the opened windows, tabs, buffers, etc.
+I use the mappings in~\cite{vimrc} because they keep the sessions
+in a reasonable directory and makes it easier to remember and tweak
+the standard commands to deal with sessions.
+More information in \tttt{:h sessions}.
+One might use \tttt{:h views} to keep the state of one window,
+but sessions keep all the states from all windows.
+This entails a strategy to deal with Vim that is similar to
+the use of Byobu/Tmux/Screen.
+The main limitation I found to this approach is that
+Vim is not keeping track of the terminals opened.
+If you open a terminal inside Vim with the \tttt{:term}
+command, you will save the session as usual, but when loading
+you get dummy empty windows for them and an error message.
+Screen and Tmux are the most popular terminal multiplexers.
+Byobu, built on top of them, has awesome keyboard shortcuts
+for managing sessions, windows and splits of terminals.
+Byoby/Screen/Tmux keep the state for future load.
+
+Because browsing the interface in Vim is fast,
+and it favors copy and paste of text,
+I tend to keeep a tab with some terminals:
+one with an IPython shell, another two for compiling latex
+and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
+One enters normal mode in terminal with \tttt{C-W N}
+and it is very corfortable to copy and browse the bash history of the
+terminal.
+I found the mappings on~\cite{vimrc} very helpful for directing
+editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
+which I make available across terminal-job and normal modes.
+But I've been thinking on using \ttt{C-} commands also
+for tabs (not only for splits).
+
+A good strategy I find is to have selective visual cues of the state to make persistent or hide
+and a mapping to toggle each of them.
+Now I toggle byobu/screen/tmux bar with \ttt{F5},
+status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
+I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
+Numbering is always there.
+I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
+Instead of keeping the status bar, I use \ttt{C-g} to know about the
+file and \ttt{gC-g} to know more and rarily.
+It seems not possible to remove the statusbar between horizontal splits.
+After asking around and experimenting, I realized that it seems
+reasonable to keep at least one line dividing the windows, so
+if it comes to it, I just \tttt{set statusline=-}.
+Unfortunatelly, as far as I could dig, one will need to enter
+Vim code to enable a horizontal split without losing a line.
+My ideal of this feature would be to have a visual cue of the
+first and/or last line of the windows in the lines-number
+column, or complete the spaces and empty chars with \$\$\$\$ or so.
+
+Autocommands are the standard way to define event-trigered routines in Vim.
+These are often related to particular file types, but are also often
+in defining the automated behavior of Vim.
+In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
+in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
+
+\subsection{Using Vim's modes}\label{modes}
+Vim has some basic and fully implemented modes of usage:
+\begin{itemize}
+  \item Normal mode: used for changing
+  the position of the cursor or the text displayed
+  at the window.
+  A core goal of the normal mode it to allow fast
+  navigation of the document while allowing
+  the typist to maintain its fingers on the home row
+  (i.e. on the center of the keyboard).
+  The mode is also used for manupulating text
+  (e.g. copy, paste, delete, change case) and
+    changing to other modes.
+  \item Insert mode: for inserting text.
+  \item Command-line mode: for entering Ex commands.
+  \item Ex mode: similar to command-line mode,
+  but more specialized for running various Ex commands.
+  \item Visual mode: for making, manipulating and navigating
+  selections of texts.
+  \item Select mode: similar to visual mode but
+  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
+\end{itemize}
+
+There is another basic mode, but it is not fully implemented:
+the Terminal-Job mode.
+There are seven additional modes which are mostly subordinate 
+to the basic modes and will be described when convenient.
+The manual page for Vim modes can be accessed by typing
+\texttt{:h vim-mode}.
+Some of the modes are now further considered for
+an overview of the Vim usage possibilites.
+
+\subsubsection{Normal mode}\label{normal}
+Sometimes also called navigation or command mode,
+the normal mode is most powerful for
+navigating, manipulating texts and changing to other modes.
+
+The simplest of these three is changing to other modes:
+type any of these letters to change to insert mode:
+\texttt{iIaAoOsScC}. More on the transition between
+normal and insert mode on Section~\ref{navIn}.
+Type any of these characters to change to command-line mode:
+\texttt{:/?}.
+Type \texttt{Q} to enter Ex mode.
+Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
+
+For most basic and naive navigation, one should check Section~\ref{minimum}.
+Most often, one uses:
+\begin{itemize}
+  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
+and whole page down and up, although these commands might
+be set to scroll a different number of lines.
+  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
+  \item \texttt{(w,b,e)} to move to the next, previous and next end-of-word.
+    There are motions to iterate over sequences of characters separated
+    by special characters (e.g. punctuation and parenthesis) as
+    specified by the output of \texttt{:se iskeyword}.
+    To iterate over space-separated tokens, use \texttt{W,B,E}).
+    To move to the end of last word, one might use \texttt{be}
+    or \texttt{ge}.
+The \texttt{),(} commands iterate through sentences,
+		\tttt{\},\{} through text blocks separated by empty lines.
+  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
+	  \tttt{;} and \tttt{,} for next and previous found character.
+  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
+  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
+    jump list.
+  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
+    the first non-blank character of the line.
+    A mark is registered by the user in any cursor position
+    by typing \texttt{mX}, where X is any letter.
+    If X is lowercase, the mark is local to the buffer (the file),
+    if it is uppercase or numeric, it is global to the Vim session
+    (cross buffers).
+\end{itemize}
+
+There are many more facilities to navigate Vim
+as explained in \texttt{:h navigation}.
+
+For changing the text, usual commands include:
+\begin{itemize}
+  \item \texttt{d\{motion\}} to delete the characters
+    involved in the motion command.
+  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
+  \item \texttt{x,X} to delete the character under or before the cursor.
+  \item \texttt{$\sim$} to swap the case of a character.
+  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
+\end{itemize}
+
+There are way more commands to change the texts.
+Some of them are discussed in Section~\ref{navIn}
+because they involve a transition to the insertion mode.
+A thorough consideration of the commands in the normal mode
+is found by executing \texttt{:h navigation} and 
+\texttt{:h change.txt}.
+
+\subsubsection{Insertion}
+Once in the insertion mode, the character keys
+input the characters at the cursor position at the current buffer.
+One can exit insert mode by pressing \texttt{<Esc>},
+and Vim will be put in normal mode.
+Most useful commands in insert mode include:
+\begin{itemize}
+  \item \ttt{C-O} to execute one and only command in normal mode.
+	  This enter a secondary mode (see Section~\ref{modes})
+  \item \ttt{C-R} to paste a register (a variable starting with
+    '@', defined, copied or recorded through a \tttt{q} command in
+    normal mode and as covered in Section~\ref{state}).
+  \item \ttt{C-T} to indent current line.
+  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
+    the line or to the previous word.
+  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand..
+  \item \ttt{C-X} commands for scrolling the window with multiple
+    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
+\end{itemize}
+
+\subsubsection{Command-line mode}
+This mode is dedicated for writing colon, search and filter commands,
+entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
+Most useful commands in this mode include:
+\begin{itemize}
+  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
+    end of the line.
+  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
+    until the cursor.
+  \item \ttt{C-R} to paste a register as in insert mode.
+\end{itemize}
+
+\subsubsection{Terminal-Job mode}\label{terminal}
+This mode is reported to not have reached a stable usage design
+(see \tttt{:h terminal}).
+I find that it works exceptionaly well and have used it to run
+scripts in an IPython shell, compile latex files, and even open PDFs and images.
+Vim browsing of windows and text manipulation is well developed,
+so the terminal mode enables a very convenient integration,
+more traditionally achieved through Byobu/Tmux/Screen terminal
+multiplexers.
+Most useful commands in terminal-job mode include:
+\begin{itemize}
+  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the normal
+    and command-line modes.
+  \item \tttt{<C-W>\_"} to paste a register.
+  \item \tttt{i} for entering the terminal-job mode from the normal mode.
+\end{itemize}
+
+It is useful to define the same mappings for nativating splits and tabs
+for both normal and terminal-job modes, as in~\cite{vimrc}.
+
+\subsubsection{Normal $\rightarrow$ insertion}\label{navIn}
+Many commands bridge from Normal to Insertion modes:
+\tttt{csrCSR}, \tttt{iws}, etc.
+These make convenient the replacement of text and populates registers.
+The absence of a short command to insert one char is
+a known issue in Vim.
+Reasonable mappings to insert and append a char to and around
+other char are in~\cite{vimrc}.
+Vim commands are designed to couple operator and motion commands.
+There are many operator commands that take the editor from
+the normal mode to the insert mode, most of them
+favoring deletion or change, as detailed in \tttt{:h operator}.
+Motion commands are described in \tttt{:h motion}.
+
+\subsubsection{Netrw}\label{netrw}
+The standard interface of Vim for browsing file trees is Netrw.
+It starts when you open a directory, such as with \tttt{:e .<CR>}.
+It has solid support for editing remote files (such as over ssh or
+ftp) and handy e.g. mappings to open the files as splits and tabs (specially \tttt{pot}).
+Most useful commands in netrw include:
+\begin{itemize}
+  \item \tttt{d} and \tttt{\%} for creating directories and files.
+    \ttt{Delete} removes both.
+  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
+  \item \tttt{gb} and \tttt{uU} are used to load directories
+    while marked files might be copied, moved, edited, grep-ed, tagged and migrated
+    to and from the argslist as in \tttt{:h netrw-mf}.
+\end{itemize}
+Last notes:
+there is no insert mode in netrw interface;
+the commands in \tttt{:h netrw-explore} are
+convenient for opening the directory of the
+file being edited;
+further information in \tttt{:h netrw}.
+
+\subsubsection{Ex}
+One might use the normal commands \tttt{q(:,/,?)} to have a window
+with colon or search command history to be edited normaly,
+and the chosen command can be run with \ttt{CR}.
+Vimscript was largerly based on the ex editor~\cite{ex},
+and a more advanced user might use it for prototyping
+by defining mappings and settings and managing sessions and scripts
+propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
+In the default interface started with the \tttt{Q} command in normal mode,
+each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
+follow documentation from \tttt{:h Ex-mode} for further information.
+I might be loosing something, but for most
+hacking I use the command-line window accessed with \tttt{q:}.
+I find it comfortable to browse the history in a normal-like mode,
+and edit them also using the insert mode as usual.
+The mapping have to be defined as autocommands as in~\cite{vimrc}.
+
+\subsection{Standard configuration files and directories and my .vim/vimrc}
+You can check the scripts Vim loads by using the debug script mentioned
+in Section~\ref{state}.
+By default, \tttt{$\sim$/.vimrc} and \tttt{$\sim$/.vim/vimrc} files are run by Vim at the beginning of the startup.
+One might edit the vimrc file with \tttt{:e \$MYVIMRC}
+and reload it with \tttt{:source \$MYVIMRC}, but
+I use the mappings in~\cite{vimrc} to encourage a continuous enhancement of my settings.
+Any other file might be included to run at startup by
+adding a line \tttt{:source <afile.vim>} in vimrc.
+In fact, it is on vimrc that one usually specifies the plugins
+and plugin managers they use.
+Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
+or follow some patterns described on the next section to change the
+scripts and sequence of them to be loaded.
+
+\subsection{Plugins and packages}
+One can see the list of standard plugins
+with \tttt{:h standard-plugin-list} command.
+Any \tttt{plugin/**/*.vim} file inside a directory
+listed by \tttt{:se runtimepath} will be loaded
+(e.g. \tttt{.vim/plugin/something/ascript.vim}).
+There are various ways to automate the installation
+and enhnace the management of plugins.
+By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
+that downloads latest scripts from sourceforge as specified in \tttt{:h getscript-data},
+and the Vimball interface, which creates and loads a Vimball for a plugin
+with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
+that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
+The Vimball can be installed in a system by \tttt{:source <filename>.vba}
+or loading it at vim startup with \tttt{\$ vim <filename>.vba}.
+
+A Vim package is a directory that contain plugins.
+It should be located inside a \tttt{pack/} directory
+somewhere in the directories listed with \tttt{:se runtimepath}.
+The plugins found in \tttt{pack/<packName>/*/start/} are loaded
+at startup, the plugins found at
+\tttt{pack/<packName>/opt/**} are loaded with\\
+\tttt{:packdd <script\_or\_directory\_name>}.
+
+All directories vim looks for scripts are described
+in \tttt{:h vimfiles} and are basically set by
+\tttt{:se runtimepath} and conventions inside
+each path therein, such as to look for \tttt{vimrc} files
+and \tttt{plugin/} or \tttt{pack/} directories.
+Vim scripts can be loaded conditionally,
+e.g. only if a function is used as in \tttt{:h autoload-functions}.
+Example of such are filetype plugins (enabled by in a \tttt{ftplugin/<filetype>.vim} file,
+e.g. inside a plugin directory).
+There is a number of plugin managers for Vim.
+Pathogen and Vundle seem to be the most popular,
+one because of its minimalism, the other because of
+advanced features, e.g. for searching and installing plugins with
+colon commands.
+
+\subsection{Spell and spelllang (en and pt\_br)}
+One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
+or \tttt{:se spelllang=pt\_br}
+and toggle spell checking with \tttt{:setl spell!}.
+These are used so often that one might use shortcuts as in~\cite{vimrc}.
+Currently, Vim will download the files for a specific language if
+not found in system.
+
+\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}\label{script}
+In Vimscript,
+the colon commands (also Ex or command-line commands) are related through spaces,
+punctuation and keywords (see \texttt{:h script}).
+Scripting the Vim editor can also be accomplished using other languages,
+as well documented e.g. \tttt{:h python}.
+Functions are defined through colon commands and are called
+inside colon commands e.g. \tttt{:call MyFunction()}
+or \tttt{:echo MyFunction(4)}.
+Notice that functions are not commands but might be bind to them
+through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
+Executing source files is very straightforward with \tttt{:so
+<filename>.vim},
+and one can always use the Ex mode for rapid scripting.
+
+At the same time, the \tttt{:term} and terminal-job mode make
+scripting other software more convenient, as output is
+promptly navigated and copied. The shell works well for generic use,
+such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
+See Section~\ref{terminal} for further directions about the terminal.
+
+\subsection{Color}\label{visual2}
+Newer versions of Vim support 24 bit true color (aka 16 million colors)
+in terminal Vim (not only in GVim).
+The terminal must support true color, and tests are available e.g. in
+\url{https://gist.github.com/XVilka/8346728}.
+Then Vim needs to be set to use true colors with
+\tttt{:set termguicolors}.
+If using 8 or 16 bit colors, Vim uses the color pallete from the
+terminal, if using true colors each color is defined directly.
+Settings for using true colors inside Byobu/Tmux involve tweakings
+and are available in~\cite{vimrc}.
+Good color schemes to use with true colors are Gruvbox
+and Solarize.
+One might source syntax files at any time to change syntax, usually
+though linking tokens to syntax groups with
+\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
+and then relating the group to another group 
+\tttt{:highlight link <group\_name> <group\_name2>}
+or by specifying the group characteristics directly:
+\tttt{highlight <group\_name> guifg=\#ffffff}.
+If you change a syntax file, reloading a file with with \tttt{:e<CR>}
+updates the highlighting on the window with the corresponding filetype.
+A complete syntax highlighting support typically involves at least three files:
+\begin{itemize}
+  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
+    detected with e.g.
+    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
+  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
+    file type, such as: 
+    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
+    expandtab autoindent}.
+  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
+    tokens and highlighting groups; and highlighting group definitions.
+\end{itemize}
+
+This very scheme is implemented very straightforward in this
+plugin~\cite{tokipona}
+for highlighting text in the Toki-Pona language~\cite{tpLang}.
+Syntax highlighting plugins are as filetype plugins,
+but also have a \tttt{syntax/<filetype>.vim} file
+relating keywords to highlighting groups. 
+One might see every highlighting group, and their final visual results,
+with the command
+\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}.
+In my system, the \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
+directory, but \tttt{syntax/} files had to be moved to \tttt{$\sim$/.vim/syntax/}.
+
+\subsection{Fonts}
+The fonts are defined by your terminal or inside GVim.
+\ttt{C-+} and \ttt{C--} can be used to change font size.
+Some settings for fonts, such as boldface, might be set using
+the syntax highlighting facilities described in the previous section.
+
+\section{Conclusions and further work}\label{conc}
+This document seems reasonable as an overall reference of the Vim editor,
+at least for my usage and proficiency.
+Given the folkloric milestone of using Vim for 10 years,
+this article might serve as a benchmark for one to relate
+it's current use and understandings.
+As a pedagogical material, it seems to be unique in the emphasis
+on namespaces, understood as commands, variables, state-related lists, etc,
+especially in Section~\ref{namespaces},
+and the reference to the standard Vim documentation
+to achieve the DRY KISS style described in Section~\ref{intro}.
+
+\noindent {\bf Potential enhancements} to this document are:
+\begin{itemize}
+  \item The inclusion of facilities such as reading emails and connecting over ssh.
+    There is a working hack in~\cite{vimrc} for browsing over the WWW,
+    but such aspect of Vim usage might receive more attention
+    given that it is confortable to use the navigation and editing facilities
+    and the resulting integrated environment.
+  \item Updating of the information I can find about the issues discussed,
+    such as about status lines in Section~\ref{visual1}.
+  \item Include a discussion about Neovim.
+    I have never used it, but it seems to be reaching a considerable user base
+    and it might be feasable to give an account of Vim and Neovim
+    after some tests and researching the official and user base documentation.
+  \item Better cite documentation, plugins and Vim authors.
+    I prefered to keep the references inline through \tttt{:h} commands
+    and URLs, more in accordance with the style of Vim documentation,
+    but bibliographical items constitute a valuable asset for academic literature,
+    and some authors might find their work more pretiged if more
+    thoroughly cited.
+  \item An analysis of my usage, e.g. according to \url{http://www.drbunsen.org/vim-croquet/}.
+    This undertake might benefit from data from many users,
+    which favors the potential plugin for usage analytics described in the next bullet list.
+\end{itemize}
+
+\noindent {\bf Potential next steps} in using Vim:
+\begin{itemize}
+  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
+  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
+  \item Better integrate Python and Vimscript, especially for data visualization
+    and syntax highlighting management.
+    In accordance with the visualization issues described in
+    Sections~\ref{visual1} and~\ref{visual2}.
+  \item Make plugins for:
+    \begin{itemize}
+      \item listing all the mappings available in each mode and the typing combinations which are available
+        for new mappings.
+      \item Sessions, as described in Section~\ref{visual1}.
+      \item AA messages (shouts): to keep track, document and share of working sessions
+        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
+        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
+        relate AA sessions to other media, such as software repositories,
+        screencasts and images, interact with IRC channels and other social platforms.
+      \item Slick Vim: a collection of the settings, mappings and plugins I use.
+        Enhancements such as using \ttt{C-} commands also to browse tabs,
+        shortcuts to join a window into a tab,
+        and dummy minimal plugins as simplest, file type and syntax highlighting,
+        Some more elaborate tweaks should also be present, such
+        as breaking lines in sensible places for natural language texts
+        while respecting e.g. \tttt{:se textwidth}.
+      \item Dealing with .swo and .swp temporary files.
+        In summary, if the restored .swo file has the same content
+        and the file being opened,
+        the restoration phase can be omitted.
+        If the contents differ, Vim should open a tab with each file
+        in a vertical split and run \tttt{:windo :diffthis}.
+      \item Rendering images and equations.
+        These are useful for using Vim in presentations
+        or achieving a textual representation when it is mandatory,
+        such as to comply with the limitations of a platform (e.g. Vim editor). 
+        but also hold stylistic merits as ascii art is often
+        very appreciated.
+        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
+        and can directly render ascii charts from data using cues such are shape, position
+        and color.
+      \item Redirecting the commands that usually show the results in a 'more' interface,
+        which cannot be searched nor copied,
+        to a parsed and linked quickfix or location window.
+        The basic idea is to use \tttt{:redir} command to redirect the output
+        of such commands with \tttt{:se nomore}.
+      \item Slide presentations. I've been using some automation for browsing
+        slides and opening figures and some Vim users asked for the settings and
+        commands. They are very elementary use of registers being executed over
+        arbitrary but consistent textual patterns.
+      \item To facilitate the tweaking of syntax highlighting.
+        This should include ways to easily access and change the
+        syntax highlighting scheme and dump it to a file (dumpung
+        current highlighting scheme is currently not supported).
+        Should also include changing the color scheme an highlighting scheme incrementally
+        and selectively using the features described in Section~\ref{visual2}.
+      \item For bringing back all the splits after an \tttt{:only}.
+        Also for bringing back the tabs after a \tttt{:tabo}.
+      \item Run Python excepts from file in an IPython shell. 
+        The buffer number of the terminal window should be stored,
+        and than any selected lines should be run on that instance.
+        Mappings should make available all movements to catch the excepts,
+        execution of registers, remain on script or in the IPython shell.
+        Also, terminal mode can use some improvements, such as \ttt{C-O} for
+        one normal mode command.
+      \item For keeping track of the usage and making analysis
+        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}. Usage analytics.
+      \item Colorschemes for true color (gui).
+        The standard colors (e.g. blue, elflord) loose some of the distinctions,
+        e.g. SpellBad tokens are not highlighted on these colorschemes
+        it you have a functioning \tttt{se termguicolors} (standard GUI mode
+        and available in terminals since recent versions of Vim 8, see Section~\ref{visual2}). In~\cite{vimrc} are some lines that make such colorschemes after over
+        \tttt{:colorscheme blue}.
+        This design of color schemes over the standard color schemes might be
+        a very simple and effective way to make new color schemes.
+      \item Opening the editor status listings such as in Section~\ref{}
+        in a quickfix or location window where they might be searched, edited
+        and used as wanted.
+        These windows open in a more-like interface that is very restricted in what
+        they allow (e.g. no search, no persistence).
+    \end{itemize}
+\end{itemize}
+
+\subsection*{Acknowledgments}
+FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
+Vim user community.
+
+\appendix
+\section{Example of usage session}
+I usually begin by opening a file or directory
+with \tttt{\$ vim <filename>}.
+The color scheme is alternated between
+blue and GruvBox with
+\tttt{:colo gruvbox} and \tttt{:colo blue}.
+I open an vertical split and then move
+the window to a new tab using
+\tttt{:vs} and \ttt{C-W\_T}.
+I then search tokens related to
+the enhancements I want to make or
+the knowledge I want to acquire.
+I go back to the previous tab with \tttt{gr}
+and make a global replace with
+\tttt{:\%s/<this>/<that/g}.
+On adding dots to sentences,
+I record in the \tttt{"q} register
+the sequence \tttt{jA.jj},
+using it as a macro 10 times by
+\tttt{10@q}.
+I move to the other tab with \tttt{gt}
+and open a terminal window with \tttt{:term}
+for compiling latex files.
+I start another terminal with \tttt{<C-W> :term}
+for opening the resulting PDFs with evince.
+If any new idea comes to mind and I have time,
+\tttt{\textbackslash\textbackslash s} opens 
+my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
+sources it.
+If there is e.g. code or notes in other projects I am working
+on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
+in Byobu/Tmux sessions and windows.
+
+\section{My vimrc file and usage}
+In this Appendix is my vimrc file.
+Although it has comments, one should
+look for the options that (s)he does
+not understand, as a thorough explanation
+of the file is tedious and out of the scope
+of this document.
+
+In using Vim with my vimrc file~\cite{vimrc},
+I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
+and the tab line with \texttt{\textbackslash\textbackslash T}.
+Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
+The mappings for transitioning through splits and tabs
+are also used constantly.
+
+\section{Example of notes on mappings}\label{notes}
+\texttt{:h index} shows all the default mappings
+while \tttt{:map} shows the user-defined mappings.
+By considering such information, one can make
+useful observations exemplified in this Appendix.
+
+\subsection{Normal mode}
+Every letter and character in the keyboard is used.
+In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
+\ttt{BS} and \ttt{C-H} is the same as h.
+\ttt{C-J} and \ttt{C-N} is the same as j.
+\ttt{C-P} is the same as k.
+Space is same as l.
+\ttt{C-[} and \ttt{Esc} are not used,
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+\ttt{C-\_} not used.
+\tttt{+} is the same as \ttt{CR} and they are both not very useful.
+Del is same as x.
+
+Many \tttt{][} combinations are not used, e.g.
+with \tttt{abhjklfg}.
+The \tttt{\_} command might be used as a more powerful \tttt{\^},
+leaving it free for mappings.
+
+Directions, home, end, page up and down, insert, all have mappings
+in more centraly located keys.
+unbounded: g,z,[]
+
+The \ttt{C-(HJKL)} commands are redundant,
+with the exception of \ttt{C-L} which redraws the screen,
+so it is a reasonable choice to use them to move focus
+of the editor to splits in the \tttt{hjkl} directions.
+
+Many key combinations are available for new mappings through the \tttt{g}
+and \tttt{z}, and \tttt{v} commands.
+They have typical uses, e.g. \tttt{z} for folds,
+spell checking and some movements (mainly when wrap is set).
+
+\subsection{Insert mode}
+All standard char keys are used for entering chars on text.
+\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
+in moving through multiple lines.
+\ttt{C-[} is the same as \ttt{ESC}.
+\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
+\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+other combinations with \ttt{C-\textbackslash} are not used.
+
+\section{My \tttt{:version}}
+I should keep the output of \tttt{:version} executed on the system
+in which I wrote this document in this link:
+\url{}.
+It was compiled with this Makefile~\cite{makefile}
+in a 16.04 LTS Ubuntu Linux,
+and is tagged as 8.0, Included Patches 1-1173.
+
+\section{Key notation and meaning for Vim}
+The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
+mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
+(or \tttt{Meta+x}).
+\ttt{A-X} is the same as \ttt{M-X} and refer to
+Alt or Meta keys.
+\ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x}  and \ttt{C-x} are the same,
+i.e. Vim does not distinguish lower and uppercase 
+letters in \ttt{C-} commands.
+The \ttt{M-x} and \ttt{M-X} (or \ttt{M-S-x} or \ttt{M-S-X})
+are different commands,
+and they are not used by Vim's builtin keys.
+Such \ttt{M-} key combinations are potentially conflicting
+with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
+but are otherwise a large and safe set of combinations
+for one to use.
+
 In summary,
-the usage of the package and plugin might be performed
-through the following actions:
-\begin{itemize}
-  \item Installation of the plugin.
-  \item Tweaking of the syntax file by hand.
-  \item Running the \tttt{makeVimSyntax.py}
-    Python script to generate a new \tttt{tokipona.vim} syntax file
-    according to other settings.
-  \item Write a file with the \tttt{.tokipona} or \tttt{.tp} extension
-    (inside Vim) using Toki Pona words.
-    Reload the highlighting scheme using \tttt{:e} whenever you
-    change the syntax file by hand or through the Python script.
-  \item Access the used highlighted groups with \tttt{:syntax}.
-    Access all the highlighting groups with \tttt{:so
-    \$VIMRUNTIME/syntax/hitest.vim} or \tttt{:hi}.
-    Change the coloring of a set of terms by associating
-    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g.
-    Visual) such as in \tttt{:highlight link tokiponaADJECTIVE Visual}.
-\end{itemize}
-
-The syntax file has an association of TP
-words with highlighting groups.
-It also holds associations of these groups to other groups
-to use their color settings.
-Therefore, one is able to use various color schemes
-in syntax-highlighted TP texts, such as in Figure~\ref{fig:syn}.
-A Vim user might run e.g. \tttt{:colorscheme blue},
-\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox} or
-\tttt{:colorscheme elflord} to see the same text colored
-with different color schemes (association of colors to sets of tokens).
-To interfere directly on the colors chosen,
-\tttt{:highlight Normal guifg=\#00000 guibg=\#0000ff}
-will change the standard foreground (text) color to pure black
-and background to pure blue.
-See \tttt{:h gui-colors} and \tttt{:h highlight}
-for the way you might edit colors directly.
-Useful commands are given as commented lines
-in the end of the syntax file. % TTM
-One might obtain an HTML file with the colored TP text
-by using the \tttt{:TOhtml} command.
-
-% reference to figure of the texts.
-
-\subsubsection{Advanced syntax highlighting considerations}
-Standard guidelines for syntax highlighting
-depend heavily on cultural and use factors
-and have scarce scientific studies~\cite{stack}.
-There are informed projects such as Solarized~\cite{solazired},
-which present solutions for some contexts.
-Strikingly, standard guidelines for syntax highlighting were not found.
-Therefore, we considered current data visualization
-theory~\cite{dv1,dv2,dv3,dv4} to glimpse at the potential guidelines:
-\begin{itemize}
-  \item The use of blue and other high frequency colors (such as
-    violet) for the background.
-  \item The avoidance of blue and other high frequency
-  colors to fill small objects, such as letters and words.
-  \item Explore simplicity and elegance through the use of discrete
-  coloring schemes. Most of the tokens should be preferably of
-  the same color.
-    One might use only a small number of colors to achieve a clean
-    visualization of texts.
-  Also, a power-law distribution of tokens among colors might be
-    well-suited to mimic natural occurring phenomena.
-    (Unfortunately, the TP dictionary classification of words are not in a
-    power-law distribution and might be better described by a half-normal or by
-overlapping normal distributions over the three peaks: $\sim35-58,
-12-18, 1-5 occurrences$).
-  \item Physical stimuli might be related to perceptual stimuli both 
-    through a power-law or an exponential law (respectively known
-    as Steven's law and Weber-Fechner's law).
-    This might be useful e.g. for coloring sets of tokens in
-    a way that their similarity is considered.
-  \item One usually wishes to maximize contrast,
-  although taste and less wearing combinations might
-  also dictate the coloring choices.
-  \item Stipulate axes of parameters with which to set colors.
-  Wavelength or frequency is an obvious axis given the considerations
-  above.
-  Other evident parameters are hue, transparency etc.
-  \item One might consider two types of color schemes: those with a dark background,
-  which are more comfortable at first; and those with a light background,
-  which are usually impressive (and even annoying) at first,
-  but the eye gets used to it and it keeps you more stimulated.
-\item The blue color is specially related to physiological
-  stimulation of the body~\cite{blue,blue2},
-    and programmers often report that a blue background keeps them
-    awake and more concentrated (I also notice such effect,
-    and I have programmed daily for more than 10 years).
-  \item Colors have been associated to enhancements in specific tasks,
-    e.g. blue and red are respectively associated to enhanced
-    creativity and detail-oriented tasks in~\cite{blue}.
-  \item It is important to consider that the user might 
-    stay many hours
-  at a text editor (and we very often do),
-    and that the colors and formats involved are
-  prone to entail a considerable effect in the body and mental
-    activity and thus in the quality of life and work of a writer
-    (e.g. a programmer).
-  \item Ideally, one should have facilities for tuning 
-    the syntax highlighting (e.g. through keyboard shortcuts)
-  as envisioned in~\cite{vim}.
-\end{itemize}
-
-The considerable irrelevance of the morphosyntactic classes as
-described in Section~\ref{syntax} suggests that an appropriate coloring
-of words should
-either distinguish only between particles, prepositions and the rest,
-or consider the syntax to identify the nouns, verbs, adjectives, adverbs,
-prepositions and particles.
-Also, the coloring for poems might be more appealing if considered
-e.g. the counting of syllables, the repeated letters or syllables,
-and the ending syllables.
-Insights might be obtained through the observation of the
-choices made and advocated in software such as text editors
-(e.g. Vim, Emacs, Sublime),
-packages dedicated to syntax highlighting (e.g.
-pigments\footnote{\url{http://pygments.org/docs/}} and 
-highlight.js\footnote{\url{https://highlightjs.org/}},
-and
-other software (e.g. Linux terminals such as Xterm and gnome-terminal).
-Finally, the careful choice of fonts is known to have
-a relevant impact in comfort and productivity~\cite{fonts},
-e.g. sans-sheriff fonts are more promptly read and yield a cleaner
-text than a serif font, and the same applies to a monospaced font
-which is more likely to yield a cleaner text, at least for programming.
-
-One might also 
-consider the mapping the textual structures
-to sound~\cite{mass} (i.e. parametrize the synthesis of sounds e.g. 
-by the counting of specific tokens inside a script, a function or
-class or around a variable of conditional or
-loop), which might be called ``syntax sonification''.
-
-\subsection{Toki Pona Wordnet}\label{wn}
-For the achievement of a first Toki Pona Wordnet,
-each of the TP words in the dictionary
-was related to Wordnet synsets~\cite{wordnet}
-through the English lemmas.
-The most canonical (i.e. Princeton) Wordnet only contains nouns,
-adjectives, verbs, and adverbs.
-Thus, particles and prepositions were not considered.
-Numbers were considered adjectives.
-Words presented as adjectives in the dictionary
-were considered both as adjectives and adverbs.~\cite{wordnet}
-
-The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
-makes available such tentative TP Wordnets in the simplest form:
-the TP words are keys in a dictionary that returns the
-corresponding synsets.
-Three of such dictionaries are implemented:
-one where all synsets related to the lemma is
-returned (the version most consistent
-with the expositions in Section~\ref{basics}),
-with 4,052 synsets;
-another as such but excluding the prepositions,
-with 3,961 synsets;
-a last one that returns only the synsets
-registered in Wordnet with the same POS tag as the lemma
-is in the official dictionary,
-with 2,493 synsets. 
-Another immediate possibility, not implemented,
-would be to relate each Toki Pona word to the synsets
-simultaneously associated to all the English
-words bounded by a semicolon in the dictionary.
-Such collection of synsets, and its relation to the whole
-Wordnet
-(the total number of synsets in Wordnet is $117,659$),
-might guide creation of other conlangs
-(e.g. one might seek to use lemmas related to synsets that
-are very far apart, or that has the most complete neighborhood
-possible).
-% Total number of synsets counting neighbors is Y. TTM
-% Depth of TP synsets is .
-
-It is worth mentioning that TP words are very general
-and each of them might mean many things.
-Thus, the words are not very easily related e.g. by hyponymy
-or meronymy.
-Exceptions: jan (people) is a hyponym of soweli (mammal);
-kili is an hyponym of kasi;
-walo, pimeja, jelo, loje, laso
-are hyponyms of kule.
-There is at least one caveat:
-if a particular meaning of each word is chosen,
-then there might be many other of such relations,
-e.g. 
-luka (as hand) is a meronym of sijelo (as body).
-
-\section{Conclusions and further work}\label{conc}
-This document presented a potentially novel description of Toki Pona (TP)
-language in Section~\ref{basics}, and innovative and useful software routines for dealing with
-the language in Section~\ref{hacks}.
-As a minimalist conlang, TP is very convenient for cognitive
-experiments
-through linguistic semantics and
-for devising tools for analysis, creation and visualization.
-Other conlangs, and even non-planned languages might benefit from
-TP and the content presented here in many ways, e.g. one might
-synthesize TP texts and translate them as convenient,
-adapt the routines to obtain statistics of the vocabulary,
-or take advantage of the language description to devise new conlangs
-or stilistic guidelines for any language.
-
-Possible next steps are numerous:
-\begin{itemize}
-  \item Further develop the text synthesis facilities described in
-    Section~\ref{synth} might be enhanced by further employing e.g.
-    rhythm, meter, rhyme and form techniques~\cite{wikiPoetry}.
-  \item Make a syntax highlighting that colors the tokens with respect
-    to the syntatic position and function, and not in a fixed manner
-    as is implemented and exposed in Section~\ref{shigh}.
-    This might be implemented by using n-grams and
-    further techniques from Natural Language Processing
-  \item Enhance and study the TP Wordnet described in Section~\ref{wn}:
-    implement a NLTK-like Wordnet interface; check if each synset
-    is correctly associated to the TP words; seek further synsets
-    that were not found by the English lemmas in the official
-    dictionary; implement the most restricted version described in
-    Section~\ref{wn}, which relates each TP word only to the synsets
-    that are related to all the English words not separated by a
-    semicolon; find out the neighborhood of the synsets of the TP
-    Wordnet in the English Wordnet; explore the possibilities for
-    devising new conlangs by means of criteria related to Wordnet.
-%   \item Translators.
-%    This Toki Pona Wordnet would be useful e.g. for naive translators
-%    of words and text.
-%    Enhancements of such translators would use POS tagging and templates
-%    in English and Toki Pona and/or corpus and further statistical methods.
-%    An efficiency of Toki Pona can be studied in terms of the coverage of
-%    the Wordnet cloud.
-  \item Conlanging: use Wordnet for making new conlangs;
-    use insights derived from TP.
-    Maybe create conlangs for specific uses:
-    describing data, programs,
-    scientific writings, creative writing
-    (exploring the thinking process).
-    A conlang might have different modes that might 
-    be specified by a section header or tag. 
-  \item Understand how the corpus was gathered in \url{tokipona.net}.
-  \item Know about previously existing words that were used for TP
-    (e.g. suno and suwi might come from sun and sweet),
-    and about the reasons that led Sonja (and maybe other people)
-    to choose the 14 letters and the syllable structure.
-    This might require a dedicated communication with the
-    speaker community and the documentation keepers.
-  \item Corpus-based analysis.
-  \item Publication of original TP texts and translations.
-  \item Make an article written in TP
-    (extending Section~\ref{finalToki} which might be the
-    first section of a scientific document written in TP).
-    The outline might be:
-    a summary in both English and Toki Pona, for
-    facilitating the acquisition of context,
-    and a article in a canonical structure
-    (introduction and related work, materials and methods,
-    results and discussion, conclusions and further work),
-    with an Appendix explaining the content and context in English.
-    I first conceived something around complexity, statistics, physics, or computer science.
-    But one possibility is to write about linguistics, philosophy, literature
-    or psychology with the partners I write in English.
-    I can start a draft,
-    they might learn the language in a few hours or weeks
-    (with or without my help),
-    to contribute, and we can write a short paper.
-  \item Enhance the synthesis of text to yield better contextualized text
-    and stylistic traces.
-    Give the user the ability to choose the sentences:
-    generate a sentence randomly according to
-    previously written or given text,
-    to some rules input by the user,
-    to the package and the language guidelines.
-    Should output to the screen and asks to keep and discard.
-  \item Make TP variations where each TP word is replaced by a word in
-    a language with a large speaker population. 
-For example,
-suno might be sun in an English variant,
-and sol in a Portuguese variant.
-A naive version might be obtained through
-choosing the first word in the description of
-an official vocabulary, as they already exist in English and French.
-  \item We are particularly interested in software tools,
-    and one possibility is to use a conlang to describe software
-    routines.
-    In this sense, it seems profitable
-    to understand what `non-ambiguous' means in Lojban,
-    and in what sense one is able to compile and parse Lojban.
-    Such study will probably entail considerations about
-    parsing TP.
-%  \item The plugin comes with the :TokiStation command, which opens a
-    %  window with the files: ijositelen.tokipona, makevimSyntax.py,
-    %  Highlight Test (created by hitest.vim above),
-    %  syntax/tokipona.vim.  Another tab with shells: python to run
-    %  over the makeVimSyntax.py and make new syntax/tokipona.vim
-    %  files. Another with Readme, and PDF documentation. Another with
-    %  an IPython. Imported tokipona. It resets completely upon
-    %  command :TokiClose, closing all created windows. TTM
-\end{itemize}
-
-\subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3);
-Toki Pona, Python and Vim authors, developers and literature maintainers;
-Toki Pona user and speaker communities. 
-I thank Mario Alzate for introducing me to Toki Pona,
-and Silverio Guazzelli Donatti for the stimulus that resulted
-from the usage of Toki Pona for chatting.
-
-\appendix
-\section{My usage of Toki Pona}\label{mytoki}
-I use the standard sounds, but often use [z] for s.
-I often translate texts to Toki Pona (e.g. biblical excepts)
-and create new texts as poems and short stories.
-Most of them are in~\cite{tokisona}
-and/or published in the TP Facebook groups.
-I omit the li particle after subjects sina and mi,
-in accordance with the norm,
-but sometimes I use them when there are many predicates.
-E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
-In such cases, the first li is sometimes omitted.
-Also, sometimes I use li after mi and sina where I find
-that there is unwanted ambiguity, e.g.
-sina moku pona 
-which 
-might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.
-
-Names are by default transliterated,
-but I advocate that, as in other languages,
-names might be used as they are in the
-correspondent mother tongue.
-E.g. the name Erdös is used in
-English and Portuguese although the standard
-alphabet does not contain ö in such languages.
-I also tend to legitimate the use of English (or German) words
-in TP texts if it is the case,
-as happens often in scientific writing
-(kernel is a German word used in English,
-webpage and software are English words used in Portuguese).
-
-Proposed notations for numbers seem numerous.
-I tend to think that one might indicate if two numbers
-are being multiplied (pi) or are in different scales
-(such as in decimal or binary notations).
-For example, luka (pi) two might mean 52.
-Most importantly,
-52 is a reasonable notation for a simple language.
-E.g.
-'mi jo e jan sama nanpa 12',
-or
-'ona li lon e soweli 27'.
-
-Avoiding needless words...
-I've been avoiding e ni: and using only :.
-I've been omitting the subject (sometimes also the li)
-if it is the
-same as in the last sentence.
-I tend to group these words:
-noka and anpa; luka, suno, sike and lawa; pali and pana.
-These clusters of meaning hint me that an even simpler
-language is possible and maybe more optimized to the
-core goals of TP.
-
-Also, ambiguities introduced by omission of li
-and absence of a token to denote preposition,
-suggest to me the possibility of a syntax that is always uniquely
-parsed (or at least less ambiguous).
-
-I have been inclined to conceive a language with the same syntax of TP:
-<sentence> la <sentence = subject + predicate + object (each term with
-a possible preposition complement)>
-but always using particles to bound the sentence sections,
-repeating them when the function is performed in the smallest scale:
-'toki pi toki pona pi jan sona' meaning
-language in toki pona and of intellectuals vs
-'toki pi toki pona pi pi jan sona'
-meaning language in [toki pona of intelectuals].
-About the keyword for initiating a preposition complement 
-might be 'a', but it is already taken in TP.
-Maybe use 'a' before a preposition and use 'ha' or 'he'
-instead of TP a.
-In such a setting,
-one might enable pi li e a (concept-qualifier, subject-predicate,
-predicate-object, concept-preposition)
-in any slot of the template:
-<sentence> la <sentence = subject + predicate + object (each term with
-a possible preposition complement)>
-Between sentences one would use:
-la pi, la li, la e, la a, and la la.
-This is a very fractal conlang proposal.
-Maybe also have a way to discern between a noun,
-a qualifier (adjectives/adverbs) and a verb,
-and accept any of them for the subject, predicate, object, preposition,
-slots. Or assume a part of speech as default for a slot or for the
-first word in a slot.
-
-\section{Final words in Toki Pona}\label{ftp}
-toki li nasin e lawa.
-li nasin tawa (pi) toki insa.
-
-% jan li ken toki li ken pali pona e toki pona.
-toki pona li pona e nasin tan ni:
-ona li jo e nimi nanpa lili.
-ona li pona.
-pona kepeken weka 'p' li ona, a.
-
-o taso la toki pona li kalama li lukin pona.
-sitelen en nasin li open e sitelen
-suli\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
-li sona. li nasin e toki e sona e lawa e lon.
-
-toki ni li wawa tawa jan mute nasin en toki.
-wawa tawa toki pi jan sona.
-pi ilo nanpa en nanpa nasin.
-taso tawa toki sona a.
-sitelen sona, sitelen musi.
-
-ilo lon Poki~\ref{hacks} li pana e sitelen
-e sona tan sitelen,
-en nasin.
-Poki~\ref{basics} li pana e nasin pi toki pona.
-e sitelen tawa kama sona.
-
-mi wile pali e sitelen lon toki pona.
-sitelen lon nasin, sitelen sona,
-sitelen musi.
-ante e toki pona
-la ona li nasa.
-taso nasa li pona mute.
-li pona mute tawa lawa,
-tawa kama sona, tawa sitelen e toki.
-ante la toki pona o.
-toki e toki pona tawa sina.
-kama sona e toki pona sina.
-
-o pona tawa jan pi toki pona.
-tawa jan Sonja, Birns-Sprage, Kipo, Pije,
-Siwejo, Malija, Tepan, jan kulupu mute.
+one might choose for mappings to overwrite defaults key combinations,
+use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
+normal commands, \ttt{C-} and \ttt{S-} commands, or any \ttt{M-} and
+\ttt{M-S-} command), or
+create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
+More information about key notation in \tttt{:h key-notation}.
+Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings
+in \tttt{:map}.
 
@@
 \fontsize{11}{0}\selectfont
-\bibitem{wordnet}
-  Fellbaum, C. (1998). WordNet. John Wiley \& Sons, Inc..
-\bibitem{corpus}
-    Henry, J. lipu pi toki pona pi jan Jakopo. Available at         
-    \url{http://jimhenry.conlang.org/conlang/tokipona/tokipona.htm} 
-\bibitem{vim}
-    Fabbri, R. An anthropological account of the Vim editor:
-features and tweaks after 10 years of usage. Available at         
-    \url{https://github.com/ttm/vim/raw/master/article/article.pdf} 
-\bibitem{sapWho}
-  O'Neill, S. P. (2015). Sapir–Whorf Hypothesis. The International
-    Encyclopedia of Language and Social Interaction.
-\bibitem{gdoc}
-  Toki Pona Community (2017). toki pona central hub preparation.
-    Available at
-    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
+\bibitem{anPh}
+	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
+	Available at \url{https://youtu.be/oeOKYc3-nbM}
+\bibitem{anPh2}
+	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
+\bibitem{ouroWiki}
+  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
+\bibitem{vimrc}
+	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
+\bibitem{ex}
+  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
 \bibitem{tokipona}
@@
     ISBN-10: 0978292308, ISBN-13: 978-0978292300.
-\bibitem{tp3}
-  Blahuš, M. (2011). Toki pona–eine minimalistische Plansprache.
-    Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
-\bibitem{kama}
-  Knight, B. (2017) o kama sona e toki pona! Available at:
-    \url{http://tokipona.net/tp/janpije/okamasona.php}
-\bibitem{Wikipesija}
-  Toki Pona community (2017). Wikipesija. Available at:
-    \url{http://tokipona.wikia.com}
-\bibitem{conlanWikip}
-  Constructed language.
-(2017, November 1).
-In Wikipedia, The Free Encyclopedia.
-Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
-\bibitem{wikiPoetry}
-Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
-    Retrieved 19:46, December 3, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
-\bibitem{stack}
-  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
-    Software Engineering in Stack Exchange. Available at
-    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
-\bibitem{solarized}
-  Schoonover, E. Solarized color scheme. Available at
-  \url{http://ethanschoonover.com/solarized}
-\bibitem{tp4}
-  Schneider, S. (2017). nasin toki pi toki pona. Available at
-  \url{https://github.com/stefichjo/toki-pona}
-\bibitem{interview}
-  Lang, S. and Broholm, K. AFP 20 – Sonja Lang: Toki Pona, Conlanging
-    and the meaning of life (Sonja Lang interview). Available at
-    \url{https://actualfluency.com/afp-20-sonja-lang-toki-pona-conlanging-meaning-life/}
-\bibitem{memrise}
-  Toki Pona Community. (2017). Various Memrise courses on Toki Pona.
-    \url{https://www.memrise.com/courses/english/?q=toki+pona}
-\bibitem{wikiToki}
-  Toki Pona. (2017, December 11). In Wikipedia, The Free Encyclopedia.
-    Retrieved 22:48, December 14, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
-\bibitem{artlanWikip}
-  Artificial language. (2017, November 26). In Wikipedia, The Free
-  Encyclopedia. Retrieved 23:00, December 14, 2017, from
-  \url{https://en.wikipedia.org/w/index.php?title=Artificial_language&oldid=812127746}
-\bibitem{akesiWawa}
-  Fabbri, R. (2017, November 6). o awen, akesi wawa.
-  In the group toki pona taso. Available at
-  \url{https://www.facebook.com/groups/tokiponataso/permalink/1895942817391318/}
-\bibitem{janKipo}
-  Clifford, J. (2017, December 15). Considerations about Toki Pona
-  phrases.
-  In the Facebook group 'toki pona'. Available at
-  \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
-\bibitem{tpNetSH}
-  Knight, B. (2017). Online gadget for syntax highlighting Toki Pona
-  texts. . Available at
-  \url{http://tokipona.net/tp/DisplayText.aspx}
-\bibitem{blue}
-  Mehta, R., \& Zhu, R. J. (2009). Blue or red? Exploring the effect of
-  color on cognitive task performances. Science, 323(5918), 1226-1229.
-\bibitem{blue2}
-  Viola, A. U., James, L. M., Schlangen, L. J., \& Dijk, D. J. (2008).
-  Blue-enriched white light in the workplace improves self-reported
-  alertness, performance and sleep quality. Scandinavian journal of
-  work, environment \& health, 297-306.
-\bibitem{fonts}
-  Spolsky, A. J. (2008). User interface design for programmers.
-  Apress.
-\bibitem{mass}
-	Fabbri, R., Vieira, V.
-, Pessotti, A.
-C.
-S., Corrêa, D.
-C., Oliveira Jr, O.
-N.
-O.
-Musical elements in the discrete-time representation of sound (2017).
-arXiv preprint arXiv:1412.6853.
-Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
-\bibitem{tokisona}
-  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
-  Available at \url{http://tokisona.github.io/}
+\bibitem{aa1}
+	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G.
+(2013).
+AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology.
+RESI.
+From \url{https://arxiv.org/abs/1604.08255}
+\bibitem{aa2}
+	Fabbri, R. (2017).
+The Algorithmic-Autoregulation (AA) Methodology and Software:
+a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
+\bibitem{makefile}
+	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
 \end{thebibliography}

### 2017-12-16 article/article.tex
detail about 24-bit true color

--- 2017-12-15 article/article.tex
@@
 \begin{document}
-\title{Basic concepts and tools for the Toki Pona minimalist language:
-Wordnet synsets; analysis, synthesis and syntax highlighting of texts}
+\title{Basic concepts and tools for the Toki Pona minimalist and
+constructed language:
+Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
 % Anthropological computation
@@
 and different solutions are described and implemented
-  using 24bit true colors in the well-established Vim text editor~\cite{vim}.
+in the well-established Vim text editor~\cite{vim}.
 The tentative TP Wordnet is made available in three forms that
@@
 prepositions (kepeken, lon, tan)
-are usable in any position in noun and verb phrases,
+are usable after any noun or verb phrase,
 (i.e. 107 words discarding synonyms).
@@
 or nouns or verbs or adverbs, etc, and might follow
-anu noun or verb phrase,
+any noun or verb phrase,
 they are one of the three main sources of ambiguity:
@@
     variation due to its simplicity, I would advocate for
-    using en and anu wherever there is no ambiguity.
+    using en and anu/en wherever there is no ambiguity.
     E.g.: 'mi tawa mute anu moku lon tenpo lili'
@@
 
-\subsubsection{Recognizing POS tags by speaker and machine}
+\subsubsection{Recognizing POS tags by speaker and machine}\label{sec:rec}
 One might perform a POS tagging by following these rules:
@@
     Starting a sentence if it is complete,
-    after an 'e', after a pi, and after a preposition.
+    after an 'e', after a pi.
     The predicate is often a noun
@@
     might be a noun or a verb.
-  \item After a preposition,
-    there is a noun phrase.
+  \item After a preposition, one might consider that what follows is 1) always
+    a noun, or that 2) it can be a noun, adjective or verb,
+    or that 3) it is a hybrid of noun, verb and adjective.~\cite{janKipo}
   \item Notice that there is ambiguity in the structure
@@
 as being a hybrid of noun, verb and adjective.
-  \item The prepositional phrase might also be taken to be a hybrid
-    of noun, verb and adjective.
 \end{itemize}
@@
 
-\subsubsection{Additional notes}
+\subsubsection{Additional notes}\label{sec:add}
 John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
-states that there are not noun, verb or prepositional
+states that there are not noun, verb, adjective (or even prepositional)
 phrases in TP, but only phrases in a structure
@@
 The only synonyms on Toki Pona are:
-a and kin; lukin and oko; sin or kamako;
+a and kin; lukin and oko; sin or namako;
 ale or ali.
@@
 Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
-(verb), object (noun) and prepositional (noun or verb) phrases of a sentence,
+(verb), object (noun) and prepositional (see Sections~\ref{sec:rec} and~\ref{sec:add}) phrases of a sentence,
 and assume that the sentence has at most one prepositional phrase,
@@
 I have been inclined to conceive a language with the same syntax of TP:
-<sentence> la <sentence = subject + predicate + object (each term with
-a possible preposition complement)>
+'sentence' la 'sentence = subject + predicate + object (each term with
+a possible preposition complement).
 but always using particles to bound the sentence sections,
@@
 in any slot of the template:
-<sentence> la <sentence = subject + predicate + object (each term with
-a possible preposition complement)>
+'sentence' la 'sentence = subject + predicate + object (each term with
+a possible preposition complement)'
 Between sentences one would use:
@@
 sitelen en nasin li open e sitelen
-suli\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
+suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
 li sona. li nasin e toki e sona e lawa e lon.
@@
 \bibitem{janKipo}
-  Clifford, J. (2017, December 15). Considerations about Toki Pona
+  Clifford, J. et al. (2017, December 15). Considerations about Toki Pona
   phrases.

### 2017-12-16 article/pos.tex
detail about 24-bit true color

@@
             by choosing the class in the dictionary respecting the
-            precedence order: PRE, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
-\begin{tabular}{| l | c | c |}\hline
+            precedence order: PRE-VERB, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
+\begin{tabular}{ l | c c }
 POS & All  & Chosen \\\hline
-NOUN & 58  & 49 \\\hline
-ADJECTIVE & 40  & 34 \\\hline
-VERB & 15  & 13 \\\hline
-PARTICLE & 12  & 12 \\\hline
-PRE & 6  & 6 \\\hline
-PREPOSITION & 5  & 5 \\\hline
+NOUN & 58  & 49 \\
+ADJECTIVE & 40  & 34 \\
+VERB & 15  & 13 \\
+PARTICLE & 12  & 12 \\
+PRE-VERB & 6  & 6 \\
+PREPOSITION & 5  & 5 \\
 NUMBER & 4  & 1 \\\hline
-total & 140  & 120 \\\hline
+total & 140  & 120 \\
 \end{tabular}\end{center}

### 2017-12-16 article/article.tex
detail about 24-bit true color

@@
 In summary, this text holds potentially novel conceptualizations about,
-and tools and results in analyzing, synthesizing or syntax highlighting the
+and tools and results in analyzing, synthesizing and syntax highlighting the
 TP language.

### 2017-12-18 article/article.tex
adiantado ate o final dos modos, inicio do Netrw

--- 2017-12-15 article/article.tex
@@
 \begin{document}
-\title{An anthropological account of the Vim editor:\\
+\title{An anthropological account of the Vim text editor:\\
 features and tweaks after 10 years of usage}
@@
 \begin{abstract}
-The Vim editor is very rich in capabilities
+The Vim text editor is very rich in capabilities
 and thus complex.
-This article is a description of the Vim text editor
-and a set of considerations on its usage and design.
-It is the result of more than ten years of experience
-with Vim for writing and editing various types of documents,
-e.g.:
-Python, C++, JavaScript, ChucK, programs;
-and \LaTeX, Markdown, HTML, RDF, Make and other markup files;
+This article is a description of Vim
+and a set of considerations about its usage and design.
+It results from more than ten years of experience
+in using Vim for writing and editing various types of documents,
+e.g. Python, C++, JavaScript, ChucK programs;
+\LaTeX, Markdown, HTML, RDF, Make and other markup files;  % TTM
 binary files.
-It is said that it takes about ten years to master (or start mastering)
+It is commonplace, in the Vim users and developers communities,
+to say that it takes about ten years to master (or start mastering)
 this text editor, and I find that other experienced users
@@
 set of features.
-Therefore, this document exposes my insights in order
+Therefore, this document exposes my understandings in order
 to confront my usage with that of other Vim users.
@@
 Upon feedback, and maturing of my Vim usage,
-this document might be enhanced or receive additional
-material.
+this document might be enhanced and expanded.
 \end{abstract}
@@
 Vim is a very complex editor,
-considered by the Linux community to be
+most often considered to be
 matched only by Emacs.
@@
 and have been developed for decades.
-This document describes the Vim editor
-and proposes a set of enhancements of the user
-experience through simple tweaks.
-The editor has a very mature documentation
-and the contents herein presented is a
+Vim is very useful because:
+\begin{itemize}
+  \item it is meant to be a plain text (e.g. ASCII, UTF-8) editor
+  and does not (by standard) insert special characters (e.g. for
+    formatting or with binary instructions).
+  \item It has a powerful architecture and set of commands.
+  \item It is highly configurable and most often the users
+  hold a set of commands for standard settings and routines kept in
+    the vimrc~\cite{vimrc} and other configuration files.
+  \item It has been used for more than 25 years, is based on
+    established technologies, and is an established text editor.
+    Thus, it has a considerable and seasoned user
+    base, high quality (both official and unofficial) documentations, 
+    and countless publicly available scripts, most often in the form of
+    plugins.
+\end{itemize}
+
+This document describes the Vim text editor
+and proposes a set of sharpening of the user
+experience through simple tweaks
+and utilization strategies.
+The contents herein presented is a
 report on the overall understandings I
-have of Vim after a little bit more
-of ten years using it.
+have of Vim after a bit more
+than ten years using it,
+resorting to the (very mature) official documentation
+whenever possible.
 The purposes of this document are:
 \begin{itemize}
-  \item to help new users in grasping Vim essentials
+  \item to help new users grasp Vim essentials
   and convenient practices.
   \item To attain a sound overall description of the editor.
-  \item To record the view of the editor that
+  \item To record the comprehension about the editor that
   a user (me) has after $10$ years of usage.
@@
 
-Advanced users might just skim throughj
-Section~\ref{basics} and consider more carefully
+Advanced users might just skim through
+Section~\ref{basics}, where standard capabilities of
+the editor should become clear, and consider more carefully
 Section~\ref{namespaces}.
-The concluding remarks and proposed enhancements
-might also deserve some attention.
-
-I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
-I used Vim for writing and editing computer code (Python, Javascript, C++, ChucK, bash, etc), markup languages (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
-Eventually, I edited database files and other types os files.
-With Vim, I mostly write software (web and scientific),
-music, poems and short stories.
-It is very useful because:
-\begin{itemize}
-  \item it is meant to be a plain text (e.g. ascii, utf8) editor
-  and does not (by standard) insert special charaters (e.g. for formating, with binary instructions).
-  \item It has a powerful architecture and set of commands.
-  \item It is highly configurable and most often the users
-  have a set of commands for standard settings and hacks kept in the vimrc and other configuration files.
-\end{itemize}
-
-The standard capabilities of the editor
-should become clear in Section~\ref{basics}.
-Vim is constantly evolving and there are many plugins,
-some of them very popular for both general and specific
-users.
-Accordingly, there are many possible enhancements,
-and Section~\ref{conc} report the most prominent of them
-for me and potential workarounds made available as plugins.
-
+The concluding remarks and proposed enhancements in Section~\ref{conc}
+may also deserve some attention because
+Vim is constantly evolving
+and there are many possible enhancements
+(often made available e.g. as plugins).
+
+\subsection{Further remarks about this document}
 This document is written is a DRY KISS
@@
 references in help files and a nice vimrc.
-Therefore the following content should be kept
+Therefore the following content is kept
 as uncomplicated and original as possible.
@@
 there is an anthropological component
-which is evident in the use of the first person
-in sentences.
-This can be understood as anthropological computation
-or physics~\cite{anPh,anPh2} and observed to help
+which is evident in the occasional use of the first person.
+This can be understood as anthropological computer science~\cite{anPh,anPh2}
+and considered to help
 in the technological groundwork of the civil society.
+Accordingly, here are some notes about my experience:
+I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
+I used Vim for writing and editing computer code 
+(Python, Javascript, C++, ChucK, bash, etc), markup languages  % TTM
+(HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
+Eventually, I edited database files and other types of files.
+Within Vim, I mostly write (web and scientific) software,
+scientific articles, music, poems and short stories.
 
@@
 Vim was first released publicly in 1991.
-It is a cross-platform GNU licensed free and open source extended clone of Bill Joy's vi text editor.
+It is a cross-platform GNU licensed free and open source
+extended clone of Bill Joy's vi text editor.
+Vi was written in 1976 as a hard link to ex: a shorthand
+to start ex in visual mode, i.e. vi is ex.
 Vim's development is coordinated (and performed) since the beginning
 mainly by Bram Moolenaar.
-Today, current bleeding-edge version is 8.0.1257.
+Today, current bleeding-edge version is 8.0.1401.
 I found no explicit stable, alpha or beta versions.
@@
 (this might be the first one), although there are
-books, software and third-party documentation on the web.
+books, software and third-party documentation on the
+web.~\cite{wikiVim}
 
@@
 Vim's interface is text-based.
-In the GUI mode (gVim),
+In the GUI (gVim),
 there are convenient menus and toolbars
@@
 command-line mode.
-This section can be thought as a tutorial
+This section may be considered a tutorial
 that focuses on the namespaces, i.e. sets of tokens
@@
 One should see Appendix~\ref{keys}
-to understand the notation vim uses.
+to understand the notation Vim uses.
 
 \subsection{The bare minimum}\label{minimum}
-You open a file with Vim by executing
+You open a file at Vim startup by executing
 the command: \texttt{vim <filename>}.
@@
 Go back to normal mode by pressing
-\texttt{<ESC>} or \texttt{<C-C>}.
+\texttt{<ESC>}, \ttt{C-$[$} or \texttt{<C-C>}.
 You save the file by typing \texttt{:w<CR>},
@@
 \subsection{Vim help}
-You can find help for vim in various places.
+Help on using Vim is found in various places.
 The standard resource is the Vim help files.
@@
 which is considerably lengthy and complex
-and is usually not read by users for a few years.
+and is usually not read by users before a few years
+of experience.
 In learning Vim, one
@@
   \item ``Vim Adventures'' is an online RPG game for practicing
-  and memorizing Vim commands while having fun.
-  This game is quite famous among Vim users.
+  and memorizing Vim commands. This game is quite famous among Vim users.
   \item There are official and semi-official Vim sites e.g.:
@@
   One often finds these links through a search engine.
-  \item You can find many videos about Vim.
+  \item Many videos about Vim are publicly available.
   One traditional site is \url{http://vimcasts.org},
-  but you might find them by a search engine (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+  but you might find them using a search engine
+  (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
 \end{itemize}
@@
 text and writing alters text and writing.
-You have namespaces where tokens have scalar
-or complex values.
+What you end with is a collection of namespaces where tokens have scalar
+or arbitrarily complex values.
 
 \subsubsection{Commands and mappings}
-There are commands, typing sequences which trigger automated actions,
+There are commands, i.e. typing sequences which trigger automated actions,
 for each mode:
@@
 		Some keys expect a second key,
-    and have combinations not used (available for new mappings),
+    and have combinations not used (thus available for new mappings),
 		specially the z and g.
@@
 As there are colon commands that execute commands in other
-modes, e.g. \tttt{:normal ?\^def },
+modes, e.g. \tttt{:normal ?\textasciicircum def},
 the \tttt{:execute} is a way to build commands in any mode,
 e.g.
-\tttt{:execute 'normal' tabn 'gt'}.
+\\\tttt{:execute 'normal i' . string(atan(bufnr('\%')))}.
 
@@
 	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
+    More on register in Section~\ref{state} and scattered in this
+    document.
 	\item Internal variables are created with let and preferably have a prefix:
@@
 \end{itemize}
-You can echo any of such variables or use in expressions.
+You can echo any of such variables or use them in expressions.
 Notice that you will only be able to echo a \tttt{b:} variable inside
@@
 \begin{itemize}
-	\item The entered commands are accessed through \tttt{:hist a}. 
-    The tokens a / s : can be used for specific types of commands, such
+	\item All the entered commands are accessed through \tttt{:hist a}. 
+    The tokens ``\tttt{a / e : i d}'' may be used for specific types of commands, such
 		as search and colon commands.
@@
 	\item Tabs list can be reached through \tttt{:tabs}.
-		It is usual both to show and hide the tabs bar (mapping in~\cite{vimrc}).
+		It is usual both to show and hide the tabs bar (mapping
+    in~\cite{vimrc}, discussion in Section~\ref{visual1}).
 	\item Jumps are available through \tttt{:jumps}.
-		One positions the cursor at each jump through \ttt{C-o} and \ttt{C-i}.
-  \item Registers are available through \tttt{:reg} command, as variables and through shortcuts in different modes.
+		One positions the cursor at each jump through \ttt{C-O} and \ttt{C-I}.
+  \item Registers are available through the \tttt{:reg} command, as variables and through shortcuts in different modes.
     They also keep track or your copy, edition and deletion and are promptly defined
@@
     might follow the same pattern, but use another \tttt{:h event} and
-    monitor register \tttt{""} (maybe also \tttt{"0}).
-	\item An undo list can be accessed with \tttt{:changes}.
+    monitor register \tttt{""} (maybe also monitor \tttt{"0}).
+	\item An undo list can be accessed through \tttt{:changes}.
 	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
@@
 		and then using \ttt{C-]} to
-		go to the position of tag under cursor, and \tttt{:tabe tags} to open
+		go to the position of tag under cursor, and \tttt{:vs tags} to open
 		the tags file.
@@
 A file with information about the state of the editor
-can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim},
-and this script might be consulted because it has
+can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
+Also, this script might be examined because it has
 a collection of commands to access various settings of Vim.
@@
 I would specially highlight the \tttt{:syntax} command because
-it displays the tokens and related type of meanings when run
+it displays token groups and their meanings when run
 inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
@@
 \subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
-
-You can keep track of the editor state though commands and persistent visual
+You can keep track of the editor state though commands, as stated
+above.
+Also, one might rely on persistent visual
 cues, specially the tabs bar, the status line, and the line
 reserved for the command-line.
+A good strategy I find is to have selective visual cues of the state to make persistent or hide
+and a mapping to toggle each of them.
+Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
+status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
+I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
+Numbering is always there, I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
+Instead of keeping the status bar, I use \ttt{C-G} to know about the
+file and \ttt{gC-G} to know more and rarely.
+It seems not possible to remove the statusbar between horizontal splits.
+After asking in online forums and experimenting, I realized that it seems
+reasonable to keep at least one line dividing the windows, so
+if it comes to it, I just \tttt{set statusline=-}.
+Unfortunately, as far as I could dig, one will need to alter
+the Vim's source code to enable a horizontal split without losing a line.
+For me, it would be ideal for this feature to have a visual cue of the
+first and/or last line of the windows in the lines-number
+column, or complete the spaces and empty chars with \$\$\$\$ or so.
+
+%%%%%%%%%%%% TTM
+\subsubsection{On the persistence of the editor state}
 For state persistence, one might keep an undo file for each file as
@@
 editor's state, with the opened windows, tabs, buffers, etc.
-I use the mappings in~\cite{vimrc} because they keep the sessions
+The mappings in~\cite{vimrc} keep the sessions
 in a reasonable directory and makes it easier to remember and tweak
@@
 This entails a strategy to deal with Vim that is similar to
-the use of Byobu/Tmux/Screen.
+the use of
+Byobu/Tmux/Screen\footnote{\url{https://en.wikipedia.org/wiki/Byobu_(software)},
+\url{https://github.com/tmux/tmux/wiki},
+\url{https://www.gnu.org/software/screen/}},
+because one can rely on restoring the state of the windows.
 The main limitation I found to this approach is that
@@
 you get dummy empty windows for them and an error message.
-Screen and Tmux are the most popular terminal multiplexers.
-Byobu, built on top of them, has awesome keyboard shortcuts
-for managing sessions, windows and splits of terminals.
-Byoby/Screen/Tmux keep the state for future load.
-
-Because browsing the interface in Vim is fast,
-and it favors copy and paste of text,
-I tend to keeep a tab with some terminals:
-one with an IPython shell, another two for compiling latex
-and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
-One enters normal mode in terminal with \tttt{C-W N}
-and it is very corfortable to copy and browse the bash history of the
-terminal.
-I found the mappings on~\cite{vimrc} very helpful for directing
-editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
-which I make available across terminal-job and normal modes.
-But I've been thinking on using \ttt{C-} commands also
-for tabs (not only for splits).
-
-A good strategy I find is to have selective visual cues of the state to make persistent or hide
-and a mapping to toggle each of them.
-Now I toggle byobu/screen/tmux bar with \ttt{F5},
-status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
-I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
-Numbering is always there.
-I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
-Instead of keeping the status bar, I use \ttt{C-g} to know about the
-file and \ttt{gC-g} to know more and rarily.
-It seems not possible to remove the statusbar between horizontal splits.
-After asking around and experimenting, I realized that it seems
-reasonable to keep at least one line dividing the windows, so
-if it comes to it, I just \tttt{set statusline=-}.
-Unfortunatelly, as far as I could dig, one will need to enter
-Vim code to enable a horizontal split without losing a line.
-My ideal of this feature would be to have a visual cue of the
-first and/or last line of the windows in the lines-number
-column, or complete the spaces and empty chars with \$\$\$\$ or so.
-
-Autocommands are the standard way to define event-trigered routines in Vim.
+More about Terminal-Job mode in Section~\ref{terminal}.
+
+Autocommands are the standard way to define event-triggered routines in Vim.
 These are often related to particular file types, but are also often
 in defining the automated behavior of Vim.
-In~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts
-in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
+If the autocommands are placed inside a configuration file (e.g. the
+vimrc), the automated behavior is persistent across Vim instances.
+E.g. in~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
 
 \subsection{Using Vim's modes}\label{modes}
-Vim has some basic and fully implemented modes of usage:
+These are the  basic and fully implemented modes of usage in Vim:
 \begin{itemize}
@@
   at the window.
-  A core goal of the normal mode it to allow fast
+  A core goal of the normal mode it to support fast
   navigation of the document while allowing
-  the typist to maintain its fingers on the home row
+  the typist to maintain the fingers on the home row
   (i.e. on the center of the keyboard).
-  The mode is also used for manupulating text
+  The mode is also used for manipulating text
   (e.g. copy, paste, delete, change case) and
     changing to other modes.
-  \item Insert mode: for inserting text.
-  \item Command-line mode: for entering Ex commands.
+    More in Section~\ref{normal}.
+  \item Insert mode: for inserting text. More in
+    Section~\ref{sec:ins}.
+  \item Command-line mode: for entering Ex commands. More in
+    Section~\ref{sec:com}.
   \item Ex mode: similar to command-line mode,
   but more specialized for running various Ex commands.
+    More in Section~\ref{sec:ex}.
   \item Visual mode: for making, manipulating and navigating
@@
   \item Select mode: similar to visual mode but
-  favors CUA\footnote{IBM Common User Access: \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
+  favors CUA\footnote{IBM Common User Access:
+    \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
 \end{itemize}
@@
 There is another basic mode, but it is not fully implemented:
-the Terminal-Job mode.
+the Terminal-Job mode (more in Section~\ref{terminal}.
 There are seven additional modes which are mostly subordinate 
-to the basic modes and will be described when convenient.
+to the basic modes and that will be described when convenient.
 The manual page for Vim modes can be accessed by typing
@@
 Some of the modes are now further considered for
-an overview of the Vim usage possibilites.
+the achievement of
+an overview of the Vim usage possibilities.
 
@@
 navigating, manipulating texts and changing to other modes.
-
 The simplest of these three is changing to other modes:
@@
 Type \texttt{Q} to enter Ex mode.
-Type \texttt{v}, \texttt{V}, \texttt{CTRL+V} to enter visual mode.
-
-For most basic and naive navigation, one should check Section~\ref{minimum}.
+Type \texttt{v}, \texttt{V}, or \texttt{CTRL+V} to enter visual mode.
+
+For very basic and naive navigation, one should check Section~\ref{minimum}.
+There are many facilities to navigate Vim
+as explained in \texttt{:h navigation}.
 Most often, one uses:
@@
   \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
-and whole page down and up, although these commands might
+and whole page down and up, respectively, although these commands might
 be set to scroll a different number of lines.
   \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
-  \item \texttt{(w,b,e)} to move to the next, previous and next end-of-word.
+  \item \texttt{(w,b,e)} to move to the next and previous word,
+    and next end-of-word.
     There are motions to iterate over sequences of characters separated
@@
     jump list.
-  \item \texttt{'X,`X} to move the cursor to a mark bond to the alphanumeric character X: \texttt{`X} moves to the exact position while \texttt{'X} moves to
+  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X:
+    \texttt{`X} moves to the exact position while \texttt{'X} moves to
     the first non-blank character of the line.
@@
 
-There are many more facilities to navigate Vim
-as explained in \texttt{:h navigation}.
-
 For changing the text, usual commands include:
@@
   \item \texttt{x,X} to delete the character under or before the cursor.
-  \item \texttt{$\sim$} to swap the case of a character.
+  \item \texttt{$\sim$} to swap the case of a character or selection.
   \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
@@
 Some of them are discussed in Section~\ref{navIn}
-because they involve a transition to the insertion mode.
+because they involve a transition to the insert mode.
 A thorough consideration of the commands in the normal mode
-is found by executing \texttt{:h navigation} and 
-\texttt{:h change.txt}.
-
-\subsubsection{Insertion}
-Once in the insertion mode, the character keys
+is found by executing \texttt{:h navigation}, 
+\texttt{:h change.txt}, \tttt{:h index}.
+
+\subsubsection{Insert mode}\label{sec:ins}
+Once in the insert mode, the character keys
 input the characters at the cursor position at the current buffer.
-One can exit insert mode by pressing \texttt{<Esc>},
-and Vim will be put in normal mode.
+One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
+\ttt{C-C}),
+and Vim will be put in the normal mode.
 Most useful commands in insert mode include:
@@
   \item \ttt{C-O} to execute one and only command in normal mode.
-	  This enter a secondary mode (see Section~\ref{modes})
+	  This enters a secondary mode (see Section~\ref{modes})
   \item \ttt{C-R} to paste a register (a variable starting with
@@
     the line or to the previous word.
-  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand..
+  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand.
   \item \ttt{C-X} commands for scrolling the window with multiple
@@
 
-\subsubsection{Command-line mode}
-This mode is dedicated for writing colon, search and filter commands,
+\subsubsection{Normal $\rightarrow$ insert modes}\label{navIn}
+Many commands bridge from Normal to Insert modes,
+e.g. \tttt{iws} or any of these letters: \tttt{csrCSR}.
+These make convenient the replacement of text and populates registers.
+The absence of a short command to insert one char is
+a known issue in Vim.
+Reasonable mappings to insert or append a char to and around
+another char are in~\cite{vimrc}.
+Vim couples operator and motion commands by design.
+There are many operator commands that take the editor from
+the normal mode to the insert mode, most of them
+favoring deletion or change, as detailed in \tttt{:h operator}.
+Motion commands are described in \tttt{:h motion}.
+
+\subsubsection{Command-line mode}\label{sec:com}
+This mode is dedicated to writing colon, search and filter commands,
 entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
@@
     until the cursor.
-  \item \ttt{C-R} to paste a register as in insert mode.
-\end{itemize}
+  \item \ttt{C-R} to paste a register (as in insert mode).
+\end{itemize}
+
+\subsubsection{Ex mode}\label{sec:ex}
+One might use the normal commands \tttt{q(:,/,?)} to have a window
+with the colon or search command history to be edited normally,
+and the chosen command can be run with \ttt{CR}.
+Vimscript was largely based on the ex editor~\cite{ex},
+and a more advanced user might use it for prototyping
+by defining mappings and settings and managing scripts
+probably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
+In the default interface started with the \tttt{Q} command in normal mode,
+each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
+follow documentation from \tttt{:h Ex-mode} for further information.
+I might be loosing something, but for
+tweaking I use the command-line window accessed through \tttt{q:}.
+I find it comfortable to browse the history in a normal-like mode,
+edit them also using the insert mode as usual,
+and having auto-completion when pressing \ttt{TAB}.
+None of such features are available in Ex mode by default
+and
+customization by mappings have to be performed through autocommands as in~\cite{vimrc}.
 
 \subsubsection{Terminal-Job mode}\label{terminal}
-This mode is reported to not have reached a stable usage design
+This mode is reported as not having reached a stable usage design
 (see \tttt{:h terminal}).
-I find that it works exceptionaly well and have used it to run
+I find that it works exceptionally well and have used it to run
 scripts in an IPython shell, compile latex files, and even open PDFs and images.
 Vim browsing of windows and text manipulation is well developed,
-so the terminal mode enables a very convenient integration,
+so the Terminal-Job mode enables a very convenient integration
+of files being edited and bash terminals,
 more traditionally achieved through Byobu/Tmux/Screen terminal
 multiplexers.
-Most useful commands in terminal-job mode include:
-\begin{itemize}
-  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the normal
-    and command-line modes.
+Most useful commands in the terminal-Job mode include:
+\begin{itemize}
+  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the Terminal-Normal
+    and command-line modes. Terminal-Normal mode is very similar to
+    Normal mode, but one cannot change the text, cannot enter insert
+    mode, and the status line reports if the job is finished or not.
   \item \tttt{<C-W>\_"} to paste a register.
-  \item \tttt{i} for entering the terminal-job mode from the normal mode.
-\end{itemize}
-
-It is useful to define the same mappings for nativating splits and tabs
-for both normal and terminal-job modes, as in~\cite{vimrc}.
-
-\subsubsection{Normal $\rightarrow$ insertion}\label{navIn}
-Many commands bridge from Normal to Insertion modes:
-\tttt{csrCSR}, \tttt{iws}, etc.
-These make convenient the replacement of text and populates registers.
-The absence of a short command to insert one char is
-a known issue in Vim.
-Reasonable mappings to insert and append a char to and around
-other char are in~\cite{vimrc}.
-Vim commands are designed to couple operator and motion commands.
-There are many operator commands that take the editor from
-the normal mode to the insert mode, most of them
-favoring deletion or change, as detailed in \tttt{:h operator}.
-Motion commands are described in \tttt{:h motion}.
-
-\subsubsection{Netrw}\label{netrw}
+  \item \tttt{i} for entering the Terminal-Job mode from the normal mode.
+\end{itemize}
+\noindent It is useful to define the same mappings for navigating splits and tabs
+for both Normal and Terminal-Job modes, as in~\cite{vimrc}.
+
+Because browsing the interface in Vim is fast,
+and it favors copy and paste of text,
+it is often handy to keep (e.g. a tab with) some terminals:
+one with an IPython shell, another two for compiling \LaTeX
+and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
+It is very comfortable to copy and browse the terminal lines
+in the Terminal-Normal mode.
+I found the mappings on~\cite{vimrc} very helpful for directing
+editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
+which I make available across Terminal-Job and Normal modes.
+But I've been thinking on using \ttt{C-} commands also
+for tabs (not only for splits), both to avoid mixing typing sequences
+with commands, and to allow the use of the commands in all Normal,
+Insert and Terminal modes.
+
+\subsection{Netrw}\label{netrw}
 The standard interface of Vim for browsing file trees is Netrw.
@@
 further information in \tttt{:h netrw}.
-
-\subsubsection{Ex}
-One might use the normal commands \tttt{q(:,/,?)} to have a window
-with colon or search command history to be edited normaly,
-and the chosen command can be run with \ttt{CR}.
-Vimscript was largerly based on the ex editor~\cite{ex},
-and a more advanced user might use it for prototyping
-by defining mappings and settings and managing sessions and scripts
-propably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
-In the default interface started with the \tttt{Q} command in normal mode,
-each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
-follow documentation from \tttt{:h Ex-mode} for further information.
-I might be loosing something, but for most
-hacking I use the command-line window accessed with \tttt{q:}.
-I find it comfortable to browse the history in a normal-like mode,
-and edit them also using the insert mode as usual.
-The mapping have to be defined as autocommands as in~\cite{vimrc}.
 
@@
 The Vimball can be installed in a system by \tttt{:source <filename>.vba}
-or loading it at vim startup with \tttt{\$ vim <filename>.vba}.
+or loading it at Vim startup with \tttt{\$ vim <filename>.vba}.
 
@@
 
-All directories vim looks for scripts are described
+All directories Vim looks for scripts are described
 in \tttt{:h vimfiles} and are basically set by
@@
     Sections~\ref{visual1} and~\ref{visual2}.
-  \item Make plugins for:
+\end{itemize}
+\noindent Many of the next steps might entail plugins.
+This is a selection of the issues that are more prominent for me:
     \begin{itemize}
@@
         of such commands with \tttt{:se nomore}.
+        Reasonable functions (and convenient commands) for having the
+        output of such commands in a standard Vim window are
+        in~\cite{vimrc}.
       \item Slide presentations. I've been using some automation for browsing
@@
     \end{itemize}
-\end{itemize}
 
@@
 
-\section{Example of notes on mappings}\label{notes}
+\section{Example notes on mappings}\label{notes}
 \texttt{:h index} shows all the default mappings
@@
 	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
+\bibitem{wikiVim}
+  Vim (text editor). (2017, December 4). In Wikipedia, The Free
+    Encyclopedia. Retrieved 07:08, December 16, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Vim_(text_editor)&oldid=813716105}
 \bibitem{ouroWiki}

### 2017-12-18 article/article.tex
added version text file

@@
 Because browsing the interface in Vim is fast,
-and it favors copy and paste of text,
+and it is very comfortable to copy the terminal lines
+in the Terminal-Normal mode,
 it is often handy to keep (e.g. a tab with) some terminals:
-one with an IPython shell, another two for compiling \LaTeX
+e.g. one with an IPython shell, another two for compiling \LaTeX\
 and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
-It is very comfortable to copy and browse the terminal lines
-in the Terminal-Normal mode.
+There are even more convenient ways to use the terminal inside Vim.
+For example, 
+one might use \tttt{:term} with the \tttt{++hidden} and \tttt{++close}
+options to compile a \LaTeX\ file in the background without needing to further manage the
+terminal and in a non-blocking manner: \tttt{:term ++hidden pdflatex
+\% \#}.
 I found the mappings on~\cite{vimrc} very helpful for directing
@@
 It starts when you open a directory, such as with \tttt{:e .<CR>}.
-It has solid support for editing remote files (such as over ssh or
-ftp) and handy e.g. mappings to open the files as splits and tabs (specially \tttt{pot}).
+It has solid support for browsing remote file trees (such as over ssh or
+ftp) and handy e.g. mappings to open the files as splits and tabs
+(specially \tttt{pot}).
 Most useful commands in netrw include:
@@
 \end{itemize}
-Last notes:
-there is no insert mode in netrw interface;
+
+There is no insert mode in netrw interface;
 the commands in \tttt{:h netrw-explore} are
@@
 file being edited;
-further information in \tttt{:h netrw}.
+further information is in \tttt{:h netrw}.
 
@@
 One might edit the vimrc file with \tttt{:e \$MYVIMRC}
-and reload it with \tttt{:source \$MYVIMRC}, but
-I use the mappings in~\cite{vimrc} to encourage a continuous enhancement of my settings.
+and reload it with \tttt{:source \$MYVIMRC}.
+Mappings in~\cite{vimrc} include such commands
+to encourage a continuous enhancement of Vim settings
+(they have helped me to improve my settings without unnecessary 
+hassle and fast).
 Any other file might be included to run at startup by
@@
 scripts and sequence of them to be loaded.
+The vimrc from other users are most useful for one to comprehend
+and pick convenient practices and settings.
+In fact, a vimrc file is most often a collage of excerpts of vimrc
+files from other users.
 
 \subsection{Plugins and packages}
-One can see the list of standard plugins
+One can see the list of the standard plugins
 with \tttt{:h standard-plugin-list} command.
 Any \tttt{plugin/**/*.vim} file inside a directory
-listed by \tttt{:se runtimepath} will be loaded
+listed in \tttt{:se runtimepath} will be loaded
 (e.g. \tttt{.vim/plugin/something/ascript.vim}).
 There are various ways to automate the installation
-and enhnace the management of plugins.
+and enhance the management of plugins.
 By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
-that downloads latest scripts from sourceforge as specified in \tttt{:h getscript-data},
-and the Vimball interface, which creates and loads a Vimball for a plugin
-with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
+that downloads latest scripts from Sourceforge as specified in \tttt{:h getscript-data},
+and the Vimball interface, which creates and loads a Vimball for a
+plugin.
+Such a Vimball may be created with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
 that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
@@
 \tttt{:se runtimepath} and conventions inside
-each path therein, such as to look for \tttt{vimrc} files
+each directory therein, such as to look for \tttt{vimrc} files
 and \tttt{plugin/} or \tttt{pack/} directories.
@@
 e.g. only if a function is used as in \tttt{:h autoload-functions}.
-Example of such are filetype plugins (enabled by in a \tttt{ftplugin/<filetype>.vim} file,
+Example of such are filetype plugins (enabled by a \tttt{ftplugin/<filetype>.vim} file,
 e.g. inside a plugin directory).
@@
 and toggle spell checking with \tttt{:setl spell!}.
-These are used so often that one might use shortcuts as in~\cite{vimrc}.
+Depending on the activities being performed,
+these commands are used so often that one might use mappings as in~\cite{vimrc}.
 Currently, Vim will download the files for a specific language if
@@
 In Vimscript,
-the colon commands (also Ex or command-line commands) are related through spaces,
+the colon commands (also Ex commands or command-line commands) are related through spaces,
 punctuation and keywords (see \texttt{:h script}).
 Scripting the Vim editor can also be accomplished using other languages,
-as well documented e.g. \tttt{:h python}.
+as well documented e.g. in \tttt{:h python}.
 Functions are defined through colon commands and are called
@@
 through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
-Executing source files is very straightforward with \tttt{:so
-<filename>.vim},
+Executing source files is very straightforward with
+\tttt{:so <filename>.vim},
 and one can always use the Ex mode for rapid scripting.
-
 At the same time, the \tttt{:term} and terminal-job mode make
 scripting other software more convenient, as output is
-promptly navigated and copied. The shell works well for generic use,
-such as opening images and PDFs through \tttt{\$ eog} and \tttt{\$ evince}.
-See Section~\ref{terminal} for further directions about the terminal.
+promptly navigated and copied (as discussed in Section~\ref{terminal}).
 
 \subsection{Color}\label{visual2}
-Newer versions of Vim support 24 bit true color (aka 16 million colors)
-in terminal Vim (not only in GVim).
+Newer versions of Vim support 24-bit true color (aka 16 million colors)
+in terminal Vim (in gVim true color received support earlier).
 The terminal must support true color, and tests are available e.g. in
@@
 \tttt{:set termguicolors}.
-If using 8 or 16 bit colors, Vim uses the color pallete from the
-terminal, if using true colors each color is defined directly.
-Settings for using true colors inside Byobu/Tmux involve tweakings
+If using 8 or 16 bit colors, Vim uses the color palette from the
+terminal, if using true color each color is defined directly.
+Settings for using true color inside Byobu/Tmux involve tweaking
 and are available in~\cite{vimrc}.
-Good color schemes to use with true colors are Gruvbox
+Good color schemes to use with true color are Gruvbox
 and Solarize.
 One might source syntax files at any time to change syntax, usually
-though linking tokens to syntax groups with
-\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...}
-and then relating the group to another group 
+though linking tokens to syntax groups as in
+\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...},
+or
+\tttt{:syntax match <group\_name>	<pattern>},
+and then relating the group to another group: 
 \tttt{:highlight link <group\_name> <group\_name2>}
-or by specifying the group characteristics directly:
+or to group characteristics directly:
 \tttt{highlight <group\_name> guifg=\#ffffff}.
 If you change a syntax file, reloading a file with with \tttt{:e<CR>}
-updates the highlighting on the window with the corresponding filetype.
+updates the highlighting on the window with the corresponding file type.
 A complete syntax highlighting support typically involves at least three files:
@@
 
-This very scheme is implemented very straightforward in this
+This scheme is implemented very straightforward in this
 plugin~\cite{tokipona}
-for highlighting text in the Toki-Pona language~\cite{tpLang}.
-Syntax highlighting plugins are as filetype plugins,
+for highlighting text in the Toki Pona language~\cite{tpLang}.
+Syntax highlighting plugins are as file type plugins,
 but also have a \tttt{syntax/<filetype>.vim} file
-relating keywords to highlighting groups. 
+relating keywords, matches and regions to highlighting groups. 
 One might see every highlighting group, and their final visual results,
-with the command
-\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}.
-In my system, the \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
-directory, but \tttt{syntax/} files had to be moved to \tttt{$\sim$/.vim/syntax/}.
+with the commands
+\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}
+or \tttt{:so hi}.
+The \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
+directory, but \tttt{syntax/} files has to be moved to
+\tttt{$\sim$/.vim/syntax/}, unfortunately.
 
 \subsection{Fonts}
-The fonts are defined by your terminal or inside GVim.
+The fonts are defined by your terminal or inside gVim.
 \ttt{C-+} and \ttt{C--} can be used to change font size.
@@
 This document seems reasonable as an overall reference of the Vim editor,
-at least for my usage and proficiency.
+at least for my usage and level of proficiency.
 Given the folkloric milestone of using Vim for 10 years,
@@
 
-\noindent {\bf Potential enhancements} to this document are:
-\begin{itemize}
-  \item The inclusion of facilities such as reading emails and connecting over ssh.
+\noindent {\bf Potential enhancements} to this document include:
+\begin{itemize}
+  \item The discussion of facilities such as reading emails and connecting over ssh.
     There is a working hack in~\cite{vimrc} for browsing over the WWW,
     but such aspect of Vim usage might receive more attention
-    given that it is confortable to use the navigation and editing facilities
+    given that it is comfortable to navigate and edit in Vim
     and the resulting integrated environment.
@@
     I have never used it, but it seems to be reaching a considerable user base
-    and it might be feasable to give an account of Vim and Neovim
-    after some tests and researching the official and user base documentation.
+    and it might be feasible to give an account of Vim and Neovim
+    after some tests and researching the official and unofficial documentation.
   \item Better cite documentation, plugins and Vim authors.
-    I prefered to keep the references inline through \tttt{:h} commands
+    I preferred to keep the references inline through \tttt{:h} commands
     and URLs, more in accordance with the style of Vim documentation,
     but bibliographical items constitute a valuable asset for academic literature,
-    and some authors might find their work more pretiged if more
+    and some authors might find their work more respected if more
     thoroughly cited.
@@
 
-\noindent {\bf Potential next steps} in using Vim:
-\begin{itemize}
-  \item Measure the performance of text mining routines in Vim against those implemented in C or Python.
+\noindent {\bf Potential next steps} in using Vim include:
+\begin{itemize}
+  \item Measure the performance of text mining routines in Vimscript against those implemented in C or Python.
   \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
   \item Better integrate Python and Vimscript, especially for data visualization
-    and syntax highlighting management.
-    In accordance with the visualization issues described in
+    and syntax highlighting management,
+    in accordance with the visualization issues described in
     Sections~\ref{visual1} and~\ref{visual2}.
 \end{itemize}
-\noindent Many of the next steps might entail plugins.
-This is a selection of the issues that are more prominent for me:
+
+\noindent This is a selection of the issues that might entail
+plugins and that are more prominent for me:
     \begin{itemize}
@@
         for new mappings.
+        Maybe already group possibilities by criteria such as length
+        of sequence and how central are the keys.
       \item Sessions, as described in Section~\ref{visual1}.
@@
         screencasts and images, interact with IRC channels and other social platforms.
-      \item Slick Vim: a collection of the settings, mappings and plugins I use.
+      \item Slick Vim: a collection of the settings and mappings I use.
         Enhancements such as using \ttt{C-} commands also to browse tabs,
         shortcuts to join a window into a tab,
-        and dummy minimal plugins as simplest, file type and syntax highlighting,
+        and dummy minimal plugins as simplest possible, then file type
+        and then syntax highlighting,
         Some more elaborate tweaks should also be present, such
@@
         such as to comply with the limitations of a platform (e.g. Vim editor). 
-        but also hold stylistic merits as ascii art is often
+        but also hold stylistic merits as ASCII art is often
         very appreciated.
-        One can both obtain an ascii representation of a binary image (e.g. JPG, PNG),
-        and can directly render ascii charts from data using cues such are shape, position
+        One can both obtain an ASCII representation of a binary image (e.g. JPG, PNG),
+        and can directly render ASCII charts from data using cues such are shape, position
         and color.
-      \item Redirecting the commands that usually show the results in a 'more' interface,
-        which cannot be searched nor copied,
-        to a parsed and linked quickfix or location window.
+      \item Redirecting the commands that usually show the results in
+        a 'more' interface (e.g. status listings such as in
+        Section~\ref{state}),
+        which cannot be searched nor copied nor persists if one
+        returns to editing a file.
+        Ideally, it should parsed and linked quickfix or location
+        window, and the syntax highlighting maintained.
         The basic idea is to use \tttt{:redir} command to redirect the output
@@
       \item Slide presentations. I've been using some automation for browsing
-        slides and opening figures and some Vim users asked for the settings and
-        commands. They are very elementary use of registers being executed over
+        slides and opening figures. Some Vim users asked for the settings and
+        commands. They are very elementary use of registers which are executed over
         arbitrary but consistent textual patterns.
+      \item For bringing back all the splits after an \tttt{:only}.
+        Also for bringing back the tabs after a \tttt{:tabo}.
+      \item Run Python excerpts from a file being edited in an IPython shell. 
+        The buffer number of the terminal window should be stored,
+        and than any selected lines should be run on that instance.
+        Mappings should make available all movements to fetch the
+        excerpts, execute registers, remain on script or in the IPython shell.
+        Also, current Terminal-Job mode can be improved easily with
+        mappings, such as \ttt{C-O} for one normal mode command.
+      \item For keeping track of the usage and making analysis
+        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}.
+        Usage analytics.
       \item To facilitate the tweaking of syntax highlighting.
         This should include ways to easily access and change the
-        syntax highlighting scheme and dump it to a file (dumpung
-        current highlighting scheme is currently not supported).
-        Should also include changing the color scheme an highlighting scheme incrementally
-        and selectively using the features described in Section~\ref{visual2}.
-      \item For bringing back all the splits after an \tttt{:only}.
-        Also for bringing back the tabs after a \tttt{:tabo}.
-      \item Run Python excepts from file in an IPython shell. 
-        The buffer number of the terminal window should be stored,
-        and than any selected lines should be run on that instance.
-        Mappings should make available all movements to catch the excepts,
-        execution of registers, remain on script or in the IPython shell.
-        Also, terminal mode can use some improvements, such as \ttt{C-O} for
-        one normal mode command.
-      \item For keeping track of the usage and making analysis
-        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}. Usage analytics.
-      \item Colorschemes for true color (gui).
+        syntax highlighting scheme and dump it to a file (dumping
+        current highlighting scheme to a working Vimscript file
+        is currently not supported by Vim!).
+        Should also include changing the color scheme and highlighting scheme incrementally
+        and selectively using the features described in Section~\ref{visual2}
+        and in~\cite{tokipona}.
+      \item Color schemes for true color.
         The standard colors (e.g. blue, elflord) loose some of the distinctions,
-        e.g. SpellBad tokens are not highlighted on these colorschemes
-        it you have a functioning \tttt{se termguicolors} (standard GUI mode
-        and available in terminals since recent versions of Vim 8, see Section~\ref{visual2}). In~\cite{vimrc} are some lines that make such colorschemes after over
+        e.g. SpellBad tokens are not highlighted on these color schemes
+        if you have a functioning \tttt{se termguicolors} (standard GUI mode
+        and available in terminals since recent versions of Vim 8,
+        see Section~\ref{visual2}).
+        In~\cite{vimrc} are some lines that make such color schemes over
         \tttt{:colorscheme blue}.
@@
         a very simple and effective way to make new color schemes.
-      \item Opening the editor status listings such as in Section~\ref{}
-        in a quickfix or location window where they might be searched, edited
-        and used as wanted.
-        These windows open in a more-like interface that is very restricted in what
-        they allow (e.g. no search, no persistence).
+        Such color schemes should also make use of the discussions
+        in~\cite{tokipona}.
     \end{itemize}
@@
 \tttt{:colo gruvbox} and \tttt{:colo blue}.
-I open an vertical split and then move
+I open a vertical split and then move
 the window to a new tab using
 \tttt{:vs} and \ttt{C-W\_T}.
-I then search tokens related to
+I then search for tokens related to
 the enhancements I want to make or
@@
 and make a global replace with
-\tttt{:\%s/<this>/<that/g}.
+\tttt{:\%s/<this>/<that>/g}.
 On adding dots to sentences,
@@
 the sequence \tttt{jA.jj},
-using it as a macro 10 times by
+and use it as a macro 10 times by
 \tttt{10@q}.
@@
 If there is e.g. code or notes in other projects I am working
-on, I reach them through \tttt{<Alt-arrows>} as other Vim instances
+on, I reach them through \tttt{<Alt-arrows>} because they
+are in separate Vim instances
 in Byobu/Tmux sessions and windows.
+Because I stay for hours editing (\LaTeX and Python
+scripts), I change the background to red, and eventually to green.
+Because the blue color scheme does not highlight the spelling errors
+if using true color, I run
+\tttt{:hi SpellBad guifg=red guibg=lightblue}
+to see the words found wrong by the spell checker.
 
 \section{My vimrc file and usage}
-In this Appendix is my vimrc file.
-Although it has comments, one should
-look for the options that (s)he does
-not understand, as a thorough explanation
-of the file is tedious and out of the scope
-of this document.
-
 In using Vim with my vimrc file~\cite{vimrc},
@@
 are also used constantly.
+Although the file is commented, one should
+look for the help pages on the options that (s)he does
+not understand, as a thorough explanation
+of the file is tedious and out of the scope
+of this document.
 
@@
 In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
-\ttt{BS} and \ttt{C-H} is the same as h.
-\ttt{C-J} and \ttt{C-N} is the same as j.
+\ttt{BS} and \ttt{C-H} are the same as h.
+\ttt{C-J} and \ttt{C-N} are the same as j.
 \ttt{C-P} is the same as k.
-Space is same as l.
+Space is the same as l.
 \ttt{C-[} and \ttt{Esc} are not used,
-\tttt{<C-\textbackslash> a-z} is reserved for extensions,
-\ttt{C-\_} not used.
+\tttt{<C-\textbackslash> a-z} are reserved for extensions,
+and \ttt{C-\_} is not used.
 \tttt{+} is the same as \ttt{CR} and they are both not very useful.
-Del is same as x.
+Del is the same as x.
 
@@
 in more centraly located keys.
-unbounded: g,z,[]
+The letters beggin unbounded mappings (require another character to
+trigger an action without having to wait for \tttt{:se timeoutlen}): g,z,[].
 
@@
 \subsection{Insert mode}
-All standard char keys are used for entering chars on text.
+All standard character keys are used for entering text.
 \tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
 in moving through multiple lines.
-\ttt{C-[} is the same as \ttt{ESC}.
+\ttt{C-$[$} is the same as \ttt{ESC}.
 \ttt{C-J} and \ttt{C-M} are \ttt{CR}.
-\tttt{<C-\textbackslash> a-z} is reserved for extensions,
+\tttt{<C-\textbackslash> a-z} are reserved for extensions,
 other combinations with \ttt{C-\textbackslash} are not used.

### 2017-12-18 article/article.tex
terminada revisao

@@
 that carry values or triggers procedures.
-One should see Appendix~\ref{keys}
+One should see Appendix~\ref{sec:not}
 to understand the notation Vim uses.
@@
       \item For keeping track of the usage and making analysis
-        for optimizing the usage, as described in~\cite{http://www.drbunsen.org/vim-croquet/}.
+        for optimizing the usage, as described in~\url{http://www.drbunsen.org/vim-croquet/}.
         Usage analytics.
@@
 in which I wrote this document in this link:
-\url{}.
+\url{https://github.com/ttm/vim/raw/master/version.txt}.
 It was compiled with this Makefile~\cite{makefile}
@@
 
-\section{Key notation and meaning for Vim}
+\section{Key notation and meaning for Vim}\label{sec:not}
 The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
@@
 \ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x} and \ttt{C-x} are the same,
-i.e. Vim does not distinguish lower and uppercase 
+i.e. Vim does not distinguish between lower and uppercase 
 letters in \ttt{C-} commands.
@@
 and they are not used by Vim's builtin keys.
-Such \ttt{M-} key combinations are potentially conflicting
+Such \ttt{M-} key combinations potentially conflict
 with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
@@
 In summary,
-one might choose for mappings to overwrite defaults key combinations,
+one might choose mappings that overwrite default key combinations,
 use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
@@
 create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
-More information about key notation in \tttt{:h key-notation}.
+More information about key notation is in \tttt{:h key-notation}.
 Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings

### 2018-02-11 article/article.tex
terminada revisao
atualizando para repasse

@@
 \begin{document}
-\title{An anthropological account of the Vim text editor:\\
-features and tweaks after 10 years of usage}
+\title{An anthropological account of the\\ Vim text editor:\\
+features and tweaks\\ after 10 years of ouroboros}
 % Anthropological computation
@@
 e.g. Python, C++, JavaScript, ChucK programs;
-\LaTeX, Markdown, HTML, RDF, Make and other markup files;  % TTM
+\LaTeX, Markdown, HTML, RDF and other markup-like files;  % TTM
 binary files.
@@
 it might generate.
-Also, it should be useful for users of any degree of experience,
+Also, the preliminary versions of this document already proved itself
+useful for users of any degree of experience,
 including me, as a compendium of commands, namespaces and tweaks.
-Upon feedback, and maturing of my Vim usage,
-this document might be enhanced and expanded.
+This document might be enhanced and expanded
+upon feedback, maturing of my Vim usage, or derivatives potentially
+yield by or with other users.
 \end{abstract}
+{\bf keywords:} Vim, Text editor, HCI, Tutorial, Anthropological computer science
 
 \section{Introduction}\label{intro}
-Vim is a very complex editor,
+Vim is a very complex text editor,
 most often considered to be
@@
 I used Vim for writing and editing computer code 
-(Python, Javascript, C++, ChucK, bash, etc), markup languages  % TTM
+(Python, Javascript, C++, ChucK, bash, etc), markup-like content
 (HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
-Eventually, I edited database files and other types of files.
+Eventually, I edited other types of files, such as database files with
+dozens of gigabytes.
 Within Vim, I mostly write (web and scientific) software,
@@
 mainly by Bram Moolenaar.
-Today, current bleeding-edge version is 8.0.1401.
-I found no explicit stable, alpha or beta versions.
+Today, bleeding-edge version is 8.0.1497.
+There seems to be no official or explicit stable, alpha or beta releases~\cite{noOfficial}.
 I found no scientific article on Vim 
@@
 there are convenient menus and toolbars
-but all functionalities are still available though
-the command line mode.
-Vimscript is the internal language of Vim,
+but all functionalities are still available through
+the command-line mode.
+VimL (a.k.a. VimScript, Vim script, Vim language)
+is the internal language of Vim,
 and is often used for scripting by users
@@
 (e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
-Each line of a Vimscript is a command on the
+VimL is preferred among other names for the language
+because it is shorter and less ambiguous then the alternatives,
+although there seems to be no official name and the manual
+uses the terms \emph{Vim script language} and \emph{Vim script}
+(see e.g. \tttt{:h usr\_41}).
+Each line of a VimL script is a command on the
 command-line mode.
@@
 that focuses on the namespaces, i.e. sets of tokens
-that carry values or triggers procedures.
+that carry values or trigger procedures.
 One should see Appendix~\ref{sec:not}
-to understand the notation Vim uses.
+to understand the notation Vim uses to represent key combinations
+(e.g. \ttt{C-[}).
 
@@
 and \tttt{plugin/} or \tttt{pack/} directories.
-Vim scripts can be loaded conditionally,
+VimL scripts can be loaded conditionally,
 e.g. only if a function is used as in \tttt{:h autoload-functions}.
@@
 	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
+\bibitem{noOfficial}
+  Vim users email list. "Vim versions. (No explicit or official stable, alpha or beta versions of Vim)" Available at \url{https://groups.google.com/forum/#!msg/vim_use/ULCa_2Cxnl0/RmvGij8YCgAJ} 
 \bibitem{wikiVim}

### 2018-02-11 article/article.tex
atualizando para repasse

@@
 
-\subsection{Further remarks about this document}
+\subsection{Further remarks about this document}\label{furRem}
 This document is written is a DRY KISS
@@
 You open a file at Vim startup by executing
-the command: \texttt{vim <filename>}.
+the command: \texttt{\$ vim <filename>}.
 Inside Vim, you start in the normal
@@
 which is considerably lengthy and complex
-and is usually not read by users before a few years
+and is usually not read by users before they achieve a few years
 of experience.
 In learning Vim, one
-might want to run the \texttt{vimtutor} command
+might want to run the \texttt{\$ vimtutor} command
 (outside Vim) to start the Vim Tutor.
@@
   \item Many videos about Vim are publicly available.
-  One traditional site is \url{http://vimcasts.org},
-  but you might find them using a search engine
-  (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+  One might find them using a search engine
+  (e.g. \url{http://derekwyatt.org/vim/tutorials/}, and
+  \url{http://vimcasts.org}) or in Youtube and Vimeo.
 \end{itemize}
@@
 text and writing alters text and writing.
-What you end with is a collection of namespaces where tokens have scalar
-or arbitrarily complex values.
+This is achieved through a collection of namespaces where tokens have scalar
+or arbitrarily complex values, or start automated routines.
 
 \subsubsection{Commands and mappings}
-There are commands, i.e. typing sequences which trigger automated actions,
+There are commands, i.e. typing sequences which trigger actions,
 for each mode:
@@
     and have combinations not used (thus available for new mappings),
-		specially the z and g.
+		specially the z and g keys.
     See Section~\ref{normal} and Appendix~\ref{notes}
@@
 and executed by the \tttt{:execute} colon command.
-E.g. \tttt{:execute 'vs afile.txt'}.
+E.g. \tttt{:execute 'vs afile.txt'} executes the \tttt{:vs afile.txt}
+colon command..
 As there are colon commands that execute commands in other
@@
 the \tttt{:execute} is a way to build commands in any mode,
-e.g.
-\\\tttt{:execute 'normal i' . string(atan(bufnr('\%')))}.
+e.g. \tttt{:execute 'normal i'.string(atan(bufnr('\%'))).'<ESC>'}.
 
@@
 	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
-    More on register in Section~\ref{state} and scattered in this
+    More on registers in Section~\ref{state} and scattered in this
     document.
-	\item Internal variables are created with let and preferably have a prefix:
+  \item Internal variables are created with \tttt{:let} and preferably have a prefix:
 	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
@@
 		sourced Vim file, respectively.
- \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
+ \tttt{v:}, \tttt{g:} are globals, the first are predefined by Vim.
 		\tttt{a:} is for function arguments.
-		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
+		If there is no prefix, the variable is global or internal to a function if defined inside a function.
 		More about internal variables in \tttt{:h internal-variables}.
-	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
-\end{itemize}
-You can echo any of such variables or use them in expressions.
+	\item The value of a variable can be of the types: scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
+\end{itemize}
+You can \tttt{:echo} any of such variables or use them in expressions.
 Notice that you will only be able to echo a \tttt{b:} variable inside
 the buffer where it is defined.
-For all the Vimscript capabilities, including loops, conditionals,
+For all the VimL capabilities, including loops, conditionals,
 and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
@@
 	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
-  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
-	\item Tabs list can be reached through \tttt{:tabs}.
+  \item The windows are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
+	\item A tabs list can be reached through \tttt{:tabs}.
 		It is usual both to show and hide the tabs bar (mapping
@@
     monitor register \tttt{""} (maybe also monitor \tttt{"0}).
-	\item An undo list can be accessed through \tttt{:changes}.
-	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
+	\item An undo list is accessed through \tttt{:changes}.
+	\item A list with all the sourced scripts in a Vim session is displayed through \tttt{:scriptnames}.
 	\item The markers defined are listed with the \tttt{:marks} command.
@@
 		Uppercase letters are cross buffers.
-	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
+	\item Quickfix and Location lists are populated through \tttt{:vim} and \tttt{:make}
 		and variations, such as \tttt{:grep}.
@@
 		More information in \tttt{:h quickfix}.
-	\item A tags list have to be made so one can use tags.
+	\item A tags list have to be made in order to enable the use of tags.
 		Most often one will generate the tags list using the exuberant ctags,
@@
     For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
-  \item All the autocommands might be accessed through \tttt{:au}.
+  \item All the autocommands are listed with \tttt{:au}.
     They are event-triggered actions Vim performs.
@@
 
-A file with information about the state of the editor
+A window with information about the state of the editor
 can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
@@
 inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
+More about using colors in Vim for syntax highlighting in Section~\ref{visual2}.
 
@@
 You can keep track of the editor state though commands, as stated
-above.
+in last subsection.
 Also, one might rely on persistent visual
@@
 reserved for the command-line.
-A good strategy I find is to have selective visual cues of the state to make persistent or hide
+A good strategy I find is to have specialized visual cues of the state to make persistent or hide
 and a mapping to toggle each of them.
 Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
-status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
+status line and tabs bar with \tttt{<localleader>-T and B}, according to script~\cite{vimrc}.
 I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
-Numbering is always there, I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
+Numbering is always there, I rarely turn them off but keep the mappings \tttt{<leader>-n and N} to toggle just in case.
+Highlight of last search string, and underline of unrecognized words (e.g. in English of Portuguese), are toggled with \tttt{<localleader>-l and p}.
 Instead of keeping the status bar, I use \ttt{C-G} to know about the
@@
 Unfortunately, as far as I could dig, one will need to alter
-the Vim's source code to enable a horizontal split without losing a line.
-For me, it would be ideal for this feature to have a visual cue of the
-first and/or last line of the windows in the lines-number
-column, or complete the spaces and empty chars with \$\$\$\$ or so.
-
-%%%%%%%%%%%% TTM
+the Vim's source code to enable a horizontal split without losing a line,
+which, for me, seems far from the ideal.
+An example solution is to implement a visual cue for the
+first and/or last line of the windows in the line-numbers
+column, or complete the spaces and empty chars with \$\$\$\$,
+
 \subsubsection{On the persistence of the editor state}
@@
 The mappings in~\cite{vimrc} keep the sessions
-in a reasonable directory and makes it easier to remember and tweak
+in a reasonable directory and makes it easy to remember and tweak
 the standard commands to deal with sessions.
@@
 If you open a terminal inside Vim with the \tttt{:term}
-command, you will save the session as usual, but when loading
-you get dummy empty windows for them and an error message.
+command, you might save the session as usual, but when loading
+you get a dummy empty window for the terminal and an error message.
 More about Terminal-Job mode in Section~\ref{terminal}.
@@
 These are often related to particular file types, but are also often
-in defining the automated behavior of Vim.
+in defining some general automated behavior for Vim.
 If the autocommands are placed inside a configuration file (e.g. the
@@
     Section~\ref{sec:com}.
-  \item Ex mode: similar to command-line mode,
+  \item Ex mode: similar to the command-line mode,
   but more specialized for running various Ex commands.
@@
 There is another basic mode, but it is not fully implemented:
-the Terminal-Job mode (more in Section~\ref{terminal}.
+the Terminal-Job mode (more in Section~\ref{terminal}).
 There are seven additional modes which are mostly subordinate 
@@
 the normal mode is most powerful for
-navigating, manipulating texts and changing to other modes.
+navigating, manipulating text and changing to other modes.
 The simplest of these three is changing to other modes:
@@
     and next end-of-word.
-    There are motions to iterate over sequences of characters separated
+    These are motions to iterate over sequences of characters separated
     by special characters (e.g. punctuation and parenthesis) as
@@
 		\tttt{\},\{} through text blocks separated by empty lines.
-  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
+    \tttt{[} and \tttt{]} are used in combination to browse sections.
+  \item \texttt{(fX,tX,FX, TX)} to move, in the same line, to or just before any X character,
 	  \tttt{;} and \tttt{,} for next and previous found character.
-  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
+  \item Search with \tttt{/} or \tttt{?} (these are considered command-line commands).
   \item \texttt{CTRL+(o,i)} to move to an older or newer position in
     jump list.
-  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X:
+  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X.
     \texttt{`X} moves to the exact position while \texttt{'X} moves to
@@
 
-There are way more commands to change the texts.
+There are way more commands to change text.
 Some of them are discussed in Section~\ref{navIn}
@@
 Once in the insert mode, the character keys
-input the characters at the cursor position at the current buffer.
+input the characters at the cursor position in the current buffer.
 One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
@@
   \item \ttt{C-X} commands for scrolling the window with multiple
-    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
+    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities
+    (\ttt{C-X}\ttt{C-(F,I,L,O,V} for file, identifier, whole line, omni and command-line completions). 
 \end{itemize}
@@
 and the chosen command can be run with \ttt{CR}.
-Vimscript was largely based on the ex editor~\cite{ex},
+VimL is largely based on the ex editor~\cite{ex},
 and a more advanced user might use it for prototyping
@@
 follow documentation from \tttt{:h Ex-mode} for further information.
+
 I might be loosing something, but for
 tweaking I use the command-line window accessed through \tttt{q:}.
-I find it comfortable to browse the history in a normal-like mode,
+It is comfortable to browse the history in a normal-like mode,
 edit them also using the insert mode as usual,
@@
 customization by mappings have to be performed through autocommands as in~\cite{vimrc}.
+The facilities exposed in Section~\ref{runExc} might complete the picture
+that explains why this mode seems to be seldom used nowadays.
 
@@
 (see \tttt{:h terminal}).
-I find that it works exceptionally well and have used it to run
-scripts in an IPython shell, compile latex files, and even open PDFs and images.
+I find that it works exceptionally well and have often used it to run
+scripts in an IPython shell, compile latex files, and open PDFs and images.
 Vim browsing of windows and text manipulation is well developed,
@@
 \begin{itemize}
-  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the Terminal-Normal
+  \item \tttt{<C-W>N} and \tttt{<C-W>:} for entering the Terminal-Normal
     and command-line modes. Terminal-Normal mode is very similar to
@@
     mode, and the status line reports if the job is finished or not.
-  \item \tttt{<C-W>\_"} to paste a register.
-  \item \tttt{i} for entering the Terminal-Job mode from the normal mode.
+  \item \tttt{<C-W>"} to paste a register.
+  \item \tttt{i} or \tttt{a} for entering the Terminal-Job mode from the Terminal-Normal mode.
 \end{itemize}
@@
 I found the mappings on~\cite{vimrc} very helpful for directing
-editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
-which I make available across Terminal-Job and Normal modes.
-But I've been thinking on using \ttt{C-} commands also
-for tabs (not only for splits), both to avoid mixing typing sequences
-with commands, and to allow the use of the commands in all Normal,
-Insert and Terminal modes.
+editor focus to splits (\ttt{A-(h,j,k,l)}) and tabs (\ttt{A-(a,d)}),
+which I make available across Terminal-Job, Normal, and Insert modes.
+I used \ttt{C-(h,j,k,l)} for splits, and \tttt{g(r,t)} for tabs,
+for a long time (maybe years), but using \ttt{A-} makes it possible
+to use consistently the same commands across the modes without
+loosing any of the standard commands, and without using \ttt{leader}
+and \ttt{localleader}, which are often used by plugins.
 
 \subsection{Netrw}\label{netrw}
-The standard interface of Vim for browsing file trees is Netrw.
+The standard interface in Vim for browsing file trees is Netrw.
 It starts when you open a directory, such as with \tttt{:e .<CR>}.
@@
 ftp) and handy e.g. mappings to open the files as splits and tabs
-(specially \tttt{pot}).
-Most useful commands in netrw include:
+(specially \tttt{p,o,t}).
+Most useful commands in Netrw include:
 \begin{itemize}
@@
 
-There is no insert mode in netrw interface;
+There is no insert mode in Netrw interface;
 the commands in \tttt{:h netrw-explore} are
 convenient for opening the directory of the
-file being edited;
+file being edited (e.g. \tttt{:Se});
 further information is in \tttt{:h netrw}.
@@
 adding a line \tttt{:source <afile.vim>} in vimrc.
-In fact, it is on vimrc that one usually specifies the plugins
-and plugin managers they use.
+In fact, it is on the vimrc that one usually specifies the plugins
+and plugin managers to be used.
 Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
@@
 and pick convenient practices and settings.
-In fact, a vimrc file is most often a collage of excerpts of vimrc
+In fact, a vimrc file is most often a patchwork of excerpts of vimrc
 files from other users.
@@
 that downloads latest scripts from Sourceforge as specified in \tttt{:h getscript-data},
-and the Vimball interface, which creates and loads a Vimball for a
+and the Vimball interface, which creates and loads a Vimball representation of a
 plugin.
 Such a Vimball may be created with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
-that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
-The Vimball can be installed in a system by \tttt{:source <filename>.vba}
-or loading it at Vim startup with \tttt{\$ vim <filename>.vba}.
+that hold paths to files to be included in the \tttt{<filename>.vbm} Vimball.
+The Vimball can be installed in a system by \tttt{:source <filename>.vbm}
+or loading it at Vim startup with \tttt{\$ vim <filename>.vbm}.
+Vimball files used the extension \tttt{.vba} and most of the official and
+non-official documentation still use \tttt{.vba} but the extension now
+is \tttt{.vbm}.
 
@@
 It should be located inside a \tttt{pack/} directory
-somewhere in the directories listed with \tttt{:se runtimepath}.
+inside a directory listed in \tttt{:se runtimepath}.
 The plugins found in \tttt{pack/<packName>/*/start/} are loaded
@@
 
-All directories Vim looks for scripts are described
+All the directories in which Vim looks for scripts are described
 in \tttt{:h vimfiles} and are basically set by
@@
 Currently, Vim will download the files for a specific language if
-not found in system.
-
-\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}\label{script}
-In Vimscript,
+not found in the system.
+
+\subsection{Scripting, Functions, VimL and other languages (e.g.  Python)}\label{script}
+In VimL,
 the colon commands (also Ex commands or command-line commands) are related through spaces,
@@
 At the same time, the \tttt{:term} and terminal-job mode make
-scripting other software more convenient, as output is
+scripting in general more convenient (not only scripting Vim), as output is
 promptly navigated and copied (as discussed in Section~\ref{terminal}).
-
-\subsection{Color}\label{visual2}
+Command sequences to run only excepts of scripts, e.g.
+to (re)define a function or recalculate some last lines,
+are exposed in Section~\ref{runExc}.
+
+\subsection{Color for syntax highlighting}\label{visual2}
 Newer versions of Vim support 24-bit true color (aka 16 million colors)
@@
 \url{https://gist.github.com/XVilka/8346728}.
-Then Vim needs to be set to use true colors with
+Then Vim needs to be set to use true color with
 \tttt{:set termguicolors}.
 If using 8 or 16 bit colors, Vim uses the color palette from the
-terminal, if using true color each color is defined directly.
+terminal. If using true color, each color is defined directly.
 Settings for using true color inside Byobu/Tmux involve tweaking
@@
 for highlighting text in the Toki Pona language~\cite{tpLang}.
-Syntax highlighting plugins are as file type plugins,
+Syntax highlighting plugins are file type plugins,
 but also have a \tttt{syntax/<filetype>.vim} file
@@
 One might see every highlighting group, and their final visual results,
-with the commands
+with the commands\\
 \tttt{:so \$VIMRUNTIME/syntax/hitest.vim}
 or \tttt{:so hi}.
-The \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
+The\\ \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
 directory, but \tttt{syntax/} files has to be moved to
 \tttt{$\sim$/.vim/syntax/}, unfortunately.
+Routines for tweaking color schemes, available as a Vim plugin,
+is described in Section~\ref{realcolors}.
 
 \subsection{Fonts}
-The fonts are defined by your terminal or inside gVim.
+The fonts are defined by the terminal or inside gVim.
 \ttt{C-+} and \ttt{C--} can be used to change font size.
 Some settings for fonts, such as boldface, might be set using
-the syntax highlighting facilities described in the previous section.
+the syntax highlighting facilities described in the previous section,
+and are conveniently available through the mentioned plugin in Section~\ref{realcolors}.
 
@@
 and the reference to the standard Vim documentation
-to achieve the DRY KISS style described in Section~\ref{intro}.
-
-\noindent {\bf Potential enhancements} to this document include:
+to achieve the DRY KISS style described in Section~\ref{furRem}.
+
+\noindent {\bf Potential enhancements to this document} include:
 \begin{itemize}
@@
     and the resulting integrated environment.
-  \item Updating of the information I can find about the issues discussed,
+  \item Updating of the information I can find about the issues discussed here,
     such as about status lines in Section~\ref{visual1}.
@@
 \begin{itemize}
-  \item Measure the performance of text mining routines in Vimscript against those implemented in C or Python.
+  \item Measure the performance of text mining routines implemented in VimL against those implemented in C or Python.
+Compare Python performance while running inside Vim (through the Python interface) or as the standard standalone interpreter.
   \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
-  \item Better integrate Python and Vimscript, especially for data visualization
+  \item Better integrate Python and VimL, especially for data visualization
     and syntax highlighting management,
@@
 
-\noindent This is a selection of the issues that might entail
-plugins and that are more prominent for me:
+\noindent This is a selection of the {\bf issues that might entail
+plugins and that are more prominent for me}:
     \begin{itemize}
@@
         of sequence and how central are the keys.
-      \item Sessions, as described in Section~\ref{visual1}.
+      \item Sessions, as described in Section~\ref{visual1} and preliminarily implemented in~\cite{vimrc}.
       \item AA messages (shouts): to keep track, document and share of working sessions
@@
         slides and opening figures. Some Vim users asked for the settings and
-        commands. They are very elementary use of registers which are executed over
-        arbitrary but consistent textual patterns.
+        commands. They comprise a very elementary use of registers which are executed over
+        consistent textual patterns.
       \item For bringing back all the splits after an \tttt{:only}.
         Also for bringing back the tabs after a \tttt{:tabo}.
-      \item Run Python excerpts from a file being edited in an IPython shell. 
+      \item Run Python excerpts, from a file being edited, in an IPython shell. 
         The buffer number of the terminal window should be stored,
@@
         syntax highlighting scheme and dump it to a file (dumping
-        current highlighting scheme to a working Vimscript file
+        current highlighting scheme to a working VimL file
         is currently not supported by Vim!).
@@
 the window to a new tab using
-\tttt{:vs} and \ttt{C-W\_T}.
+\tttt{:vs} and \tttt{<C-W>T}.
 I then search for tokens related to
@@
 the knowledge I want to acquire.
-I go back to the previous tab with \tttt{gr}
+I go back to the previous tab with \ttt{A-a}
 and make a global replace with
@@
 I record in the \tttt{"q} register
-the sequence \tttt{jA.jj},
-and use it as a macro 10 times by
+the sequence \tttt{jA.<C-[>},
+and use it as a macro 10 times by typing
 \tttt{10@q}.
-I move to the other tab with \tttt{gt}
+I move to the other tab with \ttt{A-d}
 and open a terminal window with \tttt{:term}
@@
 If there is e.g. code or notes in other projects I am working
-on, I reach them through \tttt{<Alt-arrows>} because they
+on, I most often reach them through \tttt{<A-(arrows)>} because they
 are in separate Vim instances
 in Byobu/Tmux sessions and windows.
-Because I stay for hours editing (\LaTeX and Python
-scripts), I change the background to red, and eventually to green.
+Because I stay for hours editing (e.g. \LaTeX\ and Python
+scripts), I change the background to red, and eventually to green or to yellow using the Realcolors plugin described in Section~\ref{realcolors}.
 Because the blue color scheme does not highlight the spelling errors
-if using true color, I run
+if using true color, I would run
 \tttt{:hi SpellBad guifg=red guibg=lightblue}
-to see the words found wrong by the spell checker.
+to see the words found wrong by the spell checker,
+if this tweak was not already in my vimrc~\cite{vimrc}.
 
@@
 and the tab line with \texttt{\textbackslash\textbackslash T}.
-Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
+Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}
+(or \ttt{A-w} and \ttt{A-q}).
 The mappings for transitioning through splits and tabs
@@
 in more centraly located keys.
-The letters beggin unbounded mappings (require another character to
-trigger an action without having to wait for \tttt{:se timeoutlen}): g,z,[].
+The digits that beggin unbounded mappings (require another character to
+trigger an action without having to wait for \tttt{:se timeoutlen}): g,z,[,].
 
@@
 of the editor to splits in the \tttt{hjkl} directions.
+Althought it is a usable solution, \ttt{A-(hjkl)} might
+be more consistent with other uder-defined commands for navigation,
+and presearves the useful and standard \ttt{C-L}.
 
@@
 mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
-(or \tttt{Meta+x}).
+(or \tttt{Meta+X}).
 \ttt{A-X} is the same as \ttt{M-X} and refer to

### 2018-02-11 article/article.tex
moved plugin files to plugin dir, to allow vimball to get plugin/ and syntax/ tree

--- 2017-12-18 article/article.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{tipa}
 \newcommand{\ttt}[1] {
@@
 \newcommand{\tttt}[1]{\texttt{#1}}
+% \usepackage{fontspec}
+% \usepackage[mathletters]{ucs}
+% \newcommand{\ttt}[1] {
+% 	\texttt{<#1>}}
+% \newcommand{\tttt}[1]{\texttt{#1}}
+\usepackage{subcaption}
 
 \begin{document}
-\title{An anthropological account of the Vim text editor:\\
-features and tweaks after 10 years of usage}
+\title{Basic concepts and tools for the Toki Pona minimalist and
+constructed language:
+Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
 % Anthropological computation
-\author{Renato Fabbri\\
-\texttt{renato.fabbri@gmail.com}\\
+\author{Renato Fabbri, \texttt{renato.fabbri@gmail.com}\\
+Maria Cristina Ferreira de Oliveira, \texttt{cristina@icmc.usp.br}\vspace{0.2cm}\\
 University of São Paulo,\\
@@
 \begin{abstract}
-The Vim text editor is very rich in capabilities
-and thus complex.
-This article is a description of Vim
-and a set of considerations about its usage and design.
-It results from more than ten years of experience
-in using Vim for writing and editing various types of documents,
-e.g. Python, C++, JavaScript, ChucK programs;
-\LaTeX, Markdown, HTML, RDF, Make and other markup files;  % TTM
-binary files.
-It is commonplace, in the Vim users and developers communities,
-to say that it takes about ten years to master (or start mastering)
-this text editor, and I find that other experienced users
-have a different view of Vim and that they use a different
-set of features.
-Therefore, this document exposes my understandings in order
-to confront my usage with that of other Vim users.
-Another goal is
-to make available a reference document with which new users
-can grasp a sound overview by reading it and the discussions that
-it might generate.
-Also, it should be useful for users of any degree of experience,
-including me, as a compendium of commands, namespaces and tweaks.
-Upon feedback, and maturing of my Vim usage,
-this document might be enhanced and expanded.
+A minimalist constructed language (conlang)
+is useful for experiments and comfortable for making tools.
+The Toki Pona (TP) conlang is minimalist both in the vocabulary
+(with only 14 letters and 124 words)
+and in the $\approx10$ syntax rules.
+The language is useful for being a used and somewhat established
+minimalist conlang with at least hundreds of fluent speakers.
+In this article, we describe current concepts and resources
+for TP,
+and make available Python scripted routines for
+the analysis of the language,
+the synthesis of texts, the specification of syntax highlighting
+schemes, and the achievement of a preliminary TP
+Wordnet~\cite{wordnet}.
+We focus on the analysis of the basic vocabulary,
+as corpus analyses were found~\cite{corpus}.
+The synthesis is based on sentence templates,
+relates to context by keeping track of used words,
+and renders larger texts by using a fixed number of phonemes (e.g. for poems)
+and number of sentences, words and letters (e.g. for paragraphs).
+Syntax highlighting 
+reflects morphosyntactic classes given in the official dictionary
+and different solutions are described and implemented
+in the well-established Vim text editor~\cite{vim}.
+The tentative TP Wordnet is made available in three forms that
+reflect the choices of the synsets related to each word.
+In summary, this text holds potentially novel conceptualizations about,
+and tools and results in analyzing, synthesizing and syntax highlighting the
+TP language.
+  % subj (li), obj (e), prepositions, qualification (word concatenation), 
+  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
 \end{abstract}
+{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona
 
 \section{Introduction}\label{intro}
-Vim is a very complex editor,
-most often considered to be
-matched only by Emacs.
-They both are the standard advanced text editors
-of the free software and open source communities
-and have been developed for decades.
-Vim is very useful because:
+Toki Pona (TP) is a minimalist conlang (constructed language)
+with only 124 words (120 without the synonyms).
+Therefore, the concepts are usually very general and different,
+and, without context, the words are rarely related through
+meronymy and hyponymy.
+Such a linguistic setting is desired because of the simplicity
+which entails easier e.g. learning and tool making.
+Another reason why the minimalist language design is compelling
+is the study and harnessing of the strong and weak forms of
+the Sapir-Whorf hypothesis (linguistic relativity),
+i.e. that language influences or dictates thought
+and world experience~\cite{sapWho}.
+Accordingly, one uses a conlang as a thinking tool (or platform)
+or to make experiments about the influence language has on
+the thoughts of the `speaker' (also writer and reader).
+Toki Pona is often described as a tool to meditate,
+to simplify the thinking processes, and as a way to modify
+the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
+In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
+describes that she has seen the language been used successfully
+in the context of management, creation of texts, legal texts,
+etc.
+
+In this article, we present a conceptual overview of the language,
+which is fit both to the newcomer and to the expert
+for being considerably different
+from what we found in the literature~\cite{tpLang,kama,memrise},
+with emphasis on simplicity and flexibility.
+We also present novel software routines for analysis, synthesis,
+syntax highlighting, and the achievement of a tentative TP
+Wordnet~\cite{wordnet}.
+
+Next subsections hold a description of the general resources available,
+a historical note, and some words about natural and constructed languages.
+Section~\ref{basics} describes TP's phonology and syntax.
+Section~\ref{hacks} presents the software routines we made available
+and immediate results, such as listings and statistics of words,
+poems and short stories, coloring schemes, and preliminary TP Wordnet
+versions.
+Conclusions and further work are in Section~\ref{conc}.
+Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
+with thoughts about rule breaking and potentially new conlangs.
+Appendix~\ref{ftp} holds final words in Toki Pona.
+
+% definition, historic, status (groups, speakers, etc).
+
+\subsection{Resources on Toki Pona}
+One might organize current resources for the Toki Pona
+language in: references and learning material, corpus,
+websites, interaction groups
+(where users talk a post texts and comments), and software gadgets.
+The main references of the language are:
+  the official book ``Toki Pona: The Language of Good''~\cite{tpLang},
+    authored by Sonja Lang, the creator of the language;
+and the online book ``o kama sona e toki pona!'',
+from jan Pije~\cite{kama}.
+A number of tools and other resources for dealing with Toki Pona
+were developed by the community.
+The most complete list is supposedly~\cite{gdoc} and includes
+videos, musical pieces, artistic texts, reference documents (e.g.
+a Toki Pona - Esperanto dictionary), journalistic articles,
+and software tools.
+For a more comprehensive view of the resources available
+for the user, we suggest following the links 
+in~\cite{gdoc,wikiToki}.
+
+\subsection{Historical note}
+TP was developed as an internal and personal language
+by Sonja Lang~\cite{interview}.
+It was released as a draft in 2001 and in 2007 some documents
+reported it to have a few hundred speakers.
+The English official book~\cite{tpLang} was released only in 2014.
+In 2016, a version of the official book was released in French.
+Nowadays, one finds a number of texts about TP and written
+in it
+(e.g. in social platforms such as Facebook groups,
+microblogging, Telegram and
+IRC), and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.
+
+\subsection{Natural and constructed and artificial languages}
+A language one uses (or might use) to communicate by speaking
+and writing is called `natural language'.
+A `constructed language' (also planned or invented language) is a natural language built by someone or
+a group, such as Esperanto, Toki Pona, and Lojban.
+An artificial language is a language yield by artificial agents,
+such as by AI routines, or by humans in controlled experiments,
+and are considered e.g. within `cultural evolution' studies.
+Formal languages are defined by tokens and rules to
+operate them, they span from computer programming languages
+to math and formal models for natural languages.
+
+Constructed languages are in some traditions called artificial
+languages, but creators most often prefer to use the term
+`standardized', 'constructed'
+or 'planned' language with the argument that the conlang
+is rooted on natural languages,
+and 'artificial' is misleading.
+The preferred terms seem to be planned or constructed language
+or simply conlang.
+The construction of languages is called glossopoeia.
+TP is a conlang which might be classified
+as engineered for experiments, meditation and philosophy;
+and suitable for use as an auxiliary international language and
+as an artistic language.~\cite{conlanWikip,artlanWikip}
+% taxonomy, vocabulary, ontology, terminology,
+
+% conlangs, natural langs, programming langs, Lojban as an
+% intermediary
+% non-verbal languages, musical corporal language, 
+% non-verbal languages confront the strong form of the Sapir-Whorf hypothesis.
+
+\section{Overview of the language}\label{basics}
+This section describes very succinctly the formation of
+words and sentences in the Toki Pona language.
+It should enable a newcomer to grasp the essentials
+of the language and the experienced to
+acquire new insights.
+Furthermore, it is a solid reference of the phonological and syntax
+rules, and enables one to understand and modify the software presented
+in Section~\ref{hacks}.
+
+\subsection{Phonology}\label{phonology}
+Words in Toki Pona are written using only 14 letters:
 \begin{itemize}
-  \item it is meant to be a plain text (e.g. ASCII, UTF-8) editor
-  and does not (by standard) insert special characters (e.g. for
-    formatting or with binary instructions).
-  \item It has a powerful architecture and set of commands.
-  \item It is highly configurable and most often the users
-  hold a set of commands for standard settings and routines kept in
-    the vimrc~\cite{vimrc} and other configuration files.
-  \item It has been used for more than 25 years, is based on
-    established technologies, and is an established text editor.
-    Thus, it has a considerable and seasoned user
-    base, high quality (both official and unofficial) documentations, 
-    and countless publicly available scripts, most often in the form of
-    plugins.
+  \item Vowels a (open), e (mid front), o (mid back), i (close front),
+    u (close back).
+  \item Consonants j, k, l, m, n, p, s, t, w:
+    \begin{itemize}
+      \item Nasal: m (labial), n (coronal).
+      \item Plosive: p (labial), t (coronal), k (dorsal).
+      \item Fricative: s (coronal).
+      \item Approximant: w (labial), l (coronal), j (dorsal).
     \end{itemize}
-
-This document describes the Vim text editor
-and proposes a set of sharpening of the user
-experience through simple tweaks
-and utilization strategies.
-The contents herein presented is a
-report on the overall understandings I
-have of Vim after a bit more
-than ten years using it,
-resorting to the (very mature) official documentation
-whenever possible.
-The purposes of this document are:
+\end{itemize}
+
+There are standard guidelines for pronunciation,
+but the language allows for considerable allophonic
+variation.
+For example, /p t k s l/ might be pronounced
+[p t k s l] or [b d g z \textipa{\!R}].
+Especially for poetry, one might consider
+j and w to be vowels
+(e.g. j as 'i' and w as 'u').
+
+Syllables are of the form (C)V(n):
+an optional consonant, a vowel and an optional coronal nasal consonant
+(n).
+Non word-initial syllables must follow the pattern CV(N).
+These sequences are forbidden: ji, wu, wo, ti, nm, nn.
+
+\subsection{Syntax}\label{syntax}
+\subsubsection{Fundamental notions}
+As in other natural languages, 
+colloquial TP might have
+incomplete sentences and deviate from the
+norm.
+The basic structure of sentences is:
+`subject' (Noun) li `predicate' (Verb) e `object' (Noun).
+The li might be repeated to associate more than
+one predicate to the subject.
+The particle li is omitted if the subject is a simple mi (I or us)
+or sina (you). A discussion about issues (potential problems)
+yield by this last rule
+and how I deal with them is in Appendix~\ref{mytoki}.
+The e might be repeated to associate more than
+one object to a predicate.
+Sentences might be related through la,
+'sentence' la 'sentence', where the second sentence is
+the main sentence, and the first sentence is a condition
+to the second.
+Multiple la-s are not described in literature,
+but one might assume that the last sentence
+is a conditional to the next,
+except in cases where the context strongly suggests
+otherwise.
+
+Noun and verb phrases are (usually) built with the non-particle words.
+The first word is the noun or verb and subsequent words
+qualify the noun or verb (i.e. are adjectives and adverbs).
+The pi particle might be used to separate sequences of words
+to be evaluated before the relation yield by pi.
+As pi is often ill understood and used,
+the following structures might be handy for newbies and as a
+reference:
 \begin{itemize}
-  \item to help new users grasp Vim essentials
-  and convenient practices.
-  \item To attain a sound overall description of the editor.
-  \item To record the comprehension about the editor that
-  a user (me) has after $10$ years of usage.
-  \item To confront my usage with that of other experienced
-  users. This is helpful for me, but also for the other users
-  as they might benefit from this content and from discussions
-  that might arise from it.
-  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
+  \item No pi, `word word word':
+word $\leftarrow$ (qualifies 1) word $\leftarrow$ (qualifies 2)  word.
+  \item One pi, 'word pi word word': word $\leftarrow$ (qualifies 2) [
+      word $\leftarrow$ (qualifies 1)  word ].
+  \item Two pi-s: `word pi word word word pi word word':
+    word $\leftarrow$5 [word 2 word] 3 word $\leftarrow$4 word 1  word;
+or:
+    word $\leftarrow$5 [word 1 word] 2 word $\leftarrow$4 word 3  word.
 \end{itemize}
 
-Advanced users might just skim through
-Section~\ref{basics}, where standard capabilities of
-the editor should become clear, and consider more carefully
-Section~\ref{namespaces}.
-The concluding remarks and proposed enhancements in Section~\ref{conc}
-may also deserve some attention because
-Vim is constantly evolving
-and there are many possible enhancements
-(often made available e.g. as plugins).
-
-\subsection{Further remarks about this document}
-This document is written is a DRY KISS
-(Don't Repeat Yourself, Keep It Simple Stupid) style.
-Complex is to master the use of Vim
-and one finds sound
-references in help files and a nice vimrc.
-Therefore the following content is kept
-as uncomplicated and original as possible.
-Also, because of Vim's complexity and entailed
-bond of this document to my usage,
-there is an anthropological component
-which is evident in the occasional use of the first person.
-This can be understood as anthropological computer science~\cite{anPh,anPh2}
-and considered to help
-in the technological groundwork of the civil society.
-Accordingly, here are some notes about my experience:
-I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
-I used Vim for writing and editing computer code 
-(Python, Javascript, C++, ChucK, bash, etc), markup languages  % TTM
-(HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
-Eventually, I edited database files and other types of files.
-Within Vim, I mostly write (web and scientific) software,
-scientific articles, music, poems and short stories.
-
-% Vi and Vim
-\subsection{Historical note}
-Vim was first released publicly in 1991.
-It is a cross-platform GNU licensed free and open source
-extended clone of Bill Joy's vi text editor.
-Vi was written in 1976 as a hard link to ex: a shorthand
-to start ex in visual mode, i.e. vi is ex.
-Vim's development is coordinated (and performed) since the beginning
-mainly by Bram Moolenaar.
-Today, current bleeding-edge version is 8.0.1401.
-I found no explicit stable, alpha or beta versions.
-I found no scientific article on Vim 
-(this might be the first one), although there are
-books, software and third-party documentation on the
-web.~\cite{wikiVim}
-
-\section{Basics}\label{basics}
-Vim's interface is text-based.
-In the GUI (gVim),
-there are convenient menus and toolbars
-but all functionalities are still available though
-the command line mode.
-Vimscript is the internal language of Vim,
-and is often used for scripting by users
-although other languages might be used 
-(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
-Each line of a Vimscript is a command on the
-command-line mode.
-This section may be considered a tutorial
-that focuses on the namespaces, i.e. sets of tokens
-that carry values or triggers procedures.
-One should see Appendix~\ref{sec:not}
-to understand the notation Vim uses.
-
-\subsection{The bare minimum}\label{minimum}
-You open a file at Vim startup by executing
-the command: \texttt{vim <filename>}.
-Inside Vim, you start in the normal
-mode, and might want to move around using
-\texttt{h-j-k-l} for left-down-up-right.
-To insert characters, move your
-cursor to the desired location an press i,
-which puts Vim in the insert mode.
-Go back to normal mode by pressing
-\texttt{<ESC>}, \ttt{C-$[$} or \texttt{<C-C>}.
-You save the file by typing \texttt{:w<CR>},
-and exit Vim by typing \texttt{:q<CR>}.
-You can save and quit with \tttt{:wq<CR>}
-or \tttt{:x<CR>} or \tttt{ZZ}.
-
-\subsection{Vim help}
-Help on using Vim is found in various places.
-The standard resource is the Vim help files.
-They are accessed by typing \texttt{:h <anything><CR>}
-in normal mode.
-Examples of such \texttt{<anything>} are:
-color, navigation, :vs, vimtutor.
-Type 
-\texttt{:h usr\_toc<CR>}
-to access the official User Manual,
-which is considerably lengthy and complex
-and is usually not read by users before a few years
-of experience.
-In learning Vim, one
-might want to run the \texttt{vimtutor} command
-(outside Vim) to start the Vim Tutor.
-
-There are good resources on the Web for learning
-and tweaking Vim:
+\noindent Further notes on the usage of pi:
 \begin{itemize}
-  \item ``Vim Adventures'' is an online RPG game for practicing
-  and memorizing Vim commands. This game is quite famous among Vim users.
-  \item There are official and semi-official Vim sites e.g.:
-  \url{www.vim.org}, \url{https://www.vi-improved.org} and
-  \url{http://vim.wikia.com/}.
-  \item Many hacks, understandings and general issues
-  (e.g. how to make such a move) are asked and answered
-  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
-  One often finds these links through a search engine.
-  \item Many videos about Vim are publicly available.
-  One traditional site is \url{http://vimcasts.org},
-  but you might find them using a search engine
-  (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
+  \item In a sequence of words, without pi, the second word qualifies
+    the first, the third word qualifies the phrase yield by the first
+    two words, the fourth word qualifies the phrase yield by the first
+    three words and so on.
+  \item It is redundant to use pi before the last word in a noun or
+    verb phrase, reason why it is most often
+    omitted.
+    Its use in this case is considered an error~\cite{tpLang,kama},
+    but, as one might notice, it does not add (much) information
+    through syntax because the order of qualifications is conserved.
+    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
+    Also used in~\cite{akesiWawa}.
+  \item The book by jan Pije~\cite{kama} describes another use for pi:
+    after li to mean possession, e.g. `soweli li pi sina' (your pet).
+    This employment of pi might be regarded as correct, but is promptly written
+    as a noun phrase (e.g. `soweli sina') and is not mentioned
+    by the official book~\cite{tpLang}.
 \end{itemize}
 
-% any other commands? files?
-\subsection{Namespaces}\label{namespaces}
-Vim is a text editor ouroboros~\cite{ouroWiki} because
-text and writing alters text and writing.
-What you end with is a collection of namespaces where tokens have scalar
-or arbitrarily complex values.
-
-\subsubsection{Commands and mappings}
-There are commands, i.e. typing sequences which trigger automated actions,
-for each mode:
+All the words except the particles (li, e, la, pi, a, o, anu, en, seme, mu)
+and the words classified solely as
+prepositions (kepeken, lon, tan)
+are usable after any noun or verb phrase,
+(i.e. 107 words discarding synonyms).
+Notice that the phrase expresses a noun (in a subject or
+object) or a verb (in a predicate).
+And that the first word of the phrase is the noun or verb,
+and that subsequent words are adjectives or adverbs.
+The pi precedes a noun or a verb to be qualified and then qualify
+the phrase that comes before it.
+
+At this point, the only missing syntax rule is related to
+the prepositions: kepeken, lon, sama, tan, tawa.
+They might appear at the end of a phrase,
+should be followed by another phrase,
+and require no particle. E.g.
+`toki *tan* jan Pije li pana e sona *tawa* mi'.
+Because prepositions can be used as adjectives
+or nouns or verbs or adverbs, etc, and might follow
+any noun or verb phrase,
+they are one of the three main sources of ambiguity:
+prepositions; absence of li after sina and mi; incomplete sentences.
+E.g. '(mi [li]) pana tawa kon e ilo pi suli mute'.
+
+\subsubsection{Particles}\label{sec:par}
+Beyond the structural particles (li, e, la, pi) presented
+in last section, other particles are:
+
 \begin{itemize}
-	\item in Normal mode all keys are mapped to commands.
-		There is redundancy and additional commands
-		using Ctrl and Shift keys.
-		Some keys expect a second key,
-    and have combinations not used (thus available for new mappings),
-		specially the z and g.
-    See Section~\ref{normal} and Appendix~\ref{notes}
-    for more insights into the commands available in the
-    normal mode.
-	\item In the other modes, the sequences available for mappings are more obvious and abundant.
-		One should look at \tttt{:h index} to know about all the standard mappings
-		and use \tttt{:map} to list the user-defined mappings.
+  \item a or kin, emphasis.
+  \item o, vocative or imperative ('jan lukin sitelen o, li wawa')
+  \item taso, means however as sentence or 'only' if adjective.
+  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
+    For verbs, repeat li.
+    For object nouns, repeat e.
+    If the noun is complementing a preposition (e.g. tawa, lon),
+    one might repeat the preposition.
+    As TP is a recent language, and is particularly able to cope with
+    variation due to its simplicity, I would advocate for
+    using en and anu/en wherever there is no ambiguity.
+    E.g.: 'mi tawa mute anu moku lon tenpo lili'
+    In the official documentation that is not described and
+    might be regarded as wrong in strict canonical TP.
+  \item nanpa, denotes numbering.
+  \item seme, for questions, used next to the thing being asked for.
+    'tan seme la sina pana e sike?'.
+  \item mu, for animal noises. For me it is not a particle, as in the
+    official dictionary, but a noun.
+    I also like to use it as a verb:
+    'mi pakala e luka. mi mu mute.'
 \end{itemize}
 
-\ttt{C-\textbackslash} is often reserved for extensions,
-which makes it a safe namespace to use (while there are no
-such extensions).
-
-A colon command can be written as a string
-and executed by the \tttt{:execute} colon command.
-E.g. \tttt{:execute 'vs afile.txt'}.
-As there are colon commands that execute commands in other
-modes, e.g. \tttt{:normal ?\textasciicircum def},
-the \tttt{:execute} is a way to build commands in any mode,
+The vocabulary specifies these morphosyntactic classes:
+nouns, adjectives, verbs, pre-verbs, adverbs, prepositions, particles, and numbers.
+I find that it might help the speaker, specially the newcomer, but
+it might also suggest a deviation from what I understand and read:
+the words might be used indistinctly to be a noun
+(the first word of: a subject, a predicate when there is no object, an object,
+or a prepositional complement),
+an adjective (anything that follows an noun and is not a particle or a
+preposition),
+a verb (follows mi or sina or li),
+or an adverbs (follows a verb).
+The pre-verbs (wile, ken, awen, kama, lukin, sona),
+might be followed by a verb, but might also be understood
+as a verb qualified by the next word,
+which carries a very similar if not identical meaning
+('wile moku' might be understood both as a pre-verb followed
+by a verb and as a verb followed by an adverb).
+The pre-verb words are all also specified in the official dictionary
+to have other morphosyntactic classes,
+such as adjective, noun, and verb.
+The only exception is wile, which is specified to be only a pre-verb.
+
+Thus, the classes given in the dictionary dictate little
+in practice:
+jan kala li lape lon ni.
+Where kala, lape and ni are in this phrase
+as adjective, verb and noun,
+and are in the dictionary as noun,
+adjective and adjective.
+
+\subsubsection{Recognizing POS tags by speaker and machine}\label{sec:rec}
+One might perform a POS tagging by following these rules:
+\begin{itemize}
+  \item Noun: the first word in a noun phrase.
+    Starting a sentence if it is complete,
+    after an 'e', after a pi.
+    The predicate is often a noun
+    if the sentence has no object.
+  \item Adjective: second word on after an e and after
+    second word on in the subject phrase if present.
+  \item Verb: after a li, mi or sina.
+    If there is no object, the predicate
+    might be a noun or a verb.
+  \item After a preposition, one might consider that what follows is 1) always
+    a noun, or that 2) it can be a noun, adjective or verb,
+    or that 3) it is a hybrid of noun, verb and adjective.~\cite{janKipo}
+  \item Notice that there is ambiguity in the structure
+    introduced by the omission of li after mi and sina.
+    Also, when there is no object, a noun or a verb
+    or an adjective might be in the verb position.
+    % The prepositional complement is also not defined.
+    % (mi moku tawa pali, tawa tomo). % TTM
+    These are sources of syntactic ambiguities
+    in TP.
+    They might be resolved or minimized by using the semantics
+    of the words.
+    An initial effort in this direction might be
+    using the classes in the official dictionary to resolve
+    ambiguities whenever possible.
+    This solution is not optimal for correct POS tagging,
+    and does not solve all possible ambiguities
+    (there are words classified as noun and adjective,
+    adjective and verb, noun and verb, particle and verb).
+    Another source of ambiguity is the pre-verbs as described
+    in the literature~\cite{tpLang,kama}.
+    But I find it reasonable to understand them as verbs,
+    as described in the end of Section~\ref{sec:par}.
+  \item If there is no object, the predicate might be parsed
+as being a hybrid of noun, verb and adjective.
+\end{itemize}
+
+% Also, the prepositional complement might
+% be a noun or a verb.
+% I could not come up with a sentence
+% where it would be understood as an adjective. TTM
+
+\subsubsection{Additional notes}\label{sec:add}
+John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
+states that there are not noun, verb, adjective (or even prepositional)
+phrases in TP, but only phrases in a structure
+yield by particles and content words~\cite{janKipo}.
+The only synonyms on Toki Pona are:
+a and kin; lukin and oko; sin or namako;
+ale or ali.
+In formations such as
+toki e ni:, wile e ni:, tan ni: etc.,
+'(e) ni' might be omitted and : used alone.
+Names are by default transliterated,
+but might not be.
+These and other issues are further described in Appendix~\ref{mytoki}.
+
+% my view
+Many other considerations might be made about the language,
+such as additional (deprecated and new) words,
+situations where li is used and avoided,
+canonical and alternative employment of pi,
+counting systems,
+associations of TP word sequences to words in English,
+usage of images to represent TP words and sentences,
+and best practices in writing TP texts or for translation.
+For these and other matters, visit~\cite{tpLang,kama,tp4,gdoc}.
+
+% considerations about the official book and jan Pije
+% pictorial forms
+\section{Software for analysis, synthesis, and syntax highlighting}\label{hacks}
+In this section,
+we describe software, statistic, natural language processing,
+and visualization results:
+1)
+the analysis of the Toki Pona language through statistics
+of the vocabulary obtained by processing the official dictionary;
+2)
+the synthesis of sentences, poems and paragraphs (e.g. short stories)
+in Toki Pona;
+3)
+syntax highlighting for Toki Pona,
+including fine-tuning and theoretical
+considerations;
+4)
+an initial Wordnet~\cite{wordnet} of
+Toki Pona words.
+Corpus analysis was already found in~\cite{corpus}.
+The Python scripts (or the toolbox)
+for obtaining all the results in this sections,
+and more, is publicly available in~\cite{tokipona},
+to facilitate both the inspection of the results and the
+generation of derivatives by other interested parties.
+Such scripts were made very friendly for one to understand and alter
+at will.
+
+\subsection{Statistics of the vocabulary}\label{sec:stat}
+This section focuses on the statistics of the vocabulary
+and of the syntactic rules:
+the letters, phonemes, word sizes,
+possible combinations for words and sentences.
+% The statistics are in Appendix~\cite{listings},
+% and the next paragraph is an overview. TTM
+The script \tttt{makeStatistics.py}
+was used to obtain all the measurements and tables
+discussed herein,
+and hold further measurements and sets of words
+regarded as less suitable for this exposition,
+but is useful for a deeper investigation.
+
+As described in Section~\ref{basics},
+there are only 14 letters,
+and phonemes are also very restricted.
+There are 120 different words in the official vocabulary,
+4 of them having synonyms.
+A total of 124 tokens (or lemmas),
+not counting proper nouns (names)
+and punctuation.
+Table~\ref{foobar} shows the number of words
+related to each POS tag (or morphosyntactic class)
+specified in the dictionary.
+
+\input{pos}
+
+From the official 124 words, 
+26 of them (20.97\%)
+have only one syllable,
+85 (68.55\%) have two syllables,
+and 13 (10.48\%) have three syllables.
+No official word has four or more syllables.
+In all the 124 tokens, there are 235 syllables (68 different),
+and Table~\ref{tab:freqsy} exhibits the 10 most frequent
+syllables in the first, last, middle and all positions.
+Middle-word syllables only occur 13 times,
+and are all different, with the exception of 'la'
+which occurs twice.
+A complete list of the syllables and their frequencies (in different
+positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
+script.
+A list of all words, grouped by their size in syllables,
+and ordered alphabetically and by the number of letters,
+is in the \tttt{hlsyl\_\_} variable.
+
+\input{syls}
+
+Vowel and consonant frequencies,
+and comparisons of vowels against consonants,
+are as shown in Table~\ref{freqLet}
+for start, end and internal position.
+The limited number of consonants favors
+the naturalness of the language prosody
+as it resembles babbling.
+
+\input{vowels}
+
+Given the rules in Section~\ref{phonology}
+(of 14 letters, 5 vowels, (C)V(n) phonemes,
+forbidden ji, wu, wo, ti, nn, nm),
+96 words are possible with 1 syllable,
+8256 with 2 syllables, 710016 with 3 syllables.
+In the incident words of the official dictionary,
+no middle syllable end with the nasal consonant 'n'.
+
+Possible sentence structures are unlimited.
+For a noun or verb phrase, one might quantify the possibilities
+by considering all words but the particles that are not classified
+also as something else
+(i.e. li, e, la, pi, a, o, anu, en, seme, mu)
+and the prepositions that are not classified also as something
+else (i.e. kepeken, lon, tan).
+This yields $120-13=107 words$.
+As discussed in Section~\ref{syntax},
+I advocate that, although such words are classified in the dictionary,
+they might be used indistinctly as nouns, adjectives,
+verbs and adverbs.
+Thus, one has 107 possibilities of noun and verb phrases with
+one word, $107^2=11,449$ possibilities with two words,
+$107^3=1,225,043$ with three words and so on.
+Be $n$, $v$, $o$, and $p$ the number of words in the subject (noun), predicate
+(verb), object (noun) and prepositional (see Sections~\ref{sec:rec} and~\ref{sec:add}) phrases of a sentence,
+and assume that the sentence has at most one prepositional phrase,
+one might quantify the possibilities using the formula:
+$\delta = 107^n\times 107^v\times 10^o \times 5\times 107^p$,
+where $5$ stands for the possible prepositions.
+To account for the particles, one possibility is
+to assume for simplicity that one might use none or one particle
+(not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
+at each phrase, yielding:
+$\delta = 107^n\times 107^v\times 107^o \times 5\times 107^p\times 9^4$.
+For example, assume $n=v=o=p=1$, then
+$\delta=107^1\times 107^1\times 107^1 \times 5\times 107^1\times 9^4=
+ 4.300066\times 10^+12$,
+i.e. more then 4 trillion possible
+sentences with single word phrases (i.e. no adjectives or adverbs),
+while allowing only one particle per phrase
+and only one predicate phrase
+(e.g. 'ona li moku e soweli lon supa').
+The general case is implemented as a function
+in the script \tttt{makeStatistics.py}~\cite{tokipona}.
+
+\subsection{Synthesis of text}\label{synth}
+Such counting exercises are also useful
+for (semi-)automated writing through scripting.
+The syntax organizes the words in larger structures.
+The rhymes are very restricted and sonorities are
+bounded by the small vocabulary and simple syntax.
+There are some specific tasks for achieving texts,
+such as finding the number of syllables considering the elisions,
+or handling interaction of the writer with the script
+to choose sentences or verses or stanzas.
+The same package~\cite{tokipona}
+has capabilities for synthesizing TP text.
+On its most basic level, these routines yield
+noun, verb and prepositional phrases,
+and sentences.
+They also aim at making larger scale texts
+by assuring the use of specific words (to entail context),
+e.g. to assist the creation of short narratives,
+and by employing stylistic outlines for poems.
+
+\begin{figure}     \center
+  \begin{subfigure}{.5\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/phrase_}\vspace{-0.15cm}
+          \caption{Synthesized phrase.}
+            \label{fig:sfig1}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.9\textwidth}
+    \centering
+        \includegraphics[width=1.2\linewidth]{figs/sentence_}\vspace{-0.15cm}
+          \caption{Synthesized sentence.}
+            \label{fig:sfig2}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.8\textwidth}
+    \centering
+        \includegraphics[width=1.2\linewidth]{figs/paragraph_}\vspace{-0.15cm}
+          \caption{Synthesized paragraph.}
+            \label{fig:sfig2}
+  \end{subfigure}\\\vspace{0.3cm}
+  \begin{subfigure}{.7\textwidth}
+    \centering
+        \includegraphics[width=.8\linewidth]{figs/poem_}\vspace{-0.15cm}
+          \caption{Synthesized poem.}
+            \label{fig:sfig2}
+  \end{subfigure}
+  \caption{Toki Pona texts synthesized by the \tttt{makeTexts.py} script
+  in~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
+  and (d) a poem.
+  The unusual word combinations are convenient for exploring semantic
+  possibilities.
+  One might obtain many texts to select the excerpts
+  she finds fit for the intended use.
+  For more information, see Section~\ref{synth}.
+  The words are colored in accordance with the considerations
+  in Section~\ref{shigh}.}
+  \label{fig:syn}
+\end{figure}
+
+Figure~\ref{fig:syn} holds some synthesized texts
+with different color schemes for the syntax highlighting,
+as presented in the next section.
+The textual synthesis described here,
+and implemented in \tttt{makeTexts.py},
+might be enhanced in countless ways,
+e.g. as described in Section~\ref{conc}.
+
+Automatic and randomized synthesis of texts in TP is particularly
+useful because of the reduced vocabulary where each word is related
+to a broad semantic field.
+Ideas often make sense in unexpected ways, and thus
+the synthesis yields a procedure to 
+explore the semantic possibilities within TP.
+One might object that the resulting texts are unusual 
+and even consider that they often hold insubstantial or unsound meaning.
+I advocate that these unexpected formations are desirable
+for exploring the possibilities of the language and of thought,
+and for artistic endeavors.
+Also, to obtain texts which one finds usual or satisfactory
+in more strict or personal terms, such a person might just write
+them normally.
+
+% phrase
+% sentence
+% poem
+% text
+% each with a different color scheme
+
+% nanpa followed by numbers
+% pu, taso, kin/namako
+% particles
+% rhymes 
+% modes in which the words are only used as specified in the dictionary
+
+% \clearpage
+\subsection{Syntax highlighting}\label{shigh}
+The same package~\cite{tokipona}
+has a Vim~\cite{vim} syntax highlighting plugin
+for Toki Pona,
+routines to arrange the coloring schemes
+(i.e. to yield the text files with the specifications for which tokens
+to relate to which coloring groups),
+and instructions for installing and using
+the resulting syntax highlighting for Toki Pona texts~\cite{tokipona}.
+An online syntax highlighting gadget is found at~\cite{tpNetSH},
+and the solution here described presents a number of enhancements:
+it is capable of coloring all morphosyntactic classes;
+it might be fine-tuned in the colors and their relations to sets of
+tokens; it is designed to be used within Vim
+and might be exported as HTML; the highlighting scheme is promptly
+rendered by a script; the resulting script might be used in combination
+with virtually any color scheme.
+
+Basically, the resulting syntax highlighting
+distinguishes the words among the morphosyntactic
+classes according to the official dictionary as given in
+Table~\ref{foobar}.
+A word often belongs to more than one class,
+thus precedence might be set by the user.
+Also, some classes might be further refined (e.g. words beginning with
+vowels) or joined (e.g. distinguishing only particles and the rest,
+or particles and prepositions and the rest).
+The colors are also promptly changed according to~\cite{vim}
+and exemplified in the package documentation.
+
+Currently, the Python package synthesizes the
+syntax file through the\\ \tttt{makeVimSyntax.py} script.
+The user has control of class precedence and
+merging and further details through tweaking such routine.
+The choice of precise coloring schemes
+involves fine tuning the color scheme being
+used in Vim (such as 'blue', 'elflord' and 'gruvbox'),
+and Vim's highlighting schemes as described in~\cite{vim}.
+In summary,
+the usage of the package and plugin might be performed
+through the following actions:
+\begin{itemize}
+  \item Installation of the plugin.
+  \item Tweaking of the syntax file by hand.
+  \item Running the \tttt{makeVimSyntax.py}
+    Python script to generate a new \tttt{tokipona.vim} syntax file
+    according to other settings.
+  \item Write a file with the \tttt{.tokipona} or \tttt{.tp} extension
+    (inside Vim) using Toki Pona words.
+    Reload the highlighting scheme using \tttt{:e} whenever you
+    change the syntax file by hand or through the Python script.
+  \item Access the used highlighted groups with \tttt{:syntax}.
+    Access all the highlighting groups with \tttt{:so
+    \$VIMRUNTIME/syntax/hitest.vim} or \tttt{:hi}.
+    Change the coloring of a set of terms by associating
+    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g.
+    Visual) such as in \tttt{:highlight link tokiponaADJECTIVE Visual}.
+\end{itemize}
+
+The syntax file has an association of TP
+words with highlighting groups.
+It also holds associations of these groups to other groups
+to use their color settings.
+Therefore, one is able to use various color schemes
+in syntax-highlighted TP texts, such as in Figure~\ref{fig:syn}.
+A Vim user might run e.g. \tttt{:colorscheme blue},
+\tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox} or
+\tttt{:colorscheme elflord} to see the same text colored
+with different color schemes (association of colors to sets of tokens).
+To interfere directly on the colors chosen,
+\tttt{:highlight Normal guifg=\#00000 guibg=\#0000ff}
+will change the standard foreground (text) color to pure black
+and background to pure blue.
+See \tttt{:h gui-colors} and \tttt{:h highlight}
+for the way you might edit colors directly.
+Useful commands are given as commented lines
+in the end of the syntax file. % TTM
+One might obtain an HTML file with the colored TP text
+using the \tttt{:TOhtml} command.
+
+% reference to figure of the texts.
+
+\subsubsection{Advanced syntax highlighting considerations}
+Standard guidelines for syntax highlighting
+depend heavily on cultural and use factors
+and have scarce scientific studies~\cite{stack}.
+There are informed projects such as Solarized~\cite{solarized},
+which present solutions for some contexts.
+Strikingly, standard guidelines for syntax highlighting were not found.
+Therefore, we considered current data visualization
+theory~\cite{dv1,dv2,dv3,dv4} to glimpse at the potential guidelines:
+\begin{itemize}
+  \item The use of blue and other high frequency colors (such as
+    violet) for the background.
+  \item The avoidance of blue and other high frequency
+  colors to fill small objects, such as letters and words.
+  \item Explore simplicity and elegance through the use of discrete
+  coloring schemes. Most tokens should be preferably of
+  the same color.
+    One might use only a small number of colors to achieve a clean
+    visualization of texts.
+  Also, a power-law distribution of tokens among colors might be
+    well-suited to mimic natural occurring phenomena.
+    (Unfortunately, the TP dictionary classification of words are not in a
+    power-law distribution and might be better described by a half-normal or by
+overlapping normal distributions over the three peaks: $\sim35-58,
+12-18, 1-5$ occurrences).
+  \item Physical stimuli might be related to perceptual stimuli both 
+    through a power-law or an exponential law (respectively known
+    as Steven's law and Weber-Fechner's law).
+    This might be useful e.g. for coloring sets of tokens
+    considering their similarity.
+  \item One usually wishes to maximize contrast,
+  although taste and less wearing combinations might
+  also dictate the coloring choices.
+  \item Stipulate axes of parameters to set colors.
+  Wavelength or frequency is an obvious axis given the considerations
+    previously exposed, but they are not perceptually uniform.
+    Considering perceptually uniform color spaces in making
+    choices may be relevant.
+  \item One might consider two types of color schemes: those with a dark background,
+  which are more comfortable at first; and those with a light background,
+  which are usually impressive (and even annoying) at first,
+  but the eye gets used to it and it keeps you more stimulated.
+\item The blue color is specially related to physiological
+  stimulation of the body~\cite{blue,blue2},
+    and programmers often report that a blue background keeps them
+    awake and more concentrated (I also notice such effect,
+    and I have programmed daily for more than 10 years).
+  \item Colors have been associated to enhancements in specific tasks,
+    e.g. blue and red are respectively associated to enhanced
+    creativity and detail-oriented tasks in~\cite{blue}.
+  \item It is important to consider that the user might 
+    stay many hours
+  at a text editor (and we very often do),
+    and that the colors and formats involved are
+  prone to entail a considerable effect in the body and mental
+    activity and thus in the quality of life and work of a writer
+    (e.g. a programmer).
+  \item Ideally, one should have facilities for tuning 
+    the syntax highlighting (e.g. through keyboard shortcuts)
+  as envisioned in~\cite{vim}.
+\end{itemize}
+
+The considerable irrelevance of the morphosyntactic classes as
+described in Section~\ref{syntax} suggests that an appropriate coloring
+of words should
+either distinguish only between particles, prepositions and the rest,
+or consider the syntax to identify the nouns, verbs, adjectives, adverbs,
+prepositions and particles.
+Also, the coloring for poems might be more appealing if considered
+e.g. the counting of syllables, the repeated letters or syllables,
+and the ending syllables.
+Insights might be obtained through the
+choices made and advocated in software such as text editors
+(e.g. Vim, Emacs, Sublime),
+packages dedicated to syntax highlighting (e.g.
+pigments\footnote{\url{http://pygments.org/docs/}} and 
+highlight.js\footnote{\url{https://highlightjs.org/}},
+and
+other software (e.g. Linux terminals such as Xterm and gnome-terminal).
+Finally, the careful choice of fonts is known to have
+a relevant impact in comfort and productivity~\cite{fonts},
+e.g. sans-sheriff fonts are more promptly read and yield a cleaner
+text than a serif font, and the same applies to a monospaced font
+which is more likely to yield a cleaner text, at least for programming.
+
+One might also 
+consider the mapping the textual structures
+to sound~\cite{mass} (i.e. parametrize the synthesis of sounds e.g. 
+by the counting of specific tokens inside a script, a function or
+class or around a variable of conditional or
+loop), which might be called ``syntax sonification''.
+
+\subsection{Toki Pona Wordnet}\label{wn}
+A first Toki Pona Wordnet was constructed
+relating each of the TP words in the dictionary
+to English Wordnet synsets~\cite{wordnet}
+through the English lemmas.
+The most canonical (i.e. Princeton) Wordnet only contains nouns,
+adjectives, verbs, and adverbs.
+Thus, particles and prepositions were not considered.
+Numbers were considered adjectives.
+Words presented as adjectives in the dictionary
+were considered both as adjectives and adverbs.~\cite{wordnet}
+
+The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
+makes available such tentative TP Wordnets in the simplest form:
+the TP words are keys in a dictionary that returns the
+corresponding synsets.
+Three of such dictionaries are implemented:
+one where all synsets related to the lemma is
+returned (the version most consistent
+with the expositions in Section~\ref{basics}),
+with 4,052 synsets;
+another as such but excluding the prepositions,
+with 3,961 synsets;
+a last one that returns only the synsets
+registered in Wordnet with the same POS tag as the lemma
+is in the official dictionary,
+with 2,493 synsets. 
+Another immediate possibility, not implemented,
+would be to relate each Toki Pona word to the synsets
+simultaneously associated to all the English
+words bounded by a semicolon in the dictionary.
+Such collection of synsets, and its relation to the whole
+Wordnet
+(the total number of synsets in Wordnet is $117,659$),
+might guide creation of other conlangs
+(e.g. one might seek to use lemmas related to synsets that
+are very far apart, or that has the most complete neighborhood
+possible).
+% Total number of synsets counting neighbors is Y. TTM
+% Depth of TP synsets is .
+
+It is worth mentioning that TP words are very general
+and each of them might mean many things.
+Thus, the words are not very easily related e.g. by hyponymy
+or meronymy.
+Exceptions: jan (people) is a hyponym of soweli (mammal);
+kili is an hyponym of kasi;
+walo, pimeja, jelo, loje, laso
+are hyponyms of kule.
+There is at least one caveat:
+if a particular meaning of each word is chosen,
+then there might be many other of such relations,
 e.g. 
-\\\tttt{:execute 'normal i' . string(atan(bufnr('\%')))}.
-
-\subsubsection{Variables}
-There are some types of variables in Vim:
+luka (as hand) is a meronym of sijelo (as body).
+
+\section{Conclusions and further work}\label{conc}
+This document presented a potentially novel description of Toki Pona (TP)
+language in Section~\ref{basics}, and innovative and useful software routines for dealing with
+the language in Section~\ref{hacks}.
+As a minimalist conlang, TP is very convenient for cognitive
+experiments
+through linguistic semantics and
+for devising tools for analysis, creation and visualization.
+Other conlangs, and even non-planned languages might benefit from
+TP and the content presented here in many ways, e.g. one might
+synthesize TP texts and translate them as convenient,
+adapt the routines to obtain statistics of the vocabulary,
+or take advantage of the language description to devise new conlangs
+or stilistic guidelines for any language.
+
+Possible next steps are numerous:
 \begin{itemize}
-	\item Environment variables: names start with \$ and hold system
-		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
-	\item Option variables: names start with \& and are meant to control the behavior of the editor.
-		One might change a value through set or let, e.g.
-		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
-	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
-    More on register in Section~\ref{state} and scattered in this
-    document.
-	\item Internal variables are created with let and preferably have a prefix:
-	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
-		are local to the buffer, window, tab page, function, and
-		sourced Vim file, respectively.
- \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
-		\tttt{a:} is for function arguments.
-		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
-		More about internal variables in \tttt{:h internal-variables}.
-	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
+  \item Further develop the text synthesis facilities described in
+    Section~\ref{synth} might be enhanced by further employing e.g.
+    rhythm, meter, rhyme and form techniques~\cite{wikiPoetry}.
+  \item Make a syntax highlighting that colors the tokens with respect
+    to the syntatic position and function, and not in a fixed manner
+    as is implemented and exposed in Section~\ref{shigh}.
+    This might be implemented by using n-grams and
+    further techniques from Natural Language Processing
+  \item Enhance and study the TP Wordnet described in Section~\ref{wn}:
+    implement a NLTK-like Wordnet interface; check if each synset
+    is correctly associated to the TP words; seek further synsets
+    that were not found by the English lemmas in the official
+    dictionary; implement the most restricted version described in
+    Section~\ref{wn}, which relates each TP word only to the synsets
+    that are related to all the English words not separated by a
+    semicolon; find out the neighborhood of the synsets of the TP
+    Wordnet in the English Wordnet; explore the possibilities for
+    devising new conlangs by means of criteria related to Wordnet.
+%   \item Translators.
+%    This Toki Pona Wordnet would be useful e.g. for naive translators
+%    of words and text.
+%    Enhancements of such translators would use POS tagging and templates
+%    in English and Toki Pona and/or corpus and further statistical methods.
+%    An efficiency of Toki Pona can be studied in terms of the coverage of
+%    the Wordnet cloud.
+  \item Conlanging: use Wordnet for making new conlangs;
+    use insights derived from TP.
+    Maybe create conlangs for specific uses:
+    describing data, programs,
+    scientific writings, creative writing
+    (exploring the thinking process).
+    A conlang might have different modes that might 
+    be specified by a section header or tag. 
+  \item Understand how the corpus was gathered in \url{tokipona.net}.
+  \item Know about previously existing words that were used for TP
+    (e.g. suno and suwi might come from sun and sweet),
+    and about the reasons that led Sonja (and maybe other people)
+    to choose the 14 letters and the syllable structure.
+    This might require a dedicated communication with the
+    speaker community and the documentation keepers.
+  \item Corpus-based analysis.
+  \item Publication of original TP texts and translations.
+  \item Make an article written in TP
+    (extending Section~\ref{ftp} which might be the
+    first section of a scientific document written in TP).
+    The outline might be:
+    a summary in both English and Toki Pona, for
+    facilitating the acquisition of context,
+    and a article in a canonical structure
+    (introduction and related work, materials and methods,
+    results and discussion, conclusions and further work),
+    with an Appendix explaining the content and context in English.
+    I first conceived something around complexity, statistics, physics, or computer science.
+    But one possibility is to write about linguistics, philosophy, literature
+    or psychology with the partners I write in English.
+    I can start a draft,
+    they might learn the language in a few hours or weeks
+    (with or without my help),
+    to contribute, and we can write a short paper.
+  \item Enhance the synthesis of text to yield better contextualized text
+    and stylistic traces.
+    Give the user the ability to choose the sentences:
+    generate a sentence randomly according to
+    previously written or given text,
+    to some rules input by the user,
+    to the package and the language guidelines.
+    Should output to the screen and asks to keep and discard.
+  \item Make TP variations where each TP word is replaced by a word in
+    a language with a large speaker population. 
+For example,
+suno might be sun in an English variant,
+and sol in a Portuguese variant.
+A naive version might be obtained through
+choosing the first word in the description of
+an official vocabulary, as they already exist in English and French.
+  \item We are particularly interested in software tools,
+    and one possibility is to use a conlang to describe software
+    routines.
+    In this sense, it seems profitable
+    to understand what `non-ambiguous' means in Lojban,
+    and in what sense one is able to compile and parse Lojban.
+    Such study will probably entail considerations about
+    parsing TP.
+%  \item The plugin comes with the :TokiStation command, which opens a
+    %  window with the files: ijositelen.tokipona, makevimSyntax.py,
+    %  Highlight Test (created by hitest.vim above),
+    %  syntax/tokipona.vim.  Another tab with shells: python to run
+    %  over the makeVimSyntax.py and make new syntax/tokipona.vim
+    %  files. Another with Readme, and PDF documentation. Another with
+    %  an IPython. Imported tokipona. It resets completely upon
+    %  command :TokiClose, closing all created windows. TTM
 \end{itemize}
-You can echo any of such variables or use them in expressions.
-Notice that you will only be able to echo a \tttt{b:} variable inside
-the buffer where it is defined.
-For all the Vimscript capabilities, including loops, conditionals,
-and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
-Classes are possible only in rudimentary forms, e.g. through dictionaries,
-but the language is otherwise overall quite powerful,
-specially in dealing with text and editor behavior, as expected.
-
-\subsubsection{State lists}\label{state}
-Vim keeps a number of useful lists which expresses the state of the editor:
-\begin{itemize}
-	\item All the entered commands are accessed through \tttt{:hist a}. 
-    The tokens ``\tttt{a / e : i d}'' may be used for specific types of commands, such
-		as search and colon commands.
-	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
-  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
-	\item Tabs list can be reached through \tttt{:tabs}.
-		It is usual both to show and hide the tabs bar (mapping
-    in~\cite{vimrc}, discussion in Section~\ref{visual1}).
-	\item Jumps are available through \tttt{:jumps}.
-		One positions the cursor at each jump through \ttt{C-O} and \ttt{C-I}.
-  \item Registers are available through the \tttt{:reg} command, as variables and through shortcuts in different modes.
-    They also keep track or your copy, edition and deletion and are promptly defined
-    by recording a typing sequence with the \tttt{q} normal command.
-    Vim keeps only the last edition, in register \tttt{".}.
-    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
-    A hack to keep the latest deletions and copies in the standard register
-    might follow the same pattern, but use another \tttt{:h event} and
-    monitor register \tttt{""} (maybe also monitor \tttt{"0}).
-	\item An undo list can be accessed through \tttt{:changes}.
-	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
-	\item The markers defined are listed with the \tttt{:marks} command.
-		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
-		Uppercase letters are cross buffers.
-	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
-		and variations, such as \tttt{:grep}.
-		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
-		to open the Quickfix window, where the lines of occurrence are in sequence
-		and one can \ttt{CR} one of them to have the cursor in the main window active at
-		the first character of the match.
-		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
-		to use the location-list window instead of the Quickfix, which
-		is very similar, but one per window instead of one per buffer.
-		More information in \tttt{:h quickfix}.
-	\item A tags list have to be made so one can use tags.
-		Most often one will generate the tags list using the exuberant ctags,
-		which supports dozens of languages.
-		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
-		and then using \ttt{C-]} to
-		go to the position of tag under cursor, and \tttt{:vs tags} to open
-		the tags file.
-	\item The argument list holds a list of files to be edited or browsed. 
-		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
-	or using commands (e.g. \tttt{:ar ./*}).
-		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
-		one might perform actions on each file in argument list using \tttt{:argdo}.
-		All files in argument list are also in the buffers list.
-    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
-  \item All the autocommands might be accessed through \tttt{:au}.
-    They are event-triggered actions Vim performs.
-\end{itemize}
-
-A file with information about the state of the editor
-can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
-Also, this script might be examined because it has
-a collection of commands to access various settings of Vim.
-Another good list of commands to know about Vim's state is kept on
-\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
-I would specially highlight the \tttt{:syntax} command because
-it displays token groups and their meanings when run
-inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
-
-\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
-You can keep track of the editor state though commands, as stated
-above.
-Also, one might rely on persistent visual
-cues, specially the tabs bar, the status line, and the line
-reserved for the command-line.
-A good strategy I find is to have selective visual cues of the state to make persistent or hide
-and a mapping to toggle each of them.
-Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
-status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
-I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
-Numbering is always there, I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
-Instead of keeping the status bar, I use \ttt{C-G} to know about the
-file and \ttt{gC-G} to know more and rarely.
-It seems not possible to remove the statusbar between horizontal splits.
-After asking in online forums and experimenting, I realized that it seems
-reasonable to keep at least one line dividing the windows, so
-if it comes to it, I just \tttt{set statusline=-}.
-Unfortunately, as far as I could dig, one will need to alter
-the Vim's source code to enable a horizontal split without losing a line.
-For me, it would be ideal for this feature to have a visual cue of the
-first and/or last line of the windows in the lines-number
-column, or complete the spaces and empty chars with \$\$\$\$ or so.
-
-%%%%%%%%%%%% TTM
-\subsubsection{On the persistence of the editor state}
-For state persistence, one might keep an undo file for each file as
-in~\cite{vimrc}.
-Sessions are easy to manage, enabling one to save and load the
-editor's state, with the opened windows, tabs, buffers, etc.
-The mappings in~\cite{vimrc} keep the sessions
-in a reasonable directory and makes it easier to remember and tweak
-the standard commands to deal with sessions.
-More information in \tttt{:h sessions}.
-One might use \tttt{:h views} to keep the state of one window,
-but sessions keep all the states from all windows.
-This entails a strategy to deal with Vim that is similar to
-the use of
-Byobu/Tmux/Screen\footnote{\url{https://en.wikipedia.org/wiki/Byobu_(software)},
-\url{https://github.com/tmux/tmux/wiki},
-\url{https://www.gnu.org/software/screen/}},
-because one can rely on restoring the state of the windows.
-The main limitation I found to this approach is that
-Vim is not keeping track of the terminals opened.
-If you open a terminal inside Vim with the \tttt{:term}
-command, you will save the session as usual, but when loading
-you get dummy empty windows for them and an error message.
-More about Terminal-Job mode in Section~\ref{terminal}.
-
-Autocommands are the standard way to define event-triggered routines in Vim.
-These are often related to particular file types, but are also often
-in defining the automated behavior of Vim.
-If the autocommands are placed inside a configuration file (e.g. the
-vimrc), the automated behavior is persistent across Vim instances.
-E.g. in~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).
-
-\subsection{Using Vim's modes}\label{modes}
-These are the  basic and fully implemented modes of usage in Vim:
-\begin{itemize}
-  \item Normal mode: used for changing
-  the position of the cursor or the text displayed
-  at the window.
-  A core goal of the normal mode it to support fast
-  navigation of the document while allowing
-  the typist to maintain the fingers on the home row
-  (i.e. on the center of the keyboard).
-  The mode is also used for manipulating text
-  (e.g. copy, paste, delete, change case) and
-    changing to other modes.
-    More in Section~\ref{normal}.
-  \item Insert mode: for inserting text. More in
-    Section~\ref{sec:ins}.
-  \item Command-line mode: for entering Ex commands. More in
-    Section~\ref{sec:com}.
-  \item Ex mode: similar to command-line mode,
-  but more specialized for running various Ex commands.
-    More in Section~\ref{sec:ex}.
-  \item Visual mode: for making, manipulating and navigating
-  selections of texts.
-  \item Select mode: similar to visual mode but
-  favors CUA\footnote{IBM Common User Access:
-    \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
-\end{itemize}
-
-There is another basic mode, but it is not fully implemented:
-the Terminal-Job mode (more in Section~\ref{terminal}.
-There are seven additional modes which are mostly subordinate 
-to the basic modes and that will be described when convenient.
-The manual page for Vim modes can be accessed by typing
-\texttt{:h vim-mode}.
-Some of the modes are now further considered for
-the achievement of
-an overview of the Vim usage possibilities.
-
-\subsubsection{Normal mode}\label{normal}
-Sometimes also called navigation or command mode,
-the normal mode is most powerful for
-navigating, manipulating texts and changing to other modes.
-The simplest of these three is changing to other modes:
-type any of these letters to change to insert mode:
-\texttt{iIaAoOsScC}. More on the transition between
-normal and insert mode on Section~\ref{navIn}.
-Type any of these characters to change to command-line mode:
-\texttt{:/?}.
-Type \texttt{Q} to enter Ex mode.
-Type \texttt{v}, \texttt{V}, or \texttt{CTRL+V} to enter visual mode.
-
-For very basic and naive navigation, one should check Section~\ref{minimum}.
-There are many facilities to navigate Vim
-as explained in \texttt{:h navigation}.
-Most often, one uses:
-\begin{itemize}
-  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
-and whole page down and up, respectively, although these commands might
-be set to scroll a different number of lines.
-  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
-  \item \texttt{(w,b,e)} to move to the next and previous word,
-    and next end-of-word.
-    There are motions to iterate over sequences of characters separated
-    by special characters (e.g. punctuation and parenthesis) as
-    specified by the output of \texttt{:se iskeyword}.
-    To iterate over space-separated tokens, use \texttt{W,B,E}).
-    To move to the end of last word, one might use \texttt{be}
-    or \texttt{ge}.
-The \texttt{),(} commands iterate through sentences,
-		\tttt{\},\{} through text blocks separated by empty lines.
-  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
-	  \tttt{;} and \tttt{,} for next and previous found character.
-  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
-  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
-    jump list.
-  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X:
-    \texttt{`X} moves to the exact position while \texttt{'X} moves to
-    the first non-blank character of the line.
-    A mark is registered by the user in any cursor position
-    by typing \texttt{mX}, where X is any letter.
-    If X is lowercase, the mark is local to the buffer (the file),
-    if it is uppercase or numeric, it is global to the Vim session
-    (cross buffers).
-\end{itemize}
-
-For changing the text, usual commands include:
-\begin{itemize}
-  \item \texttt{d\{motion\}} to delete the characters
-    involved in the motion command.
-  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
-  \item \texttt{x,X} to delete the character under or before the cursor.
-  \item \texttt{$\sim$} to swap the case of a character or selection.
-  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
-\end{itemize}
-
-There are way more commands to change the texts.
-Some of them are discussed in Section~\ref{navIn}
-because they involve a transition to the insert mode.
-A thorough consideration of the commands in the normal mode
-is found by executing \texttt{:h navigation}, 
-\texttt{:h change.txt}, \tttt{:h index}.
-
-\subsubsection{Insert mode}\label{sec:ins}
-Once in the insert mode, the character keys
-input the characters at the cursor position at the current buffer.
-One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
-\ttt{C-C}),
-and Vim will be put in the normal mode.
-Most useful commands in insert mode include:
-\begin{itemize}
-  \item \ttt{C-O} to execute one and only command in normal mode.
-	  This enters a secondary mode (see Section~\ref{modes})
-  \item \ttt{C-R} to paste a register (a variable starting with
-    '@', defined, copied or recorded through a \tttt{q} command in
-    normal mode and as covered in Section~\ref{state}).
-  \item \ttt{C-T} to indent current line.
-  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
-    the line or to the previous word.
-  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand.
-  \item \ttt{C-X} commands for scrolling the window with multiple
-    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
-\end{itemize}
-
-\subsubsection{Normal $\rightarrow$ insert modes}\label{navIn}
-Many commands bridge from Normal to Insert modes,
-e.g. \tttt{iws} or any of these letters: \tttt{csrCSR}.
-These make convenient the replacement of text and populates registers.
-The absence of a short command to insert one char is
-a known issue in Vim.
-Reasonable mappings to insert or append a char to and around
-another char are in~\cite{vimrc}.
-Vim couples operator and motion commands by design.
-There are many operator commands that take the editor from
-the normal mode to the insert mode, most of them
-favoring deletion or change, as detailed in \tttt{:h operator}.
-Motion commands are described in \tttt{:h motion}.
-
-\subsubsection{Command-line mode}\label{sec:com}
-This mode is dedicated to writing colon, search and filter commands,
-entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
-Most useful commands in this mode include:
-\begin{itemize}
-  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
-    end of the line.
-  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
-    until the cursor.
-  \item \ttt{C-R} to paste a register (as in insert mode).
-\end{itemize}
-
-\subsubsection{Ex mode}\label{sec:ex}
-One might use the normal commands \tttt{q(:,/,?)} to have a window
-with the colon or search command history to be edited normally,
-and the chosen command can be run with \ttt{CR}.
-Vimscript was largely based on the ex editor~\cite{ex},
-and a more advanced user might use it for prototyping
-by defining mappings and settings and managing scripts
-probably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
-In the default interface started with the \tttt{Q} command in normal mode,
-each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
-follow documentation from \tttt{:h Ex-mode} for further information.
-I might be loosing something, but for
-tweaking I use the command-line window accessed through \tttt{q:}.
-I find it comfortable to browse the history in a normal-like mode,
-edit them also using the insert mode as usual,
-and having auto-completion when pressing \ttt{TAB}.
-None of such features are available in Ex mode by default
-and
-customization by mappings have to be performed through autocommands as in~\cite{vimrc}.
-
-\subsubsection{Terminal-Job mode}\label{terminal}
-This mode is reported as not having reached a stable usage design
-(see \tttt{:h terminal}).
-I find that it works exceptionally well and have used it to run
-scripts in an IPython shell, compile latex files, and even open PDFs and images.
-Vim browsing of windows and text manipulation is well developed,
-so the Terminal-Job mode enables a very convenient integration
-of files being edited and bash terminals,
-more traditionally achieved through Byobu/Tmux/Screen terminal
-multiplexers.
-Most useful commands in the terminal-Job mode include:
-\begin{itemize}
-  \item \tttt{<C-W>\_N} and \tttt{<C-W>\_:} for entering the Terminal-Normal
-    and command-line modes. Terminal-Normal mode is very similar to
-    Normal mode, but one cannot change the text, cannot enter insert
-    mode, and the status line reports if the job is finished or not.
-  \item \tttt{<C-W>\_"} to paste a register.
-  \item \tttt{i} for entering the Terminal-Job mode from the normal mode.
-\end{itemize}
-\noindent It is useful to define the same mappings for navigating splits and tabs
-for both Normal and Terminal-Job modes, as in~\cite{vimrc}.
-
-Because browsing the interface in Vim is fast,
-and it is very comfortable to copy the terminal lines
-in the Terminal-Normal mode,
-it is often handy to keep (e.g. a tab with) some terminals:
-e.g. one with an IPython shell, another two for compiling \LaTeX\
-and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
-There are even more convenient ways to use the terminal inside Vim.
-For example, 
-one might use \tttt{:term} with the \tttt{++hidden} and \tttt{++close}
-options to compile a \LaTeX\ file in the background without needing to further manage the
-terminal and in a non-blocking manner: \tttt{:term ++hidden pdflatex
-\% \#}.
-I found the mappings on~\cite{vimrc} very helpful for directing
-editor focus to splits (\ttt{C-hjkl}) and tabs (\tttt{gr}, \tttt{gt}),
-which I make available across Terminal-Job and Normal modes.
-But I've been thinking on using \ttt{C-} commands also
-for tabs (not only for splits), both to avoid mixing typing sequences
-with commands, and to allow the use of the commands in all Normal,
-Insert and Terminal modes.
-
-\subsection{Netrw}\label{netrw}
-The standard interface of Vim for browsing file trees is Netrw.
-It starts when you open a directory, such as with \tttt{:e .<CR>}.
-It has solid support for browsing remote file trees (such as over ssh or
-ftp) and handy e.g. mappings to open the files as splits and tabs
-(specially \tttt{pot}).
-Most useful commands in netrw include:
-\begin{itemize}
-  \item \tttt{d} and \tttt{\%} for creating directories and files.
-    \ttt{Delete} removes both.
-  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
-  \item \tttt{gb} and \tttt{uU} are used to load directories
-    while marked files might be copied, moved, edited, grep-ed, tagged and migrated
-    to and from the argslist as in \tttt{:h netrw-mf}.
-\end{itemize}
-
-There is no insert mode in netrw interface;
-the commands in \tttt{:h netrw-explore} are
-convenient for opening the directory of the
-file being edited;
-further information is in \tttt{:h netrw}.
-
-\subsection{Standard configuration files and directories and my .vim/vimrc}
-You can check the scripts Vim loads by using the debug script mentioned
-in Section~\ref{state}.
-By default, \tttt{$\sim$/.vimrc} and \tttt{$\sim$/.vim/vimrc} files are run by Vim at the beginning of the startup.
-One might edit the vimrc file with \tttt{:e \$MYVIMRC}
-and reload it with \tttt{:source \$MYVIMRC}.
-Mappings in~\cite{vimrc} include such commands
-to encourage a continuous enhancement of Vim settings
-(they have helped me to improve my settings without unnecessary 
-hassle and fast).
-Any other file might be included to run at startup by
-adding a line \tttt{:source <afile.vim>} in vimrc.
-In fact, it is on vimrc that one usually specifies the plugins
-and plugin managers they use.
-Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
-or follow some patterns described on the next section to change the
-scripts and sequence of them to be loaded.
-The vimrc from other users are most useful for one to comprehend
-and pick convenient practices and settings.
-In fact, a vimrc file is most often a collage of excerpts of vimrc
-files from other users.
-
-\subsection{Plugins and packages}
-One can see the list of the standard plugins
-with \tttt{:h standard-plugin-list} command.
-Any \tttt{plugin/**/*.vim} file inside a directory
-listed in \tttt{:se runtimepath} will be loaded
-(e.g. \tttt{.vim/plugin/something/ascript.vim}).
-There are various ways to automate the installation
-and enhance the management of plugins.
-By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
-that downloads latest scripts from Sourceforge as specified in \tttt{:h getscript-data},
-and the Vimball interface, which creates and loads a Vimball for a
-plugin.
-Such a Vimball may be created with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
-that hold paths to files to be included in the \tttt{<filename>.vba} Vimball.
-The Vimball can be installed in a system by \tttt{:source <filename>.vba}
-or loading it at Vim startup with \tttt{\$ vim <filename>.vba}.
-
-A Vim package is a directory that contain plugins.
-It should be located inside a \tttt{pack/} directory
-somewhere in the directories listed with \tttt{:se runtimepath}.
-The plugins found in \tttt{pack/<packName>/*/start/} are loaded
-at startup, the plugins found at
-\tttt{pack/<packName>/opt/**} are loaded with\\
-\tttt{:packdd <script\_or\_directory\_name>}.
-
-All directories Vim looks for scripts are described
-in \tttt{:h vimfiles} and are basically set by
-\tttt{:se runtimepath} and conventions inside
-each directory therein, such as to look for \tttt{vimrc} files
-and \tttt{plugin/} or \tttt{pack/} directories.
-Vim scripts can be loaded conditionally,
-e.g. only if a function is used as in \tttt{:h autoload-functions}.
-Example of such are filetype plugins (enabled by a \tttt{ftplugin/<filetype>.vim} file,
-e.g. inside a plugin directory).
-There is a number of plugin managers for Vim.
-Pathogen and Vundle seem to be the most popular,
-one because of its minimalism, the other because of
-advanced features, e.g. for searching and installing plugins with
-colon commands.
-
-\subsection{Spell and spelllang (en and pt\_br)}
-One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
-or \tttt{:se spelllang=pt\_br}
-and toggle spell checking with \tttt{:setl spell!}.
-Depending on the activities being performed,
-these commands are used so often that one might use mappings as in~\cite{vimrc}.
-Currently, Vim will download the files for a specific language if
-not found in system.
-
-\subsection{Scripting, Functions, Vimscript and other languages (e.g.  Python)}\label{script}
-In Vimscript,
-the colon commands (also Ex commands or command-line commands) are related through spaces,
-punctuation and keywords (see \texttt{:h script}).
-Scripting the Vim editor can also be accomplished using other languages,
-as well documented e.g. in \tttt{:h python}.
-Functions are defined through colon commands and are called
-inside colon commands e.g. \tttt{:call MyFunction()}
-or \tttt{:echo MyFunction(4)}.
-Notice that functions are not commands but might be bind to them
-through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
-Executing source files is very straightforward with
-\tttt{:so <filename>.vim},
-and one can always use the Ex mode for rapid scripting.
-At the same time, the \tttt{:term} and terminal-job mode make
-scripting other software more convenient, as output is
-promptly navigated and copied (as discussed in Section~\ref{terminal}).
-
-\subsection{Color}\label{visual2}
-Newer versions of Vim support 24-bit true color (aka 16 million colors)
-in terminal Vim (in gVim true color received support earlier).
-The terminal must support true color, and tests are available e.g. in
-\url{https://gist.github.com/XVilka/8346728}.
-Then Vim needs to be set to use true colors with
-\tttt{:set termguicolors}.
-If using 8 or 16 bit colors, Vim uses the color palette from the
-terminal, if using true color each color is defined directly.
-Settings for using true color inside Byobu/Tmux involve tweaking
-and are available in~\cite{vimrc}.
-Good color schemes to use with true color are Gruvbox
-and Solarize.
-One might source syntax files at any time to change syntax, usually
-though linking tokens to syntax groups as in
-\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...},
+
+\subsection*{Acknowledgments}
+FAPESP (project 2017/05838-3);
+Toki Pona, Python and Vim authors, developers and literature maintainers;
+Toki Pona user and speaker communities. 
+I thank Mario Alzate for introducing me to Toki Pona,
+and Silverio Guazzelli Donatti for the stimulus that resulted
+from the usage of Toki Pona for chatting.
+
+\appendix
+\section{My usage of Toki Pona}\label{mytoki}
+I use the standard sounds, but often use [z] for s.
+I often translate texts to Toki Pona (e.g. biblical excepts)
+and create new texts as poems and short stories.
+Most of them are in~\cite{tokisona}
+and/or published in the TP Facebook groups.
+I omit the li particle after subjects sina and mi,
+in accordance with the norm,
+but sometimes I use them when there are many predicates.
+E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
+In such cases, the first li is sometimes omitted.
+Also, sometimes I use li after mi and sina where I find
+that there is unwanted ambiguity, e.g.
+sina moku pona 
+which 
+might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.
+
+Names are by default transliterated,
+but I advocate that, as in other languages,
+names might be used as they are in the
+correspondent mother tongue.
+E.g. the name Erdös is used in
+English and Portuguese although the standard
+alphabet does not contain ö in such languages.
+I also tend to legitimate the use of English (or German) words
+in TP texts if it is the case,
+as happens often in scientific writing
+(kernel is a German word used in English,
+webpage and software are English words used in Portuguese).
+
+Proposed notations for numbers seem numerous.
+I tend to think that one might indicate if two numbers
+are being multiplied (pi) or are in different scales
+(such as in decimal or binary notations).
+For example, luka (pi) two might mean 52.
+Most importantly,
+52 is a reasonable notation for a simple language.
+E.g.
+'mi jo e jan sama nanpa 12',
 or
-\tttt{:syntax match <group\_name>	<pattern>},
-and then relating the group to another group: 
-\tttt{:highlight link <group\_name> <group\_name2>}
-or to group characteristics directly:
-\tttt{highlight <group\_name> guifg=\#ffffff}.
-If you change a syntax file, reloading a file with with \tttt{:e<CR>}
-updates the highlighting on the window with the corresponding file type.
-A complete syntax highlighting support typically involves at least three files:
-\begin{itemize}
-  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
-    detected with e.g.
-    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
-  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
-    file type, such as: 
-    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
-    expandtab autoindent}.
-  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
-    tokens and highlighting groups; and highlighting group definitions.
-\end{itemize}
-
-This scheme is implemented very straightforward in this
-plugin~\cite{tokipona}
-for highlighting text in the Toki Pona language~\cite{tpLang}.
-Syntax highlighting plugins are as file type plugins,
-but also have a \tttt{syntax/<filetype>.vim} file
-relating keywords, matches and regions to highlighting groups. 
-One might see every highlighting group, and their final visual results,
-with the commands
-\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}
-or \tttt{:so hi}.
-The \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
-directory, but \tttt{syntax/} files has to be moved to
-\tttt{$\sim$/.vim/syntax/}, unfortunately.
-
-\subsection{Fonts}
-The fonts are defined by your terminal or inside gVim.
-\ttt{C-+} and \ttt{C--} can be used to change font size.
-Some settings for fonts, such as boldface, might be set using
-the syntax highlighting facilities described in the previous section.
-
-\section{Conclusions and further work}\label{conc}
-This document seems reasonable as an overall reference of the Vim editor,
-at least for my usage and level of proficiency.
-Given the folkloric milestone of using Vim for 10 years,
-this article might serve as a benchmark for one to relate
-it's current use and understandings.
-As a pedagogical material, it seems to be unique in the emphasis
-on namespaces, understood as commands, variables, state-related lists, etc,
-especially in Section~\ref{namespaces},
-and the reference to the standard Vim documentation
-to achieve the DRY KISS style described in Section~\ref{intro}.
-
-\noindent {\bf Potential enhancements} to this document include:
-\begin{itemize}
-  \item The discussion of facilities such as reading emails and connecting over ssh.
-    There is a working hack in~\cite{vimrc} for browsing over the WWW,
-    but such aspect of Vim usage might receive more attention
-    given that it is comfortable to navigate and edit in Vim
-    and the resulting integrated environment.
-  \item Updating of the information I can find about the issues discussed,
-    such as about status lines in Section~\ref{visual1}.
-  \item Include a discussion about Neovim.
-    I have never used it, but it seems to be reaching a considerable user base
-    and it might be feasible to give an account of Vim and Neovim
-    after some tests and researching the official and unofficial documentation.
-  \item Better cite documentation, plugins and Vim authors.
-    I preferred to keep the references inline through \tttt{:h} commands
-    and URLs, more in accordance with the style of Vim documentation,
-    but bibliographical items constitute a valuable asset for academic literature,
-    and some authors might find their work more respected if more
-    thoroughly cited.
-  \item An analysis of my usage, e.g. according to \url{http://www.drbunsen.org/vim-croquet/}.
-    This undertake might benefit from data from many users,
-    which favors the potential plugin for usage analytics described in the next bullet list.
-\end{itemize}
-
-\noindent {\bf Potential next steps} in using Vim include:
-\begin{itemize}
-  \item Measure the performance of text mining routines in Vimscript against those implemented in C or Python.
-  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
-  \item Better integrate Python and Vimscript, especially for data visualization
-    and syntax highlighting management,
-    in accordance with the visualization issues described in
-    Sections~\ref{visual1} and~\ref{visual2}.
-\end{itemize}
-
-\noindent This is a selection of the issues that might entail
-plugins and that are more prominent for me:
-    \begin{itemize}
-      \item listing all the mappings available in each mode and the typing combinations which are available
-        for new mappings.
-        Maybe already group possibilities by criteria such as length
-        of sequence and how central are the keys.
-      \item Sessions, as described in Section~\ref{visual1}.
-      \item AA messages (shouts): to keep track, document and share of working sessions
-        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
-        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
-        relate AA sessions to other media, such as software repositories,
-        screencasts and images, interact with IRC channels and other social platforms.
-      \item Slick Vim: a collection of the settings and mappings I use.
-        Enhancements such as using \ttt{C-} commands also to browse tabs,
-        shortcuts to join a window into a tab,
-        and dummy minimal plugins as simplest possible, then file type
-        and then syntax highlighting,
-        Some more elaborate tweaks should also be present, such
-        as breaking lines in sensible places for natural language texts
-        while respecting e.g. \tttt{:se textwidth}.
-      \item Dealing with .swo and .swp temporary files.
-        In summary, if the restored .swo file has the same content
-        and the file being opened,
-        the restoration phase can be omitted.
-        If the contents differ, Vim should open a tab with each file
-        in a vertical split and run \tttt{:windo :diffthis}.
-      \item Rendering images and equations.
-        These are useful for using Vim in presentations
-        or achieving a textual representation when it is mandatory,
-        such as to comply with the limitations of a platform (e.g. Vim editor). 
-        but also hold stylistic merits as ASCII art is often
-        very appreciated.
-        One can both obtain an ASCII representation of a binary image (e.g. JPG, PNG),
-        and can directly render ASCII charts from data using cues such are shape, position
-        and color.
-      \item Redirecting the commands that usually show the results in
-        a 'more' interface (e.g. status listings such as in
-        Section~\ref{state}),
-        which cannot be searched nor copied nor persists if one
-        returns to editing a file.
-        Ideally, it should parsed and linked quickfix or location
-        window, and the syntax highlighting maintained.
-        The basic idea is to use \tttt{:redir} command to redirect the output
-        of such commands with \tttt{:se nomore}.
-        Reasonable functions (and convenient commands) for having the
-        output of such commands in a standard Vim window are
-        in~\cite{vimrc}.
-      \item Slide presentations. I've been using some automation for browsing
-        slides and opening figures. Some Vim users asked for the settings and
-        commands. They are very elementary use of registers which are executed over
-        arbitrary but consistent textual patterns.
-      \item For bringing back all the splits after an \tttt{:only}.
-        Also for bringing back the tabs after a \tttt{:tabo}.
-      \item Run Python excerpts from a file being edited in an IPython shell. 
-        The buffer number of the terminal window should be stored,
-        and than any selected lines should be run on that instance.
-        Mappings should make available all movements to fetch the
-        excerpts, execute registers, remain on script or in the IPython shell.
-        Also, current Terminal-Job mode can be improved easily with
-        mappings, such as \ttt{C-O} for one normal mode command.
-      \item For keeping track of the usage and making analysis
-        for optimizing the usage, as described in~\url{http://www.drbunsen.org/vim-croquet/}.
-        Usage analytics.
-      \item To facilitate the tweaking of syntax highlighting.
-        This should include ways to easily access and change the
-        syntax highlighting scheme and dump it to a file (dumping
-        current highlighting scheme to a working Vimscript file
-        is currently not supported by Vim!).
-        Should also include changing the color scheme and highlighting scheme incrementally
-        and selectively using the features described in Section~\ref{visual2}
-        and in~\cite{tokipona}.
-      \item Color schemes for true color.
-        The standard colors (e.g. blue, elflord) loose some of the distinctions,
-        e.g. SpellBad tokens are not highlighted on these color schemes
-        if you have a functioning \tttt{se termguicolors} (standard GUI mode
-        and available in terminals since recent versions of Vim 8,
-        see Section~\ref{visual2}).
-        In~\cite{vimrc} are some lines that make such color schemes over
-        \tttt{:colorscheme blue}.
-        This design of color schemes over the standard color schemes might be
-        a very simple and effective way to make new color schemes.
-        Such color schemes should also make use of the discussions
-        in~\cite{tokipona}.
-    \end{itemize}
-
-\subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
-Vim user community.
-
-\appendix
-\section{Example of usage session}
-I usually begin by opening a file or directory
-with \tttt{\$ vim <filename>}.
-The color scheme is alternated between
-blue and GruvBox with
-\tttt{:colo gruvbox} and \tttt{:colo blue}.
-I open a vertical split and then move
-the window to a new tab using
-\tttt{:vs} and \ttt{C-W\_T}.
-I then search for tokens related to
-the enhancements I want to make or
-the knowledge I want to acquire.
-I go back to the previous tab with \tttt{gr}
-and make a global replace with
-\tttt{:\%s/<this>/<that>/g}.
-On adding dots to sentences,
-I record in the \tttt{"q} register
-the sequence \tttt{jA.jj},
-and use it as a macro 10 times by
-\tttt{10@q}.
-I move to the other tab with \tttt{gt}
-and open a terminal window with \tttt{:term}
-for compiling latex files.
-I start another terminal with \tttt{<C-W> :term}
-for opening the resulting PDFs with evince.
-If any new idea comes to mind and I have time,
-\tttt{\textbackslash\textbackslash s} opens 
-my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
-sources it.
-If there is e.g. code or notes in other projects I am working
-on, I reach them through \tttt{<Alt-arrows>} because they
-are in separate Vim instances
-in Byobu/Tmux sessions and windows.
-Because I stay for hours editing (\LaTeX and Python
-scripts), I change the background to red, and eventually to green.
-Because the blue color scheme does not highlight the spelling errors
-if using true color, I run
-\tttt{:hi SpellBad guifg=red guibg=lightblue}
-to see the words found wrong by the spell checker.
-
-\section{My vimrc file and usage}
-In using Vim with my vimrc file~\cite{vimrc},
-I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
-and the tab line with \texttt{\textbackslash\textbackslash T}.
-Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}.
-The mappings for transitioning through splits and tabs
-are also used constantly.
-Although the file is commented, one should
-look for the help pages on the options that (s)he does
-not understand, as a thorough explanation
-of the file is tedious and out of the scope
-of this document.
-
-\section{Example notes on mappings}\label{notes}
-\texttt{:h index} shows all the default mappings
-while \tttt{:map} shows the user-defined mappings.
-By considering such information, one can make
-useful observations exemplified in this Appendix.
-
-\subsection{Normal mode}
-Every letter and character in the keyboard is used.
-In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
-\ttt{BS} and \ttt{C-H} are the same as h.
-\ttt{C-J} and \ttt{C-N} are the same as j.
-\ttt{C-P} is the same as k.
-Space is the same as l.
-\ttt{C-[} and \ttt{Esc} are not used,
-\tttt{<C-\textbackslash> a-z} are reserved for extensions,
-and \ttt{C-\_} is not used.
-\tttt{+} is the same as \ttt{CR} and they are both not very useful.
-Del is the same as x.
-
-Many \tttt{][} combinations are not used, e.g.
-with \tttt{abhjklfg}.
-The \tttt{\_} command might be used as a more powerful \tttt{\^},
-leaving it free for mappings.
-
-Directions, home, end, page up and down, insert, all have mappings
-in more centraly located keys.
-The letters beggin unbounded mappings (require another character to
-trigger an action without having to wait for \tttt{:se timeoutlen}): g,z,[].
-
-The \ttt{C-(HJKL)} commands are redundant,
-with the exception of \ttt{C-L} which redraws the screen,
-so it is a reasonable choice to use them to move focus
-of the editor to splits in the \tttt{hjkl} directions.
-
-Many key combinations are available for new mappings through the \tttt{g}
-and \tttt{z}, and \tttt{v} commands.
-They have typical uses, e.g. \tttt{z} for folds,
-spell checking and some movements (mainly when wrap is set).
-
-\subsection{Insert mode}
-All standard character keys are used for entering text.
-\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
-in moving through multiple lines.
-\ttt{C-$[$} is the same as \ttt{ESC}.
-\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
-\tttt{<C-\textbackslash> a-z} are reserved for extensions,
-other combinations with \ttt{C-\textbackslash} are not used.
-
-\section{My \tttt{:version}}
-I should keep the output of \tttt{:version} executed on the system
-in which I wrote this document in this link:
-\url{https://github.com/ttm/vim/raw/master/version.txt}.
-It was compiled with this Makefile~\cite{makefile}
-in a 16.04 LTS Ubuntu Linux,
-and is tagged as 8.0, Included Patches 1-1173.
-
-\section{Key notation and meaning for Vim}\label{sec:not}
-The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
-mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
-(or \tttt{Meta+x}).
-\ttt{A-X} is the same as \ttt{M-X} and refer to
-Alt or Meta keys.
-\ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x} and \ttt{C-x} are the same,
-i.e. Vim does not distinguish between lower and uppercase 
-letters in \ttt{C-} commands.
-The \ttt{M-x} and \ttt{M-X} (or \ttt{M-S-x} or \ttt{M-S-X})
-are different commands,
-and they are not used by Vim's builtin keys.
-Such \ttt{M-} key combinations potentially conflict
-with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
-but are otherwise a large and safe set of combinations
-for one to use.
-
-In summary,
-one might choose mappings that overwrite default key combinations,
-use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
-normal commands, \ttt{C-} and \ttt{S-} commands, or any \ttt{M-} and
-\ttt{M-S-} command), or
-create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
-More information about key notation is in \tttt{:h key-notation}.
-Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings
-in \tttt{:map}.
+'ona li lon e soweli 27'.
+
+Avoiding needless words...
+I've been avoiding e ni: and using only :.
+I've been omitting the subject (sometimes also the li)
+if it is the
+same as in the last sentence.
+I tend to group these words:
+noka and anpa; luka, suno, sike and lawa; pali and pana.
+These clusters of meaning hint me that an even simpler
+language is possible and maybe more optimized to the
+core goals of TP.
+
+Also, ambiguities introduced by omission of li
+and absence of a token to denote preposition,
+suggest to me the possibility of a syntax that is always uniquely
+parsed (or at least less ambiguous).
+
+I have been inclined to conceive a language with the same syntax of TP:
+'sentence' la 'sentence = subject + predicate + object (each term with
+a possible preposition complement).
+but always using particles to bound the sentence sections,
+repeating them when the function is performed in the smallest scale:
+'toki pi toki pona pi jan sona' meaning
+language in toki pona and of intellectuals vs
+'toki pi toki pona pi pi jan sona'
+meaning language in [toki pona of intelectuals].
+About the keyword for initiating a preposition complement 
+might be 'a', but it is already taken in TP.
+Maybe use 'a' before a preposition and use 'ha' or 'he'
+instead of TP a.
+In such a setting,
+one might enable pi li e a (concept-qualifier, subject-predicate,
+predicate-object, concept-preposition)
+in any slot of the template:
+'sentence' la 'sentence = subject + predicate + object (each term with
+a possible preposition complement)'
+Between sentences one would use:
+la pi, la li, la e, la a, and la la.
+This is a very fractal conlang proposal.
+Maybe also have a way to discern between a noun,
+a qualifier (adjectives/adverbs) and a verb,
+and accept any of them for the subject, predicate, object, preposition,
+slots. Or assume a part of speech as default for a slot or for the
+first word in a slot.
+
+\section{Final words in Toki Pona}\label{ftp}
+toki li nasin e lawa.
+li nasin tawa (pi) toki insa.
+
+% jan li ken toki li ken pali pona e toki pona.
+toki pona li pona e nasin tan ni:
+ona li jo e nimi nanpa lili.
+ona li pona.
+pona kepeken weka 'p' li ona, a.
+
+o taso la toki pona li kalama li lukin pona.
+sitelen en nasin li open e sitelen
+suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
+li sona. li nasin e toki e sona e lawa e lon.
+
+toki ni li wawa tawa jan mute nasin en toki.
+wawa tawa toki pi jan sona.
+pi ilo nanpa en nanpa nasin.
+taso tawa toki sona a.
+sitelen sona, sitelen musi.
+
+ilo lon Poki~\ref{hacks} li pana e sitelen
+e sona tan sitelen,
+en nasin.
+Poki~\ref{basics} li pana e nasin pi toki pona.
+e sitelen tawa kama sona.
+
+mi wile pali e sitelen lon toki pona.
+sitelen lon nasin, sitelen sona,
+sitelen musi.
+ante e toki pona
+la ona li nasa.
+taso nasa li pona mute.
+li pona mute tawa lawa,
+tawa kama sona, tawa sitelen e toki.
+ante la toki pona o.
+toki e toki pona tawa sina.
+kama sona e toki pona sina.
+
+o pona tawa jan pi toki pona.
+tawa jan Sonja, Birns-Sprage, Kipo, Pije,
+Siwejo, Malija, Tepan, jan kulupu mute.
 
@@
 \fontsize{11}{0}\selectfont
-\bibitem{anPh}
-	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
-	Available at \url{https://youtu.be/oeOKYc3-nbM}
-\bibitem{anPh2}
-	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
-\bibitem{wikiVim}
-  Vim (text editor). (2017, December 4). In Wikipedia, The Free
-    Encyclopedia. Retrieved 07:08, December 16, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Vim_(text_editor)&oldid=813716105}
-\bibitem{ouroWiki}
-  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
-\bibitem{vimrc}
-	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
-\bibitem{ex}
-  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
-\bibitem{tokipona}
-	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
+\bibitem{wordnet}
+  Fellbaum, C. (1998). WordNet. John Wiley \& Sons, Inc..
+\bibitem{corpus}
+    Henry, J. lipu pi toki pona pi jan Jakopo. Available at         
+    \url{http://jimhenry.conlang.org/conlang/tokipona/tokipona.htm} 
+\bibitem{vim}
+    Fabbri, R. An anthropological account of the Vim editor:
+features and tweaks after 10 years of usage. Available at         
+    \url{https://github.com/ttm/vim/raw/master/article/article.pdf} 
+\bibitem{sapWho}
+  O'Neill, S. P. (2015). Sapir–Whorf Hypothesis. The International
+    Encyclopedia of Language and Social Interaction.
 \bibitem{tpLang}
@@
     ISBN-10: 0978292308, ISBN-13: 978-0978292300.
-\bibitem{aa1}
-	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G.
-(2013).
-AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology.
-RESI.
-From \url{https://arxiv.org/abs/1604.08255}
-\bibitem{aa2}
-	Fabbri, R. (2017).
-The Algorithmic-Autoregulation (AA) Methodology and Software:
-a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
-\bibitem{makefile}
-	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
+\bibitem{kama}
+  Knight, B. (2017) o kama sona e toki pona! Available at:
+    \url{http://tokipona.net/tp/janpije/okamasona.php}
+\bibitem{tp3}
+  Blahuš, M. (2011). Toki pona–eine minimalistische Plansprache.
+    Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
+\bibitem{tp4}
+  Schneider, S. (2017). nasin toki pi toki pona. Available at
+  \url{https://github.com/stefichjo/toki-pona}
+\bibitem{interview}
+  Lang, S. and Broholm, K. AFP 20 – Sonja Lang: Toki Pona, Conlanging
+    and the meaning of life (Sonja Lang interview). Available at
+    \url{https://actualfluency.com/afp-20-sonja-lang-toki-pona-conlanging-meaning-life/}
+\bibitem{memrise}
+  Toki Pona Community. (2017). Various Memrise courses on Toki Pona.
+    \url{https://www.memrise.com/courses/english/?q=toki+pona}
+\bibitem{gdoc}
+  Toki Pona Community (2017). toki pona central hub preparation.
+    Available at
+    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
+\bibitem{wikiToki}
+  Toki Pona. (2017, December 11). In Wikipedia, The Free Encyclopedia.
+    Retrieved 22:48, December 14, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
+\bibitem{conlanWikip}
+  Constructed language.
+(2017, November 1).
+In Wikipedia, The Free Encyclopedia.
+Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
+\bibitem{artlanWikip}
+  Artificial language. (2017, November 26). In Wikipedia, The Free
+  Encyclopedia. Retrieved 23:00, December 14, 2017, from
+  \url{https://en.wikipedia.org/w/index.php?title=Artificial_language&oldid=812127746}
+\bibitem{akesiWawa}
+  Fabbri, R. (2017, November 6). o awen, akesi wawa.
+  In the group toki pona taso. Available at
+  \url{https://www.facebook.com/groups/tokiponataso/permalink/1895942817391318/}
+\bibitem{janKipo}
+  Clifford, J. et al. (2017, December 15). Considerations about Toki Pona
+  phrases.
+  In the Facebook group 'toki pona'. Available at
+  \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
+\bibitem{tokipona}
+	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
+\bibitem{tpNetSH}
+  Knight, B. (2017). Online gadget for syntax highlighting Toki Pona
+  texts. . Available at
+  \url{http://tokipona.net/tp/DisplayText.aspx}
+\bibitem{stack}
+  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
+    Software Engineering in Stack Exchange. Available at
+    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
+\bibitem{solarized}
+  Schoonover, E. Solarized color scheme. Available at
+  \url{http://ethanschoonover.com/solarized}
+\bibitem{dv1}
+	Munzner, T. (2014). Visualization analysis and design. CRC press.
+\bibitem{dv2}
+  Telea, A. C. (2014). Data visualization: principles and practice.
+  CRC Press.
+\bibitem{dv3}
+  Ward, M. O., Grinstein, G., \& Keim, D. (2010). Interactive data
+  visualization: foundations, techniques, and applications. CRC Press.
+\bibitem{dv4}
+  Ware, C. (2012). Information visualization: perception for design.
+  Elsevier.
+\bibitem{blue}
+  Mehta, R., \& Zhu, R. J. (2009). Blue or red? Exploring the effect of
+  color on cognitive task performances. Science, 323(5918), 1226-1229.
+\bibitem{blue2}
+  Viola, A. U., James, L. M., Schlangen, L. J., \& Dijk, D. J. (2008).
+  Blue-enriched white light in the workplace improves self-reported
+  alertness, performance and sleep quality. Scandinavian journal of
+  work, environment \& health, 297-306.
+\bibitem{fonts}
+  Spolsky, A. J. (2008). User interface design for programmers.
+  Apress.
+\bibitem{mass}
+	Fabbri, R., Vieira, V.
+, Pessotti, A.
+C.
+S., Corrêa, D.
+C., Oliveira Jr, O.
+N.
+O.
+Musical elements in the discrete-time representation of sound (2017).
+arXiv preprint arXiv:1412.6853.
+Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
+\bibitem{wikiPoetry}
+Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
+    Retrieved 19:46, December 3, 2017, from
+    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
+\bibitem{tokisona}
+  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
+  Available at \url{http://tokisona.github.io/}
+\bibitem{Wikipesija}
+  Toki Pona community (2017). Wikipesija. Available at:
+    \url{http://tokipona.wikia.com}
 \end{thebibliography}

### 2018-02-11 article/pos.tex
moved plugin files to plugin dir, to allow vimball to get plugin/ and syntax/ tree

@@
 \caption{POS tags incident and chosen.
-            The official dictionary often relates tokens
+            The official dictionary often relates one token
             to more than one POS tag.
@@
             Section~\ref{shigh}, for example,
-            a token has to have an established tag to have
+            a token needs to have an established tag to have
             a defined color.

### 2018-02-11 article/article.tex
version used to deliver Toki Pona's Vim plugin v0.01b

@@
 \title{Basic concepts and tools for the Toki Pona minimalist and
-constructed language:
+constructed language: description of the language and main issues;
 Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}

### 2018-02-11 article/pos.tex
version used to deliver Toki Pona's Vim plugin v0.01b

--- 2017-12-16 article/pos.tex
@@
 \begin{center}
-\caption{POS tags incident and chosen.
+\caption{"POS tags incident and chosen.
             The official dictionary often relates tokens
             to more than one POS tag.
-            For the current version of the text highlighting plugin described in
-            Section~\ref{shigh}, for example,
+            For the text highlighting Plugin, for example,
             a token has to have an established tag to have
             a defined color.
-            On the 'Chosen' column, the tokens were regarded only once
-            by choosing the class in the dictionary respecting the
-            precedence order: PRE-VERB, VERB, PREPOSITION, PARTICLE, ADJECTIVE, NOUN, NUMBER.}\label{foobar}
-\begin{tabular}{ l | c c }
+            On the Chosen column, the tokens were regarded only once
+            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}\label{foobar}
+\begin{tabular}{| l | c | c |}\hline
 POS & All  & Chosen \\\hline
-NOUN & 58  & 49 \\
-ADJECTIVE & 40  & 34 \\
-VERB & 15  & 13 \\
-PARTICLE & 12  & 12 \\
-PRE-VERB & 6  & 6 \\
-PREPOSITION & 5  & 5 \\
+NOUN & 58  & 49 \\\hline
+ADJECTIVE & 40  & 37 \\\hline
+VERB & 15  & 15 \\\hline
+PARTICLE & 12  & 12 \\\hline
+PREPOSITION & 5  & 5 \\\hline
 NUMBER & 4  & 1 \\\hline
-total & 140  & 120 \\
+PRE & 0  & 0 \\\hline
+total & 134  & 120 \\\hline
 \end{tabular}\end{center}

### 2018-02-11 tokipona/makeVimSyntax.py

@@ -132,24 +132,42 @@ lines = []
 for class_ in words:
     words_ = ' '.join(words[class_])
     group = colors[class_]
-    lines.append('\nsyntax keyword '+'tokipona'+class_+' '+words_)
+    if class_ != 'tokiponaComment':
+        lines.append('\nsyntax keyword '+'tokipona'+class_+' '+words_)
+    else:
+        lines.append('\nsyntax match '+'tokipona'+class_+' '+words_)
     lines.append('highlight link '+'tokipona'+class_+' '+group)
 
 coloring = '\n'.join(lines)
 header = '''" Vim syntax file
-" Language:	toki pona
-" Maintainer:	Renato Fabbri <fabbri@usp.br>
-" Last Change:	2016 Apr 30
-" Remark:	toki pona is a conlang, not a programming language
+" Language:     toki pona
+" Maintainer:   Renato Fabbri <renato.fabbri@gmail.com>
+" Last Change:  2018 Fev 11
+" URL:          https://github.com/ttm/tokipona
+" Remark:       toki pona is a conlang, not a programming language
 
 if exists("b:current_syntax")
     finish
 endif
 
+" define here so loaded files might test for it (unlet at the end)
+if !exists("main_syntax")
+  let main_syntax = 'tokipona'
+endif
+
 syntax clear
 syntax case ignore\n\n'''
+footer = '''
+
+syntax match tokiponaComment /#.*$/
+highlight link tokiponaComment StatusLine
+
+let b:current_syntax = "tokipona"
+if main_syntax == 'tokipona'
+  unlet main_syntax
+endif'''
 with open('../syntax/tokipona.vim','w') as f:
-    f.write(header+coloring)
+    f.write(header+coloring+footer)
 
 # class_count has the number of occurrences of each pos tag.
 # each of them in the chosen form and in any of the classifications

### 2018-02-12 setup.py
starting to write files to upload package to pypi

@@ -1,0 +1,149 @@
+from setuptools import setup, find_packages  # Always prefer setuptools over distutils
+from codecs import open  # To use a consistent encoding
+from os import path
+
+here = path.abspath(path.dirname(__file__))
+
+# Get the long description from the relevant file
+with open(path.join(here, 'README.md'), encoding='utf-8') as f:
+    long_description = f.read()
+
+setup(
+    name='tokipona',
+
+    # Versions should comply with PEP440.  For a discussion on single-sourcing
+    # the version across setup.py and the project code, see
+    # https://packaging.python.org/en/latest/single_source_version.html
+    # version='0.1.dev0',
+    # according to https://semver.org/
+    version='0.01b',
+
+    description='A package for dealing with toki pona: vim syntax highlighting, tokipona wordnets, analysis of the vocabulary, synthesis of texts',
+    long_description=long_description,
+
+    # The project's main homepage.
+    url='https://github.com/ttm/tokipona',
+
+    # Author details
+    author='Renato Fabbri',
+    author_email='renato.fabbri@gmail.com',
+
+    # Choose your license
+    license='MIT',
+
+    # See https://pypi.python.org/pypi?%3Aaction=list_classifiers
+    classifiers=[
+        # How mature is this project? Common values are
+        #   3 - Alpha
+        #   4 - Beta
+        #   5 - Production/Stable
+        'Development Status :: 4 - Beta',
+
+        # Indicate who your project is intended for
+        'Intended Audience :: Science/Research',
+        'Intended Audience :: Developers',
+        'Intended Audience :: Education',
+        'Intended Audience :: Religion',
+        'Intended Audience :: Other Audience',
+        'License :: OSI Approved :: GNU Lesser General Public License v3 or later (LGPLv3+)',
+        'Natural Language :: English'…
+        'Natural Language :: Toki Pona',
+        'Operating System :: Unix',
+        'Topic :: Documentation',
+        'Topic :: Documentation :: Sphinx',
+        'Topic :: Education',
+        'Topic :: Education :: Computer Aided Instruction (CAI)',
+        'Topic :: Education :: Testing',
+        'Topic :: Games/Entertainment',
+        'Topic :: Multimedia',
+        'Topic :: Multimedia :: Graphics',
+        'Topic :: Multimedia :: Graphics :: Editors',
+        'Topic :: Multimedia :: Graphics :: Presentation',
+        'Topic :: Multimedia :: Graphics :: Viewers',
[…]

### 2018-02-12 setup.py
beta release candidate

@@ -46,7 +46,7 @@ setup(
         'Intended Audience :: Religion',
         'Intended Audience :: Other Audience',
         'License :: OSI Approved :: GNU Lesser General Public License v3 or later (LGPLv3+)',
-        'Natural Language :: English'…
+        'Natural Language :: English',
         'Natural Language :: Toki Pona',
         'Operating System :: Unix',
         'Topic :: Documentation',
@@ -114,7 +114,7 @@ setup(
     # requirements files see:
     # https://packaging.python.org/en/latest/requirements.html
     # TODO: test with virtualenv to know the dependencies
-    install_requires=['numpy', 'scipy', 'nltk'],
+    install_requires=['numpy', 'scipy', 'nltk', 'percolation'],
 
     # List additional groups of dependencies here (e.g. development dependencies).
     # You can install these using the following syntax, for example:

### 2018-02-12 tokipona/__init__.py

@@ -1,1 +1,21 @@
-from utils import *
+from . import utils
+def stats():
+    from . import makeStatistics as stats
+    return stats
+def synthesis():
+    from . import makeTexts as synthesis
+    return synthesis
+def syntax():
+    from . import makeVimSyntax as syntax
+    return syntax
+def wordnet():
+    from . import makeWordnet as wordnet
+    return wordnet
+# del makeStatistics, makeTexts, makeVimSyntax, makeWordnet
+
+__doc__ = """The main modules of this tokipona package are:
+- stats: for statistics of the official vocabulary.
+- synthesis: for synthesizing phrases, sentences, paragraphs, short narratives and poems.
+- syntax: to tweak and synthesize the Vim syntax file for Toki Pona.
+- wordnet: for thea achievement and usage of (very preliminary) Toki Pona wordnets.
+- utils: utilities for considering the space of all possible syllables and words given the Toki Pona rules"""

### 2018-02-12 tokipona/utils.py

@@ -1,5 +1,7 @@
-import nltk as k
+import nltk as k, os
 from percolation.rdf import c
+
+__doc__ = """Utilitiy variables and functions"""
 
 
 vowels = 'aeiou'
@@ -7,31 +9,48 @@ vowels_ = 'AEIOU'
 consonants = 'jklmnpstw'
 invalid_syllables = 'ji','ti','wo','wu'
 
-with file('rules.txt','rb') as f:
+tpath = os.path.abspath(__file__).replace(os.path.basename(__file__), '')
+
+with open(tpath + 'rules.txt','rb') as f:
     notes = f.read()
 
-def printTextsAccents():
+def printTextsAccents(text):
     '''Print toki pona text with accents in volgals of first syllabes'''
-    tokens = k.wordpunt_tokenize(text)
-    for token in i:
+    tokens = k.wordpunct_tokenize(text)
+    tokens_ = []
+    for token in tokens:
         if token[0] in vowels:
             token = vowels_[vowels.index(token[0])]+token[1:]
         else:
-            token = token[0]+vowels_[vowels.index(token[1])] +\
-                    token[2:]
+            token = token[0]+vowels_[vowels.index(token[1])] + token[2:]
         tokens_ += [token]
-    text_ = ' '.join(token_)
+    text_ = ' '.join(tokens_)
     print(text_)
     return text_
 
 
-def representAccents(text, 'jklmnpstw'):
-    '''exchange consonants such as k->g, p->b, t->d and s->z'''
+def representAccents(text):
+    '''Exchange consonants such as k->g, p->b, t->d and s->z
+    
+    Note
+    ----
+    Not implemented
+    
+    '''
     pass
 
 
 def consonantSyllables(consonant):
-    if consonant =! 'j':
+    '''Return all the syllables given the initial consonant.
+    
+    Notes
+    -----
+    TODO:
+    Needs revision. lj is not prohibited, but should it be allowed?
+    Send issues to discussion groups.
+    '''
+    global vowels_
+    if consonant != 'j':
         vowels_ = vowels + 'j'
     syllables = [consonant+vowel for vowel in vowels_ if
             consonant+vowel not in invalid_syllables]
@@ -40,6 +59,14 @@ def consonantSyllables(consonant):
 
 
 def allConsonantSyllables():
+    '''Return all possible syllables given the rules.
+
+    Notes
+    -----
+    TODO:
+    send syllables list to discussion groups.
+
+    '''
     # need to include the y. E.g. Enya, ijo.
     all_syllables = []
[…]

### 2018-02-12 tests/exampleRoutine.py
beta candidate almost approved

@@ -1,0 +1,48 @@
+# to import changes made in local package files,
+# e.g. if installed with pip3 install -e ~/repos/tokipona/
+import sys
+keys=tuple(sys.modules.keys())
+for key in keys:
+    if "tokipona" in key:
+        del sys.modules[key]
+
+### Basic usage
+import tokipona as t
+
+### the main functionalities are:
+
+# 1 - make a syntax highlighting file for Vim
+sh = t.syntax()
+
+# 2 - analyze the official Toki Pona vocabulary
+st = t.stats()
+
+# 3 - obtain (very preliminary) Toki Pona wordnet synsets
+wn = t.wordnet()
+
+# 4 - synthesize texts in Toki Pona
+sy = t.synthesis()
+
+print( sy.createParagraph() )
+print( sy.createPhrase() )
+print( sy.createPoem() )
+print( sy.createSentence() )
+
+# sh, st, wn, sy have objects related to each of the
+# functions as further described in this README.
+
+# 5 - further routines and variables in utils:
+v, c = t.utils.vowels, t.utils.consonants
+str_ = "TP vowels are: {}; TP consonants are: {}".format(v, c)
+print(str_)
+
+possible_words = t.utils.allTokiPonaPossibleWords(n=2)
+print("all possible words with two syllables given the TP rules are {}".format(possible_words))
+
+##############################################
+### check files in tokipona/
+# for many other functionalities and convenient variables.
+
+# Tweak at will.
+##############################################
+

### 2018-02-12 setup.py
finished beta release

@@ -16,7 +16,7 @@ setup(
     # https://packaging.python.org/en/latest/single_source_version.html
     # version='0.1.dev0',
     # according to https://semver.org/
-    version='0.01b',
+    version='0.1b1',
 
     description='A package for dealing with toki pona: vim syntax highlighting, tokipona wordnets, analysis of the vocabulary, synthesis of texts',
     long_description=long_description,
@@ -47,7 +47,6 @@ setup(
         'Intended Audience :: Other Audience',
         'License :: OSI Approved :: GNU Lesser General Public License v3 or later (LGPLv3+)',
         'Natural Language :: English',
-        'Natural Language :: Toki Pona',
         'Operating System :: Unix',
         'Topic :: Documentation',
         'Topic :: Documentation :: Sphinx',

### 2018-05-19 aux/article/article.tex
before cleaning aa

\documentclass{article}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	%pagebackref=true,
	pdfcreator={LaTeX with abnTeX2},
	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
	colorlinks=true,       		% false: boxed links; true: colored links
	linkcolor=blue,          	% color of internal links
	citecolor=blue,        		% color of links to bibliography
	filecolor=magenta,      		% color of file links
	urlcolor=blue,
	allbordercolors=black,
	bookmarksdepth=4
}
\usepackage[utf8]{inputenc}
\newcommand{\ttt}[1] {
	\texttt{<#1>}}
\newcommand{\tttt}[1]{\texttt{#1}}

\begin{document}
\title{An anthropological account of the\\ Vim text editor:\\
features and tweaks\\ after 10 years of ouroboros}
% Anthropological computation
\author{Renato Fabbri\\
\texttt{renato.fabbri@gmail.com}\\
University of São Paulo,\\
Institute of Mathematical and Computer Sciences\\
São Carlos, SP, Brazil
}
\maketitle
\begin{abstract}
The Vim text editor is very rich in capabilities
and thus complex.
This article is a description of Vim
and a set of considerations about its usage and design.
It results from more than ten years of experience
in using Vim for writing and editing various types of documents,
e.g. Python, C++, JavaScript, ChucK programs;
\LaTeX, Markdown, HTML, RDF and other markup-like files;  % TTM
binary files.
It is commonplace, in the Vim users and developers communities,
to say that it takes about ten years to master (or start mastering)
this text editor, and I find that other experienced users
have a different view of Vim and that they use a different
set of features.
Therefore, this document exposes my understandings in order
to confront my usage with that of other Vim users.
Another goal is
to make available a reference document with which new users
can grasp a sound overview by reading it and the discussions that
it might generate.
Also, the preliminary versions of this document already proved itself
useful for users of any degree of experience,
including me, as a compendium of commands, namespaces and tweaks.
This document might be enhanced and expanded
upon feedback, maturing of my Vim usage, or derivatives potentially
yield by or with other users.
\end{abstract}
{\bf keywords:} Vim, Text editor, HCI, Tutorial, Anthropological computer science

\section{Introduction}\label{intro}
Vim is a very complex text editor,
most often considered to be
matched only by Emacs.
They both are the standard advanced text editors
of the free software and open source communities
and have been developed for decades.
Vim is very useful because:
\begin{itemize}
  \item it is meant to be a plain text (e.g. ASCII, UTF-8) editor
  and does not (by standard) insert special characters (e.g. for
    formatting or with binary instructions).
  \item It has a powerful architecture and set of commands.
  \item It is highly configurable and most often the users
  hold a set of commands for standard settings and routines kept in
    the vimrc~\cite{vimrc} and other configuration files.
  \item It has been used for more than 25 years, is based on
    established technologies, and is an established text editor.
    Thus, it has a considerable and seasoned user
    base, high quality (both official and unofficial) documentations, 
    and countless publicly available scripts, most often in the form of
    plugins.
\end{itemize}

This document describes the Vim text editor
and proposes a set of sharpening of the user
experience through simple tweaks
and utilization strategies.
The contents herein presented is a
report on the overall understandings I
have of Vim after a bit more
than ten years using it,
resorting to the (very mature) official documentation
whenever possible.
The purposes of this document are:
\begin{itemize}
  \item to help new users grasp Vim essentials
  and convenient practices.
  \item To attain a sound overall description of the editor.
  \item To record the comprehension about the editor that
  a user (me) has after $10$ years of usage.
  \item To confront my usage with that of other experienced
  users. This is helpful for me, but also for the other users
  as they might benefit from this content and from discussions
  that might arise from it.
  \item To propose some enhancements to Vim through simple tweaks and potential plugins.
\end{itemize}

Advanced users might just skim through
Section~\ref{basics}, where standard capabilities of
the editor should become clear, and consider more carefully
Section~\ref{namespaces}.
The concluding remarks and proposed enhancements in Section~\ref{conc}
may also deserve some attention because
Vim is constantly evolving
and there are many possible enhancements
(often made available e.g. as plugins).

\subsection{Further remarks about this document}\label{furRem}
This document is written is a DRY KISS
(Don't Repeat Yourself, Keep It Simple Stupid) style.
Complex is to master the use of Vim
and one finds sound
references in help files and a nice vimrc.
Therefore the following content is kept
as uncomplicated and original as possible.
Also, because of Vim's complexity and entailed
bond of this document to my usage,
there is an anthropological component
which is evident in the occasional use of the first person.
This can be understood as anthropological computer science~\cite{anPh,anPh2}
and considered to help
in the technological groundwork of the civil society.
Accordingly, here are some notes about my experience:
I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writing and editing computer code 
(Python, Javascript, C++, ChucK, bash, etc), markup-like content
(HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited other types of files, such as database files with
dozens of gigabytes.
Within Vim, I mostly write (web and scientific) software,
scientific articles, music, poems and short stories.

% Vi and Vim
\subsection{Historical note}
Vim was first released publicly in 1991.
It is a cross-platform GNU licensed free and open source
extended clone of Bill Joy's vi text editor.
Vi was written in 1976 as a hard link to ex: a shorthand
to start ex in visual mode, i.e. vi is ex.
Vim's development is coordinated (and performed) since the beginning
mainly by Bram Moolenaar.
Today, bleeding-edge version is 8.0.1497.
There seems to be no official or explicit stable, alpha or beta releases~\cite{noOfficial}.
I found no scientific article on Vim 
(this might be the first one), although there are
books, software and third-party documentation on the
web.~\cite{wikiVim}

\section{An overview of the basics}\label{basics}
Vim's interface is text-based.
In the GUI (gVim),
there are convenient menus and toolbars
but all functionalities are still available through
the command-line mode.
VimL (a.k.a. VimScript, Vim script, Vim language)
is the internal language of Vim,
and is often used for scripting by users
although other languages might be used 
(e.g. Python, Perl, Lua, Racker, Ruby and Tcl). 
VimL is preferred among other names for the language
because it is shorter and less ambiguous then the alternatives,
although there seems to be no official name and the manual
uses the terms \emph{Vim script language} and \emph{Vim script}
(see e.g. \tttt{:h usr\_41}).
Each line of a VimL script is a command on the
command-line mode.
This section may be considered a tutorial
that focuses on the namespaces, i.e. sets of tokens
that carry values or trigger procedures.
One should see Appendix~\ref{sec:not}
to understand the notation Vim uses to represent key combinations
(e.g. \ttt{C-[}).

\subsection{The bare minimum}\label{minimum}
You open a file at Vim startup by executing
the command: \texttt{\$ vim <filename>}.
Inside Vim, you start in the normal
mode, and might want to move around using
\texttt{h-j-k-l} for left-down-up-right.
To insert characters, move your
cursor to the desired location an press i,
which puts Vim in the insert mode.
Go back to normal mode by pressing
\texttt{<ESC>}, \ttt{C-$[$} or \texttt{<C-C>}.
You save the file by typing \texttt{:w<CR>},
and exit Vim by typing \texttt{:q<CR>}.
You can save and quit with \tttt{:wq<CR>}
or \tttt{:x<CR>} or \tttt{ZZ}.

\subsection{Vim help}
Help on using Vim is found in various places.
The standard resource is the Vim help files.
They are accessed by typing \texttt{:h <anything><CR>}
in normal mode.
Examples of such \texttt{<anything>} are:
color, navigation, :vs, vimtutor.
Type 
\texttt{:h usr\_toc<CR>}
to access the official User Manual,
which is considerably lengthy and complex
and is usually not read by users before they achieve a few years
of experience.
In learning Vim, one
might want to run the \texttt{\$ vimtutor} command
(outside Vim) to start the Vim Tutor.

There are good resources on the Web for learning
and tweaking Vim:
\begin{itemize}
  \item ``Vim Adventures'' is an online RPG game for practicing
  and memorizing Vim commands. This game is quite famous among Vim users.
  \item There are official and semi-official Vim sites e.g.:
  \url{www.vim.org}, \url{https://www.vi-improved.org} and
  \url{http://vim.wikia.com/}.
  \item Many hacks, understandings and general issues
  (e.g. how to make such a move) are asked and answered
  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
  One often finds these links through a search engine.
  \item Many videos about Vim are publicly available.
  One might find them using a search engine
  (e.g. \url{http://derekwyatt.org/vim/tutorials/}, and
  \url{http://vimcasts.org}) or in Youtube and Vimeo.
\end{itemize}

% any other commands? files?
\subsection{Namespaces}\label{namespaces}
Vim is a text editor ouroboros~\cite{ouroWiki} because
text and writing alters text and writing.
This is achieved through a collection of namespaces where tokens have scalar
or arbitrarily complex values, or start automated routines.

\subsubsection{Commands and mappings}
There are commands, i.e. typing sequences which trigger actions,
for each mode:
\begin{itemize}
	\item in Normal mode all keys are mapped to commands.
		There is redundancy and additional commands
		using Ctrl and Shift keys.
		Some keys expect a second key,
    and have combinations not used (thus available for new mappings),
		specially the z and g keys.
    See Section~\ref{normal} and Appendix~\ref{notes}
    for more insights into the commands available in the
    normal mode.
	\item In the other modes, the sequences available for mappings are more obvious and abundant.
		One should look at \tttt{:h index} to know about all the standard mappings
		and use \tttt{:map} to list the user-defined mappings.
\end{itemize}

\ttt{C-\textbackslash} is often reserved for extensions,
which makes it a safe namespace to use (while there are no
such extensions).
   `E 
   $ `E $

AAOA colon command can be written as a string
and executed by the \tttt{:execute} colon command.
E.g. \tttt{:execute 'vs afile.txt'} executes the \tttt{:vs afile.txt}
colon command..
As there are colon commands that execute commands in other
modes, e.g. \tttt{:normal ?\textasciicircum def},
the \tttt{:execute} is a way to build commands in any mode,
e.g. \tttt{:execute 'normal i'.string(atan(bufnr('\%'))).'<ESC>'}.

\subsubsection{Variables}
TThere are some types of variables in Vim:
\begin{itemize}
	\item Environment variables: names start with \$ and hold system
		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
	\item Option variables: names start with \& and are meant to control the behavior of the editor.
		One might change a value through set or let, e.g.
		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
    More on registers in Section~\ref{state} and scattered in this
    document.
  \item Internal variables are created with \tttt{:let} and preferably have a prefix:
	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
		are local to the buffer, window, tab page, function, and
		sourced Vim file, respectively.
 \tttt{v:}, \tttt{g:} are globals, the first are predefined by Vim.
		\tttt{a:} is for function arguments.
		If there is no prefix, the variable is global or internal to a function if defined inside a function.
		More about internal variables in \tttt{:h internal-variables}.
	\item The value of a variable can be of the types: scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
\end{itemize}
You can \tttt{:echo} any of such variables or use them in expressions.
Notice that you will only be able to echo a \tttt{b:} variable inside
the buffer where it is defined.
For all the VimL capabilities, including loops, conditionals,
and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
Classes are possible only in rudimentary forms, e.g. through dictionaries,
but the language is otherwise overall quite powerful,
specially in dealing with text and editor behavior, as expected.

\subsubsection{State lists}\label{state}
rVoVVim keeps a number of useful lists which expresses the state of the editor:
\begin{itemize}
	\item All the entered commands are accessed through \tttt{:hist a}. 
    The tokens ``\tttt{a / e : i d}'' may be used for specific types of commands, such
		as search and colon commands.
	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
  \item The windows are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
	\item A tabs list can be reached through \tttt{:tabs}.
		It is usual both to show and hide the tabs bar (mapping
    in~\cite{vimrc}, discussion in Section~\ref{visual1}).
	\item Jumps are available through \tttt{:jumps}.
		One positions the cursor at each jump through \ttt{C-O} and \ttt{C-I}.
  \item Registers are available through the \tttt{:reg} command, as variables and through shortcuts in different modes.
    They also keep track or your copy, edition and deletion and are promptly defined
    by recording a typing sequence with the \tttt{q} normal command.
    Vim keeps only the last edition, in register \tttt{".}.
    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
    A hack to keep the latest deletions and copies in the standard register
    might follow the same pattern, but use another \tttt{:h event} and
    monitor register \tttt{""} (maybe also monitor \tttt{"0}).
	\item An undo list is accessed through \tttt{:changes}.
	\item A list with all the sourced scripts in a Vim session is displayed through \tttt{:scriptnames}.
	\item The markers defined are listed with the \tttt{:marks} command.
		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
		Uppercase letters are cross buffers.
	\item Quickfix and Location lists are populated through \tttt{:vim} and \tttt{:make}
		and variations, such as \tttt{:grep}.
		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
		to open the Quickfix window, where the lines of occurrence are in sequence
		and one can \ttt{CR} one of them to have the cursor in the main window active at
		the first character of the match.
		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
		to use the location-list window instead of the Quickfix, which
		is very similar, but one per window instead of one per buffer.
		More information in \tttt{:h quickfix}.
	\item A tags list have to be made in order to enable the use of tags.
		Most often one will generate the tags list using the exuberant ctags,
		which supports dozens of languages.
		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
		and then using \ttt{C-]} to
		go to the position of tag under cursor, and \tttt{:vs tags} to open
		the tags file.
	\item The argument list holds a list of files to be edited or browsed. 
		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
	or using commands (e.g. \tttt{:ar ./*}).
		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
		one might perform actions on each file in argument list using \tttt{:argdo}.
		All files in argument list are also in the buffers list.
    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
  \item All the autocommands are listed with \tttt{:au}.
    They are event-triggered actions Vim performs.
\end{itemize}

A window with information about the state of the editor
can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
Also, this script might be examined because it has
a collection of commands to access various settings of Vim.
Another good list of commands to know about Vim's state is kept on
\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
I would specially highlight the \tttt{:syntax} command because
it displays token groups and their meanings when run
inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
More about using colors in Vim for syntax highlighting in Section~\ref{visual2}.

\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
You can keep track of the editor state though commands, as stated
in last subsection.
Also, one might rely on persistent visual
cues, specially the tabs bar, the status line, and the line
reserved for the command-line.
A good strategy I find is to have specialized visual cues of the state to make persistent or hide
and a mapping to toggle each of them.
Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
status line and tabs bar with \tttt{<localleader>-T and B}, according to script~\cite{vimrc}.
I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
Numbering is always there, I rarely turn them off but keep the mappings \tttt{<leader>-n and N} to toggle just in case.
Highlight of last search string, and underline of unrecognized words (e.g. in English of Portuguese), are toggled with \tttt{<localleader>-l and p}.
Instead of keeping the status bar, I use \ttt{C-G} to know about the
file and \ttt{gC-G} to know more and rarely.
It seems not possible to remove the statusbar between horizontal splits.
After asking in online forums and experimenting, I realized that it seems
reasonable to keep at least one line dividing the windows, so
if it comes to it, I just \tttt{set statusline=-}.
Unfortunately, as far as I could dig, one will need to alter
the Vim's source code to enable a horizontal split without losing a line,
which, for me, seems far from the ideal.
An example solution is to implement a visual cue for the
first and/or last line of the windows in the line-numbers
column, or complete the spaces and empty chars with \$\$\$\$,

\subsubsection{On the persistence of the editor state}
For state persistence, one might keep an undo file for each file as
in~\cite{vimrc}.
Sessions are easy to manage, enabling one to save and load the
editor's state, with the opened windows, tabs, buffers, etc.
The mappings in~\cite{vimrc} keep the sessions
in a reasonable directory and makes it easy to remember and tweak
the standard commands to deal with sessions.
More information in \tttt{:h sessions}.
One might use \tttt{:h views} to keep the state of one window,
but sessions keep all the states from all windows.
This entails a strategy to deal with Vim that is similar to
the use of
Byobu/Tmux/Screen\footnote{\url{https://en.wikipedia.org/wiki/Byobu_(software)},
\url{https://github.com/tmux/tmux/wiki},
\url{https://www.gnu.org/software/screen/}},
because one can rely on restoring the state of the windows.
The main limitation I found to this approach is that
Vim is not keeping track of the terminals opened.
If you open a terminal inside Vim with the \tttt{:term}
command, you might save the session as usual, but when loading
you get a dummy empty window for the terminal and an error message.
More about Terminal-Job mode in Section~\ref{terminal}.

Autocommands are the standard way to define event-triggered routines in Vim.
These are often related to particular file types, but are also often
in defining some general automated behavior for Vim.
If the autocommands are placed inside a configuration file (e.g. the
vimrc), the automated behavior is persistent across Vim instances.
E.g. in~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).

\subsection{Using Vim's modes}\label{modes}
These are the  basic and fully implemented modes of usage in Vim:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A core goal of the normal mode it to support fast
  navigation of the document while allowing
  the typist to maintain the fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manipulating text
  (e.g. copy, paste, delete, change case) and
    changing to other modes.
    More in Section~\ref{normal}.
  \item Insert mode: for inserting text. More in
    Section~\ref{sec:ins}.
  \item Command-line mode: for entering Ex commands. More in
    Section~\ref{sec:com}.
  \item Ex mode: similar to the command-line mode,
  but more specialized for running various Ex commands.
    More in Section~\ref{sec:ex}.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access:
    \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

There is another basic mode, but it is not fully implemented:
the Terminal-Job mode (more in Section~\ref{terminal}).
There are seven additional modes which are mostly subordinate 
to the basic modes and that will be described when convenient.
The manual page for Vim modes can be accessed by typing
\texttt{:h vim-mode}.
Some of the modes are now further considered for
the achievement of
an overview of the Vim usage possibilities.

\subsubsection{Normal mode}\label{normal}
Sometimes also called navigation or command mode,
the normal mode is most powerful for
navigating, manipulating text and changing to other modes.
The simplest of these three is changing to other modes:
type any of these letters to change to insert mode:
\texttt{iIaAoOsScC}. More on the transition between
normal and insert mode on Section~\ref{navIn}.
Type any of these characters to change to command-line mode:
\texttt{:/?}.
Type \texttt{Q} to enter Ex mode.
Type \texttt{v}, \texttt{V}, or \texttt{CTRL+V} to enter visual mode.

For very basic and naive navigation, one should check Section~\ref{minimum}.
There are many facilities to navigate Vim
as explained in \texttt{:h navigation}.
Most often, one uses:
\begin{itemize}
  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
and whole page down and up, respectively, although these commands might
be set to scroll a different number of lines.
  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
  \item \texttt{(w,b,e)} to move to the next and previous word,
    and next end-of-word.
    These are motions to iterate over sequences of characters separated
    by special characters (e.g. punctuation and parenthesis) as
    specified by the output of \texttt{:se iskeyword}.
    To iterate over space-separated tokens, use \texttt{W,B,E}).
    To move to the end of last word, one might use \texttt{be}
    or \texttt{ge}.
The \texttt{),(} commands iterate through sentences,
		\tttt{\},\{} through text blocks separated by empty lines.
    \tttt{[} and \tttt{]} are used in combination to browse sections.
  \item \texttt{(fX,tX,FX, TX)} to move, in the same line, to or just before any X character,
	  \tttt{;} and \tttt{,} for next and previous found character.
  \item Search with \tttt{/} or \tttt{?} (these are considered command-line commands).
  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
    jump list.
  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X.
    \texttt{`X} moves to the exact position while \texttt{'X} moves to
    the first non-blank character of the line.
    A mark is registered by the user in any cursor position
    by typing \texttt{mX}, where X is any letter.
    If X is lowercase, the mark is local to the buffer (the file),
    if it is uppercase or numeric, it is global to the Vim session
    (cross buffers).
\end{itemize}

For changing the text, usual commands include:
\begin{itemize}
  \item \texttt{d\{motion\}} to delete the characters
    involved in the motion command.
  \item \texttt{dd,D} to delete a line or from the cursor to the end of the line.
  \item \texttt{x,X} to delete the character under or before the cursor.
  \item \texttt{$\sim$} to swap the case of a character or selection.
  \item \texttt{gu\{motion\},gU\{motion\},g$\sim$\{motion\}} to make lowercase, uppercase or switch the case of the characters involved in the motion.
\end{itemize}

There are way more commands to change text.
Some of them are discussed in Section~\ref{navIn}
because they involve a transition to the insert mode.
A thorough consideration of the commands in the normal mode
is found by executing \texttt{:h navigation}, 
\texttt{:h change.txt}, \tttt{:h index}.

\subsubsection{Insert mode}\label{sec:ins}
Once in the insert mode, the character keys
input the characters at the cursor position in the current buffer.
One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
\ttt{C-C}),
and Vim will be put in the normal mode.
Most useful commands in insert mode include:
\begin{itemize}
  \item \ttt{C-O} to execute one and only command in normal mode.
	  This enters a secondary mode (see Section~\ref{modes})
  \item \ttt{C-R} to paste a register (a variable starting with
    '@', defined, copied or recorded through a \tttt{q} command in
    normal mode and as covered in Section~\ref{state}).
  \item \ttt{C-T} to indent current line.
  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
    the line or to the previous word.
  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand.
  \item \ttt{C-X} commands for scrolling the window with multiple
    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities
    (\ttt{C-X}\ttt{C-(F,I,L,O,V} for file, identifier, whole line, omni and command-line completions). 
\end{itemize}

\subsubsection{Normal $\rightarrow$ insert modes}\label{navIn}
Many commands bridge from Normal to Insert modes,
e.g. \tttt{iws} or any of these letters: \tttt{csrCSR}.
These make convenient the replacement of text and populates registers.
The absence of a short command to insert one char is
a known issue in Vim.
Reasonable mappings to insert or append a char to and around
another char are in~\cite{vimrc}.
Vim couples operator and motion commands by design.
There are many operator commands that take the editor from
the normal mode to the insert mode, most of them
favoring deletion or change, as detailed in \tttt{:h operator}.
Motion commands are described in \tttt{:h motion}.

\subsubsection{Command-line mode}\label{sec:com}
This mode is dedicated to writing colon, search and filter commands,
entered through typing \tttt{:}, \tttt{?}, \tttt{/} and \tttt{!} in normal or visual modes.
Most useful commands in this mode include:
\begin{itemize}
  \item \ttt{C-B} and \ttt{C-E} to move cursor to the beginning and
    end of the line.
  \item \ttt{C-W} and \ttt{C-U} to delete last word or everything
    until the cursor.
  \item \ttt{C-R} to paste a register (as in insert mode).
\end{itemize}

\subsubsection{Ex mode}\label{sec:ex}
One might use the normal commands \tttt{q(:,/,?)} to have a window
with the colon or search command history to be edited normally,
and the chosen command can be run with \ttt{CR}.
VimL is largely based on the ex editor~\cite{ex},
and a more advanced user might use it for prototyping
by defining mappings and settings and managing scripts
probably in a \tttt{plugin/} folder of a directory in \tttt{:echo \&runtimepath}.
In the default interface started with the \tttt{Q} command in normal mode,
each command is input without entering : again. Use \tttt{:vi} to exit Ex mode,
follow documentation from \tttt{:h Ex-mode} for further information.

I might be loosing something, but for
tweaking I use the command-line window accessed through \tttt{q:}.
It is comfortable to browse the history in a normal-like mode,
edit them also using the insert mode as usual,
and having auto-completion when pressing \ttt{TAB}.
None of such features are available in Ex mode by default
and
customization by mappings have to be performed through autocommands as in~\cite{vimrc}.
The facilities exposed in Section~\ref{runExc} might complete the picture
that explains why this mode seems to be seldom used nowadays.

\subsubsection{Terminal-Job mode}\label{terminal}
This mode is reported as not having reached a stable usage design
(see \tttt{:h terminal}).
I find that it works exceptionally well and have often used it to run
scripts in an IPython shell, compile latex files, and open PDFs and images.
Vim browsing of windows and text manipulation is well developed,
so the Terminal-Job mode enables a very convenient integration
of files being edited and bash terminals,
more traditionally achieved through Byobu/Tmux/Screen terminal
multiplexers.
Most useful commands in the terminal-Job mode include:
\begin{itemize}
  \item \tttt{<C-W>N} and \tttt{<C-W>:} for entering the Terminal-Normal
    and command-line modes. Terminal-Normal mode is very similar to
    Normal mode, but one cannot change the text, cannot enter insert
    mode, and the status line reports if the job is finished or not.
  \item \tttt{<C-W>"} to paste a register.
  \item \tttt{i} or \tttt{a} for entering the Terminal-Job mode from the Terminal-Normal mode.
\end{itemize}
\noindent It is useful to define the same mappings for navigating splits and tabs
for both Normal and Terminal-Job modes, as in~\cite{vimrc}.

Because browsing the interface in Vim is fast,
and it is very comfortable to copy the terminal lines
in the Terminal-Normal mode,
it is often handy to keep (e.g. a tab with) some terminals:
e.g. one with an IPython shell, another two for compiling \LaTeX\
and opening PDF files (e.g. with \tttt{\$ evince <filename.pdf>}).
There are even more convenient ways to use the terminal inside Vim.
For example, 
one might use \tttt{:term} with the \tttt{++hidden} and \tttt{++close}
options to compile a \LaTeX\ file in the background without needing to further manage the
terminal and in a non-blocking manner: \tttt{:term ++hidden pdflatex
\% \#}.
I found the mappings on~\cite{vimrc} very helpful for directing
editor focus to splits (\ttt{A-(h,j,k,l)}) and tabs (\ttt{A-(a,d)}),
which I make available across Terminal-Job, Normal, and Insert modes.
I used \ttt{C-(h,j,k,l)} for splits, and \tttt{g(r,t)} for tabs,
for a long time (maybe years), but using \ttt{A-} makes it possible
to use consistently the same commands across the modes without
loosing any of the standard commands, and without using \ttt{leader}
and \ttt{localleader}, which are often used by plugins.

\subsection{Netrw}\label{netrw}
The standard interface in Vim for browsing file trees is Netrw.
It starts when you open a directory, such as with \tttt{:e .<CR>}.
It has solid support for browsing remote file trees (such as over ssh or
ftp) and handy e.g. mappings to open the files as splits and tabs
(specially \tttt{p,o,t}).
Most useful commands in Netrw include:
\begin{itemize}
  \item \tttt{d} and \tttt{\%} for creating directories and files.
    \ttt{Delete} removes both.
  \item \tttt{mf} and \tttt{mb} for marking files and bookmarking directories.
  \item \tttt{gb} and \tttt{uU} are used to load directories
    while marked files might be copied, moved, edited, grep-ed, tagged and migrated
    to and from the argslist as in \tttt{:h netrw-mf}.
\end{itemize}

There is no insert mode in Netrw interface;
the commands in \tttt{:h netrw-explore} are
convenient for opening the directory of the
file being edited (e.g. \tttt{:Se});
further information is in \tttt{:h netrw}.

\subsection{Standard configuration files and directories and my .vim/vimrc}
You can check the scripts Vim loads by using the debug script mentioned
in Section~\ref{state}.
By default, \tttt{$\sim$/.vimrc} and \tttt{$\sim$/.vim/vimrc} files are run by Vim at the beginning of the startup.
One might edit the vimrc file with \tttt{:e \$MYVIMRC}
and reload it with \tttt{:source \$MYVIMRC}.
Mappings in~\cite{vimrc} include such commands
to encourage a continuous enhancement of Vim settings
(they have helped me to improve my settings without unnecessary 
hassle and fast).
Any other file might be included to run at startup by
adding a line \tttt{:source <afile.vim>} in vimrc.
In fact, it is on the vimrc that one usually specifies the plugins
and plugin managers to be used.
Use an \tttt{after/} folder of a directory in \tttt{:se runtimepath}
or follow some patterns described on the next section to change the
scripts and sequence of them to be loaded.
The vimrc from other users are most useful for one to comprehend
and pick convenient practices and settings.
In fact, a vimrc file is most often a patchwork of excerpts of vimrc
files from other users.

\subsection{Plugins and packages}
One can see the list of the standard plugins
with \tttt{:h standard-plugin-list} command.
Any \tttt{plugin/**/*.vim} file inside a directory
listed in \tttt{:se runtimepath} will be loaded
(e.g. \tttt{.vim/plugin/something/ascript.vim}).
There are various ways to automate the installation
and enhance the management of plugins.
By default, one has the GETSCRIPT interface (see \tttt{:h getscript}),
that downloads latest scripts from Sourceforge as specified in \tttt{:h getscript-data},
and the Vimball interface, which creates and loads a Vimball representation of a
plugin.
Such a Vimball may be created with \tttt{:[range]MkVimball <filename> path}, where range specifies lines
that hold paths to files to be included in the \tttt{<filename>.vbm} Vimball.
The Vimball can be installed in a system by \tttt{:source <filename>.vbm}
or loading it at Vim startup with \tttt{\$ vim <filename>.vbm}.
Vimball files used the extension \tttt{.vba} and most of the official and
non-official documentation still use \tttt{.vba} but the extension now
is \tttt{.vbm}.

A Vim package is a directory that contain plugins.
It should be located inside a \tttt{pack/} directory
inside a directory listed in \tttt{:se runtimepath}.
The plugins found in \tttt{pack/<packName>/*/start/} are loaded
at startup, the plugins found at
\tttt{pack/<packName>/opt/**} are loaded with\\
\tttt{:packdd <script\_or\_directory\_name>}.

All the directories in which Vim looks for scripts are described
in \tttt{:h vimfiles} and are basically set by
\tttt{:se runtimepath} and conventions inside
each directory therein, such as to look for \tttt{vimrc} files
and \tttt{plugin/} or \tttt{pack/} directories.
VimL scripts can be loaded conditionally,
e.g. only if a function is used as in \tttt{:h autoload-functions}.
Example of such are filetype plugins (enabled by a \tttt{ftplugin/<filetype>.vim} file,
e.g. inside a plugin directory).
There is a number of plugin managers for Vim.
Pathogen and Vundle seem to be the most popular,
one because of its minimalism, the other because of
advanced features, e.g. for searching and installing plugins with
colon commands.

\subsection{Spell and spelllang (en and pt\_br)}
One might set the spelling language with\\ \tttt{:se spelllang=en\_us}
or \tttt{:se spelllang=pt\_br}
and toggle spell checking with \tttt{:setl spell!}.
Depending on the activities being performed,
these commands are used so often that one might use mappings as in~\cite{vimrc}.
Currently, Vim will download the files for a specific language if
not found in the system.

\subsection{Scripting, Functions, VimL and other languages (e.g.  Python)}\label{script}
In VimL,
the colon commands (also Ex commands or command-line commands) are related through spaces,
punctuation and keywords (see \texttt{:h script}).
Scripting the Vim editor can also be accomplished using other languages,
as well documented e.g. in \tttt{:h python}.
Functions are defined through colon commands and are called
inside colon commands e.g. \tttt{:call MyFunction()}
or \tttt{:echo MyFunction(4)}.
Notice that functions are not commands but might be bind to them
through colon commands e.g. \tttt{nnoremap gF :call MyFunction()<CR>}.
Executing source files is very straightforward with
\tttt{:so <filename>.vim},
and one can always use the Ex mode for rapid scripting.
At the same time, the \tttt{:term} and terminal-job mode make
scripting in general more convenient (not only scripting Vim), as output is
promptly navigated and copied (as discussed in Section~\ref{terminal}).
Command sequences to run only excepts of scripts, e.g.
to (re)define a function or recalculate some last lines,
are exposed in Section~\ref{runExc}.

\subsection{Color for syntax highlighting}\label{visual2}
Newer versions of Vim support 24-bit true color (aka 16 million colors)
in terminal Vim (in gVim true color received support earlier).
The terminal must support true color, and tests are available e.g. in
\url{https://gist.github.com/XVilka/8346728}.
Then Vim needs to be set to use true color with
\tttt{:set termguicolors}.
If using 8 or 16 bit colors, Vim uses the color palette from the
terminal. If using true color, each color is defined directly.
Settings for using true color inside Byobu/Tmux involve tweaking
and are available in~\cite{vimrc}.
Good color schemes to use with true color are Gruvbox
and Solarize.
One might source syntax files at any time to change syntax, usually
though linking tokens to syntax groups as in
\tttt{:syntax keyword <group\_name> <token1> <token2> <token3> ...},
or
\tttt{:syntax match <group\_name>	<pattern>},
and then relating the group to another group: 
\tttt{:highlight link <group\_name> <group\_name2>}
or to group characteristics directly:
\tttt{highlight <group\_name> guifg=\#ffffff}.
If you change a syntax file, reloading a file with with \tttt{:e<CR>}
updates the highlighting on the window with the corresponding file type.
A complete syntax highlighting support typically involves at least three files:
\begin{itemize}
  \item \tttt{ftdetect/<filetype>.vim}, where the file type is
    detected with e.g.
    \tttt{:autocmd BufNewFile,BufRead *.<file\_extension> setfiletype <filetype>}
  \item \tttt{ftplugin/<filetype>.vim} with general settings for the
    file type, such as: 
    \tttt{:set tabstop=2 softtabstop=2 shiftwidth=2 textwidth=70
    expandtab autoindent}.
  \item A \tttt{syntax/<filetype>.vim} file, with bindings between
    tokens and highlighting groups; and highlighting group definitions.
\end{itemize}

This scheme is implemented very straightforward in this
plugin~\cite{tokipona}
for highlighting text in the Toki Pona language~\cite{tpLang}.
Syntax highlighting plugins are file type plugins,
but also have a \tttt{syntax/<filetype>.vim} file
relating keywords, matches and regions to highlighting groups. 
One might see every highlighting group, and their final visual results,
with the commands\\
\tttt{:so \$VIMRUNTIME/syntax/hitest.vim}
or \tttt{:so hi}.
The\\ \tttt{ftdetect/} and \tttt{ftplugin/} folders load as expected in the \tttt{plugin/}
directory, but \tttt{syntax/} files has to be moved to
\tttt{$\sim$/.vim/syntax/}, unfortunately.
Routines for tweaking color schemes, available as a Vim plugin,
is described in Section~\ref{realcolors}.

\subsection{Fonts}
The fonts are defined by the terminal or inside gVim.
\ttt{C-+} and \ttt{C--} can be used to change font size.
Some settings for fonts, such as boldface, might be set using
the syntax highlighting facilities described in the previous section,
and are conveniently available through the mentioned plugin in Section~\ref{realcolors}.

\section{Results and discussion}\label{res}
The framework presented on the previous section yields
diverse results which are exemplified in this section
though selected examples.
Moreover, there are resulting effects in my performance
for research and development in Universities, public, private and
unregulated instances.
These include:
\begin{itemize}
  \item Fast and comfortable browsing and edition of diverse types text files.
  \item Continuous personalization of the Vim editor by means of configuration and scripting.
  \item Convenient and proper integration of text editing with other languages and the operational system (in my case, an Ubuntu GNU/Linux).
  \item Assimilation of advanced text processing techniques.
  \item Contributing to the Vim community by participation in online discussions and by making software available (e.g. Vim plugins).
\end{itemize}

Most importantly, true to the title and core purpose of the present document,
this section reports on scientific
and techological results which I am able to state after
10 years using Vim.

\subsection{The overview on the titanic Vim text editor}
In truth, this result is herein achieved by an orchestration of
all Sections~\ref{intro},~\ref{basics},~\ref{res}, and~\ref{conc}.
The benefits aimed for, and style used, are scattered in all the text,
but emphasis should be given so that this result is not overlooked.

If a programmer undertakes the endeavor
of learning Vim, and the learning curve is
as steep as is reported by both this document
the the Vim folklore,
it is of interest to know what kind of abilities,
knowledge, and framework at hand to expect in medium and long
terms.
The thorough consideration of the understand I have of
Vim features, all contained in the four sections of this article,
is very likely to help is such a complicated forecast.

Additionaly,
the ways to use Vim, and the capabilities one might achieve,
are blurry even for most advanced users.
This document is my attempt, \emph{after the emblematic milestone of using Vim for more then 10 years}, to make things nitid.
A newcommer or intermeddiate-level user
will need to read manual pages, use web search engines,
and use Vim for large periods, in order to develop
fluency with the concept and usage of Vim.
And an advanced user will most likely still maintain
a handmade vimrc and suitable collection of notes and scripts.
Nevertheless, they might benefit in numerous ways from the collections of
e.g. insights, technical notes, scripts;
reported in a culturaly elaborated,
and seemingly scientifically sound, milestone.
This yields context that entails e.g. the scope of experience
and fluency,
and the cost and benefit ratio given by the power
and use dificulty of the editor.

A less obvious merit of this undertake,
at least for me,
is that it might be of use to more advanced
HCI research, specially for text editors.
This seems to be the first scientific article on Vim,
and it might trigger other studies e.g. with user analytics,
or in comparing reports such as this.
Needless to say, The use of text editors is obviously very widespread.

There are even less trivial potentially relevant merits.
For example, the anthropological physics concept already received
some attention from the scientific community~\cite{anPhy,anPhy2,thesis}.
While still on the initial versions of this article,
I received a feedback from an previsouly unknown very experienced
researcher and software developer, which mentioned having
more then 20 years using, and made appointments accordingly.
He gave emphasis on the ``anthropological computer science''
concept here introduced, with remarks on the suitability of the
approach, that the term should have been coined earlier,
and that he intended to make an ``anthropological computer science account of C++''.

\subsection{Syntax highlighting experiments}\label{visual3}
By being configurable and scriptable, one is able not
only to load default or installed color schemes,
but to access and change individual colors arbitrarily.
This control over the colors are very timidly harnessed,
given 1) the effect color is believed to have,
and has been reported to have~\cite{}, on our performance;
2) the obliterated mechanisms by which one changes colorschemes
and individual colors in Vim; and 3) the large spans of time
a researcher or a programmer uses a text editor potentially every day.

Counterintuitively, the color scheme explorations are scarce,
and the available options most often resort to very standard settings.
For example, there is no colorscheme with a red background
among the builting Vim colorschemes (e.g. zellner, blue, etc)
nor among the usable colorschemes available online,
at least among the ones I found.
Also, there seems not to be dedicated scientific documents
with appropriate best practices or guidelines for syntax highlighting.
We have been gathering theoretical notes from visual perception, design,
and data visualization literatures, which we useful for
deriving at least an initial account of color schemes
for syntax highlighting.
These considerations should be systematized and sent
to a peer-reviewed journal soon.

conclusions: intend to write an article about all the SH issues.

\subsubsection{The Realcolors Vim plugin}
One outcome of such considerations about color schemes
and syntax highlighting is a plugin dedicated
to loading, tweaking and experimenting with color schemes in Vim.
With this plugin, it is easy to 1) load special and experimental color schemes, such as the ones prioritizing passive pink, or blood red, or a bright blue, which hold empirical or historical evidence of having an effect e.g. on mood or performance; 2) inspect the syntax groups of the token under the cursor; 3) access and change the foreground or background color of the prevailing syntax group of the token under the cursor, change bold and italic.
More information about the realcolors plugin might be found at
Vim.org~\footnote{\url{http://www.vim.org/scripts/script.php?script_id=5650}} and the public online Git repository~\footnote{https://github.com/ttm/vim}.

\subsection{The tokipona Vim plugin and Python package}
The first use of the syntax highlighting capabilities of Vim
though scripting was for studying and appreciating
the Toki Pona minimalist conlang~\cite{tpBook}.
Such developments lead to the development
of the tokipona Python package~\footnote{}
and the tokipona Vim plugin~\footnote{}.
The Python package
holds a number of routines for dealing with Toki Pona,
which includes the synthesis of Vim syntax files from
statistical analyses and the official Toki Pona dictionary.
The Vim plugin enables syntax highlighting for Toki Pona texts
and further accesses routines in the Python package.
Figure~\ref{fig:tp} exhibits an example of
syntax highlighting performed in Toki Pona text
thought the tokipona Python package and Vim plugin,
and contextualizes it with other sections of this document.

\subsection{Considerations about the keys available for mappings}
As expressed in Sections~\ref{map} and~\ref{exmapapendix},

Bram uses \_ and says that we don't use them.
I find it more comfortable then \^.
Also, \_ is more powerful: it does the same as \^,
but might receive a number N before, and jump to
the first char N-1 lines below.

\subsection{Experiments with VimL and the PRV Vim plugin}
PRV stands for Python, RDF, and Vim.
It is an attempt to account for my workflow through
writing text files in Vim.
It mingles with vimrc and usage manners described in next subsection.
By using Python to extend Vim's scripting capabilities, and RDF
to represent data and state as most convenient,
the thesis is that Vim will suffice as my platform for research and
development.
It is a somewhat usual problem, to which I visited and used a number
of solutions,
with care to incorporate suitable routines and design strategies.
Vimwiki\footnote{https://vimwiki.github.io/} is being most useful,
and might render \LaTeX and PDF files with pandoc.
This is of crucial importance, for tying notes and final documents,
such as articles fit for peer-review, is a recognized benefit and
often-pursued goal in worldwide Academic and general scientific
development.

After preliminary implementations for taking arbitrary types of notes,
and storing them as standalone binaries, RDF, custom plain text, and
database files (e.g. mongoDB),
current implementation is taking shape.
To render notes to share in the cloud, the ideal format is RDF.
To keep local notes of diverse types and formats,
it seems sufficient to have 1) one command for one string notes,
e.g. \tttt{:N todo make a banana cake colorscheme for Realcolors
plugin};
2) an efficient file-based solution, such as achieved though Vimwiki.

Such a tree of notes and related documents should have a facilitated
backup, being all in one directory tree. (vimwiki)
To avoid sharing unwanted material, one should add
\tttt{:se cm=blowfish2} and an autocommand to set a key
whenever the file is to be encripted.
Or, a less defensive solution, one just types \tttt{:se
key=aPhraseyouRemember} whenever there is sensitive content.
As Vimwiki maintains a source tree for all .wiki (and other) files,
it suffices to add a \tttt{ftplugin/<filetype>.vim} directive
to set the key. E.g. \tttt{set key=aPhraseyouRemmember} inside
\tttt{~/.vim/ftplugin/wiki.vim} should enable encryption for all
\tttt{*.wiki} files.

Another feature of PRV is a bot to enable a number of interaction
dynamics.
For example, the bot might learn the sentence structures in the
notes and synthesize sentences in general or given some reference
texts.
A user then might request sentences for some words or files which
have current research focus, or to help the researcher interconnect
the concepts which have received notes or are at least in the
vocabulary.
Current PRV bot include just returning Fortunes as made available by
Bram Moolinaar~\footnote{},
and rudimentary chat with user.
It is a variant of this very simple bot~\footnote{}, made available
online by xx in 2006.

A comparison of VimL against other languages
will need to be addressed PRV to work it 
This entails that system-specific measurements of 
e.g. text-mining routines in both VimL and Python
will have to be made available.
It seems a honest inquiry: to find out which, if any,
of the natural language routines of NLTK (Python)
might perform better using VimL.
Furthermore,
VimL seems quite powerful for object oriented
programming with funcrefs, partials, lambda functions and closures.
What are the limits and overall drawbacks of maintaining
methods and variables in dictionary key-values pairs?
The usual inheritance techniques are very trivial with dictionaries.
In this sense, dict.key access to attributes
seems to already be a convenience big enough to
supervene e.g. Python object oriented facilities.
What are the limitations inconveniences 
of VimL for object oriented programming?
Might one be able to compare VimL and Python to
C and C++?

In Python, \_xx and \_\_xx attributes are like
any other to the interpreter.
By convention, they are protected and private attributes of a class.
I advocate the same usage for object-like Vim dictionaries.

TODO: I will ask on the list if there is anything like that in VimL.

let job = job\_start(["evince", " ", "~/repos/vim/ttmvim/article/article.pdf"]) 
jobs, channels, timers, etc etc

TODO:
find smallest available tokens in each namespace:
latex, javascript, vim, python.
Propose a technique based on making often made operations
for them with these tokens.

make a system to load files and directories with
globpath(g:mydirs, tokipona)

make a decent menu to choose the colorschemes, categorized by:
realcolors
builtin
third party

show normal background and foreground colors and optionally other main
colors, open the syntax file with a mapping, etc.
Implement this with the colorpicker shared by v1z

\subsection{Usage manners}
  * also put massetes such as :@*<CR> in a visual selection

\subsection{Potential enhancements to Vim UX}

\section{Conclusions and further work}\label{conc}
This document seems reasonable as an overall reference of the Vim editor,
at least for my usage and level of proficiency.
Given the folkloric milestone of using Vim for 10 years,
this article might serve as a benchmark for one to relate
it's current use and understandings.
As a pedagogical material, it seems to be unique in the emphasis
on namespaces, understood as commands, variables, state-related lists, etc,
especially in Section~\ref{namespaces},
and the reference to the standard Vim documentation
to achieve the DRY KISS style described in Section~\ref{furRem}.

\noindent {\bf Potential enhancements to this document} include:
\begin{itemize}
  \item The discussion of facilities such as reading emails and connecting over ssh.
    There is a working hack in~\cite{vimrc} for browsing over the WWW,
    but such aspect of Vim usage might receive more attention
    given that it is comfortable to navigate and edit in Vim
    and the resulting integrated environment.
  \item Updating of the information I can find about the issues discussed here,
    such as about status lines in Section~\ref{visual1}.
  \item Include a discussion about Neovim.
    I have never used it, but it seems to be reaching a considerable user base
    and it might be feasible to give an account of Vim and Neovim
    after some tests and researching the official and unofficial documentation.
  \item Better cite documentation, plugins and Vim authors.
    I preferred to keep the references inline through \tttt{:h} commands
    and URLs, more in accordance with the style of Vim documentation,
    but bibliographical items constitute a valuable asset for academic literature,
    and some authors might find their work more respected if more
    thoroughly cited.
  \item An analysis of my usage, e.g. according to \url{http://www.drbunsen.org/vim-croquet/}.
    This undertake might benefit from data from many users,
    which favors the potential plugin for usage analytics described in the next bullet list.
\end{itemize}

\noindent {\bf Potential next steps} in using Vim include:
\begin{itemize}
  \item Measure the performance of text mining routines implemented in VimL against those implemented in C or Python. Compare Python performance while running inside Vim (through the Python interface) or as the standard standalone interpreter.
  \item Enhance the HTTP browsing capabilities of~\cite{vimrc}.
  \item Better integrate Python and VimL, especially for data visualization
    and syntax highlighting management,
    in accordance with the visualization issues described in
    Sections~\ref{visual1} and~\ref{visual2}.
\end{itemize}

\noindent This is a selection of the {\bf issues that might entail
plugins and that are more prominent for me}:
    \begin{itemize}
      \item listing all the mappings available in each mode and the typing combinations which are available
        for new mappings.
        Maybe already group possibilities by criteria such as length
        of sequence and how central are the keys.
      \item Sessions, as described in Section~\ref{visual1} and preliminarily implemented in~\cite{vimrc}.
      \item AA messages (shouts): to keep track, document and share of working sessions
        as in~\cite{aa1,aa2}, with capabilities to manage AA sessions,
        send visual or sonic cues for temporal marks, use Vim state to build AA shouts,
        relate AA sessions to other media, such as software repositories,
        screencasts and images, interact with IRC channels and other social platforms.
      \item Slick Vim: a collection of the settings and mappings I use.
        Enhancements such as using \ttt{C-} commands also to browse tabs,
        shortcuts to join a window into a tab,
        and dummy minimal plugins as simplest possible, then file type
        and then syntax highlighting,
        Some more elaborate tweaks should also be present, such
        as breaking lines in sensible places for natural language texts
        while respecting e.g. \tttt{:se textwidth}.
      \item Dealing with .swo and .swp temporary files.
        In summary, if the restored .swo file has the same content
        and the file being opened,
        the restoration phase can be omitted.
        If the contents differ, Vim should open a tab with each file
        in a vertical split and run \tttt{:windo :diffthis}.
      \item Rendering images and equations.
        These are useful for using Vim in presentations
        or achieving a textual representation when it is mandatory,
        such as to comply with the limitations of a platform (e.g. Vim editor). 
        but also hold stylistic merits as ASCII art is often
        very appreciated.
        One can both obtain an ASCII representation of a binary image (e.g. JPG, PNG),
        and can directly render ASCII charts from data using cues such are shape, position
        and color.
      \item Redirecting the commands that usually show the results in
        a 'more' interface (e.g. status listings such as in
        Section~\ref{state}),
        which cannot be searched nor copied nor persists if one
        returns to editing a file.
        Ideally, it should parsed and linked quickfix or location
        window, and the syntax highlighting maintained.
        The basic idea is to use \tttt{:redir} command to redirect the output
        of such commands with \tttt{:se nomore}.
        Reasonable functions (and convenient commands) for having the
        output of such commands in a standard Vim window are
        in~\cite{vimrc}.
      \item Slide presentations. I've been using some automation for browsing
        slides and opening figures. Some Vim users asked for the settings and
        commands. They comprise a very elementary use of registers which are executed over
        consistent textual patterns.
      \item For bringing back all the splits after an \tttt{:only}.
        Also for bringing back the tabs after a \tttt{:tabo}.
      \item Run Python excerpts, from a file being edited, in an IPython shell. 
        The buffer number of the terminal window should be stored,
        and than any selected lines should be run on that instance.
        Mappings should make available all movements to fetch the
        excerpts, execute registers, remain on script or in the IPython shell.
        Also, current Terminal-Job mode can be improved easily with
        mappings, such as \ttt{C-O} for one normal mode command.
      \item For keeping track of the usage and making analysis
        for optimizing the usage, as described in~\url{http://www.drbunsen.org/vim-croquet/}.
        Usage analytics.
      \item To facilitate the tweaking of syntax highlighting.
        This should include ways to easily access and change the
        syntax highlighting scheme and dump it to a file (dumping
        current highlighting scheme to a working VimL file
        is currently not supported by Vim!).
        Should also include changing the color scheme and highlighting scheme incrementally
        and selectively using the features described in Section~\ref{visual2}
        and in~\cite{tokipona}.
      \item Color schemes for true color.
        The standard colors (e.g. blue, elflord) loose some of the distinctions,
        e.g. SpellBad tokens are not highlighted on these color schemes
        if you have a functioning \tttt{se termguicolors} (standard GUI mode
        and available in terminals since recent versions of Vim 8,
        see Section~\ref{visual2}).
        In~\cite{vimrc} are some lines that make such color schemes over
        \tttt{:colorscheme blue}.
        This design of color schemes over the standard color schemes might be
        a very simple and effective way to make new color schemes.
        Such color schemes should also make use of the discussions
        in~\cite{tokipona}.
    \end{itemize}

\subsection*{Acknowledgments}
FAPESP (project 2017/05838-3); Vim developers and documentation maintainers;
Vim user community.

\appendix
\section{Example of usage session}
I usually begin by opening a file or directory
with \tttt{\$ vim <filename>}.
The color scheme is alternated between
blue and GruvBox with
\tttt{:colo gruvbox} and \tttt{:colo blue}.
I open a vertical split and then move
the window to a new tab using
\tttt{:vs} and \tttt{<C-W>T}.
I then search for tokens related to
the enhancements I want to make or
the knowledge I want to acquire.
I go back to the previous tab with \ttt{A-a}
and make a global replace with
\tttt{:\%s/<this>/<that>/g}.
On adding dots to sentences,
I record in the \tttt{"q} register
the sequence \tttt{jA.<C-[>},
and use it as a macro 10 times by typing
\tttt{10@q}.
I move to the other tab with \ttt{A-d}
and open a terminal window with \tttt{:term}
for compiling latex files.
I start another terminal with \tttt{<C-W> :term}
for opening the resulting PDFs with evince.
If any new idea comes to mind and I have time,
\tttt{\textbackslash\textbackslash s} opens 
my vimrc~\cite{vimrc} for editing and \tttt{\textbackslash s}
sources it.
If there is e.g. code or notes in other projects I am working
on, I most often reach them through \tttt{<A-(arrows)>} because they
are in separate Vim instances
in Byobu/Tmux sessions and windows.
Because I stay for hours editing (e.g. \LaTeX\ and Python
scripts), I change the background to red, and eventually to green or to yellow using the Realcolors plugin described in Section~\ref{realcolors}.
Because the blue color scheme does not highlight the spelling errors
if using true color, I would run
\tttt{:hi SpellBad guifg=red guibg=lightblue}
to see the words found wrong by the spell checker,
if this tweak was not already in my vimrc~\cite{vimrc}.

\section{My vimrc file and usage}
In using Vim with my vimrc file~\cite{vimrc},
I mostly toggle the status line with \texttt{\textbackslash\textbackslash B}
and the tab line with \texttt{\textbackslash\textbackslash T}.
Save and close windows with \texttt{\textbackslash w} and \texttt{\textbackslash q}
(or \ttt{A-w} and \ttt{A-q}).
The mappings for transitioning through splits and tabs
are also used constantly.
Although the file is commented, one should
look for the help pages on the options that (s)he does
not understand, as a thorough explanation
of the file is tedious and out of the scope
of this document.

\section{Example notes on mappings}\label{notes}
\texttt{:h index} shows all the default mappings
while \tttt{:map} shows the user-defined mappings.
By considering such information, one can make
useful observations exemplified in this Appendix.

\subsection{Normal mode}
Every letter and character in the keyboard is used.
In Normal mode: \ttt{TAB} is the same as \ttt{C-I},
\ttt{BS} and \ttt{C-H} are the same as h.
\ttt{C-J} and \ttt{C-N} are the same as j.
\ttt{C-P} is the same as k.
Space is the same as l.
\ttt{C-[} and \ttt{Esc} are not used,
\tttt{<C-\textbackslash> a-z} are reserved for extensions,
and \ttt{C-\_} is not used.
\tttt{+} is the same as \ttt{CR} and they are both not very useful.
Del is the same as x.

Many \tttt{][} combinations are not used, e.g.
with \tttt{abhjklfg}.
The \tttt{\_} command might be used as a more powerful \tttt{\^},
leaving it free for mappings.

Directions, home, end, page up and down, insert, all have mappings
in more centraly located keys.
The digits that beggin unbounded mappings (require another character to
trigger an action without having to wait for \tttt{:se timeoutlen}): g,z,[,].

The \ttt{C-(HJKL)} commands are redundant,
with the exception of \ttt{C-L} which redraws the screen,
so it is a reasonable choice to use them to move focus
of the editor to splits in the \tttt{hjkl} directions.
Althought it is a usable solution, \ttt{A-(hjkl)} might
be more consistent with other uder-defined commands for navigation,
and presearves the useful and standard \ttt{C-L}.

Many key combinations are available for new mappings through the \tttt{g}
and \tttt{z}, and \tttt{v} commands.
They have typical uses, e.g. \tttt{z} for folds,
spell checking and some movements (mainly when wrap is set).

\subsection{Insert mode}
All standard character keys are used for entering text.
\tttt{<C-G>(j,k)} can be achieved by \tttt{<C-O>(j,k)} which is more powerful
in moving through multiple lines.
\ttt{C-$[$} is the same as \ttt{ESC}.
\ttt{C-J} and \ttt{C-M} are \ttt{CR}.
\tttt{<C-\textbackslash> a-z} are reserved for extensions,
other combinations with \ttt{C-\textbackslash} are not used.

\section{My \tttt{:version}}
I should keep the output of \tttt{:version} executed on the system
in which I wrote this document in this link:
\url{https://github.com/ttm/vim/raw/master/version.txt}.
It was compiled with this Makefile~\cite{makefile}
in a 16.04 LTS Ubuntu Linux,
and is tagged as 8.0, Included Patches 1-1173.

\section{Key notation and meaning for Vim}\label{sec:not}
The tokens \ttt{C-X}, \ttt{S-X} and \ttt{M-X}
mean \tttt{Ctrl+x}, \tttt{Shift+x} and \tttt{Alt+X}
(or \tttt{Meta+X}).
\ttt{A-X} is the same as \ttt{M-X} and refer to
Alt or Meta keys.
\ttt{C-X}, \ttt{C-S-X}, \ttt{C-S-x} and \ttt{C-x} are the same,
i.e. Vim does not distinguish between lower and uppercase 
letters in \ttt{C-} commands.
The \ttt{M-x} and \ttt{M-X} (or \ttt{M-S-x} or \ttt{M-S-X})
are different commands,
and they are not used by Vim's builtin keys.
Such \ttt{M-} key combinations potentially conflict
with shortcuts of other programs (e.g. \ttt{M-F10} for a terminal menu),
but are otherwise a large and safe set of combinations
for one to use.

In summary,
one might choose mappings that overwrite default key combinations,
use combinations not used by Vim (e.g. some of the \tttt{g} and \tttt{z}
normal commands, \ttt{C-} and \ttt{S-} commands, or any \ttt{M-} and
\ttt{M-S-} command), or
create new mappings using \tttt{<leader>} and \tttt{<localleader>}.
More information about key notation is in \tttt{:h key-notation}.
Default mappings are in \tttt{:h index} and user-defined (and plugin-defined) mappings
in \tttt{:map}.

\begin{thebibliography}{99}
\fontsize{11}{0}\selectfont
\bibitem{anPh}
	Anthropological physics and social psychology in the critical research of networks. Complex Networks Digital Campus (CS-DC'15).
	Available at \url{https://youtu.be/oeOKYc3-nbM}
\bibitem{anPh2}
	Fabbri, R. What are you and I? [anthropological physics fundamentals], 2015. Available at \url{https://www.academia.edu/10356773/What_are_you_and_I_anthropological_physics_fundamentals_}
\bibitem{noOfficial}
  Vim users email list. "Vim versions. (No explicit or official stable, alpha or beta versions of Vim)" Available at \url{https://groups.google.com/forum/#!msg/vim_use/ULCa_2Cxnl0/RmvGij8YCgAJ} 
\bibitem{wikiVim}
  Vim (text editor). (2017, December 4). In Wikipedia, The Free
    Encyclopedia. Retrieved 07:08, December 16, 2017, from
    \url{https://en.wikipedia.org/w/index.php?title=Vim_(text_editor)&oldid=813716105}
\bibitem{ouroWiki}
  Ouroboros. (2017, November 2). In Wikipedia, The Free Encyclopedia. Retrieved 22:19, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ouroboros&oldid=808392809}
\bibitem{vimrc}
	Fabbri, R. (2017). A reasonable vimrc file. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/vimrc} 
\bibitem{ex}
  Ex (text editor). (2017, March 22). In Wikipedia, The Free Encyclopedia. Retrieved 22:22, November 9, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Ex_(text_editor)&oldid=771621020}
\bibitem{tokipona}
	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
\bibitem{tpLang}
	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
\bibitem{aa1}
	Fabbri, R., Fabbri, R., Vieira, V., Penalva, D., Shiga, D., Mendonça, M., Negrao, A., Zambianchi, L., \& Thumé, G. (2013). AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology. RESI. From \url{https://arxiv.org/abs/1604.08255}
\bibitem{aa2}
	Fabbri, R. (2017).
The Algorithmic-Autoregulation (AA) Methodology and Software:
a collective focus on self-transparency. ENMC2017. From \url{https://github.com/ttm/ensaaio/raw/master/emc/article.pdf} 
\bibitem{makefile}
	Fabbri, R. (2017). The Makefile with which I compiled Vim for this article. Available at \url{https://raw.githubusercontent.com/ttm/vim/master/Makefile} 
\end{thebibliography}
\end{document}

### 2018-06-30 article/article.tex
passada autoria da cristima para agradecimentos

@@
 \author{Renato Fabbri, \texttt{renato.fabbri@gmail.com}\\
-Maria Cristina Ferreira de Oliveira, \texttt{cristina@icmc.usp.br}\vspace{0.2cm}\\
 University of São Paulo,\\
@@
 \subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3);
+FAPESP (project 2017/05838-3, coordinated by Prof. Dr. Maria Cristina Ferreira de Oliveira);
 Toki Pona, Python and Vim authors, developers and literature maintainers;

### 2018-07-01 article/acm/sample-acmlarge.tex
arrumacao inicial do repositorio com formatacao p acm do artigo

\documentclass[acmlarge]{acmart}
\usepackage[utf8]{inputenc}

\usepackage{booktabs} % For formal tables
\usepackage{tipa}
\newcommand{\tttt}[1]{\texttt{#1}}
\usepackage{subcaption}

\usepackage[ruled]{algorithm2e} % For algorithms
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmJournal{TALLIP}
% \acmVolume{X}
% \acmNumber{X}
% \acmArticle{XX}
% \acmYear{2018}
% \acmMonth{X}
% \acmArticleSeq{XX}
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2010}
\acmMonth{3}
\acmArticleSeq{11}

%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{0000001.0000001}

% Paper history
\received{July 2018}
% \received{March 2009}
\received[accepted]{XXX XXXX}

% Document starts
\begin{document}
% Title portion
\title{Basic concepts and tools for the Toki Pona minimalist and
constructed language}
\titlenote{Description of the language and main issues;
Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}

\author{Renato Fabbri}
\authornote{This is the corresponding author}
\orcid{0000-0002-9699-629X}
\affiliation{%
  \institution{Institute of Mathematical and Computer Sciences, University of São Paulo}
  \streetaddress{Avenida Trabalhador São-carlense, 400 - Centro}
  \city{São Carlos}
  \state{SP}
  \postcode{668}
  \country{BR}}
\email{renato.fabbri@gmail.com}

\begin{abstract}
A minimalist constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimalist both in the vocabulary
(with only 14 letters and 124 words)
  and in the (about) ten syntax rules.
The language is useful for being a used and somewhat established
minimalist conlang with at least hundreds of fluent speakers.
In this article, we describe current concepts and resources
for TP,
and make available Python scripted routines for
the analysis of the language,
the synthesis of texts, the specification of syntax highlighting
schemes, and the achievement of a preliminary TP
Wordnet.
We focus on the analysis of the basic vocabulary,
as corpus analyses were found.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor.
The tentative TP Wordnet is made available in three forms that
reflect the choices of the synsets related to each word.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing and syntax highlighting the
TP language.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010405.10010469</concept_id>
  <concept_desc>Applied computing~Arts and humanities</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010405.10010497</concept_id>
  <concept_desc>Applied computing~Document management and text processing</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10003456.10010927.10003619</concept_id>
  <concept_desc>Social and professional topics~Cultural characteristics</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc{Applied computing~Arts and humanities}
\ccsdesc[500]{Applied computing~Document management and text processing}
\ccsdesc[300]{Social and professional topics~Cultural characteristics}

%
% End generated code
%

\keywords{Constructed languages, natural language processing, syntax highlighting, wordnet, Toki Pona}

% DO NOT use this command unless you want to change
% the default behavior
% \authorsaddresses{Authors' addresses: G.~Zhou, Computer Science
%   Department, College of William and Mary, 104 Jameson Rd,
%   Williamsburg, PA 23185, US, \path{gzhou@wm.edu}; V.~B\'eranger,
%   Inria Paris-Rocquencourt, Rocquencourt, France; A.~Patel, Rajiv
%   Gandhi University, Rono-Hills, Doimukh, Arunachal Pradesh, India;
%   H.~Chan, Tsinghua University, 30 Shuangqing Rd, Haidian Qu, Beijing
%   Shi, China; T.~Yan, Eaton Innovation Center, Prague, Czech Republic;
%   T.~He, C.~Huang, J.~A.~Stankovic University of Virginia, School of
%   Engineering Charlottesville, VA 22903, USA; T. F. Abdelzaher,
%   (Current address) NASA Ames Research Center, Moffett Field,
%   California 94035.}

\maketitle

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{G. Zhou et al.}

\input{samplebody-journals}

\end{document}

### 2018-07-01 article/acm/samplebody-journals.tex
arrumacao inicial do repositorio com formatacao p acm do artigo

--- 2018-02-11 article/article.tex
@@
-\documentclass{article}
-\usepackage{graphicx}
-\usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{hyperref}
-\hypersetup{
-	%pagebackref=true,
-	pdfcreator={LaTeX with abnTeX2},
-	pdfkeywords={abnt}{latex}{abntex}{USPSC}{trabalho acadêmico}, 
-	colorlinks=true,       		% false: boxed links; true: colored links
-	linkcolor=blue,          	% color of internal links
-	citecolor=blue,        		% color of links to bibliography
-	filecolor=magenta,      		% color of file links
-	urlcolor=blue,
-	allbordercolors=black,
-	bookmarksdepth=4
-}
-\usepackage[utf8]{inputenc}
-\usepackage{tipa}
-\newcommand{\ttt}[1] {
-	\texttt{<#1>}}
-\newcommand{\tttt}[1]{\texttt{#1}}
-% \usepackage{fontspec}
-% \usepackage[mathletters]{ucs}
-% \newcommand{\ttt}[1] {
-% 	\texttt{<#1>}}
-% \newcommand{\tttt}[1]{\texttt{#1}}
-\usepackage{subcaption}
-
-\begin{document}
-\title{Basic concepts and tools for the Toki Pona minimalist and
-constructed language:
-Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
-% Anthropological computation
-\author{Renato Fabbri, \texttt{renato.fabbri@gmail.com}\\
-Maria Cristina Ferreira de Oliveira, \texttt{cristina@icmc.usp.br}\vspace{0.2cm}\\
-University of São Paulo,\\
-Institute of Mathematical and Computer Sciences\\
-São Carlos, SP, Brazil
-}
-\maketitle
-\begin{abstract}
-A minimalist constructed language (conlang)
-is useful for experiments and comfortable for making tools.
-The Toki Pona (TP) conlang is minimalist both in the vocabulary
-(with only 14 letters and 124 words)
-and in the $\approx10$ syntax rules.
-The language is useful for being a used and somewhat established
-minimalist conlang with at least hundreds of fluent speakers.
-In this article, we describe current concepts and resources
-for TP,
-and make available Python scripted routines for
-the analysis of the language,
-the synthesis of texts, the specification of syntax highlighting
-schemes, and the achievement of a preliminary TP
-Wordnet~\cite{wordnet}.
-We focus on the analysis of the basic vocabulary,
-as corpus analyses were found~\cite{corpus}.
-The synthesis is based on sentence templates,
-relates to context by keeping track of used words,
-and renders larger texts by using a fixed number of phonemes (e.g. for poems)
-and number of sentences, words and letters (e.g. for paragraphs).
-Syntax highlighting 
-reflects morphosyntactic classes given in the official dictionary
-and different solutions are described and implemented
-in the well-established Vim text editor~\cite{vim}.
-The tentative TP Wordnet is made available in three forms that
-reflect the choices of the synsets related to each word.
-In summary, this text holds potentially novel conceptualizations about,
-and tools and results in analyzing, synthesizing and syntax highlighting the
-TP language.
-  % subj (li), obj (e), prepositions, qualification (word concatenation), 
-  % sent (la), particles, mi-sina, pi, punctuation is freestyle, pre-verb,
-\end{abstract}
-{\bf keywords:} Constructed languages, Natural Language Processing, Syntax highlighting, Wordnet, Toki Pona
-
 \section{Introduction}\label{intro}
@@
 
-\subsection*{Acknowledgments}
-FAPESP (project 2017/05838-3);
-Toki Pona, Python and Vim authors, developers and literature maintainers;
-Toki Pona user and speaker communities. 
-I thank Mario Alzate for introducing me to Toki Pona,
-and Silverio Guazzelli Donatti for the stimulus that resulted
-from the usage of Toki Pona for chatting.
-
 \appendix
@@
 
-\begin{thebibliography}{99}
-\fontsize{11}{0}\selectfont
-\bibitem{wordnet}
-  Fellbaum, C. (1998). WordNet. John Wiley \& Sons, Inc..
-\bibitem{corpus}
-    Henry, J. lipu pi toki pona pi jan Jakopo. Available at         
-    \url{http://jimhenry.conlang.org/conlang/tokipona/tokipona.htm} 
-\bibitem{vim}
-    Fabbri, R. An anthropological account of the Vim editor:
-features and tweaks after 10 years of usage. Available at         
-    \url{https://github.com/ttm/vim/raw/master/article/article.pdf} 
-\bibitem{sapWho}
-  O'Neill, S. P. (2015). Sapir–Whorf Hypothesis. The International
-    Encyclopedia of Language and Social Interaction.
-\bibitem{tpLang}
-	Lang, S. (2014). Toki Pona: the language of good. Tawhid Publishing.
-    ISBN-10: 0978292308, ISBN-13: 978-0978292300.
-\bibitem{kama}
-  Knight, B. (2017) o kama sona e toki pona! Available at:
-    \url{http://tokipona.net/tp/janpije/okamasona.php}
-\bibitem{tp3}
-  Blahuš, M. (2011). Toki pona–eine minimalistische Plansprache.
-    Spracherfindung und ihre Ziele. Beiträge der, 20, 51-56.
-\bibitem{tp4}
-  Schneider, S. (2017). nasin toki pi toki pona. Available at
-  \url{https://github.com/stefichjo/toki-pona}
-\bibitem{interview}
-  Lang, S. and Broholm, K. AFP 20 – Sonja Lang: Toki Pona, Conlanging
-    and the meaning of life (Sonja Lang interview). Available at
-    \url{https://actualfluency.com/afp-20-sonja-lang-toki-pona-conlanging-meaning-life/}
-\bibitem{memrise}
-  Toki Pona Community. (2017). Various Memrise courses on Toki Pona.
-    \url{https://www.memrise.com/courses/english/?q=toki+pona}
-\bibitem{gdoc}
-  Toki Pona Community (2017). toki pona central hub preparation.
-    Available at
-    \url{https://docs.google.com/document/d/1Dzs-imNeZ8TMgdHUiiungJ4Yf97CJk9ylhQPXjWLsJU}
-\bibitem{wikiToki}
-  Toki Pona. (2017, December 11). In Wikipedia, The Free Encyclopedia.
-    Retrieved 22:48, December 14, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Toki_Pona&oldid=814813420}
-\bibitem{conlanWikip}
-  Constructed language.
-(2017, November 1).
-In Wikipedia, The Free Encyclopedia.
-Retrieved 10:22, November 18, 2017, from \url{https://en.wikipedia.org/w/index.php?title=Constructed_language&oldid=808264974}
-\bibitem{artlanWikip}
-  Artificial language. (2017, November 26). In Wikipedia, The Free
-  Encyclopedia. Retrieved 23:00, December 14, 2017, from
-  \url{https://en.wikipedia.org/w/index.php?title=Artificial_language&oldid=812127746}
-\bibitem{akesiWawa}
-  Fabbri, R. (2017, November 6). o awen, akesi wawa.
-  In the group toki pona taso. Available at
-  \url{https://www.facebook.com/groups/tokiponataso/permalink/1895942817391318/}
-\bibitem{janKipo}
-  Clifford, J. et al. (2017, December 15). Considerations about Toki Pona
-  phrases.
-  In the Facebook group 'toki pona'. Available at
-  \url{https://www.facebook.com/groups/sitelen/permalink/1597077093680003/}
-\bibitem{tokipona}
-	Fabbri, R. (2017). A Toki Pona Python Package and Vim Syntax Highlighting. Available at \url{https://github.com/ttm/tokipona} 
-\bibitem{tpNetSH}
-  Knight, B. (2017). Online gadget for syntax highlighting Toki Pona
-  texts. . Available at
-  \url{http://tokipona.net/tp/DisplayText.aspx}
-\bibitem{stack}
-  Various authors, 2011-2016. Syntax-highlighting color scheme studies.
-    Software Engineering in Stack Exchange. Available at
-    \url{https://softwareengineering.stackexchange.com/questions/89936/syntax-highlighting-color-scheme-studies}
-\bibitem{solarized}
-  Schoonover, E. Solarized color scheme. Available at
-  \url{http://ethanschoonover.com/solarized}
-\bibitem{dv1}
-	Munzner, T. (2014). Visualization analysis and design. CRC press.
-\bibitem{dv2}
-  Telea, A. C. (2014). Data visualization: principles and practice.
-  CRC Press.
-\bibitem{dv3}
-  Ward, M. O., Grinstein, G., \& Keim, D. (2010). Interactive data
-  visualization: foundations, techniques, and applications. CRC Press.
-\bibitem{dv4}
-  Ware, C. (2012). Information visualization: perception for design.
-  Elsevier.
-\bibitem{blue}
-  Mehta, R., \& Zhu, R. J. (2009). Blue or red? Exploring the effect of
-  color on cognitive task performances. Science, 323(5918), 1226-1229.
-\bibitem{blue2}
-  Viola, A. U., James, L. M., Schlangen, L. J., \& Dijk, D. J. (2008).
-  Blue-enriched white light in the workplace improves self-reported
-  alertness, performance and sleep quality. Scandinavian journal of
-  work, environment \& health, 297-306.
-\bibitem{fonts}
-  Spolsky, A. J. (2008). User interface design for programmers.
-  Apress.
-\bibitem{mass}
-	Fabbri, R., Vieira, V.
-, Pessotti, A.
-C.
-S., Corrêa, D.
-C., Oliveira Jr, O.
-N.
-O.
-Musical elements in the discrete-time representation of sound (2017).
-arXiv preprint arXiv:1412.6853.
-Available at \url{https://github.com/ttm/mass/raw/master/doc/article.pdf}
-\bibitem{wikiPoetry}
-Poetry. (2017, November 19). In Wikipedia, The Free Encyclopedia.
-    Retrieved 19:46, December 3, 2017, from
-    \url{https://en.wikipedia.org/w/index.php?title=Poetry&oldid=811036139}
-\bibitem{tokisona}
-  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
-  Available at \url{http://tokisona.github.io/}
-\bibitem{Wikipesija}
-  Toki Pona community (2017). Wikipesija. Available at:
-    \url{http://tokipona.wikia.com}
-\end{thebibliography}
-\end{document}
+\begin{acks}
+
+The author thanks
+FAPESP (project 2017/05838-3, coordinated by Prof. Dr. Maria Cristina Ferreira de Oliveira);
+Toki Pona, Python and Vim authors, developers and literature maintainers;
+Toki Pona user and speaker communities; 
+Mario Alzate for introducing me to Toki Pona,
+and Silverio Guazzelli Donatti for the stimulus that resulted
+from the usage of Toki Pona for chatting.
+
+\end{acks}
+
+% Bibliography
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{sample-bibliography}

### 2018-07-01 article/acm/sample-acmlarge.tex
terminado de arrumar a bibliografia no novo formato
arrumado artigo ateh historico

@@
 \usepackage{booktabs} % For formal tables
-\usepackage{tipa}
 \newcommand{\tttt}[1]{\texttt{#1}}
 \usepackage{subcaption}
+\usepackage{tipa}
 
@@
 % \acmArticleSeq{XX}
-\acmVolume{9}
-\acmNumber{4}
-\acmArticle{39}
-\acmYear{2010}
-\acmMonth{3}
-\acmArticleSeq{11}
+\acmVolume{0}
+\acmNumber{0}
+\acmArticle{00}
+\acmYear{2018}
+\acmMonth{7}
+\acmArticleSeq{00}
 
@@
 (with only 14 letters and 124 words)
-  and in the (about) ten syntax rules.
+and in the (about) 10 syntax rules.
 The language is useful for being a used and somewhat established
 minimalist conlang with at least hundreds of fluent speakers.
-In this article, we describe current concepts and resources
+This article exposes current concepts and resources
 for TP,
-and make available Python scripted routines for
+and makes available Python scripted routines for
 the analysis of the language,
@@
 Wordnet.
-We focus on the analysis of the basic vocabulary,
+Focus is on the analysis of the basic vocabulary,
 as corpus analyses were found.
@@
 in the well-established Vim text editor.
-The tentative TP Wordnet is made available in three forms that
-reflect the choices of the synsets related to each word.
+The tentative TP Wordnet is made available in three patterns
+  of relations between synsets and word lemmas.
 In summary, this text holds potentially novel conceptualizations about,

### 2018-07-01 article/acm/samplebody-journals.tex
arrumado artigo ateh historico

@@
 Therefore, the concepts are usually very general and different,
-and, without context, the words are rarely related through
+and, without context, the words are not frequently related through
 meronymy and hyponymy.
 Such a linguistic setting is desired because of the simplicity
-which entails easier e.g. learning and tool making.
+which entails facilitated e.g. learning and tool making.
 Another reason why the minimalist language design is compelling
@@
 the Sapir-Whorf hypothesis (linguistic relativity),
-i.e. that language influences or dictates thought
+i.e. that language dictates or at least influences thought
 and world experience~\cite{sapWho}.
@@
 the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
-In~\cite{interview}, Sonja Lang (the creator of Toki Pona),
+In~\cite{interview}, Sonja Lang (the creator of TP),
 describes that she has seen the language been used successfully
@@
 
-In this article, we present a conceptual overview of the language,
+This article provides a conceptual overview of the language,
 which is fit both to the newcomer and to the expert
 for being considerably different
-from what we found in the literature~\cite{tpLang,kama,memrise},
-with emphasis on simplicity and flexibility.
-We also present novel software routines for analysis, synthesis,
+from what was found in the literature~\cite{tpLang,kama,memrise},
+here with emphasis on simplicity and flexibility.
+Novel software routines for analysis, synthesis,
 syntax highlighting, and the achievement of a tentative TP
-Wordnet~\cite{wordnet}.
+Wordnet~\cite{wordnet} are also herein presented.
 
@@
 a historical note, and some words about natural and constructed languages.
-Section~\ref{basics} describes TP's phonology and syntax.
+Section~\ref{basics} describes the TP phonology and syntax.
 Section~\ref{hacks} presents the software routines we made available
@@
 Conclusions and further work are in Section~\ref{conc}.
-Appendix~\ref{mytoki} holds considerations about my usage of Toki Pona,
+Appendix~\ref{mytoki} holds considerations about the author's
+usage of Toki Pona,
 with thoughts about rule breaking and potentially new conlangs.
@@
 
-% definition, historic, status (groups, speakers, etc).
-
 \subsection{Resources on Toki Pona}
-One might organize current resources for the Toki Pona
+One might categorize current resources for the Toki Pona
 language in: references and learning material, corpus,
@@
 For example, /p t k s l/ might be pronounced
-[p t k s l] or [b d g z \textipa{\!R}].
+% [p t k s l] or [b d g z !R].
+[p t k s l] or [b d g z \textipa{\!R}].% ſ].
 Especially for poetry, one might consider
@@
     \centering
-        \includegraphics[width=1.2\linewidth]{figs/sentence_}\vspace{-0.15cm}
+        \includegraphics[width=1\linewidth]{figs/sentence_}\vspace{-0.15cm}
           \caption{Synthesized sentence.}
@@
     \centering
-        \includegraphics[width=1.2\linewidth]{figs/paragraph_}\vspace{-0.15cm}
+        \includegraphics[width=\linewidth]{figs/paragraph_}\vspace{-0.15cm}
           \caption{Synthesized paragraph.}
@@
   \end{subfigure}\\\vspace{0.3cm}
-  \begin{subfigure}{.7\textwidth}
+  \begin{subfigure}{.6\textwidth}
     \centering

### 2018-07-01 article/article.tex
arrumado artigo ateh historico

@@
 (with only 14 letters and 124 words)
-and in the $\approx10$ syntax rules.
+and in the about 10 syntax rules.
 The language is useful for being a used and somewhat established
 minimalist conlang with at least hundreds of fluent speakers.
-In this article, we describe current concepts and resources
+This document exposes current concepts and resources
 for TP,
-and make available Python scripted routines for
+and makes available Python scripted routines for
 the analysis of the language,
@@
 schemes, and the achievement of a preliminary TP
-Wordnet~\cite{wordnet}.
-We focus on the analysis of the basic vocabulary,
-as corpus analyses were found~\cite{corpus}.
+Wordnet.
+Focus is on the analysis of the basic vocabulary,
+as corpus analyses were found.
 The synthesis is based on sentence templates,
@@
 and different solutions are described and implemented
-in the well-established Vim text editor~\cite{vim}.
+in the well-established Vim text editor.
 The tentative TP Wordnet is made available in three forms that
-reflect the choices of the synsets related to each word.
+  of relations between synsets and word lemmas.
 In summary, this text holds potentially novel conceptualizations about,

### 2018-07-01 article/acm/samplebody-journals.tex
terminado historico

@@
 TP was developed as an internal and personal language
-by Sonja Lang~\cite{interview}.
-It was released as a draft in 2001 and in 2007 some documents
+by Sonja Lang in late 1990s~\cite{interview}.
+It was released as a draft in 2001, and in 2007 some documents
 reported it to have a few hundred speakers.
@@
 Nowadays, one finds a number of texts about TP and written
-in it
-(e.g. in social platforms such as Facebook groups,
-microblogging, Telegram and
-IRC), and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.
+in TP (e.g. in social platforms such as Facebook groups,
+microblogging, Telegram and IRC),
+and other diverse uses of TP e.g. for artificial intelligence and software tools~\cite{gdoc}.
 
@@
 a group, such as Esperanto, Toki Pona, and Lojban.
-An artificial language is a language yield by artificial agents,
+An `artificial language' is most usually a term used
+to designate a language yield by artificial agents,
 such as by AI routines, or by humans in controlled experiments,
@@
 
-Constructed languages are in some traditions called artificial
+Constructed languages (aka. conlangs) are in some traditions called artificial
 languages, but creators most often prefer to use the term
 `standardized', 'constructed'
-or 'planned' language with the argument that the conlang
-is rooted on natural languages,
-and 'artificial' is misleading.
-The preferred terms seem to be planned or constructed language
-or simply conlang.
+or 'planned' language with the argument that the conlangs
+are rooted on natural languages,
+and 'artificial' is thus misleading.
+The preferred terms seem to be \emph{planned} or \emph{constructed} language
+or simply \emph{conlang}.
 The construction of languages is called glossopoeia.
-TP is a conlang which might be classified
-as engineered for experiments, meditation and philosophy;
+TP is might be considered
+engineered for experiments, meditation and philosophy;
 and suitable for use as an auxiliary international language and

### 2018-07-02 article/acm/samplebody-journals.tex
terminada visao geral da linguagem

@@
 by Sonja Lang in late 1990s~\cite{interview}.
-It was released as a draft in 2001, and in 2007 some documents
+A description was released as a draft in 2001, and in 2007 some documents
 reported it to have a few hundred speakers.
-The English official book~\cite{tpLang} was released only in 2014.
-In 2016, a version of the official book was released in French.
+The English official book~\cite{tpLang} was published only in 2014.
+In 2016, a version of the official book was published in French.
 Nowadays, one finds a number of texts about TP and written
@@
 and writing is called `natural language'.
-A `constructed language' (also planned or invented language) is a natural language built by someone or
-a group, such as Esperanto, Toki Pona, and Lojban.
+A `constructed language' (also planned or invented language),
+e.g. Esperanto, Toki Pona, and Lojban, is a natural language built by someone or
+a group.
 An `artificial language' is most usually a term used
@@
 
-Constructed languages (aka. conlangs) are in some traditions called artificial
+Constructed languages (aka. conlangs) are in some contexts called artificial
 languages, but creators most often prefer to use the term
@@
 or 'planned' language with the argument that the conlangs
-are rooted on natural languages,
+are rooted on \emph{natural} languages,
 and 'artificial' is thus misleading.
@@
 The construction of languages is called glossopoeia.
-TP is might be considered
-engineered for experiments, meditation and philosophy;
-and suitable for use as an auxiliary international language and
+TP is engineered for experiments, meditation and philosophy;
+and useful at least as an auxiliary international language and
 as an artistic language.~\cite{conlanWikip,artlanWikip}
@@
 For example, /p t k s l/ might be pronounced
-% [p t k s l] or [b d g z !R].
-[p t k s l] or [b d g z \textipa{\!R}].% ſ].
+[p t k s l] or [b d g z !R].
+% [p t k s l] or [b d g z \textipa{\!R}].% ſ].
 Especially for poetry, one might consider
@@
 yield by this last rule
-and how I deal with them is in Appendix~\ref{mytoki}.
-The e might be repeated to associate more than
+and how the author deals with them is in Appendix~\ref{mytoki}.
+The particle e might be repeated to associate more than
 one object to a predicate.
@@
 Multiple la-s are not described in literature,
-but one might assume that the last sentence
-is a conditional to the next,
-except in cases where the context strongly suggests
+but one might assume that the last sentences
+are conditions to the next,
+except in cases where the context suggests
 otherwise.
@@
 The first word is the noun or verb and subsequent words
-qualify the noun or verb (i.e. are adjectives and adverbs).
+qualify the noun or verb (i.e. yield adjectives and adverbs).
 The pi particle might be used to separate sequences of words
@@
     through syntax because the order of qualifications is conserved.
-    It adds as an emphasis because of greater length of the written segment, as a preparation: 'jan lili pi mama' (mommy's child).
+    It adds as an emphasis because of greater length of the segment, as a preparation: 'jan lili pi mama' (mommy's child).
     Also used in~\cite{akesiWawa}.
@@
     after li to mean possession, e.g. `soweli li pi sina' (your pet).
-    This employment of pi might be regarded as correct, but is promptly written
+    This employment of pi might thus be regarded as correct, but is promptly written
     as a noun phrase (e.g. `soweli sina') and is not mentioned
@@
 And that the first word of the phrase is the noun or verb,
-and that subsequent words are adjectives or adverbs.
+and that subsequent words are (used as) adjectives or adverbs.
 The pi precedes a noun or a verb to be qualified and then qualify
 the phrase that comes before it.
+These are reasonable expectations and not strict conventions.
 
@@
 They might appear at the end of a phrase,
-should be followed by another phrase,
+might also be followed by another phrase,
 and require no particle. E.g.
@@
 any noun or verb phrase,
-they are one of the three main sources of ambiguity:
+they are one of the four main sources of ambiguity:
+small vocabulary;
 prepositions; absence of li after sina and mi; incomplete sentences.
@@
   \item taso, means however as sentence or 'only' if adjective.
-  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases.
+  \item anu, en: 'or' and 'and'. Used for nouns in noun phrases when subject or after pi.
     For verbs, repeat li.
@@
     As TP is a recent language, and is particularly able to cope with
-    variation due to its simplicity, I would advocate for
+    variation due to its simplicity, one might advocate for
     using en and anu/en wherever there is no ambiguity.
-    E.g.: 'mi tawa mute anu moku lon tenpo lili'
-    In the official documentation that is not described and
-    might be regarded as wrong in strict canonical TP.
+    E.g.: 'mi tawa mute anu moku lon tenpo lili',
+    in the official documentation,
+    might otherwise be regarded as wrong in strict canonical TP.
   \item nanpa, denotes numbering.
@@
     'tan seme la sina pana e sike?'.
-  \item mu, for animal noises. For me it is not a particle, as in the
-    official dictionary, but a noun.
-    I also like to use it as a verb:
-    'mi pakala e luka. mi mu mute.'
+  \item mu, for animal noises. It is not only a particle, as in the
+    official dictionary, but also a noun and a verb:
+    'mi pakala e luka. mi mu mute'.
 \end{itemize}
@@
 nouns, adjectives, verbs, pre-verbs, adverbs, prepositions, particles, and numbers.
-I find that it might help the speaker, specially the newcomer, but
-it might also suggest a deviation from what I understand and read:
-the words might be used indistinctly to be a noun
+Such classification helps the speaker, specially the newcomer, but
+it also suggests a deviation from TP usage:
+the words might be used indistinctly as a noun
 (the first word of: a subject, a predicate when there is no object, an object,
@@
 by a verb and as a verb followed by an adverb).
-The pre-verb words are all also specified in the official dictionary
-to have other morphosyntactic classes,
+The pre-verb words are all
+have other morphosyntactic classes,
+in the official dictionary,
 such as adjective, noun, and verb.
-The only exception is wile, which is specified to be only a pre-verb.
+The only exception is wile, which is specified to be only a pre-verb,
+but is trivially used as a verb:
+'mi wile e moku e telo'.
 
@@
 
+\subsubsection{'li' as the verb 'to be'}
+The particle li very often relates phrases as
+the ver 'to be'. E.g. 'ona li pona' (that is beautiful)
+and 'jan nasin li jan wawa' (someone with techniques is someone strong).
+
 \subsubsection{Recognizing POS tags by speaker and machine}\label{sec:rec}
-One might perform a POS tagging by following these rules:
+One might perform a reasonable POS tagging by following these rules:
 \begin{itemize}
@@
     or that 3) it is a hybrid of noun, verb and adjective.~\cite{janKipo}
+  \item If there is no object, the predicate might be parsed
+as being a hybrid of noun, verb and adjective.
   \item Notice that there is ambiguity in the structure
@@
     in TP.
-    They might be resolved or minimized by using the semantics
+    They might be resolved or minimized considering the semantics
     of the words.
@@
     Another source of ambiguity is the pre-verbs as described
-    in the literature~\cite{tpLang,kama}.
-    But I find it reasonable to understand them as verbs,
-    as described in the end of Section~\ref{sec:par}.
-  \item If there is no object, the predicate might be parsed
-as being a hybrid of noun, verb and adjective.
+    in the literature~\cite{tpLang,kama} and
+    in the end of Section~\ref{sec:par}.
 \end{itemize}
@@
 
-\subsubsection{Additional notes}\label{sec:add}
-John Clifford (a notable ``toki-ponist'', a.k.a. jan Kipo)
-states that there are not noun, verb, adjective (or even prepositional)
-phrases in TP, but only phrases in a structure
-yield by particles and content words~\cite{janKipo}.
+\subsubsection{Additional notes: synonyms, e ni, names, structure, other}\label{sec:add}
 The only synonyms on Toki Pona are:
@@
 These and other issues are further described in Appendix~\ref{mytoki}.
+John Clifford (a notable ``toki-ponist'', aka. jan Kipo)
+states that there are not noun, verb, adjective (or even prepositional)
+phrases in TP, but only phrases in a structure
+yield by particles and content words~\cite{janKipo}.
 

### 2018-07-02 tokipona/__init__.py
adiantando escrita da classe

@@ -1,4 +1,4 @@
-from . import utils
+from . import utils, statistics
 def stats():
     from . import makeStatistics as stats
     return stats

### 2018-07-02 tokipona/makeStatistics.py

@@ -168,14 +168,14 @@ p.mediaRendering.tables.writeTex(
             p.mediaRendering.tables.makeTabular(
                 ["POS"]+labels+["total"], 
                 [["All", "Chosen"]] + list(zip(pos.values(), pos_.values())) + [[sum(pos.values()),120]]),
-            """"POS tags incident and chosen.
+            """"POS tags incident and chosen as preferential e.g. in text synthesis.
             The official dictionary often relates tokens
             to more than one POS tag.
             For the text highlighting Plugin, for example,
             a token has to have an established tag to have
             a defined color.
             On the Chosen column, the tokens were regarded only once
-            by choosing the first classification in the dictionary in """+str(precedence)+"."),
+            by choosing the first occurrence of """+str(precedence)+" in the official dictionary."),
         "../article/pos.tex")
 
 
@@ -342,12 +342,12 @@ p.mediaRendering.tables.writeTex(
             """Frequency of syllables in Toki Pona
             considering all 235 syllables of the 124 tokens,
             only the first or last syllables or only the middle
-            syllable.
+            syllables.
             In parenthesis are the count and percentage of the
             corresponding syllable. For more information and a
             complete list of syllables, see Section~\\ref{sec:stat}.
             """,
-            'freqSyl'),
+            'tab:freqSyl'),
         "../article/syls.tex")
 
 p.mediaRendering.tables.doubleLines("../article/syls.tex",

### 2018-07-02 tokipona/statistics.py

@@ -1,0 +1,15 @@
+# import ezodf
+import pandas as pd
+import numpy as n
+import os
+
+class TP:
+    def __init__(self):
+        os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+        xl_file = pd.ExcelFile('../data/dictionary.xlsx')
+        dfs = {sheet_name: xl_file.parse(sheet_name)
+               for sheet_name in xl_file.sheet_names}
+        df = dfs['Sheet1']
+        table = n.vstack((df.keys(), df.values))
+        self.me = locals()
+

### 2018-07-02 pack/prv/opt/tokipona/aux/article/acm/pos.tex

\begin{table*}[h!]
\begin{center}
\caption{"POS tags incident and chosen as preferential e.g. in text synthesis.
            The official dictionary often relates tokens
            to more than one POS tag.
            For the text highlighting Plugin, for example,
            a token has to have an established tag to have
            a defined color.
            On the Chosen column, the tokens were regarded only once
            by choosing the first occurrence of ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
            in the official dictionary.}\label{foobar}
\begin{tabular}{| l | c | c |}\hline
POS & All  & Chosen \\\hline
NOUN & 58  & 49 \\\hline
ADJECTIVE & 40  & 34 \\\hline
VERB & 15  & 13 \\\hline
PARTICLE & 12  & 12 \\\hline
PRE & 6  & 6 \\\hline
PREPOSITION & 5  & 5 \\\hline
NUMBER & 4  & 1 \\\hline
total & 140  & 120 \\\hline
\end{tabular}\end{center}
\end{table*}

### 2018-07-02 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex

--- 2018-07-02 article/acm/samplebody-journals.tex
@@
 1)
-the analysis of the Toki Pona language through statistics
+the analysis of the Toki Pona (TP) language through statistics
 of the vocabulary obtained by processing the official dictionary;
@@
 the synthesis of sentences, poems and paragraphs (e.g. short stories)
-in Toki Pona;
+in TP;
 3)
-syntax highlighting for Toki Pona,
+syntax highlighting for TP,
 including fine-tuning and theoretical
@@
 an initial Wordnet~\cite{wordnet} of
-Toki Pona words.
+TP words.
 Corpus analysis was already found in~\cite{corpus}.
-The Python scripts (or the toolbox)
-for obtaining all the results in this sections,
+The Python scripts (i.e. the toolbox)
+for obtaining all the results in this section,
 and more, is publicly available in~\cite{tokipona},
@@
 generation of derivatives by other interested parties.
-Such scripts were made very friendly for one to understand and alter
-at will.
+Such scripts are friendly to understand and alter.
 
@@
 and hold further measurements and sets of words
-regarded as less suitable for this exposition,
-but is useful for a deeper investigation.
+regarded as less suitable for this exposition than
+for a deeper investigation.
 
@@
 In all the 124 tokens, there are 235 syllables (68 different),
-and Table~\ref{tab:freqsy} exhibits the 10 most frequent
+and Table~\ref{tab:freqSyl} exhibits the 10 most frequent
 syllables in the first, last, middle and all positions.

### 2018-07-02 pack/prv/opt/tokipona/aux/article/acm/syls.tex

--- 2017-12-15 article/syls.tex
@@
             only the first or last syllables or only the middle
-            syllable.
+            syllables.
             In parenthesis are the count and percentage of the
@@
             complete list of syllables, see Section~\ref{sec:stat}.
-            }\label{tab:freqsy}
+            }\label{tab:freqSyl}
 \begin{tabular}{  l | c   c   c   c  }

### 2018-07-02 pack/prv/opt/tokipona/aux/article/pos.tex

--- 2018-07-02 pack/prv/opt/tokipona/aux/article/acm/pos.tex
@@
 \begin{center}
-\caption{"POS tags incident and chosen as preferential e.g. in text synthesis.
+\caption{"POS tags incident and chosen.
             The official dictionary often relates tokens
@@
             On the Chosen column, the tokens were regarded only once
-            by choosing the first occurrence of ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
-            in the official dictionary.}\label{foobar}
+            by choosing the first classification in the dictionary in ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'].}\label{foobar}
 \begin{tabular}{| l | c | c |}\hline

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/tests/basicStuf.py
finished tokipona merge into prv

@@ -1,0 +1,8 @@
+import sys
+keys=tuple(sys.modules.keys())
+for key in keys:
+    if "tokipona" in key:
+        del sys.modules[key]
+    if "t" in key:
+        del sys.modules[key]
+import tokipona as t

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/setup.py
feitas classes com estruturas de dados basicas e para sintese do arquivo de realce de sintaxe

@@ -9,7 +9,7 @@ with open(path.join(here, 'README.md'), encoding='utf-8') as f:
     long_description = f.read()
 
 setup(
-    name='tokipona',
+    name='t0kipona',
 
     # Versions should comply with PEP440.  For a discussion on single-sourcing
     # the version across setup.py and the project code, see
@@ -22,7 +22,7 @@ setup(
     long_description=long_description,
 
     # The project's main homepage.
-    url='https://github.com/ttm/tokipona',
+    url='https://github.com/ttm/prv',
 
     # Author details
     author='Renato Fabbri',
@@ -105,7 +105,7 @@ setup(
 
     # You can just specify the packages manually here if your project is
     # simple. Or you can use find_packages().
-    packages=["tokipona"],
+    packages=["t0kipona"],
     #packages=find_packages(exclude=['contrib', 'docs', 'tests*']),
 
     # List run-time dependencies here.  These will be installed by pip when your

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -1,0 +1,190 @@
+# import ezodf
+import pandas as pd
+import numpy as n
+import os
+tdir = os.path.dirname(os.path.abspath(__file__))
+
+class TP:
+    def __init__(self):
+        self._initAuxVars()
+        self._initDicts()
+        self._loadDictionary()
+        self._getClasses()
+        self._analyseLemmas()
+        self._posStats()
+
+#     def _uniqueCombinations(self, comb_pos_):
+#         comb_pos_ = comb_pos_[:]
+#         for i in comb_pos_:
+#             same_size = [j for j in comb_pos_ if (len(j) == len(i) and j != i)]
+#             for comb in same_size:
+#                 if sum([pos in i for pos in comb]) == len(i):
+#                     comb_pos_.remove(i)
+#                     return uniqueCombinations(comb_pos_)
+#         return comb_pos_
+
+    def _posStats(self):
+        self.comb_pos=list(self.word_classes.values())
+        self.comb_pos_=[]
+        for i in self.comb_pos:
+            if i not in self.comb_pos_:
+                self.comb_pos_.append(i)
+        # self.uniqueCombinations(comb_pos_)
+
+        self.pos_comb = {}
+        for comb in self.comb_pos_:
+            c = str(comb)
+            self.pos_comb[c] = []
+
+        self.comb_pos__ = [set(i) for i in self.comb_pos_]
+        for word in self.word_classes:
+            comb = self.word_classes[word]
+            c = set(comb)
+            for cc in self.comb_pos_:
+                if c == set(cc):
+                    c_ = str(cc)
+                    self.pos_comb[c_].append(word)
+
+
+    def _analyseLemmas(self):
+        self.word_classes = {}
+        for row in self.table:
+            # Better choose the class
+            classes = [i for i in row[1].split() if i.isupper() and i not in ("I,","O!")]
+            classes = ["PRE" if "PRE-VERB" in i else i for i in classes]
+            # print(classes)
+            if self.precedence=="first":
+                class_ = classes[0]
+            else:
+                class__ = classes
+                not_chosen = 1
+                for pos in self.precedence:
+                    if pos in class__ and not_chosen:
+                        class_ = pos
+                        self.class_count[pos] += 1
+                        self.class_count_[pos] += 1
+                        not_chosen = 0
[…]

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/tests/basicStuff.py

@@ -1,6 +1,6 @@
 import sys
 keys=tuple(sys.modules.keys())
 for key in keys:
-    if "tokipona" in key:
+    if "t0kipona" in key:
         del sys.modules[key]
-import tokipona as t
+import t0kipona as t

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/tokipona/statistics.py

@@ -1,15 +1,0 @@
-# import ezodf
-import pandas as pd
-import numpy as n
-import os
-
-class TP:
-    def __init__(self):
-        os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
-        xl_file = pd.ExcelFile('../data/dictionary.xlsx')
-        dfs = {sheet_name: xl_file.parse(sheet_name)
-               for sheet_name in xl_file.sheet_names}
-        df = dfs['Sheet1']
-        table = n.vstack((df.keys(), df.values))
-        self.me = locals()
-

### 2018-07-02 pack/prv/opt/tokipona/aux/article/acm/pos.tex
terminada passagem de rotinas das tabelas do artigo para classes

@@
                     On the Chosen column, the tokens were regarded only once
-            by choosing the first occurrence of ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
-            in the official dictionary.}\label{foobar}
-\begin{tabular}{| l | c | c |}\hline
+                    by choosing the first occurrence of ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER'] in the official dictionary.}\label{foobar}
+\begin{tabular}{  l | c   c  }
 POS & All  & Chosen \\\hline
-NOUN & 58  & 49 \\\hline
-ADJECTIVE & 40  & 34 \\\hline
-VERB & 15  & 13 \\\hline
-PARTICLE & 12  & 12 \\\hline
-PRE & 6  & 6 \\\hline
-PREPOSITION & 5  & 5 \\\hline
+NOUN & 58  & 49 \\
+ADJECTIVE & 40  & 34 \\
+VERB & 15  & 13 \\
+PARTICLE & 12  & 12 \\
+PRE & 6  & 6 \\
+PREPOSITION & 5  & 5 \\
 NUMBER & 4  & 1 \\\hline
-total & 140  & 120 \\\hline
+total & 140  & 120 \\
 \end{tabular}\end{center}

### 2018-07-02 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -1,6 +1,5 @@
 # import ezodf
-import pandas as pd
-import numpy as n
+import pandas as pd, numpy as n
 import os
 tdir = os.path.dirname(os.path.abspath(__file__))
 
@@ -181,10 +180,269 @@ class TPTabFig(TP):
         tf.mkAll()
 
     Check all tf.mk* methods for individual
-    tabels and figuras.
+    tabels and figures.
 
     """
     def __init__(self):
         TP.__init__(self)
+        self.tpwords = list(self.word_classes.keys())
+        self.vowels = 'aeiou'
+        self.consonants = 'jklmnpstw'
         print(self.__doc__)
+
+    def mkPosTable(self):
+        from collections import OrderedDict
+        pos_ = OrderedDict(sorted(self.class_count_.items(), key=lambda t: -t[1]))
+        pos = OrderedDict(sorted(self.class_count.items(), key=lambda t: -t[1]))
+
+        import percolation as p
+        labels=list(pos.keys())
+        fpath = tdir.split("plugin")[0] + 'aux/article/acm/pos.tex'
+        p.mediaRendering.tables.writeTex(
+                p.mediaRendering.tables.encapsulateTable(
+                    p.mediaRendering.tables.makeTabular(
+                        ["POS"]+labels+["total"], 
+                        [["All", "Chosen"]] + list(zip(pos.values(), pos_.values())) + [[sum(pos.values()),120]]),
+                    """"POS tags incident and chosen as preferential e.g. in text synthesis.
+                    The official dictionary often relates tokens
+                    to more than one POS tag.
+                    For the text highlighting Plugin, for example,
+                    a token has to have an established tag to have
+                    a defined color.
+                    On the Chosen column, the tokens were regarded only once
+                    by choosing the first occurrence of """+str(self.precedence)+" in the official dictionary."),
+                fpath)
+        p.mediaRendering.tables.doubleLines(fpath,
+                hlines=[], vlines=[], hlines_=[0,2,3,4,5,6,7,9],
+                vlines_=[0,2,3])
+
+    def mkLetters(self):
+        tpwords = self.tpwords
+        tpwords_ = ''.join(tpwords)
+
+        vowels = self.vowels
+        consonants = self.consonants
+        vowels_ = ''.join(v for v in tpwords_ if v in vowels)
+        consonants_ = ''.join(v for v in tpwords_ if v in consonants)
+
+        fracv =  [100*tpwords_.count(i)/len(tpwords_) for i in vowels]
[…]

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
finalizada revisao ateh sintese de texto

@@
 A complete list of the syllables and their frequencies (in different
-positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
+positions) is in the \tttt{tf.hsyls} variable after executing the \tttt{basic/stardardAll.py}
 script.
@@
 and ordered alphabetically and by the number of letters,
-is in the \tttt{hlsyl\_\_} variable.
+is in the \tttt{tf.hlsyl\_\_} variable.
 
@@
 are as shown in Table~\ref{freqLet}
-for start, end and internal position.
+for start, end and internal positions.
 The limited number of consonants favors
-the naturalness of the language prosody
+the language prosody to appear natural
 as it resembles babbling.
@@
 
-Given the rules in Section~\ref{phonology}
+Within the rules exposed in Section~\ref{phonology}
 (of 14 letters, 5 vowels, (C)V(n) phonemes,
-forbidden ji, wu, wo, ti, nn, nm),
+forbidden ji, wu, wo, ti, nn, nm syllables),
 96 words are possible with 1 syllable,
@@
 else (i.e. kepeken, lon, tan).
-This yields $120-13=107 words$.
+This yields $120-13=107$ words.
 As discussed in Section~\ref{syntax},
-I advocate that, although such words are classified in the dictionary,
+the author advocates that, although words have specific classes in the dictionary,
 they might be used indistinctly as nouns, adjectives,
@@
 To account for the particles, one possibility is
-to assume for simplicity that one might use none or one particle
+to assume for simplicity the use of at most one particle
 (not li, e, la, pi, resulting in $8$ particles and $9$ possibilities)
@@
 and only one predicate phrase
-(e.g. 'ona li moku e soweli lon supa').
-The general case is implemented as a function
-in the script \tttt{makeStatistics.py}~\cite{tokipona}.
+(e.g. 'ona li moku e soweli lon supa').~\cite{tokipona}.
 

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/vowels.tex
terminada passagem de rotinas das tabelas do artigo para classes
finalizada revisao ateh sintese de texto

--- 2017-12-15 article/vowels.tex
@@
-\begin{table*}[h]
+\begin{table*}[h!]
 \scriptsize
 \begin{center}
-\caption{Frequency of letters in Toki Pona.
+\caption{Frequency of letters in Toki Pona. Columns
                     freq, freq\_I, freq\_L and freq\_M are
@@
                     while in middle position 'l' is the most frequent consonant.
-            Many other conclusions can be drawn from this table and are
+                    Many other conclusions may be drawn from this table and are
                     useful e.g. for exploring sonorities in poems.}\label{freqLet}
-\hspace*{-1.3cm}
 \begin{tabular}{  l | c   c   c | c   c   c | c   c   c | c   c   c  }

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -268,7 +268,7 @@ class TPTabFig(TP):
                         ["letter"]+[i for i in vowels]+[i for i in consonants], 
                         ll,
                         True),
-                    """Frequency of letters in Toki Pona.
+                    """Frequency of letters in Toki Pona. Columns
                     freq, freq\\_I, freq\\_L and freq\\_M are
                     the frequencies of the letters in any, initial, last and middle
                     positions.
@@ -284,7 +284,7 @@ class TPTabFig(TP):
                     almost 20\\% of the words end with 'n'.
                     On the initial position, 's' is the most frequent consonant,
                     while in middle position 'l' is the most frequent consonant.
-                    Many other conclusions can be drawn from this table and are
+                    Many other conclusions may be drawn from this table and are
                     useful e.g. for exploring sonorities in poems.""",
                     'freqLet'),
                 fpath)
@@ -311,9 +311,9 @@ class TPTabFig(TP):
         hsyls = [self._histSyl(i) for i in syls]
 
         lsyl = [len(i) for i in syllables]
-        hlsyl = [100*lsyl.count(i)/len(lsyl) for i in (1, 2, 3)]
+        self.hsyl = hlsyl = [100*lsyl.count(i)/len(lsyl) for i in (1, 2, 3)]
         hlsyl_ = [lsyl.count(i) for i in (1, 2, 3)]
-        hlsyl__ = [[w for w in self.tpwords if len(self._getSyllables(w)) == i] for i in (1, 2, 3)]
+        self.hsyl__ = hlsyl__ = [[w for w in self.tpwords if len(self._getSyllables(w)) == i] for i in (1, 2, 3)]
         [i.sort() for i in hlsyl__]
         [i.sort(key=len) for i in hlsyl__]
 
@@ -445,4 +445,8 @@ class TPTabFig(TP):
         nwords = [len(i) for i in (words1, words2, words3)]
         return nwords
 
+    def mkAll(self):
+        self.mkPosTable()
+        self.mkSyllables()
+        self.mkLetters()
 

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada a classe de sintese de fragmentos e poemas em toki pona

@@
 to choose sentences or verses or stanzas.
-The same package~\cite{tokipona}
+The package~\cite{tokipona}, herein presented, also
 has capabilities for synthesizing TP text.
@@
 in more strict or personal terms, such a person might just write
-them normally.
+them normally or use the synthesized texts as raw material.
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/__init__.py

@@ -1,4 +1,4 @@
-from . import utils, statistics
+from . import utils, statistics, texts
 def stats():
     from . import makeStatistics as stats
     return stats

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -169,8 +169,52 @@ syntax case ignore\n\n'''
         with open(dpath, 'w') as f:
             f.write(header+coloring)
 
+class TPTextBasic(TP):
+    def __init__(self):
+        TP.__init__(self)
+        self.tpwords = list(self.word_classes.keys())
+        self.vowels = 'aeiou'
+        self.consonants = 'jklmnpstw'
 
-class TPTabFig(TP):
+    def _firstSyllable(self, token):
+        if len(token) <= 2:
+            sy = token
+            tk = ''
+        elif token[0] in self.vowels:
+            if token[1] == 'n':
+                if token[2] in self.vowels:
+                    sy = token[0]
+                    tk = token[1:]
+                else:
+                    sy = token[:2]
+                    tk = token[2:]
+            else:
+                sy = token[0]
+                tk = token[1:]
+        else:
+            if token[2] == 'n':
+                if len(token) == 3 or token[3] in self.consonants:
+                    sy = token[:3]
+                    tk = token[3:]
+                else:
+                    sy = token[:2]
+                    tk = token[2:]
+            else:
+                sy = token[:2]
+                tk = token[2:]
+        return sy, tk
+
+    def _getSyllables(self, token):
+        tk = token
+        syl = []
+        while tk:
+            sy, tk = self._firstSyllable(tk)
+            syl.append(sy)
+        return syl
+
+
+
+class TPTabFig(TPTextBasic):
     """
     TP tables and figures.
 
@@ -184,10 +228,7 @@ class TPTabFig(TP):
 
     """
     def __init__(self):
-        TP.__init__(self)
-        self.tpwords = list(self.word_classes.keys())
-        self.vowels = 'aeiou'
-        self.consonants = 'jklmnpstw'
+        TPTextBasic.__init__(self)
         print(self.__doc__)
 
     def mkPosTable(self):
@@ -345,42 +386,6 @@ class TPTabFig(TP):
         p.mediaRendering.tables.doubleLines(fpath,
                 hlines=[], vlines=[], hlines_=[0,2,3,4,5,6,7,8,9,10,11],
                 vlines_=[0,2,3,4,5])
-
-    def _firstSyllable(self, token):
-        if len(token) <= 2:
-            sy = token
-            tk = ''
-        elif token[0] in self.vowels:
-            if token[1] == 'n':
-                if token[2] in self.vowels:
-                    sy = token[0]
-                    tk = token[1:]
-                else:
-                    sy = token[:2]
[…]

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/texts.py

@@ -1,0 +1,290 @@
+from . import statistics
+import numpy as n
+np = n
+
+class TPSynth(statistics.TPTextBasic):
+    def __init__(self):
+        statistics.TPTextBasic.__init__(self)
+        self._declareWords()
+
+    def _declareWords(self):
+        self.plain_words = list(self.word_classes.keys())
+        self.prepositions = list(self.words['PREPOSITION'])
+        forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
+        for f in forbiden:
+            self.plain_words.remove(f)
+        self.words_nsy = [
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 1],
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 2],
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 3]
+                ]
+
+    def _countSyllables(self, token):
+        t = token.split(' ')
+        nsy = 0
+        vowel = 0
+        for i in t:
+            nsy += len(self._getSyllables(i))
+            if i[0] in self.vowels and vowel:
+                nsy -= 1
+            if i[-1] in self.vowels:
+                vowel = 1
+            else:
+                vowel = 0
+        return nsy
+
+    def createPhrase(self, nsy=0, nwords=0, vowel=False):
+        """
+        Create a Toki Pona phrase.
+
+        Such phrase might be used as a noun or verb phrase,
+        and as a subject, predicate, object or prepositional phrase.
+
+        Parameters
+        ----------
+        nsy : integer
+            The number of syllables of the phrase.
+            If nsy=0, the phrase will have a random number of syllables
+        nwords : integer
+            The number of words in the phrase.
+            If nwords != 0, nsy is ignored.
+        vowel : boolean
+
+        Returns
+        -------
+        ph : string
+            A phrase in Toki Pona.
+
+        See Also
+        --------
+        createSentence : creates a sentence in Toki Pona.
+
+        Notes
+        -----
+        If a word ends with a vowel and the next starts with a vowel,
+        an elision occurs and such two syllables are regarded as one.
+
+        References
+        ----------
+        .. [1] Fabbri, Renato, "Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts." arXiv preprint arXiv:abs/XXX.XXXX (2017)
+
[…]

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
revisada parte de sh

@@
 has a Vim~\cite{vim} syntax highlighting plugin
-for Toki Pona,
+for Toki Pona.
+The Python scripts hold
 routines to arrange the coloring schemes
-(i.e. to yield the text files with the specifications for which tokens
-to relate to which coloring groups),
-and instructions for installing and using
-the resulting syntax highlighting for Toki Pona texts~\cite{tokipona}.
+(i.e. to specify which tokens
+are related to which coloring groups)
+which are stored in the \tttt{syntax/tokipona.vim} file.
 An online syntax highlighting gadget is found at~\cite{tpNetSH},
@@
 Table~\ref{foobar}.
-A word often belongs to more than one class,
-thus precedence might be set by the user.
-Also, some classes might be further refined (e.g. words beginning with
-vowels) or joined (e.g. distinguishing only particles and the rest,
-or particles and prepositions and the rest).
+Also, some classes might be further defined (e.g. words beginning with
+vowels) or joined (e.g. distinguishing only particles against rest,
+or particles and prepositions against the rest).
 The colors are also promptly changed according to~\cite{vim}
-and exemplified in the package documentation.
+and exemplified in the package documentation,
+specially the \tttt{tokipona/syntax/tokipona.vim} syntax file
 
 Currently, the Python package synthesizes the
-syntax file through the\\ \tttt{makeVimSyntax.py} script.
-The user has control of class precedence and
+syntax file through the\\ \tttt{syntax.TPSynHigh} class.
+The user has control of class precedence,
 merging and further details through tweaking such routine.
@@
 and Vim's highlighting schemes as described in~\cite{vim}.
-In summary,
-the usage of the package and plugin might be performed
-through the following actions:
+The usage of the package and plugin might be performed,
+in summary, through the following actions:
 \begin{itemize}
-  \item Installation of the plugin.
-  \item Tweaking of the syntax file by hand.
-  \item Running the \tttt{makeVimSyntax.py}
-    Python script to generate a new \tttt{tokipona.vim} syntax file
-    according to other settings.
+  \item Installation of the plugin, so that
+    syntax highlighting of TP texts will be performed in any \tttt{*.tp} or \tttt{*.tokipona} file.
+  \item Tweaking of the syntax file by hand. Or
+  \item Instantiating the \tttt{syntax.TPSynHigh} class and then executing the \tttt{mkSynHighFile()} method
+    of the resulting object.
+    This generates a new \tttt{tokipona.vim} syntax file, at the right location in Vim folders,
+    according to object settings.
   \item Write a file with the \tttt{.tokipona} or \tttt{.tp} extension
@@
     Reload the highlighting scheme using \tttt{:e} whenever you
-    change the syntax file by hand or through the Python script.
+    change the syntax file by hand or through the Python routines.
   \item Access the used highlighted groups with \tttt{:syntax}.
-    Access all the highlighting groups with \tttt{:so
-    \$VIMRUNTIME/syntax/hitest.vim} or \tttt{:hi}.
+    Access all the highlighting groups with \tttt{:so \$VIMRUNTIME/syntax/hitest.vim} or \tttt{:hi}.
     Change the coloring of a set of terms by associating
-    a used group (e.g. tokiponaADJECTIVE) to an existing group (e.g.
-    Visual) such as in \tttt{:highlight link tokiponaADJECTIVE Visual}.
+    a used group (e.g. \tttt{tpADJECTIVE}) to an existing group (e.g.
+    Visual) such as in \tttt{:highlight link tpADJECTIVE Visual}.
+    This may also be achieved changing \tttt{syntax.TPSynHigh().colors}
+    attribute and then executing \tttt{mkSynHighFile()}.
 \end{itemize}
 
+\noindent Such procedures are exemplified in \tttt{basic/syntaxhighlighting.py}.
+
 The syntax file has an association of TP
-words with highlighting groups.
-It also holds associations of these groups to other groups
+tokens with highlighting groups (HIGs).
+It also holds associations of these HIGs to other HIGs 
 to use their color settings.
-Therefore, one is able to use various color schemes
-in syntax-highlighted TP texts, such as in Figure~\ref{fig:syn}.
-A Vim user might run e.g. \tttt{:colorscheme blue},
+Thus, one may change: associations of tokens to HIGs;
+associations of regex patterns to HIGs;
+associations of HIGs to other HIGs;
+whole color schemes or just sections of it.
+The last of these tasks is promptly performed using 
+the \tttt{space-c-s} and \tttt{space-c-c} normal Vim commands
+supplied by the color plugin of the PRV framework~\cite{vim}.
+The other tasks are more easily performed by
+scripting Vim, Python import and package source code,
+depending on the user ability and intended tasks.
+
+The use of different color schemes within the same syntax
+file is exemplified in Figure~\ref{fig:syn}.
+By copying only the \tttt{syntax/tokipona.vim} and \tttt{ftdetect/tokipona.vim} files,
+a user might benefit from the standard Vim commands.
+E.g. \tttt{:colorscheme blue},
 \tttt{:colorscheme solarize}, \tttt{:colorscheme gruvbox} or
@@
 for the way you might edit colors directly.
-Useful commands are given as commented lines
-in the end of the syntax file. % TTM
 One might obtain an HTML file with the colored TP text

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/basic/standardAll.py
primeira versao operante do pacote rearrumado

@@ -1,0 +1,9 @@
+import sys
+keys=tuple(sys.modules.keys())
+for key in keys:
+    if "t0kipona" in key:
+        del sys.modules[key]
+import t0kipona as t
+
+tf = t.statistics.TPTabFig()
+tf.mkAll()

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/__init__.py

@@ -1,17 +1,9 @@
-from . import utils, statistics, texts
-def stats():
-    from . import makeStatistics as stats
-    return stats
-def synthesis():
-    from . import makeTexts as synthesis
-    return synthesis
-def syntax():
-    from . import makeVimSyntax as syntax
-    return syntax
-def wordnet():
-    from . import makeWordnet as wordnet
-    return wordnet
-# del makeStatistics, makeTexts, makeVimSyntax, makeWordnet
+from .corpus import TPCorpus
+from .texts import TPTextBasic
+from .statistics import TPStats
+from .synth import TPSynth, TPTabFig
+
+from . import utils
 
 __doc__ = """The main modules of this tokipona package are:
 - stats: for statistics of the official vocabulary.

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/corpus.py

@@ -1,0 +1,43 @@
+class TPCorpus:
+    def __init__(self):
+        self._initAuxVars()
+        self._initDicts()
+        self._loadDictionary()
+        self._getClasses()
+
+    def _loadDictionary(self):
+        dpath = tdir.split("plugin")[0] + 'aux/dictionary.xlsx'
+        # os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
+        xl_file = pd.ExcelFile(dpath)
+        dfs = {sheet_name: xl_file.parse(sheet_name)
+               for sheet_name in xl_file.sheet_names}
+        df = dfs['Sheet1']
+        self.table = n.vstack((df.keys(), df.values))
+
+    def _getClasses(self):
+        self.classes = set()
+        for row in self.table:
+            classes = " ".join(i for i in row[1].split() if i.isupper())
+            # print(row[0], classes)
+            self.classes.add(classes)
+
+    def _changePrecedence(self):
+        if self.precedence == 'first':
+            self.precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
+        else:
+            self.precedence = 'first'
+
+    def _initAuxVars(self):
+        self.precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
+        self.synonyms = 0
+
+    def _initDicts(self):
+        self.words = {}
+        for class_ in self.precedence:
+            self.words[class_] = set()
+        self.words_all = self.words.copy()
+
+        self.class_count = dict.fromkeys(self.precedence, 0)
+        self.class_count_ = dict.fromkeys(self.precedence, 0)
+
+

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -1,26 +1,22 @@
 # import ezodf
+from .corpus import TPCorpus
+# from .synth import TPTabFig
 import pandas as pd, numpy as n
 import os
 tdir = os.path.dirname(os.path.abspath(__file__))
 
-class TP:
+
+class TPStats(TPCorpus):
+    """
+    very preliminary countings and data structures for statistics.
+
+    counting of lemma relations to POS (part-of-speech) tags.
+
+    """
     def __init__(self):
-        self._initAuxVars()
-        self._initDicts()
-        self._loadDictionary()
-        self._getClasses()
+        TPCorpus.__init__(self)
         self._analyseLemmas()
         self._posStats()
-
-#     def _uniqueCombinations(self, comb_pos_):
-#         comb_pos_ = comb_pos_[:]
-#         for i in comb_pos_:
-#             same_size = [j for j in comb_pos_ if (len(j) == len(i) and j != i)]
-#             for comb in same_size:
-#                 if sum([pos in i for pos in comb]) == len(i):
-#                     comb_pos_.remove(i)
-#                     return uniqueCombinations(comb_pos_)
-#         return comb_pos_
 
     def _posStats(self):
         self.comb_pos=list(self.word_classes.values())
@@ -43,7 +39,6 @@ class TP:
                 if c == set(cc):
                     c_ = str(cc)
                     self.pos_comb[c_].append(word)
-
 
     def _analyseLemmas(self):
         self.word_classes = {}
@@ -84,374 +79,3 @@ class TP:
             if "PRE" in self.word_classes[i]:
                 if "VERB" in self.word_classes[i]:
                     self.word_classes[i].remove("VERB")
-
-    def _loadDictionary(self):
-        dpath = tdir.split("plugin")[0] + 'aux/dictionary.xlsx'
-        # os.system('unoconv -f xlsx -o ../data/dictionary.xlsx ../data/dictionary.ods ')
-        xl_file = pd.ExcelFile(dpath)
-        dfs = {sheet_name: xl_file.parse(sheet_name)
-               for sheet_name in xl_file.sheet_names}
-        df = dfs['Sheet1']
-        self.table = n.vstack((df.keys(), df.values))
-
-    def _getClasses(self):
-        self.classes = set()
-        for row in self.table:
-            classes = " ".join(i for i in row[1].split() if i.isupper())
-            # print(row[0], classes)
-            self.classes.add(classes)
-
-    def _changePrecedence(self):
-        if self.precedence == 'first':
[…]

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/syntax.py

@@ -1,0 +1,51 @@
+from .corpus import TP
+
+class TPSynHigh(TP):
+    """
+    Syntax highlighting for Toki Pona in Vim
+
+    Basic usage:
+        sh = TPSynHigh()
+        sh.mkSynHighFile()
+    
+    For different results, tweak sh.colors
+    before executing sh.mkSynHighFile()
+
+    """
+    def __init__(self):
+        TP.__init__(self)
+        self._linkClassSyn()
+        print(self.__doc__)
+
+    def _linkClassSyn(self):
+        self.colors = {'ADJECTIVE': 'Comment', 'NOUN': 'Constant',
+                  'NUMBER': 'Identifier',
+                  'PARTICLE': 'Statement', 'PRE': 'PreProc',
+                  'PREPOSITION': 'Type', 'VERB': 'Special'}
+
+    def mkSynHighFile(self):
+        lines = []
+        for class_ in self.words:
+            words_ = ' '.join(self.words[class_])
+            group = self.colors[class_]
+            lines.append('\nsyntax keyword '+'tokipona'+class_+' '+words_)
+            lines.append('highlight link '+'tokipona'+class_+' '+group)
+
+        coloring = '\n'.join(lines)
+        header = '''" Vim syntax file
+" Language:	toki pona
+" Maintainer:	Renato Fabbri <fabbri@usp.br>
+" Last Change:	2016 Apr 30
+" Remark:	toki pona is a conlang, not a programming language
+
+if exists("b:current_syntax")
+    finish
+endif
+let b:current_syntax = "tokipona"
+
+syntax clear
+syntax case ignore\n\n'''
+        dpath = tdir.split("plugin")[0] + 'syntax/tokipona.vim'
+        with open(dpath, 'w') as f:
+            f.write(header+coloring)
+

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/synth.py

@@ -1,0 +1,542 @@
+from .texts import TPTextBasic
+import numpy as n
+np = n
+
+class TPSynth(TPTextBasic):
+    def __init__(self):
+        TPTextBasic.__init__(self)
+        self._declareWords()
+
+    def _declareWords(self):
+        self.plain_words = list(self.word_classes.keys())
+        self.prepositions = list(self.words['PREPOSITION'])
+        forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
+        for f in forbiden:
+            self.plain_words.remove(f)
+        self.words_nsy = [
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 1],
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 2],
+                [i for i in self.plain_words if len(self._getSyllables(i)) == 3]
+                ]
+
+    def _countSyllables(self, token):
+        t = token.split(' ')
+        nsy = 0
+        vowel = 0
+        for i in t:
+            nsy += len(self._getSyllables(i))
+            if i[0] in self.vowels and vowel:
+                nsy -= 1
+            if i[-1] in self.vowels:
+                vowel = 1
+            else:
+                vowel = 0
+        return nsy
+
+    def createPhrase(self, nsy=0, nwords=0, vowel=False):
+        """
+        Create a Toki Pona phrase.
+
+        Such phrase might be used as a noun or verb phrase,
+        and as a subject, predicate, object or prepositional phrase.
+
+        Parameters
+        ----------
+        nsy : integer
+            The number of syllables of the phrase.
+            If nsy=0, the phrase will have a random number of syllables
+        nwords : integer
+            The number of words in the phrase.
+            If nwords != 0, nsy is ignored.
+        vowel : boolean
+
+        Returns
+        -------
+        ph : string
+            A phrase in Toki Pona.
+
+        See Also
+        --------
+        createSentence : creates a sentence in Toki Pona.
+
+        Notes
+        -----
+        If a word ends with a vowel and the next starts with a vowel,
+        an elision occurs and such two syllables are regarded as one.
+
+        References
+        ----------
+        .. [1] Fabbri, Renato, "Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts." arXiv preprint arXiv:abs/XXX.XXXX (2017)
+
[…]

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/texts.py

@@ -1,290 +1,47 @@
 from . import statistics
-import numpy as n
-np = n
+# from .synth import TPTabFig
+from .statistics import TPStats
 
-class TPSynth(statistics.TPTextBasic):
+class TPTextBasic(TPStats):
     def __init__(self):
-        statistics.TPTextBasic.__init__(self)
-        self._declareWords()
+        TP.__init__(self)
+        self.tpwords = list(self.word_classes.keys())
+        self.vowels = 'aeiou'
+        self.consonants = 'jklmnpstw'
 
-    def _declareWords(self):
-        self.plain_words = list(self.word_classes.keys())
-        self.prepositions = list(self.words['PREPOSITION'])
-        forbiden = ['li', 'e', 'la', 'pi', 'a', 'o', 'anu', 'en', 'seme', 'mu', 'kepeken', 'lon', 'tan']
-        for f in forbiden:
-            self.plain_words.remove(f)
-        self.words_nsy = [
-                [i for i in self.plain_words if len(self._getSyllables(i)) == 1],
-                [i for i in self.plain_words if len(self._getSyllables(i)) == 2],
-                [i for i in self.plain_words if len(self._getSyllables(i)) == 3]
-                ]
-
-    def _countSyllables(self, token):
-        t = token.split(' ')
-        nsy = 0
-        vowel = 0
-        for i in t:
-            nsy += len(self._getSyllables(i))
-            if i[0] in self.vowels and vowel:
-                nsy -= 1
-            if i[-1] in self.vowels:
-                vowel = 1
+    def _firstSyllable(self, token):
+        if len(token) <= 2:
+            sy = token
+            tk = ''
+        elif token[0] in self.vowels:
+            if token[1] == 'n':
+                if token[2] in self.vowels:
+                    sy = token[0]
+                    tk = token[1:]
+                else:
+                    sy = token[:2]
+                    tk = token[2:]
             else:
-                vowel = 0
-        return nsy
-
-    def createPhrase(self, nsy=0, nwords=0, vowel=False):
-        """
-        Create a Toki Pona phrase.
-
-        Such phrase might be used as a noun or verb phrase,
-        and as a subject, predicate, object or prepositional phrase.
-
-        Parameters
-        ----------
-        nsy : integer
-            The number of syllables of the phrase.
-            If nsy=0, the phrase will have a random number of syllables
-        nwords : integer
-            The number of words in the phrase.
-            If nwords != 0, nsy is ignored.
[…]

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada arrumacao minima nos modulos do pacote

@@
 \noindent Such procedures are exemplified in \tttt{basic/syntaxhighlighting.py}.
-
-The syntax file has an association of TP
-tokens with highlighting groups (HIGs).
-It also holds associations of these HIGs to other HIGs 
-to use their color settings.
-Thus, one may change: associations of tokens to HIGs;
+One may change: associations of tokens to HIGs;
 associations of regex patterns to HIGs;
@@
 depending on the user ability and intended tasks.
-
 The use of different color schemes within the same syntax
@@
   \item The use of blue and other high frequency colors (such as
-    violet) for the background.
-  \item The avoidance of blue and other high frequency
-  colors to fill small objects, such as letters and words.
-  \item Explore simplicity and elegance through the use of discrete
-  coloring schemes. Most tokens should be preferably of
-  the same color.
-    One might use only a small number of colors to achieve a clean
+    violet) for the background and the avoidance of their use
+  to fill small objects, such as letters and words.
+  \item Explore simplicity and elegance in the  coloring schemes.
+    Most tokens should be preferably of the same color, i.e.
+    a small number of colors might be explored to achieve a clean
     visualization of texts.
-  Also, a power-law distribution of tokens among colors might be
+  \item A power-law distribution of tokens among colors might be
     well-suited to mimic natural occurring phenomena.
-    (Unfortunately, the TP dictionary classification of words are not in a
-    power-law distribution and might be better described by a half-normal or by
-overlapping normal distributions over the three peaks: $\sim35-58,
-12-18, 1-5$ occurrences).
+    The TP dictionary morphosyntactic classification of words
+    resemble a long-tailed distribution: as shown in Table~\ref{foobar},
+    few classes hold the vast majority of words.
   \item Physical stimuli might be related to perceptual stimuli both 
@@
     This might be useful e.g. for coloring sets of tokens
-    considering their similarity.
+    considering their similarity, or correctly choosing values
+    for e.g. the hue or luminosity channels.
   \item One usually wishes to maximize contrast,
@@
     Considering perceptually uniform color spaces in making
-    choices may be relevant.
-  \item One might consider two types of color schemes: those with a dark background,
-  which are more comfortable at first; and those with a light background,
+    choices may be relevant, such as CIElab and CIEluv.
+  \item Consider two types of color schemes: of those with a dark background,
+  which are more comfortable at first; and of those with a light background,
   which are usually impressive (and even annoying) at first,
-  but the eye gets used to it and it keeps you more stimulated.
+  but the eye adapts in a few minutes, keeps you more stimulated,
+    and is suitable for bright contexts (e.g. in daylight) .
 \item The blue color is specially related to physiological
@@
     and programmers often report that a blue background keeps them
-    awake and more concentrated (I also notice such effect,
-    and I have programmed daily for more than 10 years).
+    awake and more concentrated (the author also notices such effect).
   \item Colors have been associated to enhancements in specific tasks,
@@
     creativity and detail-oriented tasks in~\cite{blue}.
-  \item It is important to consider that the user might 
-    stay many hours
-  at a text editor (and we very often do),
+  \item It is important to consider that \emph{a text editor user} might 
+    stay many hours using the tool (and they very often do),
     and that the colors and formats involved are
-  prone to entail a considerable effect in the body and mental
-    activity and thus in the quality of life and work of a writer
+  thus prone to entail a considerable effect in the body and mental
+    activity, the quality of life, and work results of a writer
     (e.g. a programmer).
@@
     the syntax highlighting (e.g. through keyboard shortcuts)
-  as envisioned in~\cite{vim}.
+  as envisioned and implemented in~\cite{vim}.
 \end{itemize}

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/__init__.py

@@ -2,6 +2,7 @@ from .corpus import TPCorpus
 from .texts import TPTextBasic
 from .statistics import TPStats
 from .synth import TPSynth, TPTabFig
+from .syntax import TPSynHigh
 
 from . import utils
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/corpus.py

@@ -1,3 +1,7 @@
+import os
+tdir = os.path.dirname(os.path.abspath(__file__))
+import pandas as pd, numpy as n
+
 class TPCorpus:
     def __init__(self):
         self._initAuxVars()

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -1,7 +1,6 @@
 # import ezodf
 from .corpus import TPCorpus
 # from .synth import TPTabFig
-import pandas as pd, numpy as n
 import os
 tdir = os.path.dirname(os.path.abspath(__file__))
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/syntax.py

@@ -1,6 +1,6 @@
-from .corpus import TP
+from .corpus import TPCorpus, tdir
 
-class TPSynHigh(TP):
+class TPSynHigh(TPCorpus):
     """
     Syntax highlighting for Toki Pona in Vim
 
@@ -13,7 +13,7 @@ class TPSynHigh(TP):
 
     """
     def __init__(self):
-        TP.__init__(self)
+        TPCorpus.__init__(self)
         self._linkClassSyn()
         print(self.__doc__)
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/synth.py

@@ -1,4 +1,4 @@
-from .texts import TPTextBasic
+from .texts import TPTextBasic, tdir
 import numpy as n
 np = n
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/texts.py

@@ -1,10 +1,10 @@
 from . import statistics
 # from .synth import TPTabFig
-from .statistics import TPStats
+from .statistics import TPStats, tdir
 
 class TPTextBasic(TPStats):
     def __init__(self):
-        TP.__init__(self)
+        TPStats.__init__(self)
         self.tpwords = list(self.word_classes.keys())
         self.vowels = 'aeiou'
         self.consonants = 'jklmnpstw'

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada revisao do artigo de toda a parte de syntax highlighting

@@
 
-The considerable irrelevance of the morphosyntactic classes as
-described in Section~\ref{syntax} suggests that an appropriate coloring
+The considerable irrelevance of the morphosyntactic classes,
+described in Section~\ref{syntax}, suggests that an appropriate coloring
 of words should
@@
 prepositions and particles.
-Also, the coloring for poems might be more appealing if considered
+Also, the coloring of poems might be more appealing if considered
 e.g. the counting of syllables, the repeated letters or syllables,
 and the ending syllables.
-Insights might be obtained through the
+Further insights might be obtained through the
 choices made and advocated in software such as text editors
@@
 pigments\footnote{\url{http://pygments.org/docs/}} and 
-highlight.js\footnote{\url{https://highlightjs.org/}},
+highlight.js\footnote{\url{https://highlightjs.org/}}),
 and
@@
 which is more likely to yield a cleaner text, at least for programming.
-
 One might also 

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada revisao de wordnet e acrescentados antonimos

@@
 through the English lemmas.
-The most canonical (i.e. Princeton) Wordnet only contains nouns,
+The canonical (i.e. Princeton) Wordnet only contains nouns,
 adjectives, verbs, and adverbs.
-Thus, particles and prepositions were not considered.
+Thus, particles were not considered.
 Numbers were considered adjectives.
 Words presented as adjectives in the dictionary
-were considered both as adjectives and adverbs.~\cite{wordnet}
-
-The \tttt{makeWordnet.py} script, available in~\cite{tokipona},
-makes available such tentative TP Wordnets in the simplest form:
+were considered both as adjectives and adverbs.
+Prepositions were considered in all classes.~\cite{wordnet}
+
+The \tttt{TPWordnet} class~\cite{tokipona},
+provides such tentative TP Wordnets in their simplest form:
 the TP words are keys in a dictionary that returns the
@@
 
-It is worth mentioning that TP words are very general
+TP words are very general
 and each of them might mean many things.
@@
 e.g. 
-luka (as hand) is a meronym of sijelo (as body).
+  lawa, luka, noka, sinpin, monsi, lukin, mute, kute, palisa, and lupa
+  might be considered meronyms of sijelo.
+  Antonyms are also specially useful in TP: suno and mun, pona and jaki or ike, sinpin and munsi, lawa and noka, mije and meli, sike and palisa, pana and kama jo, pimeja and walo, weka and poka, sama and ante, ali and ala, anu and e, selo and insa.
 

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
feita implementacao mais decente via classe da wordnet de tp

@@
 with the expositions in Section~\ref{basics}),
-with 4,052 synsets;
+with 4,027 synsets;
 another as such but excluding the prepositions,
-with 3,961 synsets;
+with 3,929 synsets;
 a last one that returns only the synsets
@@
 is in the official dictionary,
-with 2,493 synsets. 
+with 2,462 synsets. 
 Another immediate possibility, not implemented,

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/__init__.py

@@ -3,6 +3,7 @@ from .texts import TPTextBasic
 from .statistics import TPStats
 from .synth import TPSynth, TPTabFig
 from .syntax import TPSynHigh
+from .wordnet import TPWordnet
 
 from . import utils
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/corpus.py

@@ -5,7 +5,6 @@ import pandas as pd, numpy as n
 class TPCorpus:
     def __init__(self):
         self._initAuxVars()
-        self._initDicts()
         self._loadDictionary()
         self._getClasses()
 
@@ -34,14 +33,4 @@ class TPCorpus:
     def _initAuxVars(self):
         self.precedence = ['PRE', 'VERB', 'PREPOSITION', 'PARTICLE', 'ADJECTIVE', 'NOUN', 'NUMBER']
         self.synonyms = 0
-
-    def _initDicts(self):
-        self.words = {}
-        for class_ in self.precedence:
-            self.words[class_] = set()
-        self.words_all = self.words.copy()
-
-        self.class_count = dict.fromkeys(self.precedence, 0)
-        self.class_count_ = dict.fromkeys(self.precedence, 0)
-
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/statistics.py

@@ -14,6 +14,7 @@ class TPStats(TPCorpus):
     """
     def __init__(self):
         TPCorpus.__init__(self)
+        self._initDicts()
         self._analyseLemmas()
         self._posStats()
 
@@ -78,3 +79,14 @@ class TPStats(TPCorpus):
             if "PRE" in self.word_classes[i]:
                 if "VERB" in self.word_classes[i]:
                     self.word_classes[i].remove("VERB")
+
+    def _initDicts(self):
+        self.words = {}
+        for class_ in self.precedence:
+            self.words[class_] = set()
+        self.words_all = self.words.copy()
+
+        self.class_count = dict.fromkeys(self.precedence, 0)
+        self.class_count_ = dict.fromkeys(self.precedence, 0)
+
+

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/wordnet.py

@@ -1,0 +1,89 @@
+# import ezodf
+from .corpus import TPCorpus
+
+class TPWordnet(TPCorpus):
+    def __init__(self):
+        TPCorpus.__init__(self)
+        self._mkDataStr()
+        self._init()
+
+    def _mkDataStr(self):
+        self.words = {}
+        for row in self.table:
+            all_pos = row[1].split("\n")
+            lemma_pos = []
+            for pos in all_pos:
+                class_ = [i for i in pos.split() if i.isupper() and i not in ("I,","O!")][0]
+                en_words = [i for i in pos.replace(" or ", ' ').split() if i.islower() or i in ("I,","O!")]
+                for enw in en_words:
+                    if enw in ('to','a','of','and','in','take','on','form','at','for'):
+                        continue
+                    w = enw.replace(";", "").replace(',', '').replace('(', '').replace(')', '').replace('?', '')
+                    lemma_pos.append( (w, class_) )
+
+            words_ = row[0].split()
+            if 'or' in words_:
+                words_.remove("or")
+            for w in words_:
+                self.words[w] = lemma_pos
+
+    def _init(self):
+        import nltk as k
+        self.pos = {'NOUN': 'n',
+                'VERB': 'v',
+                'PRE-VERB': 'v', # pre-verbs are also verbs
+                'ADJECTIVE': 'asr', # adjectives are also adverbs and satellite adjectives
+                'NUMBER': 'asr' # numbers are qualifiers
+                }
+
+        self.wordnet = {}
+        self.wordnet_ = {}
+        self.wordnet__ = {}
+        for word in self.words:
+            self.wordnet[word] = set()
+            self.wordnet_[word] = set()
+            self.wordnet__[word] = set()
+        for word in self.words:
+            # if word in ('li', 'e', 'la', 'pi'):
+            #     continue
+            synsets = []
+            synsets_ = []
+            synsets__ = []
+            for enw in self.words[word]:
+                class_ = enw[1]
+                if class_ in ('PARTICLE'):
+                    continue
+                w = enw[0]
+                ss_ = k.wordnet.wordnet.synsets(w)
+                if class_ in ('PREPOSITION'):
+                    synsets += ss_
+                    continue
+                c = self.pos[class_]
+                synsets += ss_
+                synsets_ += ss_
+
+                ss = k.wordnet.wordnet.synsets(w, pos=c)
+                synsets__ += ss
[…]

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/wordnet.py
feita implementacao mais decente via classe da wordnet de tp

@@ -2,6 +2,23 @@
 from .corpus import TPCorpus
 
 class TPWordnet(TPCorpus):
+    """
+    Toki Pona (hacky) wordnet.
+
+    After initialization, the most important attributes are three
+    dictionaries relating Toki Pona words to Princeton Wordnet:
+        wordnet, relates each word to all synsets with the same lemma
+        wordnet_, prepositions are excluded
+        wordnet__, only synsets of the same morphosyntactic class as
+            in the official Toki Pona dictionary
+
+    TODO
+    ----
+    - make a better mask to Wordnet, as made in NLTK for Portuguese.
+    - remove/enhance relarions yield by compound terms
+        (e.g. 'have information on')
+    """
+
     def __init__(self):
         TPCorpus.__init__(self)
         self._mkDataStr()
@@ -86,4 +103,3 @@ class TPWordnet(TPCorpus):
         # TODO:
         # total number of synsets
         # total number of synsets counting neighbors
-

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada revisao do manuscritp de tp

@@
 through linguistic semantics and
-for devising tools for analysis, creation and visualization.
+for devising tools for analysis, creation and visualization,
+a fact to which this document is evidence.
 Other conlangs, and even non-planned languages might benefit from
-TP and the content presented here in many ways, e.g. one might
+TP and this content in many ways, e.g. one might
 synthesize TP texts and translate them as convenient,
@@
 or take advantage of the language description to devise new conlangs
-or stilistic guidelines for any language.
-
-Possible next steps are numerous:
+or stylistic guidelines for any language.
+
+Some of the many potential next steps should be highlighted:
 \begin{itemize}
-  \item Further develop the text synthesis facilities described in
+  \item The text synthesis facilities described in
     Section~\ref{synth} might be enhanced by further employing e.g.
     rhythm, meter, rhyme and form techniques~\cite{wikiPoetry}.
-  \item Make a syntax highlighting that colors the tokens with respect
-    to the syntatic position and function, and not in a fixed manner
-    as is implemented and exposed in Section~\ref{shigh}.
-    This might be implemented by using n-grams and
-    further techniques from Natural Language Processing
-  \item Enhance and study the TP Wordnet described in Section~\ref{wn}:
-    implement a NLTK-like Wordnet interface; check if each synset
+  \item Syntax highlighting that colors the tokens with respect
+    to the syntactic positions and functions, and not in a fixed manner
+    as is implemented and exposed in Section~\ref{shigh},
+    may be implemented by using n-grams and
+    other techniques from Natural Language Processing.
+  \item The TP Wordnet(s) described in Section~\ref{wn} should be may
+    be enhanced:
+    implement a NLTK-like Wordnet interface; double-check if each synset
     is correctly associated to the TP words; seek further synsets
@@
     that are related to all the English words not separated by a
-    semicolon; find out the neighborhood of the synsets of the TP
-    Wordnet in the English Wordnet; explore the possibilities for
-    devising new conlangs by means of criteria related to Wordnet.
+    semicolon; scrutinize the neighborhood of the synsets of the TP
+    Wordnet in the English Wordnet.
 %   \item Translators.
@@
   \item Conlanging: use Wordnet for making new conlangs;
-    use insights derived from TP.
-    Maybe create conlangs for specific uses:
+    use insights derived from TP Wordnet.
+    Create conlangs for specific uses:
     describing data, programs,
@@
     (exploring the thinking process).
-    A conlang might have different modes that might 
-    be specified by a section header or tag. 
-  \item Understand how the corpus was gathered in \url{tokipona.net}.
-  \item Know about previously existing words that were used for TP
+    A conlang might have different modes that are
+    specified by section headers or tags. 
+  \item Explore TP relation to other languages
     (e.g. suno and suwi might come from sun and sweet),
-    and about the reasons that led Sonja (and maybe other people)
+    and the reasons that led Sonja (and maybe other people)
     to choose the 14 letters and the syllable structure.
@@
     speaker community and the documentation keepers.
+  \item Understand how the corpus was gathered in \url{tokipona.net}.
   \item Corpus-based analysis.
   \item Publication of original TP texts and translations.
-  \item Make an article written in TP
+  \item Write academic articles in TP
     (extending Section~\ref{ftp} which might be the
@@
     a summary in both English and Toki Pona, for
-    facilitating the acquisition of context,
-    and a article in a canonical structure
-    (introduction and related work, materials and methods,
+    facilitated acquisition of context,
+    and an article in a canonical structure
+    (e.g. introduction and related work, materials and methods,
     results and discussion, conclusions and further work),
     with an Appendix explaining the content and context in English.
-    I first conceived something around complexity, statistics, physics, or computer science.
-    But one possibility is to write about linguistics, philosophy, literature
-    or psychology with the partners I write in English.
-    I can start a draft,
-    they might learn the language in a few hours or weeks
-    (with or without my help),
-    to contribute, and we can write a short paper.
-  \item Enhance the synthesis of text to yield better contextualized text
-    and stylistic traces.
-    Give the user the ability to choose the sentences:
-    generate a sentence randomly according to
-    previously written or given text,
-    to some rules input by the user,
-    to the package and the language guidelines.
-    Should output to the screen and asks to keep and discard.
+    Maybe something about complexity, statistics, physics,
+    or computer science;
+    or linguistics, philosophy, literature
+    or psychology.
+    Partners may learn the language in a few hours or weeks
+    (with or without help).
+    This is a very challenging step.
   \item Make TP variations where each TP word is replaced by a word in
@@
 an official vocabulary, as they already exist in English and French.
-  \item We are particularly interested in software tools,
-    and one possibility is to use a conlang to describe software
+  \item Explore a minimalistic conlang to describe software
     routines.
-    In this sense, it seems profitable
+    In this sense, it seems advantageous
     to understand what `non-ambiguous' means in Lojban,
     and in what sense one is able to compile and parse Lojban.
-    Such study will probably entail considerations about
-    parsing TP.
+    Consider the possibilities of and reasons for parsing TP.
 %  \item The plugin comes with the :TokiStation command, which opens a
@@
 \appendix
-\section{My usage of Toki Pona}\label{mytoki}
-I use the standard sounds, but often use [z] for s.
-I often translate texts to Toki Pona (e.g. biblical excepts)
-and create new texts as poems and short stories.
-Most of them are in~\cite{tokisona}
+\section{Current usage of Toki Pona by the author}\label{mytoki}
+Disclaimer: this whole section holds personal views and usages of
+the Toki Pona (TP) language.
+It is reasonable to use the standard sounds, but often [z] instead of s
+is more comfortable.
+Translation into Toki Pona (e.g. of biblical excepts),
+and the creation of new texts as poems and short stories,
+is both challenging and pleasant.
+Texts are shared e.g. in~\cite{tokisona}
 and/or published in the TP Facebook groups.
-I omit the li particle after subjects sina and mi,
+The li particle is most often not used after subjects sina and mi,
 in accordance with the norm,
-but sometimes I use them when there are many predicates.
+but, even so, they are useful when there are many predicates.
 E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
 In such cases, the first li is sometimes omitted.
-Also, sometimes I use li after mi and sina where I find
+Also, li is found after mi and sina where one finds
 that there is unwanted ambiguity, e.g.
@@
 Names are by default transliterated,
-but I advocate that, as in other languages,
+but one should notice that, as in other languages,
 names might be used as they are in the
@@
 alphabet does not contain ö in such languages.
-I also tend to legitimate the use of English (or German) words
-in TP texts if it is the case,
+The use of e.g. English (or German) words
+in TP texts is found when needed,
 as happens often in scientific writing
@@
 Proposed notations for numbers seem numerous.
-I tend to think that one might indicate if two numbers
+E.g. one might indicate if two numbers
 are being multiplied (pi) or are in different scales
-(such as in decimal or binary notations).
-For example, luka (pi) two might mean 52.
+(such as in decimal or binary notations):
+luka (pi) two mean 52.
 Most importantly,
@@
 Avoiding needless words...
-I've been avoiding e ni: and using only :.
-I've been omitting the subject (sometimes also the li)
+'e ni:' may most often be written as ':'.
+Subject phrase may be omitted (sometimes also the li)
 if it is the
 same as in the last sentence.
-I tend to group these words:
+One may vary TP by grouping e.g. these words:
 noka and anpa; luka, suno, sike and lawa; pali and pana.
-These clusters of meaning hint me that an even simpler
-language is possible and maybe more optimized to the
-core goals of TP.
 
@@
 and absence of a token to denote preposition,
-suggest to me the possibility of a syntax that is always uniquely
+suggest the possibility of a syntax that is always uniquely
 parsed (or at least less ambiguous).
 
-I have been inclined to conceive a language with the same syntax of TP:
-'sentence' la 'sentence = subject + predicate + object (each term with
-a possible preposition complement).
+It is tempting to conceive a language with the same syntax of TP:
+'sentence' la 'sentence = subject + predicate + object' (each term with
+a possible prepositional complement),
 but always using particles to bound the sentence sections,
@@
 meaning language in [toki pona of intelectuals].
-About the keyword for initiating a preposition complement 
+The keyword for initiating a preposition complement 
 might be 'a', but it is already taken in TP.
-Maybe use 'a' before a preposition and use 'ha' or 'he'
+One possibility is to use 'a' before a preposition and use 'ha' or 'he'
 instead of TP a.
 In such a setting,
-one might enable pi li e a (concept-qualifier, subject-predicate,
+one might enable pi, li, e, a (concept-qualifier, subject-predicate,
 predicate-object, concept-preposition)
-in any slot of the template:
-'sentence' la 'sentence = subject + predicate + object (each term with
-a possible preposition complement)'
-Between sentences one would use:
+in any slot of the syntax template.
+Between sentences one may use:
 la pi, la li, la e, la a, and la la.
@@
 and accept any of them for the subject, predicate, object, preposition,
-slots. Or assume a part of speech as default for a slot or for the
+slots. Or assume a part-of-speech as default for a slot or for the
 first word in a slot.
@@
 \section{Final words in Toki Pona}\label{ftp}
-toki li nasin e lawa.
+toki li nasin e lawa.\\
 li nasin tawa (pi) toki insa.
 
-% jan li ken toki li ken pali pona e toki pona.
-toki pona li pona e nasin tan ni:
-ona li jo e nimi nanpa lili.
-ona li pona.
+% jan li ken toki li ken pali pona e toki pona.\\
+toki pona li pona e nasin tan ni:\\
+ona li jo e nimi nanpa lili.\\
+ona li pona.\\
 pona kepeken weka 'p' li ona, a.
 
-o taso la toki pona li kalama li lukin pona.
-sitelen en nasin li open e sitelen
-suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
+o taso la toki pona li kalama li lukin pona.\\
+sitelen en nasin li open e sitelen\\
+suli~\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.\\
 li sona. li nasin e toki e sona e lawa e lon.
 
-toki ni li wawa tawa jan mute nasin en toki.
-wawa tawa toki pi jan sona.
-pi ilo nanpa en nanpa nasin.
-taso tawa toki sona a.
+toki ni li wawa tawa jan mute nasin en toki.\\
+wawa tawa toki pi jan sona.\\
+pi ilo nanpa en nanpa nasin.\\
+taso tawa toki sona a.\\
 sitelen sona, sitelen musi.
 
-ilo lon Poki~\ref{hacks} li pana e sitelen
-e sona tan sitelen,
-en nasin.
-Poki~\ref{basics} li pana e nasin pi toki pona.
+ilo lon Poki~\ref{hacks} li pana e sitelen\\
+e sona tan sitelen,\\
+en nasin.\\
+Poki~\ref{basics} li pana e nasin pi toki pona.\\
 e sitelen tawa kama sona.
 
-mi wile pali e sitelen lon toki pona.
-sitelen lon nasin, sitelen sona,
-sitelen musi.
-ante e toki pona
-la ona li nasa.
-taso nasa li pona mute.
-li pona mute tawa lawa,
-tawa kama sona, tawa sitelen e toki.
-ante la toki pona o.
-toki e toki pona tawa sina.
+mi wile pali e sitelen lon toki pona.\\
+sitelen lon nasin, sitelen sona,\\
+sitelen musi.\\
+ante e toki pona\\
+la ona li nasa.\\
+taso nasa li pona mute.\\
+li pona mute tawa lawa,\\
+tawa kama sona, tawa sitelen e toki.\\
+ante la toki pona o.\\
+toki e toki pona tawa sina.\\
 kama sona e toki pona sina.
 
-o pona tawa jan pi toki pona.
-tawa jan Sonja, Birns-Sprage, Kipo, Pije,
+o pona tawa jan pi toki pona.\\
+tawa jan Sonja, Birns-Sprage, Kipo, Pije,\\
 Siwejo, Malija, Tepan, jan kulupu mute.

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
finished review of tp section

@@
 Toki Pona user and speaker communities; 
-Mario Alzate for introducing me to Toki Pona,
+Mario Alzate Lopez for introducing me to Toki Pona,
 and Silverio Guazzelli Donatti for the stimulus that resulted

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
feito roteiro basico de uso do pacote em python

@@
 % and the next paragraph is an overview. TTM
-The script \tttt{makeStatistics.py}
+The class \tttt{TPTabFig}
 was used to obtain all the measurements and tables
@@
 regarded as less suitable for this exposition than
-for a deeper investigation.
+for a deeper investigation.\cite{prv}
 

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/basic/standardAll.py

@@ -1,9 +1,0 @@
-import sys
-keys=tuple(sys.modules.keys())
-for key in keys:
-    if "t0kipona" in key:
-        del sys.modules[key]
-import t0kipona as t
-
-tf = t.statistics.TPTabFig()
-tf.mkAll()

### 2018-07-03 pack/prv/opt/tokipona/plugin/t0kipona/basic/tbBasic.py

@@ -1,0 +1,37 @@
+#####
+# this is just a hack to reload the module files when they change
+import sys
+keys=tuple(sys.modules.keys())
+for key in keys:
+    if "t0kipona" in key:
+        del sys.modules[key]
+import t0kipona as t
+#####
+
+#######
+# to synthesize TP text
+s = t.TPSynth()
+# use s.createPhrase(), s.createSentence(), s.createParagraph() and
+# s.createPoem()
+
+#######
+# to use the preliminary TP wordnets
+w = t.TPWordnet()
+# use w.wordnet, w.wordnet_ and w.wordnet__, each with less synsets
+
+#######
+# to make the syntax highlighting file:
+sh = t.TPSynHigh()
+sh.mkSynHighFile()  # this creates the Vim syntax highlighting
+
+#######
+# this is the most complete object with data on Toki Pona (TP),
+# including statistics, although other classes have other data.
+tf = t.TPTabFig()
+tf.mkAll()  # this renders all the figures and tables of the TP article
+
+#######
+# also, check the functions and variables in
+# t.utils
+
+# Ter Jul  3 19:59:02 -03 2018

### 2018-07-03 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
terminada versao candidata p submissao

@@
 and useful at least as an auxiliary international language and
-as an artistic language.~\cite{conlanWikip,artlanWikip}
+as an artistic language.~\cite{conlanWikip,wikiArtLang}
 % taxonomy, vocabulary, ontology, terminology,
@@
 % and the next paragraph is an overview. TTM
-The class \tttt{TPTabFig}
+The \tttt{TPTabFig} Python class 
 was used to obtain all the measurements and tables
@@
 A complete list of the syllables and their frequencies (in different
-positions) is in the \tttt{tf.hsyls} variable after executing the \tttt{basic/stardardAll.py}
-script.
+positions) is in the \tttt{tf.hsyls} variable after executing the \tttt{basic/tbBasic.py} script.
 A list of all words, grouped by their size in syllables,
@@
   \end{subfigure}
-  \caption{Toki Pona texts synthesized by the \tttt{makeTexts.py} script
-  in~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
+  \caption{Toki Pona texts synthesized by an instantiated \tttt{TPSynth} Python class~\cite{tokipona}: (a) a phrase, (b) a sentence, (c) a paragraph,
   and (d) a poem.
@@
 The textual synthesis described here,
-and implemented in \tttt{makeTexts.py},
+and implemented in the \tttt{TPSynth} class,
 might be enhanced in countless ways,
@@
 and exemplified in the package documentation,
-specially the \tttt{tokipona/syntax/tokipona.vim} syntax file
+specially the \tttt{syntax/tokipona.vim} syntax file
 
 Currently, the Python package synthesizes the
-syntax file through the\\ \tttt{syntax.TPSynHigh} class.
+syntax file through the\\ \tttt{TPSynHigh} class.
 The user has control of class precedence,
@@
   \item Tweaking of the syntax file by hand. Or
-  \item Instantiating the \tttt{syntax.TPSynHigh} class and then executing the \tttt{mkSynHighFile()} method
+  \item Instantiating the \tttt{TPSynHigh} class and then executing the \tttt{mkSynHighFile()} method
     of the resulting object.
@@
     Visual) such as in \tttt{:highlight link tpADJECTIVE Visual}.
-    This may also be achieved changing \tttt{syntax.TPSynHigh().colors}
-    attribute and then executing \tttt{mkSynHighFile()}.
+    This may also be achieved changing \tttt{TPSynHigh().colors}
+    attribute before executing \tttt{mkSynHighFile()}.
 \end{itemize}
 
-\noindent Such procedures are exemplified in \tttt{basic/syntaxhighlighting.py}.
+\noindent Such procedures are facilitated by the \tttt{basic/tpBasic.py}.
 One may change: associations of tokens to HIGs;
@@
 \bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{unsrt}
 \bibliography{sample-bibliography}

### 2018-07-05 pack/prv/opt/tokipona/aux/article/acm/sample-acmsmall.tex
passado artigo p template correto p tallip (acmsmall)

--- 2018-07-01 article/acm/sample-acmlarge.tex
@@
-\documentclass[acmlarge]{acmart}
+\documentclass[format=acmsmall, review=false, screen=true]{acmart}
 \usepackage[utf8]{inputenc}
@@
 
-%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
-%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
-
 % Copyright
 %\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
+\setcopyright{acmlicensed}
 %\setcopyright{rightsretained}
 %\setcopyright{usgov}
-\setcopyright{usgovmixed}
+%\setcopyright{usgovmixed}
 %\setcopyright{cagov}
@@
 \begin{document}
-% Title portion
-\title{Basic concepts and tools for the Toki Pona minimalist and
-constructed language}
-\titlenote{Description of the language and main issues;
-Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}
+% Title portion. Note the short title for running heads
+\title[A Multifrequency MAC for Wireless Sensor]{Basic concepts and tools for the Toki Pona minimal and constructed language: description of the language and main issues; analysis of the vocabulary; text synthesis and syntax highlighting; Wordnet synsets}
 
@@
 \begin{abstract}
-A minimalist constructed language (conlang)
+A minimal constructed language (conlang)
 is useful for experiments and comfortable for making tools.
-The Toki Pona (TP) conlang is minimalist both in the vocabulary
-(with only 14 letters and 124 words)
+The Toki Pona (TP) conlang is minimal both in the vocabulary
+(with only 14 letters and 124 lemmas)
 and in the (about) 10 syntax rules.
 The language is useful for being a used and somewhat established
-minimalist conlang with at least hundreds of fluent speakers.
+minimal conlang with at least hundreds of fluent speakers.
 This article exposes current concepts and resources
 for TP,
-and makes available Python scripted routines for
+  and makes available Python (and Vim) scripted routines for
 the analysis of the language,
-the synthesis of texts, the specification of syntax highlighting
-schemes, and the achievement of a preliminary TP
-Wordnet.
+synthesis of texts, syntax highlighting schemes,
+  and the achievement of a preliminary TP Wordnet.
 Focus is on the analysis of the basic vocabulary,
@@
 
-% DO NOT use this command unless you want to change
-% the default behavior
-% \authorsaddresses{Authors' addresses: G.~Zhou, Computer Science
-%   Department, College of William and Mary, 104 Jameson Rd,
-%   Williamsburg, PA 23185, US, \path{gzhou@wm.edu}; V.~B\'eranger,
-%   Inria Paris-Rocquencourt, Rocquencourt, France; A.~Patel, Rajiv
-%   Gandhi University, Rono-Hills, Doimukh, Arunachal Pradesh, India;
-%   H.~Chan, Tsinghua University, 30 Shuangqing Rd, Haidian Qu, Beijing
-%   Shi, China; T.~Yan, Eaton Innovation Center, Prague, Czech Republic;
-%   T.~He, C.~Huang, J.~A.~Stankovic University of Virginia, School of
-%   Engineering Charlottesville, VA 22903, USA; T. F. Abdelzaher,
-%   (Current address) NASA Ames Research Center, Moffett Field,
-%   California 94035.}
-
 \maketitle
@@
 % The default list of authors is too long for headers.
-\renewcommand{\shortauthors}{G. Zhou et al.}
+\renewcommand{\shortauthors}{Fabbri, R.}
 

### 2018-07-05 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
finalizado artigo para publicacao
passado artigo p template correto p tallip (acmsmall)

@@
 \section{Introduction}\label{intro}
-Toki Pona (TP) is a minimalist conlang (constructed language)
+Toki Pona (TP) is a minimal conlang (constructed language)
 with only 124 words (120 without the synonyms).
@@
 which entails facilitated e.g. learning and tool making.
-Another reason why the minimalist language design is compelling
+Another reason why the minimal language design is compelling
 is the study and harnessing of the strong and weak forms of
@@
 For example, /p t k s l/ might be pronounced
-[p t k s l] or [b d g z !R].
-% [p t k s l] or [b d g z \textipa{\!R}].% ſ].
+% [p t k s l] or [b d g z !R].
+[p t k s l] or [b d g z \textipa{\!R}].% ſ].
 Especially for poetry, one might consider
@@
 the language in Section~\ref{hacks}.
-As a minimalist conlang, TP is very convenient for cognitive
+As a minimal conlang, TP is very convenient for cognitive
 experiments
@@
 
+At the moment, the framework in~\cite{prv} is being used, by the author, for TP chatter bots
+and for semantic explorations of domain knowledge.
 Some of the many potential next steps should be highlighted:
@@
 an official vocabulary, as they already exist in English and French.
-  \item Explore a minimalistic conlang to describe software
+  \item Explore a minimal conlang to describe software
     routines.
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliographystyle{unsrt}
 \bibliography{sample-bibliography}

### 2018-07-05 pack/prv/opt/bot/plugin/b0t/b0t/__init__.py
conversacao pt e en para toki pona jah operante no b0t

@@ -1,4 +1,5 @@
 __doc__ = "chat agents capable of writing poems and articles though computational intelligence"
 
 from .n4ive import *
+from .tokipona import TPBot
 

### 2018-07-05 pack/prv/opt/bot/plugin/b0t/b0t/n4ive.py

@@ -1,4 +1,4 @@
-import os, types, random
+import os, types, random, string
 import nltk as k, networkx as x
 from collections import Counter
 
@@ -118,7 +118,8 @@ class SimplestMarkovReality(Reality):
         # choose a random bigram or trigram
         # choose a word to begin sentence
         # concatenate bi/tri until the phrase is complete
-        words_ = words.split(' ')
+        words_ =  k.wordpunct_tokenize(words)
+        words_ = [i for i in words_ if i not in string.punctuation]
         if len(words_) > 0:
             word = random.choice(words_)
             while word not in self.vocab:

### 2018-07-05 pack/prv/opt/bot/plugin/b0t/b0t/tokipona.py

@@ -1,0 +1,37 @@
+import string, random
+import nltk as k
+from t0kipona import TPWordnetPP, TPSynth
+
+class TPBot:
+    def __init__(self):
+        self.wn = TPWordnetPP()
+        self.s = TPSynth()
+
+    def tpTalk(self, words=''):
+        words_ = k.wordpunct_tokenize(words)
+        words_ = [i for i in words_ if i not in string.punctuation]
+        # find the synsets for each w in word_
+        # find the tp words which have such synsets
+        # choose a random word or the word that has more synsets in the wS
+        ss = [set(k.wordnet.wordnet.synsets(w, lang='por')) for w in words_]
+        [i.update(set(k.wordnet.wordnet.synsets(w))) for i, w in zip(ss, words_)]
+        wnmask = dict(zip(words_, ss))
+        tpw = []
+        i = 0
+        self.wordsss = {}
+        for ss_ in ss:
+            for ss__ in ss_:
+                self.wordsss[ss__] = words_[i]
+                if ss__ in self.wn.wnss:
+                    tpw.append(self.wn.wnss[ss__])
+            i += 1
+        print(tpw)
+        tpw_ = list(set(tpw))
+        w_ = random.choice(tpw_)
+        self.dd = locals()
+        return self.s.createParagraph(1, [w_])
+        # percorrer hiperonimos p achar os synsets do toki pona
+
+
+
+

### 2018-07-05 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/__init__.py

@@ -3,7 +3,7 @@ from .texts import TPTextBasic
 from .statistics import TPStats
 from .synth import TPSynth, TPTabFig
 from .syntax import TPSynHigh
-from .wordnet import TPWordnet
+from .wordnet import TPWordnet, TPWordnetPP
 
 from . import utils
 

### 2018-07-05 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/synth.py

@@ -71,6 +71,8 @@ class TPSynth(TPTextBasic):
         """
         phrase = ''
         if nwords:
+            if nsy == 0:
+                nsy = n.random.randint(1,7)
             for i in range(nwords):
                 w = n.random.choice(self.plain_words)
                 phrase += w + ' '
@@ -113,6 +115,7 @@ class TPSynth(TPTextBasic):
                             elif self._countSyllables(' '.join(p_)) > nsy:
                                 chosen = n.random.randint(0, len(p_))
                                 while p_[chosen] == 'pi':
+
                                     chosen = n.random.randint(0, len(p_))
                                 p_.pop(chosen)
                         else:

### 2018-07-05 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/wordnet.py

@@ -103,3 +103,14 @@ class TPWordnet(TPCorpus):
         # TODO:
         # total number of synsets
         # total number of synsets counting neighbors
+
+class TPWordnetPP(TPWordnet):
+    def __init__(self):
+        TPWordnet.__init__(self)
+        self._mkSSMask()
+
+    def _mkSSMask(self):
+        self.wnss = {}
+        for w in self.wordnet:
+            for ss in self.wordnet[w]:
+                self.wnss[ss] = w

### 2018-07-05 pack/prv/opt/bot/plugin/b0t/b0t/tokipona.py
feitos hacks no bot e aa para conversar em ptbr e en e tp

@@ -28,7 +28,7 @@ class TPBot:
         print(tpw)
         tpw_ = list(set(tpw))
         w_ = random.choice(tpw_)
-        self.dd = locals()
+        # self.dd = locals()
         return self.s.createParagraph(1, [w_])
         # percorrer hiperonimos p achar os synsets do toki pona
 

### 2018-07-05 pack/prv/opt/bot/plugin/b0t/basic/b0tBasic.py

@@ -1,0 +1,12 @@
+#####
+# this is just a hack to reload the module files when they change
+import sys
+keys=tuple(sys.modules.keys())
+for key in keys:
+    if ("t0kipona" in key) or ('b0t' in key):
+        del sys.modules[key]
+import t0kipona as t
+import b0t as b
+#####
+
+bb = b.TPBot()

### 2018-07-06 pack/prv/opt/bot/plugin/b0t/b0t/tokipona.py
removido bot de tp dos shouts de aa

@@ -1,4 +1,4 @@
-import string, random
+import string
 import nltk as k
 from t0kipona import TPWordnetPP, TPSynth
 
@@ -27,7 +27,10 @@ class TPBot:
             i += 1
         print(tpw)
         tpw_ = list(set(tpw))
-        w_ = random.choice(tpw_)
+        if tpw_:
+            w_ = self.s.randWord(tpw_)
+        else:
+            w_ = self.s.randWord()
         # self.dd = locals()
         return self.s.createParagraph(1, [w_])
         # percorrer hiperonimos p achar os synsets do toki pona

### 2018-07-06 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/syntax.py

@@ -1,6 +1,6 @@
-from .corpus import TPCorpus, tdir
+from .statistics import TPStats, tdir
 
-class TPSynHigh(TPCorpus):
+class TPSynHigh(TPStats):
     """
     Syntax highlighting for Toki Pona in Vim
 
@@ -13,7 +13,7 @@ class TPSynHigh(TPCorpus):
 
     """
     def __init__(self):
-        TPCorpus.__init__(self)
+        TPStats.__init__(self)
         self._linkClassSyn()
         print(self.__doc__)
 

### 2018-07-06 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/synth.py

@@ -1,5 +1,5 @@
 from .texts import TPTextBasic, tdir
-import numpy as n
+import numpy as n, random
 np = n
 
 class TPSynth(TPTextBasic):
@@ -99,7 +99,7 @@ class TPSynth(TPTextBasic):
         nsy_ = sum([len(self._getSyllables(i)) for i in p_])
         if len(p_) > 2 and nsy_ > 4:
             if n.random.random() < .2:  # use pi in 1/5 of sentences
-                print('\n', p, self._countSyllables(p))
+                # print('\n', p, self._countSyllables(p))
                 where = n.random.randint(len(p_)-2)
                 p_.insert(where+1, 'pi')
                 if p_[where+2][0] not in self.vowels or (p_[where+2][0] in self.vowels and p_[where][-1] in self.vowels):
@@ -125,7 +125,7 @@ class TPSynth(TPTextBasic):
                             w = n.random.choice(self.words_nsy[size-2])
                             p_[chosen] = w
                             p__ = ' '.join(p_)
-                print(' '.join(p_), self._countSyllables(' '.join(p_)))
+                # print(' '.join(p_), self._countSyllables(' '.join(p_)))
         pp = ' '.join(p_)
 
         return pp
@@ -291,6 +291,11 @@ class TPSynth(TPTextBasic):
                     w_.remove(w)
         paragraph = ". ".join(s1+s2)
         return paragraph
+
+    def randWord(self, words=[]):
+        if not words:
+            words = self.plain_words
+        return random.choice(words)
 
 
 class TPTabFig(TPTextBasic):

### 2018-07-13 pack/prv/opt/tokipona/plugin/t0kipona/t0kipona/utils.py
atualizando p atualizar dist e scipy p usar spatial e fazer layouts por forc

@@ -1,5 +1,5 @@
 import nltk as k, os
-from percolation.rdf import c
+# from percolation.rdf import c
 
 __doc__ = """Utilitiy variables and functions"""
 
@@ -118,13 +118,13 @@ def allTokiPonaExistentWords():
     return stats
 
 if __name__ == '__main__':
-    c('vowels:', vowels)
-    c('\n', 'consonants:', consonants)
-    c('\n', 'invalid_syllables (4):', invalid_syllables)
+    print('vowels:', vowels)
+    print('\n', 'consonants:', consonants)
+    print('\n', 'invalid_syllables (4):', invalid_syllables)
     all_syllables = allConsonantSyllables()
-    c('\n', 'valid syllables ({}):'.format(len(all_syllables)),
+    print('\n', 'valid syllables ({}):'.format(len(all_syllables)),
             all_syllables)
     all_possible_words = allTokiPonaPossibleWords()
-    c('\n', 'all tokipona possible words with 3 syllables',
+    print('\n', 'all tokipona possible words with 3 syllables',
             '({})'.format(len(all_possible_words)),
             all_possible_words)
